---
title: "Terrain and Hydrology"
format:
  docx:
    reference-doc: ./references/style.docx
    highlight-style: github
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Elevation

Terrain is the covariate behind almost everything else in a forest: it drives productivity, access, erosion, and where water collects. It arrives as a digital elevation model (DEM), a raster whose cell value is height above a datum. In operational work you download one, and the `elevatr` package fetches open tiles for any area:

```{r}
#| eval: false
library(elevatr)
aoi <- st_as_sf(data.frame(x = -78.143, y = 38.896),
                coords = c("x", "y"), crs = 4326)
dem <- get_elev_raster(aoi, z = 13) |> rast()
```

So the chapter runs offline and reproducibly, we build a DEM with a valley over the study plot instead. The surface tilts gently and is cut by a central channel, the kind of relief that governs drainage:

```{r}
library(terra)
library(sf)

stems <- st_as_sf(read.csv("data/scbi_stems.csv"),
                  coords = c("lon", "lat"), crs = 4326) |>
  st_transform(32617)

dem <- rast(ext(vect(stems)), resolution = 10, crs = "EPSG:32617")
xy  <- xyFromCell(dem, 1:ncell(dem))
ymid <- mean(xy[, 2])
elev <- 120 + 0.03 * (xy[, 1] - min(xy[, 1])) -
        18 * exp(-((xy[, 2] - ymid) / 70)^2)
dem  <- setValues(dem, elev)
names(dem) <- "elevation"
global(dem, c("min", "max", "mean"))
```

## Terrain derivatives

Elevation is rarely the variable of interest; what it *implies*, slope, aspect, ruggedness, is. `terrain()` derives these directly. Slope and aspect need a projected DEM so horizontal and vertical distances share units, which ours already is:

```{r}
slope  <- terrain(dem, "slope",  unit = "degrees")
aspect <- terrain(dem, "aspect", unit = "degrees")
tri    <- terrain(dem, "TRI")            # terrain ruggedness index
plot(c(slope, tri), main = c("Slope (deg)", "Ruggedness (TRI)"))
```

Slope matters for operability and erosion risk; ruggedness is a useful productivity covariate because sheltered, gentler terrain often carries taller, denser stands. A hillshade turns terrain into an intuitive relief image by simulating illumination, the standard base layer under any map:

```{r}
hs <- shade(terrain(dem, "slope",  unit = "radians"),
            terrain(dem, "aspect", unit = "radians"),
            angle = 40, direction = 315)
plot(hs, col = grey(0:100 / 100), legend = FALSE, main = "Hillshade")
```

## Flow direction

Water moves downhill, so a DEM already encodes drainage. `terrain()` computes flow direction, the neighbour each cell drains to, which is the first step of every hydrological routing algorithm and the input to catchment delineation:

```{r}
flowdir <- terrain(dem, "flowdir")
plot(flowdir, main = "Flow direction")
```

Full watershed delineation, flow accumulation, stream networks, catchment boundaries, is the domain of dedicated tools such as `whitebox`, which route flow across the accumulated grid. The principle is the one shown here: every hydrological product begins with elevation and the slope and flow fields derived from it.

## Keystone: hydrological analysis

Wetland and inundation mapping asks a direct question of a DEM: given a water level, what land floods? The Lake Chilwa inundation work answers it across a shallow basin whose area swings enormously with small changes in level. The method is a threshold on elevation, and it is the same logic whether the water is a lake, a floodplain, or a reservoir behind a proposed dam.

We treat the valley floor as the basin. Flooded cells are those below a chosen water level, and the flooded area is the deliverable:

```{r}
water_level <- min(values(dem), na.rm = TRUE) + 3   # 3 m above the channel floor
flooded <- dem < water_level
cell_ha <- prod(res(dem)) / 1e4
flooded_ha <- global(flooded, "sum", na.rm = TRUE)[1, 1] * cell_ha
round(flooded_ha, 2)                                  # hectares inundated
```

```{r}
plot(hs, col = grey(0:100 / 100), legend = FALSE, main = "Inundation at +3 m")
plot(flooded, add = TRUE, alpha = 0.5, col = c(NA, "steelblue"), legend = FALSE)
```

The value of the method is the rating curve: sweep the water level and record the area at each, and you have the relationship between stage and flooded extent that drives flood risk, wetland carbon, and reservoir planning:

```{r}
levels <- seq(min(values(dem), na.rm = TRUE) + 1,
              min(values(dem), na.rm = TRUE) + 8, by = 1)
area_at <- sapply(levels, function(L)
  global(dem < L, "sum", na.rm = TRUE)[1, 1] * cell_ha)
data.frame(water_level_m = round(levels, 1), flooded_ha = round(area_at, 1))
```

Reading down the table, a few metres of stage change a small flooded area into a large one, the non-linearity that makes shallow basins so sensitive and makes stating the assumed water level essential to any wetland-area claim. The real project adds a satellite-derived water mask to validate the DEM prediction, but the analytical core is this thresholded elevation surface.

## Exercises

1. Build the DEM and report its relief (max minus min elevation). Derive slope and describe where the steepest ground lies relative to the central valley.
2. Compute a hillshade with the sun in the northeast (`direction = 45`) and compare it with the northwest illumination used above. Why does illumination direction change which slopes look bright?
3. Recompute flooded area at water levels of +2, +4 and +6 m. Plot area against level: is the relationship linear? What does the shape tell you about the basin?
4. *(Applied.)* A project claims a 50-hectare seasonal wetland. Explain how you would use a DEM, a water-level assumption, and the rating-curve method above to test that claim, and what single assumption most affects your answer.
