---
title: "Geometry and Joins"
format:
  docx:
    reference-doc: ./references/style.docx
    highlight-style: github
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Geometry operations

The previous chapter joined tables by a shared key. This chapter joins them by **location**: which plots fall inside which harvest block, how much area lies within a riparian buffer, which parcel is nearest to a road. These are the operations that make GIS more than a database, and for an auditor they are the heart of the job, because the questions that matter are almost always spatial.

We work in a projected, metric CRS throughout so that buffers, areas, and distances come out in metres and hectares. As decided in the chapter on coordinate reference systems, that CRS is BC Albers.

```{r}
library(sf)
library(dplyr)

plots <- st_as_sf(read.csv("data/forest_plots.csv"),
                  coords = c("lon", "lat"), crs = 4326) |>
  st_transform(3005)
```

To have something to overlay the plots against, we define three management blocks that partition the study area. In practice these would come from a tenure or harvest-block layer; here we build them so the chapter is self-contained.

```{r}
make_block <- function(xmin, xmax, name) {
  p <- st_polygon(list(rbind(
    c(xmin, 49.24), c(xmax, 49.24), c(xmax, 49.34),
    c(xmin, 49.34), c(xmin, 49.24))))
  st_sf(block = name, geometry = st_sfc(p, crs = 4326))
}
blocks <- rbind(
  make_block(-124.78, -124.71, "West"),
  make_block(-124.71, -124.65, "Central"),
  make_block(-124.65, -124.58, "East")
) |> st_transform(3005)

plot(st_geometry(blocks), col = c("#e5f5e0", "#a1d99b", "#31a354"))
plot(st_geometry(plots), add = TRUE, pch = 19, cex = 0.4)
```

## Buffers

A **buffer** grows a geometry outward by a fixed distance. Buffers answer questions of influence and exclusion: the area within 30 metres of a stream that must be left unharvested, the zone around a plot that a pixel can be said to represent. Because we are in a metric CRS, the distance argument is simply metres:

```{r}
plot_buffers <- st_buffer(plots, 100)          # 100 m around each plot
st_area(plot_buffers[1])                       # square metres, one plot
```

## Overlays

Overlay operations combine two layers geometrically. **Union** merges geometries into one; **intersection** keeps only the shared area; **difference** cuts one shape out of another.

Dissolving the per-plot buffers into a single footprint, and measuring its area, uses union:

```{r}
footprint <- st_union(plot_buffers)
as.numeric(st_area(footprint)) / 10000         # hectares sampled within 100 m
```

Clipping the plots' sampling footprint to just the West block uses intersection:

```{r}
west <- blocks |> filter(block == "West")
west_footprint <- st_intersection(footprint, west)
as.numeric(st_area(west_footprint)) / 10000    # hectares, West block only
```

`st_difference()` is the complement: `st_difference(footprint, west)` would return the sampled area *outside* the West block. Together these three verbs express most of the polygon arithmetic an audit requires.

## Spatial predicates

Before joining by location it helps to see the raw relationship. **Predicates** return which features of one layer touch, contain, or fall within another. `st_intersects()` is the workhorse; `st_within()` and `st_contains()` are stricter. Asking which block each plot sits in:

```{r}
hits <- st_within(plots, blocks)               # list: for each plot, which block
head(hits, 3)
```

The result is a list because in general a feature can relate to several others. When the relationship is one-to-one, as here (each plot lies in exactly one block), a spatial join is tidier.

## Spatial joins

`st_join()` attaches the attributes of one layer to another wherever they intersect. This is the spatial analogue of `left_join()`: instead of matching on a key column, it matches on geometry. Tagging every plot with its block, then totalling carbon by block, is a complete auditing workflow in four lines:

```{r}
plots_tagged <- st_join(plots, blocks)

plots_tagged |>
  st_drop_geometry() |>
  group_by(block) |>
  summarise(n_plots = n(),
            mean_carbon = round(mean(carbon), 1))
```

We have gone from a bag of points and a tenure layer to a table of mean carbon density per management block, entirely from geometry.

## Distance and nearest

Finally, proximity. `st_distance()` returns the metric distances between features, and `st_nearest_feature()` finds, for each feature in one layer, the index of the closest feature in another. To find the plot nearest to the centre of the study area:

```{r}
centre <- st_centroid(st_union(blocks))
nearest <- st_nearest_feature(centre, plots)
plots[nearest, "plot_id"]
st_distance(centre, plots[nearest, ])
```

Nearest-feature logic underlies many audit checks: the closest road to a plot, the nearest watercourse to a proposed cut, the sample point nearest an anomaly flagged in imagery.

## Exercises

1. Draw a 250-metre buffer around every plot, dissolve them, and report the total sampled area in hectares. What fraction of the West block does it cover?
2. Spatially join the plots to the three blocks and produce a table of total carbon stock (MgC, remembering each plot represents 0.1 ha) by block. Which block holds the most carbon, and is that because of density or area?
3. Use `st_within()` to count how many plots fall in each block, then confirm the counts match your spatial-join table from the previous exercise.
4. Compute the distance from every plot to the boundary of the East block (hint: `st_boundary()` then `st_distance()`). Which plot is closest to the edge, and how might edge proximity matter for a carbon estimate?
