---
title: "LiDAR and Point Clouds"
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Point Clouds

Airborne laser scanning measures the forest in three dimensions. The sensor
fires pulses and times their returns, and a single pulse passing through a
canopy produces several: the first off the treetops, later ones off branches and
understorey, the last off the ground. The result is a point cloud of millions of
x, y, z coordinates from which both the ground surface and the vegetation above
it can be reconstructed.

The R package is `lidR` [@lidR], which reads the LAS and LAZ formats laser data
ships in. It bundles sample tiles, so this chapter runs on real airborne data
without downloading anything:

```{r}
library(lidR)
library(terra)
library(sf)

topo <- readLAS(system.file("extdata", "Topography.laz", package = "lidR"),
                filter = "-drop_z_below 0")
topo
```

Two properties govern everything that follows. Point density sets the finest
structure that can be resolved, and the classification field records what each
return is assumed to have hit. The LAS standard reserves class 2 for ground and
class 1 for unclassified, and the numbers matter because every later step keys
off them:

```{r}
data.frame(
  points     = npoints(topo),
  area_ha    = round(as.numeric(st_area(topo)) / 1e4, 2),
  density    = round(density(topo), 2),
  z_min      = round(min(topo$Z), 1),
  z_max      = round(max(topo$Z), 1)
)

table(topo$Classification)
```

Elevations here are hundreds of metres above the datum, not heights above
ground. That distinction is the subject of most of this chapter.

## Ground Classification

Separating returns that hit soil from those that hit low vegetation is the
delicate step, and everything downstream inherits its errors. Two algorithms
dominate.

Cloth simulation filtering [@zhang2016easy] inverts the point cloud and drapes a
simulated cloth over it; the cloth settles onto the ground and the points it
touches are classified. Progressive morphological filtering [@zhang2003pmf]
instead opens the surface with windows of increasing size, removing anything
that rises too fast over too short a distance.

Stripping the existing classification and running both makes the comparison
honest:

```{r}
raw <- topo
raw@data$Classification <- 0L

t_csf <- system.time(
  g_csf <- classify_ground(raw, csf(sloop_smooth = TRUE, 0.5, 1))
)
t_pmf <- system.time(
  g_pmf <- classify_ground(raw, pmf(ws = c(3, 6, 9), th = c(0.5, 1, 1.5)))
)

data.frame(
  algorithm = c("CSF", "PMF"),
  ground    = c(sum(g_csf$Classification == 2L), sum(g_pmf$Classification == 2L)),
  seconds   = round(c(t_csf[["elapsed"]], t_pmf[["elapsed"]]), 2)
)

round(100 * mean((g_csf$Classification == 2L) == (g_pmf$Classification == 2L)), 1)
```

The two agree on about nine returns in ten and disagree on the tenth, with PMF
admitting substantially more points as ground. Neither is a reference standard.
The disagreement is concentrated where it always is, under dense low vegetation
and on steep breaks of slope, and it propagates directly into terrain height and
therefore into every canopy height derived from it. A project that does not state
which algorithm produced its terrain model, and with what parameters, has not
described a reproducible method.

## Noise Removal

Laser data contains returns that are physically impossible: birds, low cloud,
multipath reflections, and points hundreds of metres below the ground. Statistical
outlier removal flags a point when its mean distance to its `k` nearest
neighbours exceeds `m` standard deviations of that statistic over the cloud:

```{r}
noisy <- classify_noise(g_csf, sor(k = 10, m = 3))
sum(noisy$Classification == LASNOISE)

clean <- filter_poi(noisy, Classification != LASNOISE)
```

A few hundred points out of seventy thousand. The count is small and the
consequence is not, because a single spurious high return becomes a spurious
treetop, and a single spurious low return drags the interpolated ground surface
down with it across the whole neighbourhood.

## Terrain Models

Interpolating the ground returns onto a grid gives the digital terrain model.
`lidR` offers inverse-distance weighting on the k nearest neighbours, a
triangulated surface, and kriging. The first is fast and robust at moderate
density:

```{r}
dtm <- rasterize_terrain(clean, res = 1, knnidw(k = 10, p = 2, rmax = 50))
global(dtm, range, na.rm = TRUE)
```

```{r}
#| fig-cap: "Hillshaded terrain model interpolated from the classified ground returns."
hs <- shade(terrain(dtm, "slope",  unit = "radians"),
            terrain(dtm, "aspect", unit = "radians"),
            angle = 40, direction = 270)

plot(hs, col = grey(0:100 / 100), legend = FALSE, mar = c(2, 2, 2, 2))
```

## Height Normalisation

Biomass relates to height above ground, not height above sea level. Normalising
subtracts the terrain from every point, so a return four metres up a slope at
800 m and one four metres up a slope at 810 m become the same number:

```{r}
norm <- normalize_height(clean, dtm)

data.frame(
  z_min = round(min(norm$Z), 2),
  z_max = round(max(norm$Z), 2)
)
```

The check on a normalisation is the residual on the ground returns themselves.
They should centre on zero, because the surface was interpolated through them:

```{r}
g <- filter_ground(norm)$Z

data.frame(
  mean_m  = round(mean(g), 3),
  sd_m    = round(sd(g), 3),
  q95_abs = round(unname(quantile(abs(g), 0.95)), 3)
)
```

Centred, with a spread of a few centimetres. A mean that departs from zero
indicates a systematic terrain bias, and a spread of metres rather than
centimetres indicates that ground classification has admitted vegetation.

### Normalisation Order

The order of operations here is not interchangeable, and getting it wrong fails
silently. Interpolating a terrain model from an already-normalised cloud
produces this:

```{r}
dtm_wrong <- rasterize_terrain(norm, res = 1, knnidw(k = 10, p = 2, rmax = 50))
global(dtm_wrong, range, na.rm = TRUE)
```

Twenty-five metres of relief have become one metre of residual noise about zero.
The arithmetic is correct and the result is useless: the ground returns of a
normalised cloud are already at height zero, so interpolating them recovers the
interpolation residual rather than the terrain.

Nothing errors. The object is a valid raster of the right extent, resolution and
projection, and it plots without complaint. Slope computed from it is noise,
aspect is undefined, and any model carrying terrain covariates loses them
entirely while continuing to fit, predict and report. The rule is that terrain
models are built before normalisation, from the classified but un-normalised
cloud, and there is no exception.

## Canopy Height Models

The canopy height model is the surface of the vegetation over the terrain. Built
from a normalised cloud it is a direct rasterisation of the highest return in
each cell, and the difficulty is the cells that receive no return at all:

```{r}
mc <- readLAS(system.file("extdata", "MixedConifer.laz", package = "lidR"),
              filter = "-drop_z_below 0")

chm_p2r     <- rasterize_canopy(mc, res = 0.5, p2r(subcircle = 0.2))
chm_pitfree <- rasterize_canopy(mc, res = 0.5, pitfree(subcircle = 0.2))

data.frame(
  algorithm = c("points-to-raster", "pit-free"),
  empty     = c(global(is.na(chm_p2r),     "sum")[1, 1],
                global(is.na(chm_pitfree), "sum")[1, 1]),
  cells     = ncell(chm_p2r)
)
```

Points-to-raster leaves several hundred empty cells: the pits that appear
wherever a pulse penetrated to the ground between crowns. The pit-free algorithm
[@khosravipour2014generating] builds a set of triangulated surfaces at
successive height thresholds and takes their maximum, which closes almost all of
them. Pits matter because local-maximum tree detection reads a pit as a crown
edge and splits one tree into two.

```{r}
#| fig-cap: "Pit-free canopy height model of the mixed conifer stand."
plot(chm_pitfree, col = height.colors(50), mar = c(2, 2, 2, 4))
```

## Individual Tree Detection

Each treetop is a local maximum: a cell higher than every other cell within some
window. The window size is the entire method, and it is not a detail.

### Window Size

```{r}
counts <- sapply(c(3, 5, 11), function(w) nrow(locate_trees(mc, lmf(ws = w))))
area_ha <- as.numeric(st_area(mc)) / 1e4

data.frame(
  window_m  = c(3, 5, 11),
  trees     = counts,
  stems_ha  = round(counts / area_ha)
)
```

A three-metre window finds nearly five times as many trees as an eleven-metre
window over the same 0.8 hectares. Neither count is a measurement of the stand.
The small window splits broad crowns into multiple maxima; the large one merges
adjacent trees and detects only dominants. The correct window depends on crown
size, which depends on height, which is why fixed windows are a starting point
rather than an answer.

### Variable Windows

Making the window a function of height is the standard refinement. A linear
function with a floor is the usual form, and the coefficients are stand-specific:

```{r}
wf <- function(x) pmax(0.07 * x + 0.6, 0.5)

ttops <- locate_trees(mc, lmf(wf), uniqueness = "bitmerge")

data.frame(
  trees    = nrow(ttops),
  stems_ha = round(nrow(ttops) / area_ha)
)
```

This is a calibrated function taken from a different forest, a regenerating
burn, and applied here it returns an implausible stem density: an order of
magnitude above what a mixed conifer stand carries. The window it opens at
typical heights in this stand is a couple of metres, far narrower than the
crowns, so it fragments them.

The lesson is not that variable windows are wrong. It is that a window function
is a calibration, valid for the stand structure it was fitted to, and
transferring one between forests without refitting produces confident nonsense.
Detected stem counts must be checked against field plots before they are used.

The `uniqueness = "bitmerge"` argument is a separate and mandatory precaution.
When a survey is processed tile by tile with overlapping buffers, sequential
tree identifiers collide between tiles and the same tree is counted twice.
Bitmerge derives the identifier from the tree's coordinates instead, so the
duplicate resolves to one tree.

## Crown Segmentation

Detection gives points. Segmentation assigns every return to a tree, which is
what yields crown area and per-tree metrics. Dalponte and Coomes' algorithm
[@dalponte2016tree] grows regions outward from each detected top, bounded by
height thresholds relative to the seed:

```{r}
ttops_fixed <- locate_trees(mc, lmf(ws = 5))
ttops_fixed$treeID <- seq_len(nrow(ttops_fixed))

seg    <- segment_trees(mc, dalponte2016(chm_pitfree, ttops_fixed))
crowns <- crown_metrics(seg, func = .stdtreemetrics, geom = "convex")

data.frame(
  crowns          = nrow(crowns),
  median_area_m2  = round(median(crowns$convhull_area), 1),
  median_height_m = round(median(crowns$Z), 1)
)
```

The `treeID` reset on the line above is not cosmetic. Identifiers produced by
`uniqueness = "bitmerge"` are 64-bit integers derived from coordinates, and
`dalponte2016()` uses the identifier to index its output raster. Passing bitmerge
identifiers to it produces an error or a corrupted segmentation depending on
version. Reset them to a simple sequence within a single tile.

```{r}
#| fig-cap: "Segmented crown polygons over the canopy height model."
plot(chm_pitfree, col = height.colors(50), mar = c(2, 2, 2, 4))
plot(st_geometry(crowns), add = TRUE, border = "black", lwd = 0.4)
```

## Dominant Height

Stand height is a mensurational variable with a specific definition, usually the
mean height of the tallest trees per unit area rather than the mean of all
trees. The raster equivalent is an upper percentile of canopy height aggregated
to a coarser cell:

```{r}
q95 <- function(x, ...) quantile(x, 0.95, na.rm = TRUE)
chm_dom <- aggregate(chm_pitfree, fact = 2, fun = q95)

v  <- terra::extract(chm_dom, vect(ttops))[, 2]
ok <- !is.na(v)

data.frame(
  r    = round(cor(ttops$Z[ok], v[ok]), 3),
  rmse = round(sqrt(mean((ttops$Z[ok] - v[ok])^2)), 2),
  bias = round(mean(ttops$Z[ok] - v[ok]), 2)
)
```

The raster surface tracks the detected tree heights closely but sits about a
metre high, the negative bias, because a 95th percentile over a cell containing
several crowns returns something near the tallest of them rather than the mean.
Upper percentiles are preferred as biomass predictors for exactly this reason:
they are dominated by the large trees that carry most of the biomass, and they
are insensitive to the understorey that laser data sees poorly.

## Area-Based Inventory

Individual-tree methods are appealing and rarely how operational inventory is
done. The area-based approach instead summarises the point cloud into metrics on
a grid matching the field plot size, regresses plot-measured biomass on those
metrics, and applies the model wall to wall.

The metrics are distributional summaries of return heights in each cell:

```{r}
mega <- readLAS(system.file("extdata", "Megaplot.laz", package = "lidR"),
                filter = "-drop_z_below 0")
mega <- normalize_height(classify_ground(mega, csf(sloop_smooth = TRUE, 0.5, 1)),
                         knnidw())

m <- pixel_metrics(mega, ~list(
  zmax  = max(Z),
  zmean = mean(Z),
  zsd   = sd(Z),
  zq75  = quantile(Z, 0.75),
  zq95  = quantile(Z, 0.95),
  cover = sum(Z > 2) / length(Z)
), res = 20)

names(m)
round(as.data.frame(global(m, "mean", na.rm = TRUE)), 2)
```

Every cell now carries a description of its vertical structure, and this is the
form in which laser data enters a biomass model. Height percentiles and cover
fraction carry most of the signal; the mean carries least, because it is dragged
down by ground and understorey returns.

The reason this matters for verification is coverage. A canopy metric exists for
every cell in the survey, where field plots sample a fraction of a per cent of
the area, so laser data turns a few dozen plots into a wall-to-wall map. That
leverage is also the risk: the model is fitted on the plots and applied
everywhere, so its error applies everywhere too, and the uncertainty chapter
treats how that error is estimated and carried into a reported number.

::: {.verification}
#### Documentation Requirements {.unnumbered .unlisted}

Laser-derived inventory is reproducible in principle and frequently not in
practice. A verifier should require, in writing, the items this chapter has
shown to change the answer.

The acquisition specification: sensor, date, pulse density, scan angle, and
stated vertical accuracy. Density sets what can be resolved and the date sets
what was standing.

The ground classification algorithm and its parameters. The two algorithms
compared above differed by around ten per cent of returns on the same tile, and
that difference moves the terrain surface and therefore every canopy height.

The normalisation order. A terrain model built after normalisation is flat, and
nothing in the output announces it.

The detection method and its calibration. A window function is fitted to a stand
structure. The one used above, transferred from another forest, overstated stem
density by roughly an order of magnitude while producing no warning of any kind.

The plot-to-metric model and its validation. Predicted biomass is a model
output, not a measurement, and a map without a cross-validated error statistic
carries no defensible uncertainty.
:::

## Exercises

1. Rebuild the canopy height model at 1 m rather than 0.5 m resolution and
   recount the empty cells under both `p2r()` and `pitfree()`. Explain why
   coarsening the grid reduces pits and what it costs.

2. Run tree detection with fixed windows of 3, 5, 7 and 11 m and plot detected
   stems per hectare against window size. At what window does the count
   stabilise, and what does the shape of that curve tell you about mean crown
   diameter in this stand?

3. Refit the variable window function so that the detected stem density falls
   within a plausible range for a mixed conifer stand. State the coefficients you
   chose and the stand assumption that justifies them.

4. Compute the 25th, 50th, 75th and 95th height percentiles from the pixel
   metrics above. Explain why upper percentiles outperform the mean as biomass
   predictors, referring to where biomass sits in the height distribution.

5. *(Applied.)* You have laser coverage of 10,000 ha and field plots on 30 ha.
   Describe, in steps, how you would calibrate a metric-based biomass model and
   produce a wall-to-wall estimate, naming the two largest sources of error and
   where each enters.
