---
title: "Disturbance and Risk"
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Disturbance Screening

A verifier's central remote-sensing task is to find where the forest changed
between two dates, and to decide what caused it. The screening is not a
substitute for plot remeasurement or the developer's monitoring report. It is an
independent layer used to direct attention: to find the clearing nobody
declared, and to test whether the loss a project did declare has the extent and
the cause it claims.

The method is spectral index differencing. Take an index at two dates, subtract,
and threshold the difference. Everything difficult is in the last step, and this
chapter is mostly about the last step.

The working data are the 2022 Sentinel-2 land-cover map of Rondonia and one NDVI
composite over the same window, both introduced in the raster chapter. The map
carries labelled disturbance classes, which makes it possible to measure how
well an index threshold actually separates cleared ground from forest rather
than assuming it does:

```{r}
library(terra)
library(sf)

sf_use_s2(FALSE)

cls  <- rast("data/rondonia_class_2022.tif")
ndvi <- rast("data/rondonia_ndvi_2022-07-16.tif") / 1e4
leg  <- read.csv("data/rondonia_legend.csv")

window <- project(as.polygons(ext(ndvi), crs = crs(ndvi)), crs(cls))
cw     <- crop(cls, window) |> project(ndvi, method = "near")
names(cw) <- "class"; names(ndvi) <- "ndvi"
```

The NDVI rasters store the index multiplied by ten thousand, which is why the
division appears above. Reading them raw yields indices in the thousands and
every threshold in this chapter silently fails to fire.

## Spectral Indices

Three indices carry most operational disturbance screening, and they answer
different questions.

$$\text{NDVI} = \frac{\text{NIR} - \text{Red}}{\text{NIR} + \text{Red}}
\qquad
\text{NDMI} = \frac{\text{NIR} - \text{SWIR}_1}{\text{NIR} + \text{SWIR}_1}
\qquad
\text{NBR} = \frac{\text{NIR} - \text{SWIR}_2}{\text{NIR} + \text{SWIR}_2}$$

NDVI tracks greenness and responds to canopy removal. NDMI tracks canopy
moisture and responds to stress before the canopy is gone, which makes it the
index for insect attack and drought. NBR contrasts near-infrared against the
longer shortwave band where char absorbs strongly, so it responds to fire.

The sign conventions differ and mixing them up inverts the map:

$$\Delta\text{NDVI} = \text{pre} - \text{post}
\qquad
\Delta\text{NDMI} = \text{pre} - \text{post}
\qquad
\Delta\text{NBR} = \text{post} - \text{pre}$$

Positive dNDVI and dNDMI mean loss. Positive dNBR means burned, following the
convention of the Monitoring Trends in Burn Severity programme. There is no deep
reason for the inconsistency; it is historical, and it has to be checked against
the source of any published threshold before that threshold is applied.

## Threshold Selection

Published severity breaks exist for each index. The ones in operational use,
drawn from the burn severity literature and from forest health screening
practice, are these:

| Index | Low | Moderate | High |
|---|---|---|---|
| dNDVI | 0.10 to 0.20 | 0.20 to 0.35 | above 0.35 |
| dNDMI | 0.15 to 0.30 | 0.30 to 0.45 | above 0.45 |
| dNBR | 0.10 to 0.27 | 0.27 to 0.44 | above 0.44 |

These are defaults, not constants. Treating them as constants is the most
common failure in disturbance screening, and the way to see why is to measure
what a threshold actually catches.

### Class Separability

The land-cover map labels each pixel, so the NDVI distribution of each class can
be read directly. This is the empirical basis a threshold ought to rest on:

```{r}
d <- as.data.frame(c(ndvi, cw), na.rm = TRUE)

tab <- do.call(rbind, lapply(leg$id, function(i) {
  v <- d$ndvi[d$class == i]
  if (!length(v)) return(NULL)
  data.frame(label  = leg$label[leg$id == i],
             forest = leg$forest[leg$id == i],
             n      = length(v),
             q05    = round(unname(quantile(v, 0.05)), 3),
             median = round(median(v), 3),
             q95    = round(unname(quantile(v, 0.95)), 3))
}))
tab
```

Three things in that table decide everything downstream.

Intact forest is tight: a median around 0.84 with the fifth percentile still
above 0.78. Cleared ground with bare soil sits far below it, around 0.51. Those
two are separable and any sensible threshold finds them.

Clear cut with regrowth is not separable. Its median sits near 0.79, inside the
forest distribution, because two or three years after clearing the vegetation
that returns is green. An index threshold cannot distinguish recovering clearing
from forest, and a project whose deforestation is a few years old will not be
caught by one.

Wetland has a median around 0.84, indistinguishable from forest, which is why
water masking is applied before thresholding rather than after.

Measuring the trade-off directly makes the cost of each published break
explicit:

```{r}
forest_ndvi <- d$ndvi[d$class %in% leg$id[leg$forest]]
cleared     <- d$ndvi[d$class %in% c(1, 2, 6)]
baseline    <- median(forest_ndvi)

sens <- do.call(rbind, lapply(c(0.10, 0.15, 0.20, 0.30), function(t) {
  cut <- baseline - t
  data.frame(dNDVI_threshold = t,
             ndvi_below      = round(cut, 3),
             pct_cleared_caught = round(100 * mean(cleared < cut), 1),
             pct_forest_flagged = round(100 * mean(forest_ndvi < cut), 1))
}))
sens
```

Read the third column. The "Low severity" break of 0.10 catches about three
quarters of genuinely cleared ground. The "Moderate" break of 0.20 catches
three fifths. The "High severity" break of 0.35, extrapolating past the last
row, catches well under half. Every one of those misses is omitted
deforestation, and omission understates emissions, which is the direction that
concerns an auditor most.

The fourth column is the other side of the trade. Tightening the threshold buys
a falling false-positive rate on forest, from 2.5 per cent down to 0.5. That is
a real gain, and it is why the tight thresholds are popular: they produce clean
maps. Clean is not the same as correct.

### Distributional Diagnostics

Since defaults cannot be trusted, the threshold has to be read off the data. The
diagnostic is the histogram of the difference image, and it supports three
readings.

```{r}
#| fig-cap: "NDVI distribution for forest and for cleared classes, with the published dNDVI breaks marked relative to the forest median."
hist(forest_ndvi, breaks = 60, xlim = c(-0.2, 1), col = "#1E844980",
     border = NA, main = "", xlab = "NDVI", ylab = "Cells")
hist(cleared, breaks = 60, col = "#D7C49C99", border = NA, add = TRUE)
abline(v = baseline - c(0.10, 0.20, 0.35), lty = c(3, 2, 1), col = "grey20")
legend("topleft", bty = "n",
       legend = c("Forest", "Cleared", "Breaks: 0.10 / 0.20 / 0.35"),
       fill = c("#1E844980", "#D7C49C99", NA), border = NA)
```

A unimodal distribution centred on zero with a narrow tail means noise and
phenology only, and the defaults can stand because nothing is there to find.

A bimodal distribution with a gap above about 0.15 means a real disturbance
signal, and the low break belongs in the gap rather than on a default.

A long unimodal right tail with no gap means the composite is contaminated, by
cloud shadow, senescence, or view-angle effects. The correct response is to
rebuild the composite with a tighter cloud filter or a shifted date window, not
to classify it.

## Composite Construction

The threshold problem is downstream of a compositing problem, and three
practical rules do more for accuracy than any amount of threshold tuning.

Use enough scenes. A five-day revisit over a two-month window yields eight or
more cloud-filtered observations per pixel, and a median composite is unstable
below about four.

Choose the season for the question. Composites built in the dormant or senescent
season carry a canopy that is naturally dry and naturally less green, which
inflates dNDMI enormously. In one verification the default moisture breaks
applied to an October to December composite flagged 38 per cent of the project
area as disturbed; rebuilding the same analysis on a July to September window
collapsed that to 4 per cent, all of it co-located with reported beetle survey
polygons. The 34-point difference was seasonal phenology, not disturbance.

Watch the processing baseline. Sentinel-2 surface reflectance carries a digital
number offset introduced in January 2022. Mixing scenes across that boundary
without harmonising propagates a spurious index difference of around 0.04 across
the entire project area, which is a third of the way to the low severity break
before any real change has occurred.

## Agent Attribution

Detecting loss is half the task. Deciding its cause determines whether it is a
reversal the project must account for, a natural disturbance that may be
excluded, or a leakage signal on land the proponent controls elsewhere.

No single index answers it. The cross-check does:

| Signal | Reading | Required corroboration |
|---|---|---|
| High dNDVI, dNBR clean | Harvest, blowdown or clearing | Edge geometry, harvest records |
| High dNDVI, high dNBR | Fire | Official burn perimeters |
| High dNDMI alone | Moisture stress | Drought index or pest survey |

The third row is the one most often over-read. Moisture stress on its own is not
disturbance, and classifying it as such without corroboration produces the false
positives described above.

### Edge Geometry

The most reliable attribution cue is the shape of the patch. Anthropogenic
clearing follows property and operational boundaries, so it has straight edges,
right angles, and fills its bounding box. Natural disturbance follows wind,
terrain and hydrology, so it is curvilinear and amorphous.

That is a testable claim, and the labelled map lets it be tested. Rectangularity,
the ratio of a patch's area to the area of its bounding box, formalises it:

```{r}
rectangularity <- function(class_ids, min_cells = 40) {
  m <- ifel(cw %in% class_ids, 1, NA)
  p <- as.polygons(patches(m, directions = 8), dissolve = TRUE) |> st_as_sf()
  p$area <- as.numeric(st_area(p))
  p <- p[p$area >= min_cells * prod(res(cw)), ]
  bb <- vapply(seq_len(nrow(p)), function(i) {
    b <- st_bbox(p[i, ]); as.numeric((b$xmax - b$xmin) * (b$ymax - b$ymin))
  }, numeric(1))
  p$area / bb
}

anthro  <- rectangularity(c(1, 2))
natural <- rectangularity(c(8, 9))

data.frame(
  origin      = c("Clear cut (bare, burned)", "Wetland, seasonally flooded"),
  patches     = c(length(anthro), length(natural)),
  median_rect = round(c(median(anthro), median(natural)), 3)
)
```

The cleared patches sit marginally above the natural ones, and the difference is
far too small to separate anything. As a summary statistic over all patches, the
edge geometry rule fails here.

That result is worth more than a confirmation would have been, because it locates
what the rule actually is. Three things defeat the aggregate test. Patches
delineated from a classified map inherit the classifier's edges, which are
already blocky at 20 m whatever produced them. Amazonian wetlands follow river
courses and are themselves linear and compact. And rectangularity collapses a
shape to one number, discarding the thing an analyst is actually reading, which
is whether the corners are square.

The rule is a visual heuristic applied to individual polygons by someone looking
at them, and it is a good one in that role: a right-angled corner on a loss patch
is strong evidence of machinery. It is not a classifier, it does not survive
being averaged, and a screening workflow that automates it into a threshold will
find nothing. Geometry raises or lowers a prior on a specific polygon. Records
settle it.

## Risk Mapping

Where disturbance is *likely* is a different question from where it happened,
and it is the one that drives a baseline. Under VM0048 and the associated
allocation tooling, a risk map distributes an expected quantity of future loss
across the landscape in proportion to modelled pressure.

The covariates are distances. Pressure falls with distance from existing
clearing, from roads, from settlements and from water, and rises on flat ground.
Two of those are computable from the map itself:

```{r}
forest_mask <- ifel(cw %in% leg$id[leg$forest], 1, NA)
edge        <- boundaries(forest_mask, inner = FALSE)
dist_edge   <- distance(edge) / 1000

cleared_mask <- ifel(cw %in% c(1, 2, 6), 1, NA)
dist_clear   <- distance(cleared_mask) / 1000

normalise <- function(x) {
  r <- unlist(global(x, range, na.rm = TRUE))
  (x - r[1]) / (r[2] - r[1])
}

risk_edge  <- 1 - normalise(dist_edge)
risk_clear <- 1 - normalise(dist_clear)
```

Both are inverted, because risk rises as distance falls. Forgetting the
inversion produces a map that is exactly wrong and looks entirely plausible.

### Index Weighting

The covariates are combined as a weighted sum. A published implementation uses
these weights:

```{r}
weights <- c(edge = 0.2, roads = 0.2, places = 0.2,
             urban = 0.1, water = 0.1, slope = 0.1)
sum(weights)
```

They sum to 0.9, not 1.0. The consequence is not that the map is wrong: the
index is rescaled to zero-one afterwards, and the allocation step below
normalises by the zonal sum, so a constant factor cancels. The consequence is
that the number cannot be audited. Nobody reading the index can tell whether the
missing tenth is a dropped covariate, a typing error, or an intentional choice,
and a verifier confronted with weights that do not sum to unity is entitled to
reject the whole construction until it is explained.

Weights should sum to one and each should be justified. With the two covariates
available here:

```{r}
risk <- 0.5 * risk_edge + 0.5 * risk_clear
risk <- normalise(risk)
names(risk) <- "risk"

global(risk, quantile, probs = c(0.05, 0.5, 0.95), na.rm = TRUE)
```

```{r}
#| fig-cap: "Deforestation risk index over the Rondonia window, with mapped forest outlined."
plot(risk, col = hcl.colors(50, "Inferno"), mar = c(2, 2, 2, 4))
```

### Risk Allocation

The risk surface is relative, not absolute. It becomes a baseline only when a
quantity of expected loss is distributed across it. The allocation rule assigns
each pixel a share of its zone's expected annual loss in proportion to its own
risk:

$$\text{AllocatedLoss}_{\text{pixel}}
= \frac{\text{risk}_{\text{pixel}}}{\sum \text{risk}_{\text{zone}}}
\times \text{AnnualLoss}_{\text{zone}}$$

In practice the fraction is computed as a per-zone factor and multiplied
through, which is the same arithmetic in a form that rasterises:

```{r}
annual_loss_ha <- 4654          # zone expected annual loss, hectares
risk_forest    <- mask(risk, forest_mask)
risk_sum       <- global(risk_forest, "sum", na.rm = TRUE)[1, 1]

loss_factor <- annual_loss_ha / risk_sum
allocated   <- risk_forest * loss_factor

data.frame(
  risk_sum        = round(risk_sum),
  loss_factor     = signif(loss_factor, 3),
  allocated_total = round(global(allocated, "sum", na.rm = TRUE)[1, 1], 1)
)
```

The allocated total reproduces the input by construction. That is the property to
check first on any allocation: it is a redistribution, not a model of how much
will be lost, and if the totals do not close the implementation is broken.
Everything about *how much* comes from the historical rate, which the baseline
chapter treats, and everything about *where* comes from the risk index above.

## Leakage Belts

Activity displaced outside a project boundary is leakage, and it is assessed on a
belt drawn around the project. The rules are prescriptive: a belt of a stated
width, then a sequence of exclusions for land where deforestation could not
plausibly be displaced, or where it would be attributed elsewhere.

The exclusions typically remove ground within a set distance of existing road
infrastructure, ground steeper than ten per cent, mapped wetlands, and protected
areas in the stricter management categories. The slope rule is worth writing
carefully, because it is a strict inequality:

```{r}
#| eval: false
slope_pct     <- tan(terrain(dem, "slope", unit = "degrees") * pi / 180) * 100
slope_invalid <- slope_pct > 10          # strictly greater, not >=
```

The combination step is where implementations fail. Each mask produces a
residual belt, the belt minus that exclusion, and the valid area is the ground
that survives *every* exclusion. That is an intersection:

```{r}
#| eval: false
valid <- Reduce(st_intersection,
                list(belt_minus_roads, belt_minus_slope,
                     belt_minus_wetland, belt_minus_protected))
```

Chaining `st_union()` across the four residuals instead, which is an easy slip
because each residual is itself called a leakage area, readmits every excluded
parcel: a pixel excluded by the slope mask survives in the roads residual, and
the union restores it. The resulting belt is larger than it should be in a
direction that suits the project. A verifier reproducing a leakage area should
check the set operation before checking the arithmetic.

::: {.verification}
#### Standards Divergence {.unnumbered .unlisted}

Two families of requirement apply to the material in this chapter, and they rest
on different philosophies. Treating them as interchangeable misrepresents both.

The Verra methodologies approach map quality through stratified area estimation.
A map is accompanied by a sample drawn across strata, the sample yields an
area-adjusted estimate with a confidence interval, and per-class accuracies must
clear stated floors: for deforestation, user's and producer's accuracy of at
least 70 per cent each, with forest above 90 and open canopy classes at 60 and
80. The map is not the estimate; the sample is. The uncertainty chapter develops
the estimator.

The ACR improved forest management standards approach disturbance as screening
rather than estimation. There is no confusion matrix requirement, because the
output is not an area claim. Defensibility comes instead from a documented
threshold justified by the histogram, corroboration across independent layers,
and cross-reference to official fire perimeters and pest survey polygons. A
finding directs a verifier to a location; the account of what happened there
comes from records and from the field.

The practical consequence is that a threshold and a sample answer different
questions. Do not report a screening result as an area estimate, and do not
offer a stratified area estimate as evidence that a particular polygon was
harvested.
:::

## Exercises

1. Recompute the threshold sensitivity table for breaks of 0.05 and 0.25. At
   what threshold does the proportion of cleared ground caught fall below half,
   and what does that imply for a project relying on the published high severity
   break?

2. Repeat the class separation table using only the clear cut with regrowth
   class against forest. Explain why no NDVI threshold separates them, and name
   the data source that would.

3. Compute rectangularity for the riparian forest class and compare it against
   both groups in the text. Riparian forest is natural but linear: what does
   that do to the edge geometry rule, and how would you amend the rule?

4. Rebuild the risk index weighting distance to existing clearing at 0.8 and
   distance to forest edge at 0.2. How much of the high risk area moves, and
   what assumption about deforestation behaviour does the reweighting encode?

5. *(Applied.)* A project reports 200 hectares of avoided deforestation, an
   overall map accuracy of 78 per cent, and a disturbance screen showing no
   unreported loss. State what each of those three claims does and does not
   support, and list the per-class accuracies you would require before accepting
   the area.
