---
title: "Uncertainty"
format:
  docx:
    reference-doc: ./references/style.docx
    highlight-style: github
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Why uncertainty

Every estimate in this book is wrong by some amount, and the discipline of verification is knowing by how much. A carbon figure without an uncertainty is not conservative, it is simply incomplete, and the crediting standards, ART-TREES, VCS, ACR, all deduct credits in proportion to how uncertain the estimate is. Reducing uncertainty is therefore worth real money, and quantifying it honestly is the auditor's core duty. This chapter builds the estimate, then its error, on the running stem data.

## A carbon estimate

We convert diameters to aboveground biomass with a published generic hardwood equation [@jenkins2003biomass], apply the IPCC default carbon fraction of 0.47 [@ipcc2006gl], and scale to the plot area:

```{r}
library(dplyr)
stems <- read.csv("data/scbi_stems.csv")

stems <- stems |>
  mutate(agb_kg   = exp(-2.48 + 2.4835 * log(dbh)),   # Jenkins generic hardwood
         carbon_Mg = 0.47 * agb_kg / 1000)

area_ha  <- 25.6
total_C  <- sum(stems$carbon_Mg)
C_per_ha <- total_C / area_ha
round(c(total_MgC = total_C, MgC_per_ha = C_per_ha), 1)
```

That point estimate is the easy part. Three distinct sources of error sit beneath it: sampling error from measuring plots rather than the whole population, model error in the allometric equation, and parameter error in constants like the carbon fraction.

## Sampling error

Divide the plot into quadrats and each becomes a sample of carbon density. The variation among them is the sampling uncertainty of the plot mean, summarised by the standard error and a confidence interval:

```{r}
stems$quadrat <- paste(cut(stems$gx, 8), cut(stems$gy, 12))
quadrat_area  <- 25.6 / (8 * 12)                 # ha per quadrat (equal-area cut)

qd <- stems |>
  group_by(quadrat) |>
  summarise(carbon_Mg = sum(carbon_Mg), .groups = "drop") |>
  mutate(carbon_ha = carbon_Mg / quadrat_area)

m  <- mean(qd$carbon_ha); se <- sd(qd$carbon_ha) / sqrt(nrow(qd))
ci <- m + c(-1.96, 1.96) * se
round(c(mean = m, se = se, lower = ci[1], upper = ci[2]), 1)
```

The half-width relative to the mean is the number the standards care about: the wider it is, the deeper the deduction.

## Monte Carlo propagation

Sampling error is only part of the story; the allometric model and the carbon fraction are uncertain too, and their errors compound. When error propagates through a non-linear function like $a\,D^{b}$, the clean way to combine sources is Monte Carlo simulation: draw the uncertain inputs from their distributions many times, recompute the whole estimate each time, and read the spread of results [@ipcc2006gl].

```{r}
set.seed(42)
n_sim <- 2000
sim <- numeric(n_sim)
for (i in seq_len(n_sim)) {
  b  <- rnorm(1, 2.4835, 0.05)            # allometric exponent uncertainty
  cf <- rnorm(1, 0.47, 0.02)              # carbon-fraction uncertainty
  agb <- exp(-2.48 + b * log(stems$dbh))
  sim[i] <- sum(cf * agb / 1000) / area_ha
}
mc_ci <- quantile(sim, c(0.025, 0.975))
round(c(mean = mean(sim), lower = mc_ci[1], upper = mc_ci[2]), 1)
```

```{r}
hist(sim, breaks = 40, col = "grey80", border = "white",
     main = "Monte Carlo carbon density", xlab = "MgC / ha")
abline(v = mc_ci, col = "firebrick", lwd = 2, lty = 2)
```

## Conservative estimates

The standards translate uncertainty into a deduction. A common rule is that when the half-width of the 90% or 95% interval exceeds a threshold (often 15% of the mean), credits are issued against the conservative lower bound rather than the mean. Computing the uncertainty percentage and the conservative estimate is a few lines:

```{r}
half_width <- (mc_ci[2] - mc_ci[1]) / 2
unc_pct    <- 100 * half_width / mean(sim)
conservative <- mean(sim) - half_width           # lower-bound crediting basis
round(c(uncertainty_pct = unc_pct,
        mean = mean(sim), conservative = conservative), 1)
```

If the uncertainty percentage clears the threshold, the gap between the mean and the conservative figure is credit the project forgoes for being imprecise, the direct financial incentive to measure better.

## Keystone: area-adjusted accuracy

The Bolivia land-cover work raises the same question for areas rather than carbon: a class area read straight off a map inherits the map's classification errors, and good practice adjusts it and attaches a confidence interval [@olofsson2014gpg]. This is the LULC counterpart of the carbon deduction, and it uses the confusion matrix from the disturbance chapter. We take mapped forest and non-forest areas and an error matrix, and compute the area-adjusted estimate:

```{r}
A    <- c(forest = 18000, nonforest = 7000)      # mapped area, ha
Atot <- sum(A); W <- A / Atot

# error matrix: rows = map class, cols = reference class
n   <- matrix(c(95, 5, 12, 88), nrow = 2, byrow = TRUE,
              dimnames = list(map = names(A), ref = names(A)))
ni  <- rowSums(n)

# Olofsson stratified estimator of the reference area proportion of "forest"
p_forest <- sum(W * n[, "forest"] / ni)
se_p     <- sqrt(sum(W^2 * (n[, "forest"]/ni) * (1 - n[, "forest"]/ni) / (ni - 1)))

adj_forest_ha <- Atot * p_forest
ci_ha         <- Atot * 1.96 * se_p
round(c(mapped_forest_ha  = A["forest"],
        adjusted_forest_ha = adj_forest_ha,
        ci_halfwidth_ha    = ci_ha), 0)
```

The adjusted area differs from the mapped area because the map commits and omits forest at rates the reference sample reveals, and the confidence interval is the honest precision of the figure. A project that reports only the mapped area, with no adjustment and no interval, has reported a number no auditor can accept. The full Bolivia workflow applies this class by class across the national legend; the estimator is the one above.

## Reported against adjusted

One year in isolation understates what this correction does. A crediting period runs for a decade, the adjustment is applied every year, and the interesting quantity is the gap between what a project reports and what it can defend. We run the same estimator across an eight-year record of mapped forest loss. Each year has its own reference sample and its own error rates, and both degrade towards the present, because recent imagery has had less interpretation effort and the map is harder to validate close to the reporting date.

```{r}
years  <- 2017:2024
Atot   <- 250000                              # ha in the accounting area
mapped <- c(1200, 2400, 4100, 3300, 3500, 3700, 6900, 9200)   # ha mapped as loss

# reference sample sizes thin out in recent years
n_chg  <- c(150, 150, 150, 150, 140, 130,  90,  60)
n_stab <- c(900, 900, 880, 880, 850, 800, 520, 340)
# user's accuracy of the loss class, and omission from the stable stratum
ua_chg  <- c(0.79, 0.80, 0.78, 0.80, 0.79, 0.77, 0.72, 0.68)
om_stab <- c(0.0022, 0.0022, 0.0026, 0.0022, 0.0026, 0.0026, 0.0030, 0.0034)

est <- t(sapply(seq_along(years), function(k) {
  W <- unname(c(mapped[k], Atot - mapped[k]) / Atot)
  p <- W[1] * ua_chg[k] + W[2] * om_stab[k]
  se <- sqrt(W[1]^2 * ua_chg[k]  * (1 - ua_chg[k])  / (n_chg[k]  - 1) +
             W[2]^2 * om_stab[k] * (1 - om_stab[k]) / (n_stab[k] - 1))
  c(adjusted = Atot * p, ci = 1.96 * Atot * se)
}))

adj <- est[, "adjusted"]; ci <- est[, "ci"]
data.frame(year = years, mapped = mapped, adjusted = round(adj),
           lower = round(adj - ci), upper = round(adj + ci),
           bias_pct = round(100 * (mapped - adj) / adj, 1))
```

Plotting the two series together with the interval around the adjusted estimate is the single most useful figure a monitoring report can carry:

```{r}
plot(years, mapped, type = "n", ylim = c(0, max(adj + ci) * 1.05),
     xlab = "", ylab = "forest loss (ha)",
     main = "Mapped against area-adjusted loss")
polygon(c(years, rev(years)), c(adj - ci, rev(adj + ci)),
        col = adjustcolor("#c4703f", alpha.f = 0.28), border = NA)
lines(years, adj, col = "#a8481f", lwd = 2.2, lty = 2)
lines(years, mapped, col = "#2f5d4e", lwd = 2.2)
legend("topleft", c("mapped (reported)", "area-adjusted", "95% interval"),
       col = c("#2f5d4e", "#a8481f", adjustcolor("#c4703f", alpha.f = 0.5)),
       lwd = c(2.2, 2.2, 8), lty = c(1, 2, 1), bty = "n", cex = 0.85)
```

Read the two lines against each other. Early in the record the adjustment pushes the estimate *up*: the mapped loss area is small, so omitted change scattered through the vast stable stratum outweighs anything the map over-called, and the defensible figure exceeds the reported one. Late in the record the sign reverses. As the map's user accuracy falls from 0.79 to 0.68, commission error dominates, and the reported figure overstates what the reference data support by thirty per cent in the final year.

The direction of the correction is therefore not something you can guess, and the common assumption that adjustment always shrinks a claim is wrong. It depends on the balance between what the map missed and what it invented, weighted by stratum area, and only a reference sample can tell you which way it falls.

```{r}
round(c(mapped_total   = sum(mapped),
        adjusted_total = sum(adj),
        overstatement_pct = 100 * (sum(mapped) - sum(adj)) / sum(adj),
        conservative_total = sum(adj - ci),
        forgone_pct = 100 * sum(ci) / sum(adj)), 1)
```

Over the whole period the reported total overstates the defensible total by around twelve per cent. That gap is the project's exposure: it is the portion of the claim that would not survive a reference-based check. The conservative basis, summing the lower bounds, is lower again by about a quarter, and that second gap is the price of imprecision rather than of bias.

Notice which of the two is easier to fix. The bias is corrected by the estimator itself, at the cost of running a reference sample, and once corrected it is gone. The interval width is set by how many reference points sit in each stratum, and it narrows only as the square root of that effort. The widening band in the last two years is not a modelling failure; it is a direct readout of thin reference data, and the only remedy is more interpretation. This is the whole argument of the chapter in one picture. Uncertainty is not a statistical afterthought appended to a result. It is a budget, it has line items, and each line has a known price.

## Exercises

1. Recompute the sampling confidence interval using a 4-by-6 quadrat grid instead of 8-by-12. Does the mean change? Does the standard error? Explain the difference.
2. In the Monte Carlo, double the uncertainty on the carbon fraction (sd 0.04) and rerun. How much does the uncertainty percentage grow, and would it now trigger a 15% deduction?
3. Report the conservative (lower-bound) carbon density and express the forgone credit as a percentage of the mean. In one sentence, state the incentive this creates.
4. *(Applied.)* Change the error matrix so the map omits more forest (change the second row to 25, 75). Recompute the adjusted forest area and its interval, and explain which way omission error pushes the adjustment.
