---
title: "Cartography"
format:
  docx:
    reference-doc: ./references/style.docx
    highlight-style: github
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Maps as evidence

A map in an audit report is not decoration; it is evidence. A well-made figure lets a reviewer see at a glance where the carbon is, where the plots fell, and whether the sampling covered the claim. A badly made one hides exactly those things. This chapter builds publication-quality maps two ways: with `tmap`, which is purpose-built for thematic cartography, and with `ggplot2`, which fits naturally if you already draw statistical graphics.

```{r}
library(sf)
library(dplyr)
library(tmap)

plots <- st_as_sf(read.csv("data/forest_plots.csv"),
                  coords = c("lon", "lat"), crs = 4326) |>
  st_transform(3005)

make_block <- function(xmin, xmax, name) {
  p <- st_polygon(list(rbind(
    c(xmin, 49.24), c(xmax, 49.24), c(xmax, 49.34),
    c(xmin, 49.34), c(xmin, 49.24))))
  st_sf(block = name, geometry = st_sfc(p, crs = 4326))
}
blocks <- rbind(make_block(-124.78, -124.71, "West"),
                make_block(-124.71, -124.65, "Central"),
                make_block(-124.65, -124.58, "East")) |> st_transform(3005)
```

## tmap grammar

`tmap` builds a map in layers, much as `ggplot2` builds a plot. Each layer starts with `tm_shape()` to name the data, followed by a drawing function: `tm_polygons()`, `tm_dots()`, `tm_lines()`, `tm_raster()`. Layers stack with `+`. The simplest useful map draws the blocks and drops the plots on top:

```{r}
tmap_mode("plot")
tm_shape(blocks) +
  tm_polygons(fill = "grey95", col = "grey40") +
tm_shape(plots) +
  tm_dots(size = 0.3)
```

## Graduated symbols

The point of a thematic map is to encode a variable. To show carbon density we colour the plots by the `carbon` column and let `tmap` build a legend. A sequential green palette reads naturally for a "more forest" variable:

```{r}
tm_shape(blocks) +
  tm_polygons(fill = "grey95", col = "grey40") +
tm_shape(plots) +
  tm_dots(fill = "carbon",
          fill.scale = tm_scale_intervals(style = "quantile", values = "brewer.greens"),
          fill.legend = tm_legend(title = "Carbon (MgC/ha)"),
          size = 0.4)
```

Classifying by quantile puts an equal number of plots in each colour class, which is a defensible default because it uses the whole palette regardless of the distribution's shape. Other schemes (`"equal"`, `"jenks"`) tell different stories from the same data, and choosing among them is a cartographic decision you should be able to justify.

## Finishing maps

A figure headed for a report needs the furniture that makes it legible on its own: a title, a scale bar, and a north arrow. These are layers too.

```{r}
tm_shape(blocks) +
  tm_polygons(fill = "grey95", col = "grey40") +
tm_shape(plots) +
  tm_dots(fill = "carbon",
          fill.scale = tm_scale_intervals(style = "quantile", values = "brewer.greens"),
          fill.legend = tm_legend(title = "Carbon\n(MgC/ha)"),
          size = 0.4) +
tm_title("Aboveground carbon, Alberni Valley plots") +
tm_scalebar(position = c("left", "bottom")) +
tm_compass(position = c("right", "top")) +
tm_layout(frame = FALSE, legend.outside = TRUE)
```

## Interactive maps

The same `tmap` code becomes an interactive, zoomable web map by switching mode. Interactive maps are invaluable in review meetings, because a stakeholder can pan to a plot and read its value directly. A scale bar and north arrow are dropped automatically in this mode.

```{r}
tmap_mode("view")
tm_shape(plots) +
  tm_dots(fill = "carbon",
          fill.scale = tm_scale_intervals(style = "quantile", values = "brewer.greens"),
          fill.legend = tm_legend(title = "Carbon (MgC/ha)"),
          size = 0.1)
```

```{r}
#| include: false
tmap_mode("plot")
```

## ggplot2 maps

If your reporting is already built on `ggplot2` [@wickham2016ggplot2], the `geom_sf()` layer draws `sf` objects directly, and the full grammar of themes, scales, and facets applies. The same carbon map in `ggplot2`:

```{r}
library(ggplot2)
ggplot() +
  geom_sf(data = blocks, fill = "grey95", colour = "grey40") +
  geom_sf(data = plots, aes(colour = carbon), size = 2) +
  scale_colour_viridis_c(option = "viridis", name = "Carbon\n(MgC/ha)") +
  labs(title = "Aboveground carbon, Alberni Valley plots") +
  theme_minimal()
```

Facetting is where `ggplot2` shines for auditing. Splitting the map by stand type shows at once whether the classes are spatially segregated, a check you would otherwise struggle to make:

```{r}
ggplot() +
  geom_sf(data = blocks, fill = "grey96", colour = "grey70") +
  geom_sf(data = plots, aes(colour = carbon), size = 1.6) +
  scale_colour_viridis_c(name = "Carbon") +
  facet_wrap(~ stand_type) +
  theme_minimal()
```

## Exercises

1. Redraw the carbon map using equal-interval and Jenks classifications instead of quantiles. How does the story change, and which would you defend in a report?
2. Make a map that shows `stand_type` as the colour of each plot and `carbon` as its size. Is a two-variable map clearer or busier than two separate maps?
3. Produce an interactive map of the plots on a satellite basemap (add `tm_basemap("Esri.WorldImagery")`). Zoom to a cluster of high-carbon plots: does the imagery corroborate them?
4. Using `ggplot2`, facet the plot map by an elevation class you create from the covariate table. Do high-carbon plots concentrate at particular elevations?
