---
title: "Areal Data"
format:
  docx:
    reference-doc: ./references/style.docx
    highlight-style: github
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Spatial dependence

Tobler's first law of geography states that everything is related to everything else, but near things are more related than distant things [@tobler1970computer]. In a forest, plots a few hundred metres apart share soil, aspect, disturbance history, and management, so their carbon densities tend to be alike. This tendency is **spatial autocorrelation**, and it has a consequence that matters directly to an audit: when nearby plots are correlated, they carry less independent information than their number suggests, so a variance calculated as if they were independent will be too small and your confidence interval too narrow. Measuring autocorrelation is therefore part of stating honest uncertainty.

```{r}
library(sf)
library(dplyr)
library(spdep)

plots <- st_as_sf(read.csv("data/forest_plots.csv"),
                  coords = c("lon", "lat"), crs = 4326) |>
  st_transform(3005)
```

## Defining neighbours

Autocorrelation is always relative to a definition of "near". In `spdep` that definition is a **neighbours list**: for each plot, which other plots count as its neighbours. A simple, robust choice is the *k* nearest neighbours. Here each plot is linked to its six closest peers, and the links are given equal weight:

```{r}
coords <- st_coordinates(plots)
knn <- knearneigh(coords, k = 6)
nb  <- knn2nb(knn)
lw  <- nb2listw(nb, style = "W")     # row-standardised weights
nb
```

The printout summarises the graph: 140 plots, each with six neighbours. Plotting the graph shows the web of relationships the statistic will use:

```{r}
plot(st_geometry(plots), pch = 19, cex = 0.4, col = "grey40")
plot(nb, coords, add = TRUE, col = "steelblue", lwd = 0.4)
```

## Global Moran's I

**Moran's I** is the standard global measure of spatial autocorrelation. It ranges from about −1 (neighbouring values systematically opposite, a checkerboard) through 0 (no spatial pattern, values scattered at random) to +1 (neighbouring values alike, smooth clusters). We test whether carbon is autocorrelated across the study area:

```{r}
moran.test(plots$carbon, lw)
```

Read the output as you would any test: the Moran's I statistic is the observed autocorrelation, the expectation under no pattern is slightly negative, and the p-value asks whether the observed value could have arisen by chance. A `moran.plot` visualises the same thing, each plot's carbon against the average of its neighbours; a positive slope is positive autocorrelation:

```{r}
moran.plot(plots$carbon, lw, xlab = "Plot carbon", ylab = "Neighbour mean carbon")
```

## Local indicators

Global Moran's I gives one number for the whole landscape, but auditors usually want to know *where* the clustering is. **Local Moran's I** decomposes the global statistic into one value per plot, identifying local clusters of high or low carbon, the hot and cold spots:

```{r}
locm <- localmoran(plots$carbon, lw)
plots$local_I <- locm[, "Ii"]
plots$local_p <- locm[, ncol(locm)]     # last column is the p-value (name varies by version)

plot(plots["local_I"], pch = 19,
     main = "Local Moran's I for carbon")
```

Cells with a high positive local I sitting among high values mark carbon-dense clusters; the same statistic flags isolated anomalies, a low-carbon plot amid high-carbon neighbours, which in an audit is exactly the kind of point worth revisiting for a field or data error.

## Exercises

1. Rebuild the neighbours list using a distance threshold (`dnearneigh`) of 1.5 km instead of six nearest neighbours. Does Moran's I for carbon change? Why might the neighbour definition matter?
2. Test for autocorrelation in `basal_area` and `stems_ha` as well as carbon. Which forest variable is most strongly clustered in space?
3. Map the local Moran's I and highlight plots that are significant at p < 0.05. Are the hot spots in particular blocks or elevations?
4. Simulate a regular (evenly spaced) point pattern and compare its mean nearest-neighbour distance to the random and clustered cases above. Where do designed inventory grids fall on this spectrum?
