---
title: "Predicting biomass wall to wall from lidar and optical metrics under spatially blocked cross-validation"
subtitle: "Chapter 9 draft, 20 September 2026"
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

::: callout-note
## Scope

Drafted 20 September 2026 on two public pairings of field plots with airborne
laser scanning, the Smithsonian Conservation Biology Institute census under
United States Geological Survey lidar, and the Quebec permanent sample plot
network under the province's open lidar. Both are committed to this repository
and need no network. Every number in the prose is computed by a chunk on this
page. The chapter begins where the plot design chapter ended, with a plot network
already designed and measured, and it ends at a fitted model and the evidence that the
model is worth trusting. Converting a diameter into mass belongs to the
allometry chapter and the error budget belongs to the chapter after it.
:::

```{r setup}
library(ranger)
library(blockCV)
library(CAST)
library(sf)
library(terra)
library(ggplot2)

set.seed(2026)

scbi <- read.csv("data/scbi_lidar_plots.csv")
chm <- rast("data/scbi_chm_2020.tif")

predictors <- c("h_mean", "h_max", "h_p95", "h_sd", "cover")
form <- as.formula(paste("ba_ha ~", paste(predictors, collapse = " + ")))

rmse <- function(observed, predicted) sqrt(mean((observed - predicted)^2))
```

## Study Data

The Smithsonian Conservation Biology Institute plot in Front Royal, Virginia,
carries the census the plot design chapter sampled, and the United States Geological
Survey flew lidar over the same ground for the VA_NShenandoah_1_2020 project.
The flight ran on 9 and 10 December 2020, which the point cloud records in its
own timestamps, and the forest is deciduous, so the laser met bare branches
rather than leaves. The census nearest that flight was measured in 2018.

Each of the census's own 20 by 20 metre quadrats is a field plot here, because
that is the support the crew worked at and the support the plot design chapter
handed forward.
For every quadrat the chapter has the basal area of the stems standing in it and
five metrics summarising the laser returns above it.

```{r scbi-summary}
n_quadrats <- nrow(scbi)
ba_mean <- mean(scbi$ba_ha)
ba_sd <- sd(scbi$ba_ha)
ba_cv <- 100 * ba_sd / ba_mean
h_range <- range(scbi$h_p95)
r_p95 <- cor(scbi$ba_ha, scbi$h_p95)
```

The `r n_quadrats` quadrats held a mean basal area of `r round(ba_mean, 1)`
square metres per hectare with a standard deviation of `r round(ba_sd, 1)`, a
coefficient of variation of `r round(ba_cv, 1)` per cent. The 95th percentile of
canopy height ran from `r round(h_range[1], 1)` to `r round(h_range[2], 1)`
metres. Its correlation with basal area was `r round(r_p95, 2)`.

```{r scbi-scatter}
#| fig-cap: "Basal area against the 95th percentile of canopy height on 640 census quadrats at the Smithsonian plot, with the fitted line."
ggplot(scbi, aes(h_p95, ba_ha)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick") +
  labs(x = "95th percentile of canopy height (m)",
       y = expression("Basal area (" * m^2 * " " * ha^-1 * ")")) +
  theme_minimal(base_size = 11)
```

## In-Sample Error

A model fitted on every quadrat and then asked to predict those same quadrats
reports the error it made on ground it has already seen.

```{r scbi-insample}
rf_all <- ranger(form, data = scbi, num.trees = 500, seed = 2026)
in_sample <- rmse(scbi$ba_ha, predict(rf_all, scbi)$predictions)
```

```{r scbi-randomcv}
folds <- sample(rep(1:10, length.out = nrow(scbi)))
pred_rf <- numeric(nrow(scbi))
pred_lm <- numeric(nrow(scbi))

for (k in 1:10) {
  train <- scbi[folds != k, ]
  test <- scbi[folds == k, ]
  pred_rf[folds == k] <- predict(ranger(form, train, num.trees = 500,
                                        seed = 2026), test)$predictions
  pred_lm[folds == k] <- predict(lm(form, train), test)
}

rmse_rf <- rmse(scbi$ba_ha, pred_rf)
rmse_lm <- rmse(scbi$ba_ha, pred_lm)
r2_rf <- 1 - rmse_rf^2 / var(scbi$ba_ha)
r2_lm <- 1 - rmse_lm^2 / var(scbi$ba_ha)
```

The random forest scored on its own training data reported a root mean square
error of `r round(in_sample, 2)` square metres per hectare. Held out of the
fitting in ten random folds, the same model reported `r round(rmse_rf, 2)`, so
the in-sample figure flattered it by a factor of
`r round(rmse_rf / in_sample, 1)`. The linear model reported
`r round(rmse_lm, 2)` against a standard deviation of `r round(ba_sd, 2)`, which
is what predicting the mean of the plots would have cost, so the explained
variance was `r round(r2_lm, 3)` for the linear model and
`r round(r2_rf, 3)` for the random forest.

Neither model has any skill, and the reason is in the flight date rather than in
the fitting. A December flight over a deciduous forest measures branches, so
canopy cover ran from `r round(min(scbi$cover), 2)` to
`r round(max(scbi$cover), 2)` where a closed summer canopy would sit near one,
and a height metric taken through leafless crowns carries little of the stem
packing that basal area counts.

The gap between `r round(in_sample, 2)` and `r round(rmse_rf, 2)` is the one a
reader is most likely to reproduce by accident, because the commonest way to
produce it is a single argument. Several widely used prediction functions take
new data through an argument named `newdata`, and passing it as `data` instead
leaves the argument unread, so the function returns the fitted values and the
reported test error is a training error wearing the name of a test.

## Spatial Cross-Validation

Random folds scatter the held-out quadrats through the plot, so almost every
test quadrat has a neighbour in the training set twenty metres away. Blocked
folds hold out whole patches of ground instead, and nearest neighbour distance
matching chooses folds so that the distances from test quadrat to training data
resemble the distances a real prediction would face.

```{r scbi-spatialcv}
scbi_sf <- st_as_sf(scbi, coords = c("x", "y"), crs = 32617, remove = FALSE)

blocks <- cv_spatial(x = scbi_sf, size = 100, k = 10, selection = "random",
                     iteration = 20, progress = FALSE, report = FALSE,
                     plot = FALSE, seed = 2026)

cv_predict <- function(fold_id) {
  pred <- numeric(nrow(scbi))
  for (k in sort(unique(fold_id))) {
    train <- scbi[fold_id != k, ]
    test_rows <- fold_id == k
    pred[test_rows] <- predict(ranger(form, train, num.trees = 500,
                                      seed = 2026), scbi[test_rows, ])$predictions
  }
  pred
}

block_id <- blocks$folds_ids
rmse_block <- rmse(scbi$ba_ha, cv_predict(block_id))
```

Blocked folds returned a root mean square error of `r round(rmse_block, 2)`
square metres per hectare against the `r round(rmse_rf, 2)` the random folds
returned, a difference of
`r round(100 * (rmse_block - rmse_rf) / rmse_rf, 1)` per cent. The gap is small
here, and that is the expected result rather than a reassuring one, because a
model with no skill has no borrowed neighbourhood information to lose. The
blocked estimate is the honest one in both cases, and the next section shows
what the two disagree by when a model does have something to lose.

## Area of Applicability

A fitted model will return a number for every cell it is handed, including cells
whose predictors sit outside anything it was trained on. The area of
applicability marks the ground where the model's cross-validated error still
describes what it is doing, by measuring how far each cell sits from the
training data in the space of the predictors, weighted by how much the model
leans on each one.

```{r scbi-aoa}
metrics <- c(aggregate(chm, 20, mean, na.rm = TRUE),
             aggregate(chm, 20, max, na.rm = TRUE),
             aggregate(chm, 20, function(x, ...) quantile(x, 0.95, na.rm = TRUE)),
             aggregate(chm, 20, sd, na.rm = TRUE),
             aggregate(chm > 2, 20, mean, na.rm = TRUE))
names(metrics) <- predictors

library(caret)
clustered <- scbi[scbi$qx < 160 & scbi$qy < 240, ]

set.seed(2026)
fit_corner <- train(clustered[, predictors], clustered$ba_ha, method = "rf",
                    importance = TRUE, ntree = 200,
                    trControl = trainControl(method = "cv", number = 5))
set.seed(2026)
fit_all <- train(scbi[, predictors], scbi$ba_ha, method = "rf",
                 importance = TRUE, ntree = 200,
                 trControl = trainControl(method = "cv", number = 5))

aoa_corner <- aoa(metrics, fit_corner, verbose = FALSE)
aoa_all <- aoa(metrics, fit_all, verbose = FALSE)

inside <- function(x) 100 * global(x$AOA, "mean", na.rm = TRUE)[[1]]
```

Building the five metrics as layers rather than as columns uses `c()` on the
aggregated rasters and names them afterwards, because naming them inside the
call returns a list of rasters rather than one raster with five layers, and the
difference only surfaces later when something asks how many layers there are.

Trained on the `r nrow(clustered)` quadrats in one corner of the plot, the model
covered `r round(inside(aoa_corner), 1)` per cent of the same plot. Trained on
all `r n_quadrats`, it covered `r round(inside(aoa_all), 1)` per cent. The
predictors here are five summaries of one canopy over 25 hectares of continuous
forest, which is about as uniform as a prediction problem gets, and
`r round(100 - inside(aoa_corner), 1)` per cent of it still sat outside the
reach of a model fitted in one corner of it.

::: callout-tip
## Statistical Agreement

Skip this box and the chapter still works. It applies the accuracy test a
crediting standard would run against these validation plots.

The American Carbon Registry's Framework for Remotely Sensed Quantification of
Forest Carbon, version 1.0 of 27 March 2026, declares a predictive model to be
in statistical agreement with its validation plots when two conditions hold at
once. Equation 4 puts the root mean square error over the mean density and
requires the result to be less than 20 per cent. Equation 5 turns the same root
mean square error into a 90 per cent confidence interval by dividing by the
square root of the plot count and multiplying by 1.645, and Equation 6 puts that
over the mean density and requires 10 per cent or less.

```{r acr-gate}
unc_rmse <- 100 * rmse_block / ba_mean
ci_pct <- 100 * (rmse_block / sqrt(n_quadrats)) * 1.645 / ba_mean

n_seq <- 2:60
forced <- min(n_seq[20 * 1.645 / sqrt(n_seq) <= 10])
```

Measured on the blocked cross-validation, the first test returned
`r round(unc_rmse, 1)` per cent against a limit of 20, so the model fails. The
second returned `r round(ci_pct, 2)` per cent against a limit of 10, so it
passes comfortably, and it passes for a reason that has nothing to do with the
model. Dividing by the square root of the plot count means that any model
sitting just inside the first test satisfies the second as soon as there are
`r forced` validation plots, since 20 per cent times 1.645 over the square root
of `r forced` is `r round(20 * 1.645 / sqrt(forced), 2)` per cent. Above that
count the two tests are not two tests. The first does all the work and the
second records how many plots were installed.
:::
