---
title: "Interpolation and Geostatistics"
format:
  docx:
    reference-doc: ./references/style.docx
    highlight-style: github
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Points to surface

An inventory measures carbon at a scatter of plots, but a landowner, a registry, or an auditor usually wants a value *everywhere*, a continuous map from which any area can be totalled. Filling the gaps between sample points with a defensible surface is **spatial interpolation**, and its most rigorous form, **kriging**, also returns a map of its own uncertainty. That second map is what distinguishes a professional carbon surface from a pretty picture: it says where the estimate is trustworthy and where the sampling was too sparse to believe.

```{r}
library(sf)
library(dplyr)
library(gstat)
library(terra)

plots <- st_as_sf(read.csv("data/forest_plots.csv"),
                  coords = c("lon", "lat"), crs = 4326) |>
  st_transform(3005)
```

## Prediction grid

Interpolation predicts onto a target grid. We build one covering the study area at 150-metre spacing, in the same metric CRS as the plots:

```{r}
grid <- st_make_grid(st_as_sfc(st_bbox(plots)),
                     cellsize = 150, what = "centers") |> st_as_sf()
template <- rast(ext(vect(grid)), resolution = 150, crs = "EPSG:3005")
nrow(grid)
```

## Inverse distance weighting

The simplest interpolator is **inverse distance weighting** (IDW): each predicted point is a weighted average of the sample plots, with nearer plots weighted more heavily. It is quick and assumption-free, but it is purely mechanical, it invents no measure of uncertainty and always smooths toward the data.

```{r}
idw_res <- idw(carbon ~ 1, plots, grid, idp = 2)
idw_r   <- rasterize(vect(idw_res), template, field = "var1.pred")

plot(idw_r, main = "IDW carbon surface (MgC/ha)")
plot(st_geometry(plots), add = TRUE, pch = 19, cex = 0.3)
```

The bullseyes around isolated high plots are IDW's signature weakness: with no model of how carbon varies in space, it cannot tell a real gradient from a lone large tree.

## The variogram

Kriging fixes this by first *learning* how similarity decays with distance, from the data themselves. The tool is the **variogram**, which plots the average squared difference between plot pairs against the distance separating them. Nearby plots differ little; distant plots differ more, until the curve flattens at the range beyond which plots are effectively independent.

```{r}
v <- variogram(carbon ~ 1, plots)
plot(v, main = "Empirical variogram of carbon")
```

We fit a model to that cloud, here a spherical model, whose three parameters have clear meanings: the **nugget** (variability at zero distance, measurement error and fine-scale noise), the **sill** (the total variance), and the **range** (the distance at which correlation vanishes).

```{r}
v_fit <- fit.variogram(v, vgm(psill = var(plots$carbon),
                              model = "Sph", range = 1500, nugget = 500))
v_fit
plot(v, v_fit, main = "Fitted spherical variogram")
```

## Ordinary kriging

With the variogram in hand, **ordinary kriging** predicts the surface as a weighted average whose weights come from the fitted spatial model rather than from raw distance. Crucially it returns two layers: the prediction and its variance.

```{r}
ok <- krige(carbon ~ 1, plots, grid, model = v_fit)
pred_r <- rasterize(vect(ok), template, field = "var1.pred")
sd_r   <- rasterize(vect(ok), template, field = "var1.var") |> sqrt()

plot(pred_r, main = "Kriged carbon (MgC/ha)")
plot(sd_r,   main = "Kriging standard error (MgC/ha)")
```

The prediction map is smoother and more physical than IDW because it respects the fitted range. The standard-error map is the one an auditor should linger on: error is lowest where plots are dense and climbs in the gaps and at the edges. It is a direct, spatial statement of where the sampling supports the claim and where it does not.

## Cross-validation

Never trust an interpolated surface without checking it. **Leave-one-out cross-validation** removes each plot in turn, predicts its value from the rest, and compares prediction to truth. The residuals tell you the surface's real accuracy:

```{r}
cv <- krige.cv(carbon ~ 1, plots, model = v_fit)
rmse <- sqrt(mean(cv$residual^2))
me   <- mean(cv$residual)
c(RMSE = round(rmse, 1), MeanError = round(me, 2))
```

A mean error near zero says the surface is unbiased; the RMSE, in the same units as carbon, is the typical error of a prediction at an unsampled location. Reported alongside the map, these two numbers turn a picture into evidence.

## Exercises

1. Refit the variogram with an exponential and a Gaussian model instead of spherical. Which fits the empirical variogram best, and how much does the choice change the kriged map?
2. Carbon is right-skewed because of a few very dense old-growth plots. Repeat the kriging on `log(carbon)` and back-transform. Does the surface or its cross-validated RMSE improve?
3. Compare the IDW and kriging surfaces cell by cell (subtract one raster from the other). Where do they disagree most, and why?
4. Coarsen the prediction grid to 500 metres and refit. How do the prediction and the standard-error map change, and what does that tell you about choosing a grid resolution for reporting?
