---
title: "Spatial Regression"
format:
  docx:
    reference-doc: ./references/style.docx
    highlight-style: github
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Spatial modelling

Kriging in the previous chapter mapped carbon from the plots' *positions* alone. But we also know something about *why* carbon varies: it responds to terrain, to elevation, slope, and ruggedness that shape moisture, exposure, and disturbance. A **spatial model** relates the field measurement to those environmental predictors, so that we can predict carbon anywhere the predictors exist, and, just as importantly, say which drivers matter. Where kriging captures pattern, modelling captures process, and the two are complementary.

The same machinery underlies species distribution and habitat-suitability modelling: fit a response to environmental covariates at sample sites, then project across a raster stack. Here the response is a continuous carbon density; the workflow is identical for a presence/absence suitability model.

```{r}
library(sf)
library(dplyr)
library(terra)
library(elevatr)

plots <- st_as_sf(read.csv("data/forest_plots.csv"),
                  coords = c("lon", "lat"), crs = 4326) |>
  st_transform(3005)
```

## Model data

Modelling needs a table of the response and its predictors at each plot, exactly the covariate extraction from the raster chapter. We rebuild the terrain covariate stack and extract it to the plots.

```{r}
aoi <- st_as_sfc(st_bbox(c(xmin = -124.78, xmax = -124.58,
                           ymin = 49.24,  ymax = 49.34),
                         crs = st_crs(4326))) |> st_sf()
dem <- get_elev_raster(aoi, z = 11, clip = "locations") |> rast() |>
  project("EPSG:3005")
names(dem) <- "elevation"

covariates <- c(dem,
                terrain(dem, "slope", unit = "degrees"),
                terrain(dem, "TRI"))
names(covariates) <- c("elevation", "slope", "tri")

dat <- plots |>
  bind_cols(terra::extract(covariates, vect(plots), ID = FALSE)) |>
  st_drop_geometry()
head(dat)
```

## Model fitting

We start with a linear model, because an auditor should reach for the simplest thing that works and be able to read every coefficient. Carbon is modelled as a function of the three terrain predictors:

```{r}
m <- lm(carbon ~ elevation + slope + tri, data = dat)
summary(m)
```

Read the coefficients as effects: the sign and size of each tells you how carbon shifts per unit of the predictor, holding the others constant, and the p-values flag which effects are distinguishable from noise. Before trusting them, check that the predictors are not collinear, since correlated predictors make coefficients unstable:

```{r}
car_ok <- try(car::vif(m), silent = TRUE)
if (!inherits(car_ok, "try-error")) car_ok
```

A quick look at the residuals confirms the model's assumptions, no strong pattern against fitted values, roughly normal residuals:

```{r}
par(mfrow = c(1, 2)); plot(m, which = c(1, 2)); par(mfrow = c(1, 1))
```

## Prediction

The payoff of a model is prediction. `terra::predict()` applies the fitted model to the covariate stack, cell by cell, producing a wall-to-wall carbon surface whose layer names match the model's predictors:

```{r}
carbon_pred <- terra::predict(covariates, m)
plot(carbon_pred, main = "Model-predicted carbon (MgC/ha)")
plot(st_geometry(plots), add = TRUE, pch = 19, cex = 0.3)
```

Because this surface is built from terrain, it shows the imprint of the landscape: carbon following the valley's form rather than smoothly averaging between plots as kriging did. Comparing the two surfaces is instructive, agreement builds confidence, and disagreement flags either a missing predictor or genuine spatial structure the terrain does not explain.

## Model validation

A model must be judged on data it did not see. Ten-fold cross-validation splits the plots into ten groups, trains on nine and predicts the tenth, and repeats, yielding an honest estimate of predictive error:

```{r}
set.seed(1)
folds <- sample(rep(1:10, length.out = nrow(dat)))
preds <- numeric(nrow(dat))
for (k in 1:10) {
  fit <- lm(carbon ~ elevation + slope + tri, data = dat[folds != k, ])
  preds[folds == k] <- predict(fit, dat[folds == k, ])
}
rmse <- sqrt(mean((dat$carbon - preds)^2))
r2   <- cor(dat$carbon, preds)^2
c(RMSE = round(rmse, 1), R2 = round(r2, 2))
```

The cross-validated RMSE is the typical error of a carbon prediction at an unsurveyed location, and the R² is the share of variance the terrain explains. On terrain alone that share is modest, which is the honest and expected result: elevation and slope set the stage, but stand history and species do much of the work. This is the cue to add better predictors, a vegetation index or a canopy-height model from LiDAR, which slot into exactly this workflow as additional layers in the covariate stack.

## Regression kriging

The strongest carbon surfaces use both chapters at once. **Regression kriging** fits the terrain model, then krige-interpolates its residuals to recover the spatial structure the model missed, and adds the two together. It marries the process-based model of this chapter with the pattern-based interpolation of the last, and is the method of choice when both a driver and residual clustering are present. Building it is a natural extension once both pieces are in hand.

## Spatial regression

Ordinary regression assumes independent residuals, but spatial data rarely oblige: neighbouring locations share unmeasured drivers, so the residuals themselves are autocorrelated. Ignoring that violates the model's assumptions and understates the uncertainty of the coefficients. The check is to test the residuals for autocorrelation, Moran's I on the residuals, using the neighbour tools from the areal-data chapter; if they cluster, a **spatial regression** model is called for.

Two classical forms account for it. A **spatial lag** model adds a term for the neighbouring values of the response, capturing spillover; a **spatial error** model places the autocorrelation in the error term, for when it reflects unobserved covariates rather than a real process. The `spatialreg` package fits both on a neighbour list:

```{r}
#| eval: false
library(spdep); library(spatialreg)

lw <- nb2listw(knn2nb(knearneigh(st_coordinates(plots), k = 6)))
lm.morantest(m, lw)                                       # residual autocorrelation?

lag <- lagsarlm(carbon ~ elevation + slope + tri, data = dat, listw = lw)   # spillover
err <- errorsarlm(carbon ~ elevation + slope + tri, data = dat, listw = lw) # correlated errors
```

The choice between them rests on whether the spillover lives in the process (lag) or only in unobserved covariates (error). Either way the discipline is the one this book returns to everywhere: test the assumption, and if it fails, use a model that respects the spatial structure rather than one that pretends it away.

## Exercises

1. Add `aspect` (transformed to "northness" as `cos(aspect × π/180)`) to the model. Does it improve the cross-validated RMSE? Why might aspect matter on Vancouver Island?
2. Fit a random forest (`ranger` or `randomForest`) with the same predictors and compare its cross-validated RMSE to the linear model. What do you gain, and what interpretability do you lose?
3. Test the linear model's residuals for spatial autocorrelation with `lm.morantest()`. If they are autocorrelated, fit a spatial lag model and compare its coefficients to the ordinary fit.
4. *(Applied.)* An enhanced forest inventory fuses field plots, LiDAR canopy metrics, and existing inventory layers into one predictive model, often an ensemble of several learners (`caretEnsemble`). Sketch, in code comments, how you would extend this chapter's model into such an ensemble and which covariate layers from earlier chapters would feed it.
