---
title: "Capstone"
format:
  docx:
    reference-doc: ./references/style.docx
    highlight-style: github
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Putting it together

Every technique in this course now serves a single deliverable: a landscape carbon estimate with defensible uncertainty. This capstone walks the whole chain, from individual tree measurements to a reported total with a confidence interval, exactly the sequence an auditor must be able to reproduce and defend. Nothing here is new machinery; it is the earlier chapters assembled into one workflow.

```{r}
library(sf)
library(dplyr)

trees <- read.csv("data/tree_measurements.csv")
plots <- read.csv("data/forest_plots.csv")
```

## Trees to plots

The inventory measured individual trees; carbon is derived, not measured. Each tree's aboveground biomass comes from an allometric equation applied to its diameter, informed by the Canadian national biomass equations [@ung2008allometric]; carbon is biomass times the IPCC default fraction of 0.47 [@ipcc2006gl]. We rebuild plot carbon from the tree table to make the chain explicit, then confirm it matches the plot summary.

```{r}
plot_carbon <- trees |>
  mutate(carbon_kg = agb_kg * 0.47) |>
  group_by(plot_id) |>
  summarise(carbon_rebuilt = sum(carbon_kg) / 1000 / 0.1)   # 0.1 ha plots

check <- plot_carbon |>
  left_join(select(plots, plot_id, reported = carbon), by = "plot_id")
head(check)
summary(check$carbon_rebuilt - check$reported)   # ~0: the chain reproduces
```

The residuals are essentially zero: our reconstruction from trees reproduces the reported plot carbon, which is the first thing an audit confirms, that the numbers trace back to the measurements.

## Plots to landscape

With carbon at each plot, the population estimate is design-based: the mean plot density estimates the mean density of the whole area, and multiplying by the area gives the total stock. We take the study area as the inventory population; in a real audit this is the certified project boundary.

```{r}
area_ha <- 12000        # forested population area, hectares (project boundary)

n        <- nrow(plots)
mean_c   <- mean(plots$carbon)
sd_c     <- sd(plots$carbon)
se_c     <- sd_c / sqrt(n)

total_MgC <- mean_c * area_ha

cat("Plots:                 ", n, "\n")
cat("Mean carbon density:   ", round(mean_c, 1), "MgC/ha\n")
cat("Standard error:        ", round(se_c, 1),  "MgC/ha\n")
cat("Total stock estimate:  ", format(round(total_MgC), big.mark = ","), "MgC\n")
```

## Reporting uncertainty

A total without an interval is not an audit result. The sampling uncertainty of the mean gives a classical confidence interval. Carbon programmes such as ART-TREES work at the 90% level, so we use a *z* of 1.645:

```{r}
z  <- 1.645
ci <- c(lower = mean_c - z * se_c, upper = mean_c + z * se_c) * area_ha
half_width_pct <- z * se_c / mean_c * 100

cat("90% CI for total stock: ",
    format(round(ci["lower"]), big.mark = ","), "to",
    format(round(ci["upper"]), big.mark = ","), "MgC\n")
cat("Uncertainty (half-width): ", round(half_width_pct, 1), "% of the mean\n")
```

Two cautions carry over from earlier chapters. First, this interval assumes the plots are independent; the autocorrelation chapter showed they may not be, in which case the effective sample size is smaller and the true interval wider. Second, this captures only *sampling* uncertainty. It ignores the error in the allometric model itself, which for biomass is often the larger term.

## Monte Carlo

The honest total combines both error sources, and the clean way to do that is simulation. We draw the mean density from its sampling distribution, apply a multiplicative allometric-model error (here a coefficient of variation of 8%, a realistic figure for aboveground biomass equations), and repeat many thousands of times to build the full distribution of the total.

```{r}
set.seed(42)
n_sim <- 10000
allometric_cv <- 0.08

sim_density   <- rnorm(n_sim, mean_c, se_c)                 # sampling error
sim_allometry <- rnorm(n_sim, 1, allometric_cv)            # model error
sim_total     <- sim_density * sim_allometry * area_ha

q <- quantile(sim_total, c(0.05, 0.5, 0.95))
hist(sim_total / 1e6, breaks = 40, col = "darkseagreen",
     main = "Simulated total carbon stock",
     xlab = "Total stock (million MgC)")
abline(v = q / 1e6, lty = c(2, 1, 2), lwd = 2)

cat("Median total:        ", format(round(q[2]), big.mark = ","), "MgC\n")
cat("90% interval:        ",
    format(round(q[1]), big.mark = ","), "to",
    format(round(q[3]), big.mark = ","), "MgC\n")
cat("Combined uncertainty:", round((q[3] - q[1]) / 2 / q[2] * 100, 1), "% of median\n")
```

The Monte Carlo interval is wider than the sampling-only one, because it carries the allometric error the classical formula omitted. That widening is not a defect; it is the difference between an uncertainty an auditor can defend and one that will not survive scrutiny.

## Reproducible audit

The reason this course insisted on scripts from the first chapter is now plain. Every number above, plot carbon, the landscape total, both intervals, was produced by code from open data, and any reviewer can re-run it, change an assumption, and see the effect. That is what reproducibility buys an auditor: not just an answer, but an answer that can be defended, revised, and trusted.

A complete inventory report assembles the pieces this book has built: a study-area map and sampling design (the vector, geometry, and cartography chapters); the plot and tree data with their processing (the setup and raster-extraction chapters); a landscape carbon surface with its own uncertainty (the interpolation and modelling chapters); a monitoring assessment of change since the last period (the time-series chapter); and this quantified total with a stated, propagated uncertainty. Rendered as a single Quarto document, it is at once the analysis and the report.

## Exercises

1. Recompute the total stock and its 90% interval using only the old-growth plots, then only the second-growth plots. How would a stratified estimate, combining the strata by area, differ from the single pooled estimate here?
2. Raise the allometric CV from 8% to 15% and re-run the Monte Carlo. How much does the combined uncertainty widen, and what does that imply about where to invest in better data?
3. The interval assumes independent plots. Using the Moran's I result from the spatial-autocorrelation chapter, argue qualitatively whether the true interval should be wider, and by roughly how much.
4. Turn this chapter into a one-page reproducible report: a title, a study-area map, the headline total with its interval, and a single monitoring map from the time-series chapter. What is the minimum an auditor must document for the result to be defensible?
