---
title: "Spatial Patterns"
aliases:
  - 09-spatial-patterns.html
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Point patterns

Every earlier chapter analysed values measured at locations: carbon at a plot, elevation at a pixel, an index through time. Point-pattern analysis asks a different question, about the locations themselves. Are the events, tree stems, seedlings, detected clearings, spread evenly, scattered at random, or clustered together? The answer characterises the process that generated them, and it is the second capstone of this course: a whole branch of spatial statistics with its own tools. We use `spatstat`, whose `ppp` object holds a point pattern inside an observation window.

The running example here is a real regeneration survey: conifer seedlings mapped after the 2015 Mt Midgeley wildfire in the Darkwoods Conservation Area of the southern Selkirk Mountains, British Columbia [@murphy2026selkirk]. We load the seedlings and build a `ppp` from one four-hectare plot, its bounding box serving as the window:

```{r}
library(spatstat)

seed <- read.csv("data/darkwoods_seedlings.csv")
p3   <- subset(seed, plot == 3)
win  <- owin(range(p3$x), range(p3$y))
pp   <- ppp(p3$x, p3$y, window = win)
pp
```

The printout reports the number of points and the window: a little over 1,600 seedlings in a 200 by 200 metre plot. That is the pattern whose structure we now interrogate.

## Spatial randomness

The reference against which every pattern is judged is complete spatial randomness (CSR): points falling independently and with equal probability everywhere, a homogeneous Poisson process. Real patterns depart from CSR in one of two directions, toward regularity, where points avoid each other as in a planted stand, or toward clustering, where points attract as in natural regeneration around surviving seed trees. The whole toolkit exists to measure that departure and test whether it is real. A plot of the seedlings already suggests the answer:

```{r}
plot(pp, main = "Mapped conifer seedlings, plot 3", pch = 20, cex = 0.3, cols = "forestgreen")
```

## Species maps

Before any statistic, plot the pattern by species across all three plots. A
composition map is the cheapest check on a survey and the fastest way to see
that the plots are not replicates of one another:

```{r}
#| fig-cap: "Mapped seedlings by species across the three surveyed plots."
#| fig-width: 10
#| fig-height: 3.6
spp_levels <- names(sort(table(seed$species), decreasing = TRUE))
spp_cols   <- setNames(
  c("#E69F00", "#009E73", "#8B4513", "#9370DB",
    "#D55E00", "#000000", "#0072B2", "#56B4E9")[seq_along(spp_levels)],
  spp_levels)

par(mfrow = c(1, 3), mar = c(2, 2, 3, 1))
for (p in sort(unique(seed$plot))) {
  s <- subset(seed, plot == p)
  plot(s$x, s$y, asp = 1, pch = 20, cex = 0.5, col = spp_cols[s$species],
       xlab = "", ylab = "", main = paste("Plot", p))
  if (p == 1) legend("topright", legend = spp_levels, col = spp_cols,
                     pch = 20, bty = "n", cex = 0.7)
}
par(mfrow = c(1, 1))
```

The colours are keyed by name, not by position. That is worth stating because the
positional alternative, building a palette and a legend independently and
trusting them to line up, is a standing source of published figures whose legend
contradicts the plot. Sorting the levels by abundance and naming the palette
makes the two impossible to desynchronise.

The three panels are not interchangeable. Lodgepole pine dominates everywhere,
but the minor species are distributed differently between plots, and the plots
themselves sit at different elevations kilometres apart. Any analysis that
pooled them without carrying plot identity would be averaging over three
different regeneration environments.

```{r}
round(prop.table(table(seed$plot, seed$species), 1), 3)
```

## Intensity

The intensity is the first-order property: the expected number of points per unit area, and how it varies across the window. A kernel density estimate smooths the points into a continuous intensity surface, the point-pattern analogue of a heat map, revealing where regeneration concentrated:

```{r}
#| fig-cap: "Kernel intensity of seedlings on plot 3, with contours, and the same surface in perspective."
#| fig-width: 10
#| fig-height: 4.6
dens <- density(pp, sigma = 8, kernel = "epanechnikov", diggle = TRUE)

par(mfrow = c(1, 2), mar = c(2, 2, 3, 4))
plot(dens, main = "Intensity (stems per square metre)", las = 1)
contour(dens, add = TRUE, col = "grey20", lwd = 0.5)
points(pp, pch = 20, cex = 0.15)
persp(dens, theta = 70, phi = 25, shade = 0.4, border = NA,
      main = "Same surface, perspective")
par(mfrow = c(1, 1))
```

Two arguments in that call are doing real work. The Epanechnikov kernel is the
minimum-variance choice and matters little in practice, but Diggle's edge
correction matters a great deal: without it the estimate falls away
artificially near the window boundary, because points outside the plot that
would have contributed were never mapped. On a 200 metre plot smoothed at 8
metres the affected band is a few per cent of the area, and it is precisely the
band where a naive reading would conclude that regeneration thins toward the
plot edge.

The contours and the perspective view show the same surface twice on purpose.
Contours support measurement, letting a reader take a value off the map.
Perspective supports recognition, making the number and separation of the peaks
immediately legible. Neither alone is sufficient for a figure whose point is that
the surface has structure.

Regeneration is patchy, not uniform: dense pockets sit near surviving seed sources and gentler, wetter ground, while other areas stay bare. That spatial unevenness is the signal, and explaining it is what the disturbance and terrain covariates in the study were for.

## Nearest neighbour

Whether points cluster or repel is the second-order property, about the interaction between them. The most direct diagnostic is the nearest-neighbour distance: how far, on average, each seedling sits from its closest neighbour, compared with what CSR predicts. Under CSR the expected mean nearest-neighbour distance is $1/(2\sqrt{\lambda})$, so the ratio of observed to expected is the Clark-Evans index: below one means clustered, above one means regular.

```{r}
ce_ratio <- function(df) {
  q <- ppp(df$x, df$y, window = owin(range(df$x), range(df$y)))
  obs <- mean(nndist(q))
  expected <- 1 / (2 * sqrt(intensity(q)))
  obs / expected
}
round(ce_ratio(p3), 3)          # < 1 indicates clustering
```

The ratio sits well below one: seedlings are closer to one another than randomness would allow, the quantitative signature of clustered regeneration.

## Ripley's K

A single number cannot capture how clustering changes with scale, points may cluster tightly at short range yet disperse at long range. Ripley's K function describes the pattern across all distances at once, counting on average how many other points lie within a growing radius of each point, against the CSR expectation. Plotting the observed K with a simulation envelope is the standard test:

```{r}
K <- envelope(pp, Kest, nsim = 19, verbose = FALSE)
plot(K, main = "Ripley's K with CSR envelope")
```

Where the observed curve rises above the envelope the pattern is more clustered than random at that distance; below it, more regular. The seedlings sit above the envelope across the mapped range, confirming clustering is real and not an artefact of a few tight groups, and the scale-explicit view is why K, and its variance-stabilised form the L function, is the workhorse of the field.

## Seedling pattern

The published study runs exactly this arc, then goes further [@murphy2026selkirk]. Across 3,886 mapped seedlings on three four-hectare plots, point-pattern analysis found significant clustering for all seven conifer species, with intensity governed by fire severity, distance to seed source, topography, and wind. The lean diagnostics above reproduce its core finding on one plot: clustered regeneration, quantified by the Clark-Evans ratio and confirmed by Ripley's K. Repeating the Clark-Evans ratio species by species shows the clustering is not confined to the dominant lodgepole pine but holds across the community:

```{r}
spp <- c("Pinus_contorta", "Larix", "Picea", "Pseudotsuga")
sapply(spp, function(s) {
  q <- subset(p3, species == s)
  if (nrow(q) < 20) return(NA)
  round(ce_ratio(q), 3)
})
```

Every species returns a ratio below one. A ratio is one number, though, and the
scale-explicit version is the per-species envelope. Simultaneous, or global,
envelopes are used rather than pointwise ones, because a pointwise band is
exceeded somewhere by chance in most realisations and invites reading
significance off any excursion:

```{r}
#| fig-cap: "Inhomogeneous K with simultaneous envelopes, by species, on plot 3."
#| fig-width: 10
#| fig-height: 3.4
set.seed(2026)
par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
for (s in c("Pinus_contorta", "Larix", "Picea")) {
  q  <- subset(p3, species == s)
  qp <- ppp(q$x, q$y, window = win)
  e  <- envelope(qp, Kinhom, nsim = 39, correction = "best",
                 global = TRUE, verbose = FALSE)
  plot(e, main = sub("_", " ", s), legend = FALSE)
}
par(mfrow = c(1, 1))
```

The inhomogeneous form divides out a varying intensity, so what remains above the
envelope is clustering over and above the trend. Reading these needs one caution,
and it is the reason the residual testing section later in this chapter exists:
the intensity is estimated from the same pattern being tested, so the estimate
absorbs part of the clustering and the test is conservative. An excursion above
the envelope here is therefore trustworthy; the absence of one is not evidence of
no clustering.

That settles the description. It does not explain anything, and description is where most spatial analysis stops. The rest of this chapter is the step past it.

## Pattern to process

Clark-Evans and Ripley's K both compare the observed pattern against complete spatial randomness. Rejecting CSR is easy and, on its own, close to uninformative: almost no ecological pattern is homogeneous Poisson, so a significant test tells you mainly that you had enough points to detect the obvious. The useful question is different. Given that intensity varies, what does it vary *with*, and is there clustering left over once you account for that?

Answering it requires a model. Two, in fact, layered.

The first-order model says the expected intensity at a location is a function of covariates measured there. Written on the log scale so intensity stays positive, it is a Poisson point process model, `ppm`, which behaves in almost every respect like a generalised linear model whose response is the pattern itself:

$$\log \lambda(u) = \beta_0 + \beta_1 z_1(u) + \cdots + \beta_k z_k(u)$$

The second-order model says the points are not independent even after the covariates are accounted for. Seedlings arrive in groups because a surviving cone-bearing tree drops a cohort around itself. A Thomas cluster process makes that mechanism explicit: parent locations follow a Poisson process of intensity $\kappa$, and each parent scatters offspring around itself with a Gaussian kernel of standard deviation $\sigma$. Fitting both layers together is what `kppm` does, and the resulting model is a Cox process: Poisson conditional on a random intensity field [@baddeley2015spatstat].

Both layers matter, but for different reasons. The first tells you what drives regeneration. The second tells you how much to believe the first.

## Covariates

The mapped seedlings carry the covariates the study extracted at each stem: elevation, slope, topographic wetness, distance to the nearest surviving seed source, and the two disturbance histories. Beetle attack is recorded in three stages, and fire in four severity classes:

```{r}
seed$mpb  <- factor(seed$mpb_class,  0:2, c("free", "grey", "red"))
seed$fire <- factor(seed$fire_class, 1:4, c("free", "low", "med", "high"))
round(prop.table(table(seed$mpb)), 3)
round(prop.table(table(seed$fire)), 3)
```

Grey-stage attack, where needles have fallen and crowns collapsed, covers under four per cent of the seedling records. Red-stage, the recent attack that still holds dead red needles, covers a quarter. Those proportions match the published study's plot summaries, which is the first check worth running on any dataset you did not extract yourself.

The three plots are separate 200 metre squares kilometres apart, so the observation window is their union, not one bounding box that would count the empty ground between them as surveyed. We buffer each plot by half a metre so no mapped seedling falls exactly on a boundary and gets silently dropped:

```{r}
mkwin  <- function(s) owin(range(s$x) + c(-0.5, 0.5), range(s$y) + c(-0.5, 0.5))
plots  <- split(seed, seed$plot)
pooled <- do.call(superimpose,
                  unname(lapply(plots, function(s) ppp(s$x, s$y, window = mkwin(s)))))
marks(pooled) <- NULL

c(points = npoints(pooled), hectares = round(area.owin(Window(pooled)) / 1e4, 2))
```

A `ppm` needs covariates as continuous surfaces, not as values attached to points, because the model must evaluate intensity everywhere in the window and not only where a seedling happens to be. The study built those surfaces from 30 metre rasters. Those rasters are not distributed with the seedling data, so we reconstruct each surface by assigning every location the value recorded at its nearest mapped seedling:

```{r}
sd_   <- do.call(rbind, plots)          # rows in the same order as the pooled pattern
as_im <- function(v) nnmark(ppp(sd_$x, sd_$y, window = Window(pooled), marks = v),
                            dimyx = 256)
covs <- list(
  elevation = as_im(sd_$elevation),
  slope     = as_im(sd_$slope),
  twi       = as_im(sd_$twi),
  distance  = as_im(sd_$distance_m),
  mpb_grey  = as_im(as.numeric(sd_$mpb == "grey")),
  mpb_red   = as_im(as.numeric(sd_$mpb == "red")),
  fire_high = as_im(as.numeric(sd_$fire == "high")),
  fire_low  = as_im(as.numeric(sd_$fire == "low")),
  fire_med  = as_im(as.numeric(sd_$fire == "med")))

plot(covs$distance, main = "Distance to nearest seed source (m)")
```

Be clear about what that reconstruction costs, because it is a real limitation and not a technicality. Nearest-neighbour interpolation from the sample points recovers a covariate faithfully where seedlings are dense and guesses where they are sparse. For a common condition the guess is harmless. For a rare one it is not: a rare class gets painted across the whole neighbourhood of the few points that carry it, which inflates its apparent extent. Grey-stage attack is exactly that rare class here, and we will see the consequence shortly. Working from the original rasters is always better. When you cannot, say so and show where it bites.

## Screening covariates

With nine candidate covariates you need a selection step, and the natural one compares each single-covariate model against an intercept-only model by likelihood ratio test. The deviance measures how much explanatory power the covariate adds:

```{r}
null   <- ppm(pooled ~ 1)
screen <- t(sapply(names(covs), function(nm) {
  m <- ppm(pooled, trend = as.formula(paste("~", nm)), covariates = covs[nm])
  a <- anova(null, m, test = "LRT")
  c(beta = unname(coef(m)[2]), deviance = a$Deviance[2], p = a$`Pr(>Chi)`[2])
}))
round(screen[order(-screen[, "deviance"]), ], 4)
```

Read the ordering, not the p-values. With nearly four thousand points almost anything clears a significance threshold, so the p-value column mostly reports sample size. The deviance column is the one that separates the covariates that matter from those that merely pass.

Distance to seed source dominates, which is the expected ecology: seed rain falls off sharply with distance from a surviving parent. Fire severity comes next, and its two coefficients point in opposite directions. Low-severity burn carries a positive coefficient and high-severity burn a strongly negative one, so light fire helps regeneration and severe fire suppresses it. Medium severity is the one covariate that fails outright, with a deviance near zero, and it is dropped. Terrain enters weakly. Elevation and wetness are significant but explain several times less than distance or fire, which is the ordering you would expect on ground where the disturbance history is recent and severe.

## Bandwidth selection

Every kernel estimate in this chapter has depended on a smoothing bandwidth chosen by hand. Choosing it by likelihood cross-validation instead removes the arbitrariness. `bw.ppl` maximises a leave-one-out likelihood over candidate bandwidths, and running it per plot shows whether the plots agree:

```{r}
bw_obj <- lapply(plots, function(s)
  bw.ppl(ppp(s$x, s$y, window = mkwin(s))))
bw <- sapply(bw_obj, as.numeric)
round(bw, 2)
```

Plotting the cross-validation criterion rather than only its optimum shows how
sharply the data identify that scale:

```{r}
#| fig-cap: "Likelihood cross-validation criterion against bandwidth for each plot. The optimum is the maximum of each curve."
#| fig-width: 10
#| fig-height: 3.4
par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
for (p in names(bw_obj)) {
  plot(bw_obj[[p]], main = paste("Plot", p), xlim = c(0, 16))
  abline(v = bw[[p]], col = "firebrick", lty = 2)
}
par(mfrow = c(1, 1))
```

All three land close together, near six and a half metres, comfortably inside the range the published study reports across plots and globally. That consistency is worth more than the number itself. Three independently surveyed plots agreeing on a scale is evidence the scale is a property of the regeneration process rather than of any one plot's layout.

The curves add a caution the single number hides. Each is flat near its
maximum, so a range of bandwidths is very nearly as well supported as the
optimum, and reporting the optimum to two decimals implies a precision the
criterion does not have. Where a curve is flat, state the range.

## Cluster process

Now fit both layers at once: the covariate trend from the screening step, and a Thomas cluster process for what the covariates do not explain.

```{r}
fit <- kppm(pooled,
            trend = ~ distance + twi + mpb_grey + mpb_red + fire_high + fire_low,
            clusters = "Thomas", covariates = covs)
round(coef(fit), 4)
```

```{r}
signif(fit$clustpar, 4)
```

The trend coefficients read as multipliers on intensity once exponentiated. Distance is negative, so intensity falls with every metre from a seed source. High-severity fire is strongly negative and low-severity fire positive, reproducing the study's central finding about severity. Red-stage beetle attack is negative, matching the published sign.

The cluster parameters describe the second layer. The fitted scale is roughly twenty metres, so offspring sit within a few tens of metres of their parent. The parent intensity $\kappa$ is about 0.00014 per square metre, which is a little under one and a half parents per hectare. Together they describe a modest number of well-separated cohorts rather than a fine-grained scatter, which is what a post-fire regeneration surface driven by scattered surviving seed trees should look like.

## Corrected standard errors

Here is the part that changes what you would conclude, and it is the reason this section exists.

A `ppm` assumes points are independent given the covariates. They are not. Seedlings arrive in sibling groups, so the pattern carries far less information than its point count suggests. Fitting an independence model to clustered data leaves the coefficients roughly unbiased but makes the standard errors far too small. The cluster model estimates its variance by a composite likelihood that accounts for the dependence [@waagepetersen2007], and comparing the two makes the size of the error visible:

```{r}
est   <- coef(fit)
naive <- sqrt(diag(vcov(as.ppm(fit))))   # Poisson, assumes independence
clust <- sqrt(diag(vcov(fit)))           # allows for the cluster process

round(cbind(estimate  = est,
            se_naive  = naive,
            se_cluster = clust,
            inflation = clust / naive,
            z_naive   = est / naive,
            z_cluster = est / clust), 3)
```

```{r}
c(significant_naive   = sum(abs(est / naive) > 1.96),
  significant_cluster = sum(abs(est / clust) > 1.96),
  terms               = length(est))
```

The correct standard errors are between six and a half and ten times the naive ones. Under the independence assumption six of the seven terms clear the conventional threshold. Under the cluster-corrected variance, one does. Every substantive conclusion you would have drawn from the naive fit evaporates.

Nothing about the data changed. The estimates are identical; only the honesty of the uncertainty changed. Because variance scales as the square of the standard error, an inflation factor of seven to ten means the naive fit treated the data as carrying fifty to a hundred times more information than it holds. Put the other way, the effective sample size is not 3,912 but a few dozen well-separated cohorts. This is the same problem that inflates confidence in any spatially autocorrelated dataset, and it is why the uncertainty chapter insists that the sampling design and the variance estimator must match.

Treat this as the general lesson. When observations cluster and your model assumes they do not, you will find significance that is not there. The failure is silent, the output looks authoritative, and the only defence is to fit a model that represents the dependence and compare.

## Residual testing

The last question is whether clustering survives the covariates. If distance to seed source and fire severity explained the pattern completely, the residual process would be Poisson and there would be nothing for the Thomas layer to do.

Testing this needs more care than it first appears, and the trap is worth seeing. The inhomogeneous K function, `Kinhom`, generalises Ripley's K by dividing out a varying intensity. If you estimate that intensity nonparametrically from the same data, the estimate absorbs the very clustering you are trying to detect, and the test loses its power or reverses sign entirely. The null must also match: simulating homogeneous CSR while computing an inhomogeneous statistic compares two different things.

The coherent test fixes both. Fit an inhomogeneous Poisson model, hold its fitted intensity fixed, and simulate from that same fitted model:

```{r}
s3    <- plots[["3"]]
win3  <- mkwin(s3)
pp3   <- ppp(s3$x, s3$y, window = win3)
c3    <- function(v) nnmark(ppp(s3$x, s3$y, window = win3, marks = v), dimyx = 256)
cv3   <- list(twi       = c3(s3$twi),
              distance  = c3(s3$distance_m),
              fire_high = c3(as.numeric(s3$fire == "high")),
              fire_low  = c3(as.numeric(s3$fire == "low")))

inhom <- ppm(pp3, trend = ~ twi + distance + fire_high + fire_low, covariates = cv3)
lam   <- predict(inhom)

set.seed(2026)
env <- envelope(inhom, Kinhom, nsim = 999, global = TRUE, verbose = FALSE,
                simulate = expression(rpoispp(lam)), lambda = lam, normpower = 2)
```

```{r}
plot(env, main = "Inhomogeneous K against the fitted trend, 999 simulations")
```

```{r}
above <- env$r[env$obs > env$hi]
round(c(radii_above = length(above), of = length(env$r),
        from_m = min(above), to_m = max(above)), 2)
```

The observed curve leaves the envelope from about seven metres outward and stays above it to the edge of the tested range. Clustering is not an artefact of the covariates. Seedlings aggregate at scales of tens of metres beyond anything distance to seed source, wetness or fire severity explains, which is precisely the residual structure the Thomas layer was fitted to absorb.

Two details in that call are the difference between a real test and a decorative one. `global = TRUE` gives a simultaneous envelope, valid to interpret across the whole range of distances at once; the default pointwise envelope is only valid at a single distance chosen in advance, and reading a pointwise band across all radii inflates the false-positive rate badly. And 999 simulations sets the finest resolvable p-value at 0.001. Nineteen simulations, enough for the illustrative envelope earlier in this chapter, cannot resolve anything below 0.05.

## Published comparison

The study fitted the same class of model on the original 30 metre rasters, with covariates this reconstruction does not have, and screened them the same way [@murphy2026selkirk]. Its likelihood-ratio results are:

| Covariate | Published $\beta$ | Published deviance | Published test |
|---|---|---|---|
| MPB-grey | -0.0233 | 0.47 | 0.492, not significant |
| MPB-red | -0.1461 | 14.85 | 0.001 |
| Fire-high | -0.6430 | 211.64 | 0.001 |
| Fire-low | 0.5682 | 276.87 | 0.001 |
| Fire-med | 0.0188 | 0.34 | 0.563, not significant |

Compare that against the screening table computed above. Fire-high, fire-low and MPB-red agree in sign and in rough magnitude, and fire-med is negligible in both. For a reconstruction working from point-extracted covariates rather than the source rasters, that is close agreement, and it is the check that gives you licence to teach from this dataset at all.

One covariate disagrees, and it disagrees loudly. Grey-stage attack comes out strongly positive here and near zero in the published model. This is the artefact promised earlier. Grey-stage is the rare class, under four per cent of records, so nearest-neighbour reconstruction paints it across whole neighbourhoods around the few seedlings that carry it, and the model reads that inflated footprint as a regeneration signal. The cluster-corrected standard error is what rescues the result: under the correct variance the grey-stage coefficient is not significant, which is the published conclusion.

That covariate has a history worth knowing, because it shows the same failure occurring in real research and being caught. When the analysis was re-run during revision, the grey-stage layer was rebuilt, and a value-threshold error entered that briefly made the covariate significant. The error was found, the layer corrected against an independent red-stage reference, and the affected models refitted. The corrected coefficient returns to non-significant, where it had been in the originally reviewed model, and the study's conclusions about red-stage attack and fire severity are unchanged by it.

Two lessons sit in that history and both outrank any coefficient. A rare binary covariate is the most fragile thing in a spatial model, because a small error in its extent is a large error in its proportion. And a result that appears only under an assumption you have not tested is not a result. Here the assumption was independence, and the test was the variance comparison in the previous section.

::: {.verification}
#### Model interrogation {.unnumbered .unlisted}

Point-process models reach verification through claims rather than through a prescribed method. A project asserting successful regeneration, adequate stocking distribution, or a disturbance attribution is making a spatial-statistical claim, and the standards require its uncertainty to be quantified honestly even where they do not name the technique.

Four questions do most of the work, in order of how often they find something.

First, what variance estimator was used, and does it allow for spatial dependence? This is the question that exposed the near order-of-magnitude error above. An analysis reporting Poisson or ordinary least squares standard errors on clustered spatial data is overstating its confidence, and by how much is not knowable without refitting.

Second, where did the covariate surfaces come from? Extracted at sample points and interpolated is not the same as measured on a grid, and rare classes suffer most. Ask for the extent of every binary covariate as an area, not as a coefficient.

Third, was the null hypothesis the right one? Rejecting complete spatial randomness demonstrates almost nothing in a landscape with any topography at all. The informative test is against a fitted trend, not against uniformity.

Fourth, is the simulation envelope global or pointwise, and how many simulations? A pointwise envelope read across a range of distances is not a valid test, and a p-value cannot be finer than one over the number of simulations plus one.

None of this requires the verifier to refit the model. Each question is answerable from a methods description, and a methods description that cannot answer them is itself the finding.
:::

## Exercises

1. Build the `ppp` for plot 1 instead of plot 3 and compute its Clark-Evans ratio. Is the regeneration there clustered, random, or regular, and how does it compare with plot 3?
2. Map the intensity surface with a larger smoothing bandwidth (`sigma = 20`) and a smaller one (`sigma = 4`). How does the choice change what you see, and what does that imply about reporting a single heat map?
3. Refit the `kppm` with `clusters = "MatClust"` instead of `"Thomas"`, which scatters offspring uniformly in a disc rather than by a Gaussian kernel. How much do the trend coefficients move, and how much do the cluster-corrected standard errors move? Which of the two layers is the more robust to the choice?
4. Drop `distance` from the trend and refit. Watch what happens to the fitted cluster scale. Explain why removing a covariate that explains large-scale variation makes the cluster process look stronger.
5. Rerun the final envelope with `global = FALSE` and count how many radii fall outside the band. Explain why that count cannot be read as a test.
6. *(Applied.)* Detected clearings in a monitoring image are also a point pattern. Explain how Ripley's K could distinguish scattered natural tree-fall from an organised, clustered harvest front, and why that distinction matters for attributing disturbance.
