---
title: "Allometry and Biomass"
format:
  docx:
    reference-doc: ./references/style.docx
    highlight-style: github
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Why allometry

A cruiser measures diameters and heights, not carbon. The step that turns a tape measurement into a tonne of CO₂ is the **allometric equation**, and it is usually the single largest source of uncertainty in a forest-carbon estimate. An auditor who cannot say which equation was used, how it was validated, and how much error it carries cannot defend the headline number. This chapter fits allometric models to reference trees, validates them the way a verifier would, converts biomass to carbon with the standard methodological factors, and propagates the whole chain's uncertainty by Monte Carlo simulation.

We work with the tree measurements behind the plot inventory. For this chapter they stand in for a set of destructively sampled reference trees: trees whose true aboveground biomass is known, against which candidate equations are tested.

```{r}
library(dplyr)
trees <- read.csv("data/tree_measurements.csv")
head(trees)
```

## Fitting models

The workhorse form relates aboveground biomass to diameter as a power law, $\text{AGB} = a \cdot D^{b}$, which is linear once both sides are logged. Fitting it on the log scale is the classical approach; fitting the power form directly by non-linear least squares avoids the back-transformation bias. We do both for Douglas-fir and compare:

```{r}
fdc <- trees |> filter(species == "Douglas-fir")

# Log-log linear model
m_log <- lm(log(agb_kg) ~ log(dbh_cm), data = fdc)

# Direct power model by non-linear least squares
m_nls <- nls(agb_kg ~ a * dbh_cm^b, data = fdc,
             start = list(a = 0.1, b = 2.4))

coef(m_nls)
```

The exponent *b* near 2.4 is the biological signal: biomass scales with roughly the square-to-cube of diameter, which is why a handful of large trees dominate a stand's carbon.

## Model validation

A verifier does not trust a fitted model; they test it. Two diagnostics matter: the **root mean squared error** (RMSE), the typical prediction error in kilograms, and the **bias** (mean residual), which reveals systematic over- or under-prediction. Systematic bias is the more dangerous of the two in carbon accounting, because it does not average out over a landscape.

```{r}
pred      <- predict(m_nls)
resid_kg  <- fdc$agb_kg - pred
rmse      <- sqrt(mean(resid_kg^2))
bias      <- mean(resid_kg)
rmse_pct  <- rmse / mean(fdc$agb_kg) * 100

c(RMSE_kg = round(rmse, 1), Bias_kg = round(bias, 2), RMSE_pct = round(rmse_pct, 1))

plot(pred, fdc$agb_kg, pch = 19, cex = 0.4,
     xlab = "Predicted AGB (kg)", ylab = "Reference AGB (kg)",
     main = "Model validation, Douglas-fir")
abline(0, 1, col = "red")
```

Points scattering evenly about the 1:1 line, with bias near zero, is the signature of a usable equation. Where the pantropical setting applies, the `BIOMASS` package implements the standard Chave equations and wood-density lookups for exactly this workflow; here, with a temperate species set, we fit local equations, which is itself the Tier-2 improvement that reduces reported uncertainty.

## Biomass to carbon

Biomass is not carbon, and aboveground biomass is not the whole tree. Two methodological constants complete the conversion, following the standard stock accounting used under VCS and IPCC guidance. The **carbon fraction** (CF, default 0.47–0.50) converts biomass to carbon; a **root-to-shoot ratio** (R) adds belowground biomass. We apply them per tree:

```{r}
CF   <- 0.47   # IPCC default carbon fraction
R    <- 0.25   # root-to-shoot ratio (belowground / aboveground)

trees <- trees |>
  mutate(agb_kg    = agb_kg,
         bgb_kg    = agb_kg * R,
         carbon_kg = (agb_kg + bgb_kg) * CF)

plot_carbon <- trees |>
  group_by(plot_id) |>
  summarise(carbon_MgC = sum(carbon_kg) / 1000, .groups = "drop")
head(plot_carbon)
```

## Monte Carlo

The single most important discipline in carbon accounting is carrying uncertainty through every step rather than reporting a point estimate. The allometric model has parameter uncertainty; the carbon fraction and root-to-shoot ratio are themselves uncertain. Monte Carlo simulation draws each uncertain quantity from its distribution many times and rebuilds the estimate, yielding the full distribution of stand carbon.

```{r}
set.seed(2025)
n_sim  <- 5000
co     <- coef(m_nls); a_hat <- co["a"]; b_hat <- co["b"]
se     <- summary(m_nls)$coefficients[, "Std. Error"]

# one representative stand of Douglas-fir stems
D <- fdc$dbh_cm

sim_carbon <- replicate(n_sim, {
  a_i  <- rnorm(1, a_hat, se["a"])
  b_i  <- rnorm(1, b_hat, se["b"])
  cf_i <- rnorm(1, 0.47, 0.02)             # carbon-fraction uncertainty
  r_i  <- rnorm(1, 0.25, 0.03)             # root-to-shoot uncertainty
  agb  <- a_i * D^b_i
  sum(agb * (1 + r_i) * cf_i) / 1000       # MgC for this stand
})

q <- quantile(sim_carbon, c(0.05, 0.5, 0.95))
hist(sim_carbon, breaks = 40, col = "darkseagreen",
     main = "Simulated stand carbon (Douglas-fir stems)",
     xlab = "Carbon (MgC)")
abline(v = q, lty = c(2, 1, 2), lwd = 2)

cat("Median:", round(q[2], 2), "MgC\n")
cat("90% interval:", round(q[1], 2), "to", round(q[3], 2), "MgC\n")
cat("Uncertainty (half-width):", round((q[3] - q[1]) / 2 / q[2] * 100, 1), "%\n")
```

The width of that interval is the honest uncertainty of the allometric step, and it is invariably wider than a naive calculation suggests. Reporting it, and showing which input drives it, is what separates a defensible carbon estimate from a fragile one. In practice a sensitivity analysis, re-running the simulation with each source of uncertainty switched off in turn, identifies where better data would most reduce the interval, and therefore where to invest.

## Exercises

1. Fit the power model for western redcedar and western hemlock as well as Douglas-fir. Compare their RMSE percentages: which species is hardest to predict, and how would that affect a mixed-stand estimate?
2. Compare the log-log linear model and the non-linear power model for one species. Do their predictions differ at the largest diameters, and why does back-transformation bias matter most there?
3. Add diameter–height as a two-predictor model (`agb_kg ~ a * dbh_cm^b * height_m^c`). Does including height reduce the RMSE enough to justify measuring it in the field?
4. Run the Monte Carlo with only the allometric parameter uncertainty, then add the carbon-fraction and root-to-shoot terms one at a time. Which contributes most to the final interval?
