---
title: "Carbon Accounting"
format:
  docx:
    reference-doc: ./references/style.docx
    highlight-style: github
execute:
  eval: true
  echo: true
  warning: false
  message: false
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Stocks and baselines

A carbon project is not paid for the carbon standing in the forest; it is paid for the carbon that would *not* be there under business as usual. Accounting therefore has two halves: measuring the **stock** on the ground, which the earlier chapters have equipped us to do, and comparing the project's trajectory against a **baseline** counterfactual. This chapter assembles a stock-based account in the style of an Improved Forest Management (IFM) methodology such as VCS VM0010, using the inventory we have carried through the book.

```{r}
library(dplyr)
plots <- read.csv("data/forest_plots.csv")
```

## Carbon pools

A tonne of tree is not a tonne of carbon, and the stem is not the whole ecosystem. Complete accounting sums several **pools**. The two that the inventory measures directly are aboveground and belowground biomass; the standard factors from IPCC guidance convert between them and to carbon [@ipcc2006gl].

```{r}
CF <- 0.47    # carbon fraction of biomass
R  <- 0.25    # root-to-shoot ratio (belowground / aboveground)

plots <- plots |>
  mutate(agc_MgCha  = carbon,                 # aboveground carbon (already MgC/ha)
         bgc_MgCha  = agc_MgCha * R,          # belowground carbon
         total_MgCha = agc_MgCha + bgc_MgCha)

plots |>
  group_by(stand_type) |>
  summarise(n = n(),
            AGC = round(mean(agc_MgCha), 1),
            BGC = round(mean(bgc_MgCha), 1),
            Total = round(mean(total_MgCha), 1))
```

Dead organic matter (deadwood and litter) and soil organic carbon are additional pools; a conservative account may exclude a pool if excluding it *understates* the project benefit, a principle of conservativeness that runs through every standard.

## Stratified stocks

Totalling carbon across a heterogeneous estate is more precise when the estate is divided into **strata** of similar forest and each stratum estimated separately. Our stands form natural strata. The stratified estimate weights each stratum's mean density by its area:

```{r}
# area of each stratum in the project (hectares) — from the tenure/type map
stratum_area <- tibble(
  stand_type = c("old-growth", "second-growth", "riparian"),
  area_ha    = c(4200, 6300, 1500)            # example project areas
)

stock <- plots |>
  group_by(stand_type) |>
  summarise(mean_density = mean(total_MgCha),
            sd_density   = sd(total_MgCha),
            n            = n(), .groups = "drop") |>
  left_join(stratum_area, by = "stand_type") |>
  mutate(stratum_stock = mean_density * area_ha)

total_stock <- sum(stock$stratum_stock)
stock
cat("Total project carbon stock:", format(round(total_stock), big.mark = ","), "MgC\n")
```

## Baseline versus project

IFM credits the carbon retained by managing more conservatively than the baseline. The simplest credible baseline is a legally permissible harvest that draws the stock down over the crediting period; the project scenario retains more of it. The **emission reduction** is the difference in stock trajectories, converted from carbon to CO₂ with the 44/12 molecular ratio.

```{r}
years        <- 0:20
baseline_rate <- -0.020   # baseline loses 2% of standing carbon per year (harvest)
project_rate  <- 0.005    # project grows 0.5% per year (deferred harvest)

baseline <- total_stock * (1 + baseline_rate)^years
project  <- total_stock * (1 + project_rate)^years

er_MgCO2 <- (project - baseline) * (44/12)     # emission reductions, tCO2e

plot(years, project,  type = "l", col = "forestgreen", lwd = 2,
     ylim = range(c(baseline, project)),
     xlab = "Year", ylab = "Carbon stock (MgC)", main = "Baseline vs project")
lines(years, baseline, col = "firebrick", lwd = 2)
legend("topleft", c("Project", "Baseline"), col = c("forestgreen", "firebrick"), lwd = 2, bty = "n")

cat("Cumulative ERs at year 20:", format(round(er_MgCO2[length(years)]), big.mark = ","), "tCO2e\n")
```

## Wood products

Not all harvested carbon is emitted at once. A share is locked into **harvested wood products** (HWP), sawn timber that stores carbon for decades. Methodologies credit the long-lived fraction, discounting the rest as emitted. A first-order treatment applies a stored fraction to the harvested volume:

```{r}
harvested_MgC <- total_stock * 0.020          # carbon harvested in year 1 (baseline rate)
hwp_stored_frac <- 0.30                        # fraction retained long-term in products
hwp_credit_MgCO2 <- harvested_MgC * hwp_stored_frac * (44/12)
cat("HWP long-term storage credit:", format(round(hwp_credit_MgCO2), big.mark = ","), "tCO2e\n")
```

## Conservative crediting

A credible account never issues its central estimate. Two deductions stand between the estimate and the credit: an **uncertainty deduction**, tied to the confidence interval quantified in the uncertainty chapters, and a **buffer** contribution to a pooled reserve against reversals such as fire. Applying an illustrative uncertainty deduction and buffer to the year-20 reductions:

```{r}
gross_er     <- er_MgCO2[length(years)]
unc_ded      <- 0.10    # uncertainty deduction (e.g., from the 90% CI half-width)
buffer       <- 0.15    # non-permanence buffer contribution
net_credits  <- gross_er * (1 - unc_ded) * (1 - buffer)

cat("Gross ERs: ", format(round(gross_er), big.mark = ","), "tCO2e\n")
cat("Net issuable credits after deductions:", format(round(net_credits), big.mark = ","), "tCO2e\n")
```

The chain from a diameter tape to an issuable credit is now complete and, because it is all code over open data, entirely reproducible. Every factor, the carbon fraction, the root-to-shoot ratio, the baseline rate, the HWP fraction, the deductions, is explicit and can be challenged, revised, and re-run, which is exactly what a verifier requires.

## Exercises

1. Recompute the total stock including a soil organic carbon pool of 80 MgC/ha applied uniformly. How much does including soil change the headline, and why do many IFM projects exclude it?
2. Vary the baseline harvest rate from 1% to 4% per year. How sensitive are the year-20 emission reductions to the baseline assumption, and what does that imply about where verification scrutiny should focus?
3. Combine this chapter with the Monte Carlo from the allometry chapter: propagate the density uncertainty into the total stock and set the uncertainty deduction from the resulting 90% interval rather than a fixed 10%.
4. Explain, in a short paragraph an auditor could paste into a report, why the buffer and uncertainty deductions are separate and what risk each addresses.
