---
title: "Verification Checks"
format:
  docx:
    reference-doc: ./references/style.docx
    highlight-style: github
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Verifying maps

Remote sensing lets a project map carbon wall to wall, but a map is a model, and a model is a claim. Monitoring, reporting and verification (MRV) frameworks, such as the ACR framework for remotely sensed quantification of forest carbon, exist to test that claim against independent field measurement before any credit is issued. This chapter builds the two checks an auditor performs most often: an accuracy assessment of a carbon map against validation plots, and a geometry-and-area quality check on the project boundary itself.

```{r}
library(sf)
library(terra)
library(dplyr)
```

## Validation sampling

Validation must be independent and unbiased, which means the plots are located by a documented random design, not chosen by eye. The standard is a random sample within the project boundary, drawn on an **equal-area** projection so that every hectare has the same chance of selection, and seeded so the draw is reproducible and auditable. Frameworks set a floor on the sample size; we use the common minimum of 45.

```{r}
# project boundary (here: the study-area rectangle; in practice the tenure polygon)
project <- st_as_sfc(st_bbox(c(xmin = -124.78, xmax = -124.58,
                               ymin = 49.24,  ymax = 49.34),
                             crs = st_crs(4326))) |> st_sf() |>
  st_make_valid()

n_points <- 45
set.seed(6933)                                   # documented, ACR-style fixed seed

# sample on an equal-area CRS, then return to lon/lat for field navigation
val_plots <- project |>
  st_transform(6933) |>                          # World Equidistant Cylindrical (equal area)
  st_sample(size = n_points, type = "random", exact = TRUE) |>
  st_transform(4326) |>
  st_sf() |>
  mutate(PointID = sprintf("P%02d", row_number()))

plot(st_geometry(project), border = "grey40")
plot(st_geometry(val_plots), add = TRUE, pch = 3, col = "firebrick")
```

## Map accuracy

With plots located, the check compares the **mapped** carbon at each plot against the **field-measured** carbon. For a reproducible demonstration we simulate a mapped aboveground-carbon surface and a set of field measurements that carry realistic scatter around it; in a real audit the map is the project's product and the field values come from the cruise.

```{r}
set.seed(1)
# a simulated mapped carbon surface over the project
r <- rast(ext(vect(project)), resolution = 0.002, crs = "EPSG:4326")
values(r) <- 150 + 60 * scale(rnorm(ncell(r)))[, 1]   # MgC/ha, mean ~150

mapped <- terra::extract(r, vect(val_plots), ID = FALSE)[, 1]
field  <- mapped + rnorm(n_points, mean = 3, sd = 22)  # field = map + bias + noise

acc <- tibble(PointID = val_plots$PointID, mapped = mapped, field = field)
head(acc)
```

The framework's acceptance test rests on two statistics: the **RMSE** as a percentage of the mean field value, and the **bias**, the mean signed difference. A representative rule accepts a map when its uncertainty is within 20% and its systematic error within 10%, the "20/10" test.

```{r}
resid  <- acc$mapped - acc$field
rmse   <- sqrt(mean(resid^2))
bias   <- mean(resid)
mean_f <- mean(acc$field)

rmse_pct <- rmse / mean_f * 100
bias_pct <- bias / mean_f * 100
# 90% confidence half-width on the mean map error
ci_hw_pct <- qt(0.95, df = n_points - 1) * (sd(resid) / sqrt(n_points)) / mean_f * 100

cat(sprintf("RMSE:  %.1f MgC/ha  (%.1f%%)\n", rmse, rmse_pct))
cat(sprintf("Bias:  %.1f MgC/ha  (%.1f%%)\n", bias, bias_pct))
cat(sprintf("90%% CI half-width on mean error: %.1f%%\n", ci_hw_pct))

pass <- rmse_pct <= 20 && abs(bias_pct) <= 10
cat("20/10 eligibility test:", if (pass) "PASS" else "FAIL", "\n")

plot(acc$mapped, acc$field, pch = 19, cex = 0.6,
     xlab = "Mapped carbon (MgC/ha)", ylab = "Field carbon (MgC/ha)",
     main = "Map vs field at validation plots")
abline(0, 1, col = "red")
```

A map that fails the test is not necessarily wrong, but it cannot be used to issue credits until it is improved or its uncertainty deduction is enlarged. Reporting the RMSE, the bias, and the confidence half-width, exactly these three numbers, is the core of a remote-sensing MRV submission.

## Geometry checks

The other routine check is on the boundary itself. Project areas arrive as shapefiles or KMZ exports that are frequently **invalid**, self-intersecting, containing slivers, or overlapping their own parts, and an invalid boundary corrupts every area total downstream. The verifier's workflow is: make the geometry valid, decompose it to expose overlaps, and quantify any artifacts.

```{r}
# a deliberately messy boundary: two overlapping blocks + a sliver
b1 <- st_polygon(list(rbind(c(0,0), c(1000,0), c(1000,1000), c(0,1000), c(0,0))))
b2 <- st_polygon(list(rbind(c(900,100), c(1900,100), c(1900,1100), c(900,1100), c(900,100))))
messy <- st_sf(id = c("A", "B"), geometry = st_sfc(b1, b2, crs = 3005))

cat("All geometries valid?", all(st_is_valid(messy)), "\n")

# make valid, then decompose overlaps with st_intersection
valid  <- st_make_valid(messy) |> st_intersection()
valid$area_ha <- round(as.numeric(st_area(valid)) * 1e-4, 3)

# the overlap appears as a piece shared by both features (n.overlaps > 1)
overlaps <- valid[valid$n.overlaps > 1, ]
cat("Overlapping area (double-counted):", sum(overlaps$area_ha), "ha\n")
```

Decomposing with `st_intersection()` turns an overlap into an explicit polygon carrying an `n.overlaps` count, so double-counted hectares, the classic way an area total is inflated, become visible and can be subtracted. The same routine flags **artifacts**: slivers below a threshold (commonly 0.4 ha) that are digitising errors rather than real forest and should be reported and removed. A short table of any artifacts, and the reconciled net area, is the deliverable an auditor attaches to the area finding.

```{r}
reconciled_area <- sum(st_area(st_union(st_make_valid(messy)))) * 1e-4
cat("Reconciled (de-duplicated) project area:",
    round(as.numeric(reconciled_area), 1), "ha\n")
```

## Exercises

1. Re-draw the 45 validation plots with three different seeds. How much do RMSE and bias move between draws, and why does fixing and documenting the seed matter for an audit?
2. Increase the simulated field bias until the 20/10 test fails. At what bias does the map become ineligible, and what would a project do to recover eligibility?
3. Add a third overlapping block to the messy boundary and reconcile the area. Confirm that the de-duplicated area is less than the naive sum of the parts.
4. Write the three-number accuracy statement (RMSE %, bias %, 90% CI half-width) as a sentence suitable for a verification report, and explain what each number tells the reader.
