---
title: "Statistical Primers"
format:
  docx:
    reference-doc: ./references/style.docx
    highlight-style: github
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

The chapters assume you can read a standard error, combine two uncertainties, and
say what a Monte Carlo simulation is doing. If any of that has gone rusty, this
appendix is where to repair it. It is deliberately short and it teaches only what
the book actually uses. Nothing here is a substitute for a statistics course, and
nothing in the chapters depends on reading it first.

## Spread, precision and intervals

Start with a distinction that inventory reports get wrong more often than any
other. The standard deviation and the standard error are not the same quantity
and they answer different questions.

The standard deviation describes the forest. It measures how much individual
observations differ from one another. A stand of uniform planted spruce has a
small standard deviation; a mixed natural stand with veterans and saplings has a
large one. Adding plots does not shrink it, because it is a property of the thing
you are measuring, not of your effort in measuring it.

The standard error describes your sample. It measures how far your sample mean is
likely to sit from the true population mean. It shrinks as you measure more,
because more measurement genuinely does pin the mean down better:

$$\text{SE} = \frac{s}{\sqrt{n}}$$

The square root is the part worth internalising. To halve a standard error you
must quadruple the sample. That single fact governs the economics of every
inventory design in this book, and it is why the cheap uncertainty reductions are
almost never "measure more plots".

Take the stem data and divide the plot into quadrats, so each quadrat becomes one
sample of carbon density:

```{r}
stems <- read.csv("data/scbi_stems.csv")
stems$agb_kg    <- exp(-2.48 + 2.4835 * log(stems$dbh))
stems$carbon_Mg <- 0.47 * stems$agb_kg / 1000

stems$quadrat <- paste(cut(stems$gx, 8), cut(stems$gy, 12))
quadrat_area  <- 25.6 / (8 * 12)

qd <- aggregate(carbon_Mg ~ quadrat, data = stems, FUN = sum)
qd$carbon_ha <- qd$carbon_Mg / quadrat_area

n  <- nrow(qd)
m  <- mean(qd$carbon_ha)
s  <- sd(qd$carbon_ha)
se <- s / sqrt(n)
round(c(n = n, mean = m, sd = s, se = se), 2)
```

The standard deviation is an order of magnitude larger than the standard error,
which is exactly as it should be with `r n` quadrats. Quadrupling the sample
would move the standard error to `r round(s / sqrt(4 * n), 2)` and leave the
standard deviation where it is.

A confidence interval turns the standard error into a statement you can report.
It is the mean plus or minus some multiple of the standard error, and the
multiple is set by how confident you want to be:

```{r}
ci <- function(mean, se, level) {
  z  <- qnorm(1 - (1 - level) / 2)
  hw <- z * se
  c(z = z, half_width = hw, rel_pct = 100 * hw / mean,
    lower = mean - hw, upper = mean + hw)
}
round(rbind(`90%` = ci(m, se, 0.90),
            `95%` = ci(m, se, 0.95)), 2)
```

The crediting standards work at 90 per cent, so the multiplier you will meet
most often is 1.645. What they consume is the half-width expressed as a
percentage of the mean, the `rel_pct` column above. That is the uncertainty
figure a deduction schedule is applied to, and here it is large enough to trigger
one under every schedule in the uncertainty chapter. Sampling error alone, on a
well-mapped 25.6 hectare plot, is not a small problem.

Two cautions. First, the interval is a statement about the mean, not about
individual trees. An interval that contains the stand mean will not contain most
of the trees, and a prediction interval for a single tree is far wider. Second,
the interval assumes the sample mean is approximately normally distributed, which
is a much weaker assumption than assuming the data are.

That second point deserves a demonstration, because forest data are about as far
from normal as ecological data get. Per-tree biomass is violently
right-skewed, and yet the mean of a sample of them is not:

```{r}
set.seed(1)
draws <- replicate(2000, mean(sample(stems$agb_kg, 96, replace = TRUE)))

par(mfrow = c(1, 2))
hist(stems$agb_kg, breaks = 60, col = "grey80", border = "white",
     main = "Per-tree biomass", xlab = "kg")
hist(draws, breaks = 40, col = "steelblue", border = "white",
     main = "Mean of 96 trees", xlab = "kg")
```

```{r}
round(c(tree_mean   = mean(stems$agb_kg),
        tree_median = median(stems$agb_kg),
        tree_max    = max(stems$agb_kg),
        mean_of_96  = mean(draws),
        sd_of_means = sd(draws)), 1)
```

The left panel is the forest: a median of `r round(median(stems$agb_kg), 1)` kg
against a maximum of `r round(max(stems$agb_kg))` kg, with `r round(100 * sum(sort(stems$agb_kg, decreasing = TRUE)[1:round(0.05 * nrow(stems))]) / sum(stems$agb_kg))`
per cent of all the biomass sitting in the largest five per cent of the stems.
The right panel is a sampling distribution, and it is close to symmetric. This is
the central limit theorem doing the work that lets you put a normal-theory
interval around a mean computed from wildly non-normal data. It is also the
reason large trees dominate inventory design: the skew does not stop the
mathematics working, but it does mean a sample that misses the big stems misses
most of the carbon.

## Combining uncertainties

Carbon estimates are built by multiplying uncertain things together. Area times
carbon density gives stock. Activity data times an emission factor gives
emissions. Each input carries its own error, and you need the error of the
result.

When the inputs are independent, relative errors combine in quadrature: square
them, add, take the square root.

$$U_{\text{total}} = \sqrt{U_1^2 + U_2^2 + \cdots + U_k^2}$$

The reason is not arbitrary. Errors have signs. Sometimes the area is high while
the factor is low, and the two partly cancel. Adding the percentages assumes
every error goes the wrong way at the same time, which is possible but not
probable, and an estimate built on that assumption is not conservative so much as
wrong:

```{r}
quad <- function(...) sqrt(sum(c(...)^2))

c(quadrature = quad(8, 12), naive_sum = 8 + 12)
```

Now notice what happens when you try to improve either input:

```{r}
rbind(
  `as measured`         = c(ad = 8, ef = 12, total = quad(8, 12)),
  `halve activity data` = c(ad = 4, ef = 12, total = quad(4, 12)),
  `halve factor`        = c(ad = 8, ef =  6, total = quad(8, 6))
)
```

Halving the smaller term buys you `r round(quad(8, 12) - quad(4, 12), 2)`
percentage points. Halving the larger buys `r round(quad(8, 12) - quad(8, 6), 2)`.
Quadrature is dominated by its largest term, and that is the single most useful
piece of guidance in uncertainty management: find the biggest contributor and
spend the money there. Refining anything else is close to wasted effort.

The formula above applies to products. Sums behave differently: when you add
carbon pools, their absolute errors combine in quadrature, so the relative error
of the total is a magnitude-weighted quadratic mean in which a large pool with
modest relative error can outweigh a small pool with a terrible one. That is the
form the ACR improved forest management equation takes, and the uncertainty
chapter works it through against the standard.

Two assumptions are buried in all of this. The inputs must be independent, and if
they are correlated, quadrature understates the total. And the combination must
be roughly linear over the range of the errors. When either fails, stop
propagating algebraically and simulate.

## Monte Carlo methods

Monte Carlo simulation replaces the algebra with brute force. Draw each uncertain
input from its distribution, recompute the whole estimate, and repeat until the
spread of results is itself stable. The spread is the answer.

The method is not an approximation to quadrature; quadrature is an approximation
to it. Where the two are both valid they agree, and it is worth seeing that
before trusting either:

```{r}
set.seed(42)
N <- 100000
area   <- rnorm(N, 1000, 80)     # 8% relative standard error
factor <- rnorm(N, 150,  18)     # 12% relative standard error
total  <- area * factor

round(c(monte_carlo = 100 * sd(total) / mean(total),
        quadrature  = quad(8, 12)), 3)
```

They agree to within a few hundredths of a percentage point, which is the
reassurance you want: when the function is a simple product of independent
normals, the closed form is correct and very much cheaper. Reach for simulation
when it is not.

Allometry is the standard case where it is not. Biomass is exponential in the
fitted coefficient, so the same coefficient uncertainty produces a different
result depending on the tree it is applied to:

```{r}
set.seed(42)
b <- rnorm(N, 2.4835, 0.05)      # uncertainty in the allometric exponent

spread <- function(D) {
  agb <- exp(-2.48 + b * log(D))
  c(dbh_cm = D,
    rel_sd_pct = 100 * sd(agb) / mean(agb),
    mean_over_median = mean(agb) / median(agb))
}
round(t(sapply(c(10, 30, 60, 100), spread)), 3)
```

Two lessons fall out. The relative uncertainty grows with diameter, so quoting a
single model uncertainty percentage for a whole inventory is meaningless without
saying which trees it applies to. And the mean exceeds the median at every
diameter, because exponentiating a symmetric distribution produces a skewed one.
That skew is the same effect the Baskerville correction addresses in the
allometry chapter, arriving here from the other direction.

The remaining question is how many iterations to run. Monte Carlo error falls
with the square root of the iteration count, the same rule that governs sample
size:

```{r}
set.seed(7)
iters    <- c(100, 1000, 10000, 100000)
converge <- sapply(iters, function(k) sd(replicate(200, mean(rnorm(k, 150, 18)))))
round(rbind(iterations = iters, sd_of_result = converge), 4)
```

Each hundredfold increase in iterations shrinks the scatter of the answer by
roughly an order of magnitude, which is the square-root rule again. This is why
ART-TREES requires that Monte Carlo simulations use the 90 per cent confidence
interval and a simulation n of 10,000. Below that the reported
uncertainty is itself noticeably uncertain, and two honest analysts running the
same model would report materially different deductions from the same data.

One operational point matters more than any of the statistics. Set the seed. A
verifier must be able to run your script and obtain your number, and an
unseeded simulation cannot be checked, only re-estimated. Every simulation in
this book is seeded for that reason.

## Where these are used

| Idea | First used in |
|---|---|
| Standard error, confidence intervals | the uncertainty chapter, sampling error |
| Skew and the sampling distribution | the allometry chapter, model fitting |
| Quadrature | the uncertainty chapter, combining sources |
| Magnitude-weighted quadratic mean | the uncertainty chapter, ACR deduction |
| Monte Carlo | the uncertainty chapter, propagation |
| Seeding and reproducibility | throughout |
