---
title: "Emission Factors"
format:
  docx:
    reference-doc: ./references/style.docx
    highlight-style: github
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## The other half of the identity

Almost every greenhouse gas estimate in land use reduces to one multiplication:

$$\text{Emissions} = \text{Activity Data} \times \text{Emission Factor}$$

Activity data is the extent of something happening: hectares deforested, hectares converted, hectares under a new management regime. The emission factor is the greenhouse gas consequence per unit of that activity. The previous chapters have been about the left-hand term, because that is where spatial analysis lives. This chapter is about the right-hand one, which is where most of the money and most of the uncertainty actually sit.

The distinction matters more than it looks. An analyst who maps deforestation to the pixel and then multiplies by a factor pulled from a default table has spent all their effort on the term they measured and none on the term they assumed. The factor is usually the larger source of error, and unlike the map it cannot be improved by better imagery.

## Tiers

The IPCC organises methods into three tiers, and the tier is a statement about where the numbers came from rather than about how sophisticated the arithmetic is [@ipcc2006gl].

**Tier 1** uses global or regional default factors published in the guidelines, applied through simple equations, with default assumptions about what happens after a land-use change. It requires no local measurement beyond the activity data. It is the fallback when nothing better exists, and its uncertainties are correspondingly wide.

**Tier 2** keeps the Tier 1 equations but substitutes country-specific or region-specific factors derived from local measurement. This is where most credible national inventories sit.

**Tier 3** replaces the equations themselves with models or direct measurement systems, calibrated and validated locally.

The tiers are not a quality ladder in the sense that Tier 1 is wrong. A well-applied Tier 1 estimate with an honest uncertainty is more defensible than a Tier 3 model whose calibration nobody can reproduce. What the tiers do determine is the size of the interval you must report, and therefore the deduction you must accept.

## Where the default factors come from

The Tier 1 defaults are not conjured. They are meta-analyses of published soil and biomass measurements, and the guidelines publish their sample sizes and confidence limits alongside the means. Reading those columns is the single most useful habit in factor selection, because the stated uncertainty varies by more than an order of magnitude across the same table.

This is an extract from the reference soil organic carbon stocks of the 2019 Refinement [@ipcc2019refinement, Table 2.3], which are derived from a global soil profile compilation [@batjes2011]. The full table covers ten climate zones by six soil classes; these are seven cells of it.

```{r}
soc_ref <- data.frame(
  climate = c("Cool temperate moist", "Cool temperate moist", "Warm temperate dry",
              "Tropical moist", "Tropical moist", "Tropical dry", "Boreal moist/dry"),
  soil    = c("HAC", "Sandy", "HAC", "HAC", "LAC", "LAC", "Sandy"),
  socref  = c(81, 51, 24, 40, 38, 19, 10),      # tonnes C / ha, 0-30 cm
  unc_pct = c(5, 13, 5, 7, 5, 10, 90),          # 95% limit, % of mean
  n       = c(334, 126, 781, 226, 326, 135, NA))
knitr::kable(soc_ref)
```

Look at the last row against the third. Warm temperate dry soils with high activity clay are stated as 24 tonnes per hectare with a 95 per cent limit of five per cent, resting on 781 measured profiles. Boreal sandy soils are stated as 10 tonnes per hectare with a limit of ninety per cent, resting on no profiles at all: that figure is a nominal error assigned because no data were available, carried forward from the 1996 guidelines.

Those two entries look identical in a spreadsheet. They are epistemically unrelated. One is a well-estimated mean; the other is a placeholder with an honesty marker attached. A project whose stratification lands it in the second cell is in a different position from one that lands in the first, and no amount of downstream care recovers the difference.

```{r}
c(worst_to_best_ratio = max(soc_ref$unc_pct) / min(soc_ref$unc_pct))
```

The stated uncertainty spans eighteenfold within a single table. Factor selection is therefore not a lookup; it is a decision about which of these positions you are willing to defend.

## The stock-difference method

Soil carbon is the pool where Tier 1 methods have changed most. The 2006 guidelines assumed that converting land oxidised its soil carbon immediately. The 2019 Refinement replaces that with a linear transition to a new equilibrium spread over a default twenty years, computed as a difference between two stock estimates [@ipcc2019refinement, Eq. 2.25]:

$$\Delta C_{Mineral} = \frac{SOC_0 - SOC_{(0-T)}}{D}$$

where $SOC_0$ is the stock in the final year of the inventory period, $SOC_{(0-T)}$ the stock at its beginning, and $D$ the time dependence of the stock change factors, twenty years under Tier 1. Each stock is itself built from a reference value and three dimensionless factors:

$$SOC = SOC_{REF} \times F_{LU} \times F_{MG} \times F_{I}$$

$F_{LU}$ is the land use factor, $F_{MG}$ the management factor, and $F_I$ the input factor, each a multiplier relative to the native reference condition. All are defined for a 30 cm depth, because that is the depth the reference stocks were compiled at, and mixing a factor from one depth convention with a stock from another is a common and invisible error.

```{r}
socref <- 38      # tropical moist, LAC
f_lu   <- 0.83    # forest to long-term cultivated
f_mg   <- 1.00    # full tillage
f_i    <- 0.94    # low input

soc_new    <- socref * f_lu * f_mg * f_i
annual_c   <- (soc_new - socref) / 20
annual_co2 <- annual_c * 44 / 12

round(c(soc_before = socref, soc_after = soc_new,
        tC_ha_yr = annual_c, tCO2e_ha_yr = annual_co2), 3)
```

The chain matters more than any single link. Three multiplicative factors mean three opportunities for a stratification error, and because they multiply, a mistake in any one of them scales the whole result rather than adding to it.

## The asymmetry nobody mentions

Tier 1 does not treat gains and losses alike. Carbon *increases* are spread over the twenty-year transition; carbon *losses* from dead organic matter on conversion are assigned to a single year [@ipcc2019refinement, s.2.3.2]. That asymmetry is deliberate and conservative: it credits slowly and debits fast.

It also has a practical consequence for anyone comparing a project's numbers against an inventory. A project reporting a gain over five years captures a quarter of the eventual stock change, not all of it, and a project reporting a conversion loss books the whole thing at once. Two projects with identical biophysical outcomes and different timings will report very different annual figures, entirely legitimately.

## Propagating the uncertainty

Each factor in the chain carries its own uncertainty, and they combine. For independent terms entering a product, the proportional uncertainties combine in quadrature:

$$U_{total} = \sqrt{U_1^2 + U_2^2 + \dots + U_n^2}$$

Take a silvopastoral project on a thousand hectares, with the uncertainty on each component taken from the guidelines and from the project's own measurement:

```{r}
u <- c(SOC_REF = 0.05,   # reference stock, well-sampled cell
       F_MG    = 0.09,   # management factor, improved tropical grassland
       F_I     = 0.30,   # input factor, high organic input
       G       = 0.63,   # biomass accumulation rate, agroforestry
       Area    = 0.10)   # activity data

u_total <- sqrt(sum(u^2))
round(100 * u_total, 1)
```

Seventy-one per cent. That figure is not unusual for a Tier 1 agroforestry estimate, and it is worth sitting with, because it means the honest interval on the claim is wider than the claim itself in one direction.

## Where the uncertainty actually lives

Quadrature has a property that governs everything about how a project should spend its money, and it is not intuitive. Because the terms are squared before summing, a term's contribution to the total is proportional to its square, so large terms do not merely dominate, they annihilate the small ones.

```{r}
round(100 * u^2 / sum(u^2), 1)      # share of total variance
```

The biomass accumulation rate contributes seventy-eight per cent of the variance. The reference soil stock contributes half of one per cent. Now ask what improving each one would buy:

```{r}
sapply(names(u), function(k) {
  v <- u; v[k] <- v[k] / 2
  round(100 * sqrt(sum(v^2)), 1)
})
```

Halving the uncertainty on the growth rate takes the project from 71 to 46 per cent. Halving the uncertainty on the reference soil stock takes it from 71.2 to 71.1. The comparison becomes starker if you imagine perfect knowledge:

```{r}
sapply(names(u), function(k) {
  v <- u; v[k] <- 0
  round(100 * sqrt(sum(v^2)), 1)
})
```

Eliminating the area uncertainty *entirely*, surveying every boundary to the centimetre and knowing the hectares exactly, moves total uncertainty from 71.2 per cent to 70.5. Eliminating the soil reference stock uncertainty moves it to 71.1.

This is the most actionable result in the chapter, and it cuts against how carbon projects usually allocate effort. Spatial precision is visible, satisfying, and easy to procure. It is also, in a project of this shape, nearly worthless. The money belongs in the growth-rate measurement, which is unglamorous fieldwork on permanent plots, and which no improvement in imagery will ever substitute for.

The general rule follows from the algebra: rank the terms by $U_i^2$, and work down the list. A term contributing under one per cent of the variance cannot be worth measuring, however cheap, because the total will not move.

## What the imprecision costs

Convert it to credits, which is the language the decision is actually made in:

```{r}
point_estimate <- 18.5 * 1000                   # tCO2e/yr over 1,000 ha
conservative   <- point_estimate * (1 - u_total)

round(c(point_estimate = point_estimate,
        conservative   = conservative,
        forgone        = point_estimate - conservative,
        forgone_pct    = 100 * u_total))
```

Reporting against the lower bound, the project claims about 5,300 tonnes where its central estimate is 18,500. Thirteen thousand tonnes a year are forgone, not because the carbon is not there, but because the project cannot demonstrate that it is. Under the deduction arithmetic of the uncertainty chapter the loss is smaller and the logic identical.

That number is also the budget available for reducing it. If halving the growth-rate uncertainty costs less than the value of the credits it recovers, the measurement pays for itself, and this calculation is how you find out before commissioning the fieldwork rather than after.

::: {.verification}
#### Double counting across timelines {.unnumbered .unlisted}

The twenty-year transition creates a bookkeeping hazard that catches experienced
practitioners. A parcel that changes management in year one is still accruing its
stock change in year fifteen. If a second intervention is credited on the same
parcel in year eight, part of the soil carbon response may be claimed twice:
once against the first transition and once against the second.

The rule is that stock change for a given parcel is computed against a single
transition timeline, not per intervention. Where interventions overlap, the
parcel carries one trajectory and the later activity is credited only for the
increment beyond what the first would have delivered anyway.

Tracking this in practice means the parcel, not the project, is the accounting
unit, and each parcel needs the date its transition clock started. Where activity
data comes from wall-to-wall mapping rather than parcel records, that date may
simply not exist, which is one of the strongest arguments for the transition
matrices of IPCC Approach 3 over the aggregate statistics of Approach 1.
:::

::: {.verification}
#### Which factors a registry will accept {.unnumbered .unlisted}

Standards differ on how much a project may lean on defaults. In general, a Tier 1
default is acceptable where it is conservative for the claim being made, and not
otherwise. A default that inflates a crediting claim will be challenged; the same
default used to inflate a *baseline* against which the project is credited is
usually accepted without argument, because it works against the proponent.

That asymmetry is worth stating plainly because it is the fastest way to predict
which of a project's factor choices a verifier will scrutinise. Work out, for each
factor, whether an error in the direction of the default increases or decreases
the credits issued. The ones that increase them are where the questions come.

The uncertainty from this chapter then feeds the deduction machinery compared at
the end of the uncertainty chapter. Under ACR it joins the other pool uncertainties in the
weighted combination of Equation 22, and is measured against the ten per cent
precision requirement. Under ART-TREES and Verra it enters the half-width of the
90 per cent interval directly. A 71 per cent uncertainty is catastrophic under any
of them, which is the practical reason Tier 1 agroforestry projects tend to move
to Tier 2 growth rates as soon as they can afford to.
:::

## Exercises

1. Recompute the stock change using a cool temperate moist HAC reference stock (81 tC/ha) instead of the tropical moist LAC value, holding the three factors constant. How much does the annual figure move, and what does that say about the cost of misclassifying climate zone?
2. Using the extract of Table 2.3 above, identify which cells you would be comfortable using in a crediting claim and which you would not. Write one sentence per cell justifying the decision on the basis of the stated uncertainty and sample size.
3. In the propagation example, suppose better fieldwork reduces the growth-rate uncertainty from 63 to 25 per cent, and that improving the input factor from 30 to 20 per cent costs the same amount. Which would you buy, and by how much does each move the total?
4. Add a sixth uncertainty term of 15 per cent to the propagation, representing a carbon fraction. Does it change the ranking of where to spend money? Explain in terms of the squared contributions rather than the raw percentages.
5. The transition period is twenty years for gains and one year for losses. Construct a numerical example of two projects with identical biophysical outcomes but different timings, and show how their reported annual figures differ.
6. *(Applied.)* A project reports 71 per cent uncertainty and a point estimate of 18,500 tCO2e per year. Using the three registry deduction rules from the uncertainty chapter, calculate the credits issued under each. Then state what you would advise the project to measure first.
