---
title: "Geometries"
format:
  docx:
    reference-doc: ./references/style.docx
    highlight-style: github
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Simple features

All vector data rest on one shared idea: a **feature** is a real-world thing that has both attributes and a geometry, and a geometry is a set of coordinates arranged into a defined shape. The international standard that formalises this is *Simple Features*, and it is the vocabulary every geospatial tool speaks [@pebesma2018sf]. Understanding the standard is the foundation of everything that follows, because once you know the geometry types and how they are represented, reading and manipulating any vector dataset becomes routine.

```{r}
library(sf)
```

## Geometry types

The standard defines seven geometry types that cover essentially all vector data. Three are **single** geometries and three are their **multi** counterparts, plus a collection:

- `POINT` — a single location.
- `LINESTRING` — an ordered path of points.
- `POLYGON` — one or more rings enclosing an area (an outer ring, optionally with holes).
- `MULTIPOINT`, `MULTILINESTRING`, `MULTIPOLYGON` — sets of the above, so one feature can be several disjoint shapes (an archipelago is one `MULTIPOLYGON`).
- `GEOMETRYCOLLECTION` — any mixture of the above in a single feature.

Our forest plots are `POINT` features; a management block is a `POLYGON`; a river network would be a `MULTILINESTRING`. Knowing which type you hold determines which operations are meaningful, you can take the length of a line and the area of a polygon, but not the area of a point.

## Building geometries

`sf` constructs each type from coordinates. A single geometry object has class `sfg`:

```{r}
pt   <- st_point(c(-124.68, 49.29))
line <- st_linestring(rbind(c(-124.72, 49.26), c(-124.66, 49.30), c(-124.60, 49.28)))
poly <- st_polygon(list(rbind(
  c(-124.74, 49.25), c(-124.60, 49.25), c(-124.60, 49.33),
  c(-124.74, 49.33), c(-124.74, 49.25))))

st_geometry_type(st_sfc(pt, line, poly))
```

A `POLYGON` is a *list* of rings because it may contain holes: the first ring is the exterior, any further rings are interior boundaries cut out of it. A `MULTIPOLYGON` nests one level deeper, a list of polygons.

```{r}
mp <- st_multipoint(rbind(c(-124.7, 49.27), c(-124.65, 49.30), c(-124.62, 49.26)))
st_geometry_type(st_sfc(mp))
```

## WKT and WKB

Geometries need a portable text or binary form so they can be stored and exchanged. **Well-Known Text** (WKT) is the human-readable representation; **Well-Known Binary** (WKB) is its compact machine form used inside databases and files. `st_as_text()` shows the WKT:

```{r}
st_as_text(pt)
st_as_text(poly)
```

Reading a WKT string back into a geometry closes the loop, which is exactly what happens every time `sf` reads a GeoPackage or a spatial database column:

```{r}
st_as_sfc("POINT (-124.68 49.29)")
```

## Dimensions

Coordinates are usually two-dimensional (XY), but the standard also allows a Z (elevation) and an M (a "measure", such as a distance along a route or a timestamp). A LiDAR return, for instance, is naturally an XYZ point. `sf` tracks this:

```{r}
p3 <- st_point(c(-124.68, 49.29, 850))   # XYZ: lon, lat, elevation
st_dimension(st_sfc(p3))
class(p3)
```

## Validity

A geometry can be structurally *invalid*, a polygon whose boundary crosses itself, a ring that does not close, and invalid geometries silently corrupt areas, overlays, and joins. Checking and repairing validity is a foundational habit, not a niche concern:

```{r}
# a bow-tie polygon that self-intersects
bad <- st_polygon(list(rbind(c(0,0), c(1,1), c(1,0), c(0,1), c(0,0))))
st_is_valid(bad)
fixed <- st_make_valid(bad)
st_is_valid(fixed)
st_geometry_type(fixed)          # repair may change the type
```

`st_make_valid()` repairs the geometry using the underlying GEOS library, sometimes changing its type in the process (a self-intersecting polygon may become a `MULTIPOLYGON`). Running a validity check the moment data arrive, before any measurement, is the single cheapest way to avoid a corrupted result.

## Exercises

1. Construct one geometry of each of the seven simple-feature types and confirm each with `st_geometry_type()`. Which types can return a meaningful `st_length()`, and which an `st_area()`?
2. Write a polygon with a hole (an exterior ring plus one interior ring) and compute its area. Confirm the hole is subtracted.
3. Convert the forest plots to WKT with `st_as_text()`, then read one point back with `st_as_sfc()`. Why is WKB, not WKT, used inside spatial databases?
4. *(Applied.)* Airborne LiDAR returns are XYZ points. Build a small `MULTIPOINT` of XYZ coordinates representing tree tops with heights in Z, and confirm the dimension is `XYZ`. How would carrying Z change how you compute a canopy height?
