---
title: "Attributes and Support"
format:
  docx:
    reference-doc: ./references/style.docx
    highlight-style: github
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Attributes

A feature carries attributes alongside its geometry, and it is tempting to treat those attributes as ordinary table columns. But an attribute value stands in a particular *relationship* to the geometry it is attached to, and getting that relationship wrong is one of the most common and least visible errors in spatial analysis. This chapter is about that relationship: what a value refers to, and what therefore happens to it when geometries are combined or split.

```{r}
library(sf)
library(dplyr)
plots <- st_as_sf(read.csv("data/forest_plots.csv"),
                  coords = c("lon", "lat"), crs = 4326)
```

## Support

The **support** of an attribute is the area (or volume, or time) over which it was measured or to which it applies. A plot's carbon density was measured over a fixed-area plot of 0.1 hectare; that 0.1 ha is its support. A stand-level average has the whole stand as its support. Two numbers can look identical yet mean different things because they have different support, and comparing or combining them without accounting for support produces nonsense. Whenever you record a value, you should be able to state its support.

## Extensive and intensive

Attributes divide into two kinds by how they behave when a geometry changes size. An **extensive** attribute is a total that scales with area, a count of stems, a tonnage of carbon; split a polygon in two and the value divides between the pieces. An **intensive** attribute is a rate or density that does not scale with area, stems per hectare, carbon per hectare, mean diameter; split the polygon and each piece keeps (roughly) the same value.

This distinction dictates how you aggregate. Our carbon column is intensive (MgC per hectare); the total carbon a plot represents is extensive (MgC over its 0.1 ha):

```{r}
plots <- plots |>
  mutate(carbon_density = carbon,          # intensive: MgC / ha
         carbon_total   = carbon * 0.1)    # extensive: MgC over the 0.1-ha plot
```

## Aggregation

To summarise carbon over a group of plots, the two attribute kinds demand different operations. The extensive total is **summed**; the intensive density is **averaged** (area-weighted when the units differ in size). Summing a density, or averaging a total, is simply wrong:

```{r}
plots |>
  st_drop_geometry() |>
  group_by(stand_type) |>
  summarise(
    total_carbon_MgC   = sum(carbon_total),               # extensive -> sum
    mean_density_MgCha = weighted.mean(carbon_density, w = 0.1)  # intensive -> (area-weighted) mean
  )
```

`sf`'s `aggregate()` applies a function while dissolving geometries, and the same rule governs the choice of function: `sum` for extensive attributes, `mean` for intensive ones. The dissolve gives the combined geometry; you supply the aggregation that respects the attribute's kind.

## Change of support

Analysis constantly moves values from one support to another. Predicting plot-support carbon onto a raster of 20-metre pixels, or up to a whole stand, is the **change-of-support problem**, and it is why an estimate's uncertainty depends on the support you report it at. A prediction for a single pixel is noisier than the same model's prediction averaged over a stand, because averaging over a larger support cancels error. Geostatistics, later in the book, makes this explicit through *block* prediction, estimating a value for an area rather than a point.

The practical rule for an auditor is to always state the support of a reported number, and never to combine values of different support without an explicit, defensible rule. A carbon density and a carbon total are not interchangeable, and treating them as if they were is how area totals silently inflate.

## Exercises

1. Classify each column of `forest_plots.csv` as extensive or intensive: `n_trees`, `stems_ha`, `basal_area`, `carbon`. For each, state whether it should be summed or averaged when combining plots.
2. Aggregate carbon to the stand-type level two ways, summing the density and correctly summing the total. Show that the first answer is meaningless and explain why.
3. The plots have 0.1-ha support. Explain what changes about the *value* and about its *uncertainty* if you predict carbon at 20-metre pixel support instead.
4. *(Applied.)* A verifier receives a stand table giving carbon *density* per stand and a separate area per stand. Write the one line that produces the extensive total carbon per stand, and explain which attribute kind each input is.
