---
title: "Point Patterns"
format:
  docx:
    reference-doc: ./references/style.docx
    highlight-style: github
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Point patterns

Most of this book analyses *values measured at locations*: carbon at a plot, elevation at a pixel. Point-pattern analysis asks a different question, about the **locations themselves**. Are the events, tree stems, disease outbreaks, detected clearings, spread evenly, scattered at random, or clustered together? The answer characterises the process that generated them, and it is a core branch of spatial statistics with its own tools. We use the `spatstat` package, whose `ppp` object represents a point pattern inside an observation window.

```{r}
library(spatstat)
```

## Complete spatial randomness

The reference against which every pattern is judged is **complete spatial randomness** (CSR): points falling independently and with equal probability everywhere, a homogeneous Poisson process. Real patterns depart from CSR in one of two directions, toward regularity (points avoid each other, as in a planted stand) or toward clustering (points attract, as in natural regeneration around seed trees). We simulate a one-hectare stand three ways to see them:

```{r}
set.seed(1)
win <- owin(c(0, 100), c(0, 100))                       # 100 x 100 m stand

csr     <- rpoispp(lambda = 0.03, win = win)            # random
clustered <- rMatClust(kappa = 0.004, scale = 6, mu = 8, win = win)  # clumped regen
regular <- rSSI(r = 6, n = 120, win = win)              # spaced (min 6 m apart)

par(mfrow = c(1, 3))
plot(csr, main = "Random"); plot(clustered, main = "Clustered"); plot(regular, main = "Regular")
par(mfrow = c(1, 1))
```

## Intensity

The **intensity** is the first-order property: the expected number of points per unit area, and how it varies across the window. A kernel density estimate smooths the points into a continuous intensity surface, revealing where events concentrate, the point-pattern analogue of a heat map:

```{r}
plot(density(clustered), main = "Estimated intensity (points/m²)")
points(clustered, pch = 20, cex = 0.4)
```

## Clustering

Whether points are clustered or regular is the second-order property, about the *interaction* between points. A quick, formal check is a quadrat test, which divides the window into cells and compares the counts to what CSR predicts:

```{r}
quadrat.test(clustered, nx = 4, ny = 4)
```

A small p-value rejects CSR; the clustered pattern shows it clearly. The average nearest-neighbour distance points the same way, clustered points sit closer to their nearest neighbour than random points, and regular points farther:

```{r}
c(clustered = mean(nndist(clustered)),
  random    = mean(nndist(csr)),
  regular   = mean(nndist(regular)))
```

## Ripley's K

A single number cannot capture how clustering changes with scale, points may cluster at short range but disperse at long range. **Ripley's K function** describes the pattern across all distances at once by counting, on average, how many other points lie within a growing radius of each point, compared with CSR. Plotting the observed K against the theoretical CSR curve, with a simulation envelope, is the standard diagnostic:

```{r}
plot(envelope(clustered, Kest, nsim = 39, verbose = FALSE),
     main = "Ripley's K with CSR envelope")
```

Where the observed curve rises above the envelope, the pattern is more clustered than random at that distance; below, more regular. This scale-explicit view is why K (and its variance-stabilised form, the L function) is the workhorse of point-pattern analysis, and it applies equally to stems in a plot and to disturbance events across a landscape.

## Exercises

1. Simulate a CSR pattern and run `quadrat.test()` on it. Does the test correctly fail to reject randomness? Repeat a few times: how often does it wrongly reject at p < 0.05?
2. Compare the mean nearest-neighbour distance for the random, clustered, and regular patterns, and relate the ordering to Tobler's idea of spatial dependence.
3. Estimate and map the intensity of the clustered pattern. How would you use an intensity surface to prioritise field checks?
4. *(Applied.)* Detected clearings in a monitoring image are a point pattern. Explain how Ripley's K could distinguish scattered natural tree-fall from an organised, clustered illegal-harvest front.
