---
title: "Coordinate Systems and the Hectare That Moves"
subtitle: "Chapter 1 redraft, 6 September 2026"
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

::: callout-note
## Redraft

A prose redraft of the `01-positional-accuracy.qmd` skeleton under the
`writing-scientific-books` skill, kept beside the original for comparison. The
argument, the code and the two open decisions are unchanged. What is new is that
the arguments are written out, every term is introduced in the sentence that
first uses it, and every number in the prose is computed by a chunk on this page.
:::

```{r setup}
library(sf)
library(dplyr)
```

## The plot

Twenty-five point six hectares of oak and hickory sit on a hillside outside
Front Royal, Virginia, and every tree on them has been measured. The plot is a
rectangle 400 metres across and 640 metres up the slope, its corners surveyed,
its stems mapped on a local grid that starts at zero in one corner. Somebody
walked it, tagged it, and wrote down where each stem stood.

```{r first-look}
stems <- read.csv("../data/scbi_stems.csv") |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

st_crs(stems)$epsg
st_crs(stems)$Name
st_bbox(stems)
```

```{r plot-facts, echo=FALSE}
raw <- read.csv("../data/scbi_stems.csv")
n_stems <- nrow(raw)
n_spp   <- length(unique(raw$species))
w_m     <- diff(range(raw$gx, na.rm = TRUE))
h_m     <- diff(range(raw$gy, na.rm = TRUE))
claimed_ha <- w_m * h_m / 1e4
```

The file holds `r format(n_stems, big.mark = ",")` stems of `r n_spp` species.
The local grid runs `r w_m` metres in one direction and `r h_m` in the other, so
the plot covers `r claimed_ha` hectares by construction, and that is the number
the plot is known by.

Hold onto that figure. By the end of this chapter the same plot will have three
different areas, none of them a survey error, and the difference between the
largest and the smallest will be larger than the amount of carbon most projects
argue about.

The bounding box above should land over Front Royal, near 78.1 degrees west and
38.9 north. If it had landed at zero the coordinates would never have been
georeferenced at all, and here is the trap in that. A map of the stems would
still have looked perfectly sensible, because the plotting would have used the
plot's own axes and the trees would have sat in their proper arrangement
relative to each other. Nothing on the screen tells you the whole thing is
sitting in the Gulf of Guinea.

## Coordinate systems

A coordinate is not a fact about the ground. It is a claim about a model of the
Earth, and the claim is only as good as the model named alongside it.

The number 38.9 means nothing on its own. It means something once you say it is
degrees of latitude on WGS 84, which is a particular mathematical model of the
Earth's shape maintained by the United States Department of Defense. A
*coordinate reference system*, meaning the pairing of that model with the rules
for turning positions on it into numbers, is what makes a coordinate readable.
Every one of them in common use carries a code from the EPSG registry, a public
catalogue originally assembled by the European Petroleum Survey Group, and
looking a code up at `epsg.io` takes about four seconds.

The data above declares 4326, which is longitude and latitude in degrees on
WGS 84. Degrees are a poor unit for measuring anything, because a degree of
longitude is 111 kilometres at the equator and nothing at all at the pole, so
the first move in almost any analysis is to reproject.

```{r reproject}
stems_utm <- st_transform(stems, 32617)   # UTM zone 17N, metres
st_bbox(stems_utm)
```

The numbers are now metres east and metres north within Universal Transverse
Mercator zone 17, one of sixty north-south strips the world is cut into so that
each can be flattened with tolerable distortion. The trees have not moved. What
changed is the ruler.

A shapefile records its system in a plain text sidecar with the extension
`.prj`, and that file holds exactly one system, described in a notation called
well-known text. It records the system the data is in now. It records nothing
about where the data has been, which is a limitation this chapter returns to
twice.

## Ellipsoid and geoid

Every flat measurement is a curved surface pressed onto a plane, and the
flattening is a choice with consequences you can compute.

The physical Earth is lumpy, so three surfaces are used to stand in for it and
they are not interchangeable. The *ellipsoid* is a smooth squashed sphere chosen
to fit the planet as a whole, and it is a pure mathematical object with no
physical existence. The *geoid* is the surface at which gravity has the same
value everywhere, roughly the shape the oceans would settle into if they ran
under the continents, and it undulates by tens of metres because rock is denser
in some places than others. The *terrain* is the ground you actually stand on.
Horizontal position is quoted against the ellipsoid and height is quoted against
the geoid, which is why height is the harder coordinate and why an elevation
that does not name its vertical datum is close to meaningless.

The oldest version of this problem was solved with a stick. Eratosthenes,
measuring the shadow at Alexandria at the moment the sun stood directly over a
well at Syene, got the circumference of the Earth to within a few per cent
around 240 BC. Latitude yields to that kind of observation, because the sun and
the stars tell you how far north you are from anywhere with a clear horizon and
a table. Longitude does not. Longitude is a question about time, in that knowing
how far east you are means knowing what the hour is somewhere else at the same
instant, and that required a clock that would keep time on a rolling deck
through months of salt and temperature swing. The British Board of Longitude
offered a prize for it in 1714 and the prize took a lifetime to claim. That is
the shortest route into what a datum is, which is an agreement about where zero
is and what shape the Earth has, arrived at because the alternative was ships
hitting rocks.

Datums are revised, and each revision moves the coordinates of ground that never
moved. North American surveying ran on the Clarke 1866 ellipsoid, then on the
North American Datum of 1983, then increasingly on WGS 84, and positions shifted
by up to about a hundred metres across those changes in parts of the continent.
The modern versions agree far more closely, but agreement is not identity. WGS 84
as most software resolves it is not one datum at all but an *ensemble*, a bundle
of successive realisations treated as interchangeable, and the EPSG registry
gives that ensemble a positional accuracy of about two metres. Two metres is
published, it is in the metadata, and almost nobody reads it before quoting a
boundary to the centimetre.

## Area measurement

Now the applied core of the chapter. The plot claims `r claimed_ha` hectares.
Measure it two ways and neither answer is that.

```{r area-two-ways}
claimed_ha <- 25.6

bbox_ha <- as.numeric(st_area(st_as_sfc(st_bbox(stems_utm)))) / 1e4
hull_ha  <- as.numeric(st_area(st_convex_hull(st_union(stems_utm)))) / 1e4

data.frame(
  geometry      = c("bounding box", "convex hull"),
  measured_ha   = round(c(bbox_ha, hull_ha), 2),
  difference_ha = round(c(bbox_ha, hull_ha) - claimed_ha, 2),
  pct_error     = round(100 * (c(bbox_ha, hull_ha) - claimed_ha) / claimed_ha, 1)
)
```

```{r area-numbers, echo=FALSE}
box_pct  <- 100 * (bbox_ha - claimed_ha) / claimed_ha
hull_pct <- 100 * (hull_ha - claimed_ha) / claimed_ha
rot_fit  <- lm(st_coordinates(stems_utm)[, 1] ~ raw$gx + raw$gy)
rot_deg  <- abs(atan2(coef(rot_fit)[3], coef(rot_fit)[2]) * 180 / pi)
```

The bounding box gives `r round(bbox_ha, 2)` hectares, which is
`r round(box_pct, 1)` per cent more than the plot has. The convex hull gives
`r round(hull_ha, 2)`, which is `r round(abs(hull_pct), 1)` per cent under and
is effectively the right answer.

The cause is a rotation of `r round(rot_deg, 1)` degrees. The plot was laid out
along the slope, not along the grid, so its long axis runs
`r round(rot_deg, 1)` degrees off UTM north. A bounding box is axis-aligned by
definition, so wrapping one round a tilted rectangle is like putting a tilted
book into an upright box. The box has to be bigger than the book, and the more
the book is tilted the more air you buy.

Neither number is a survey error. Both are correct measurements of different
things, and the difference between them is `r round(bbox_ha - hull_ha, 2)`
hectares of ground that does not exist. When a claimed area and a mapped area
disagree, the first question is not whether the survey is wrong. It is whether
the thing being measured is the boundary or a box drawn round it, because that
choice alone moved the answer by more than a hectare on a plot small enough to
walk across in ten minutes.

## Boundary verification

Area non-compliance is the most common finding an auditor raises against a
forest carbon project, and it is raised before a single tonne of carbon has been
calculated. The reason is arithmetic. Credits are a rate multiplied by an area,
so an area that is wrong by three per cent puts every downstream number wrong by
three per cent, and no amount of careful biomass modelling recovers it.

::: callout-warning
## Pending data

This section runs on the a public dataset ACR submission in
`training/area-verification`, which is outside this repository. Before this chapter can
render offline the following have to be copied into `data/` and cleared for
publication: the submitted boundary, the strata, the harvest units, and the
`slivers` and `artifacts` outputs. Anonymise the client if TÜV SÜD clearance is
not in hand.
:::

```{r area-check, eval=FALSE}
# Sequence as run in training/area-verification/area-verification.qmd
sf_use_s2(FALSE)                                  # configure the geometric plane
st_layers("../data/submission.gpkg")              # what arrived
bdry <- st_read("../data/submission.gpkg", "boundary", quiet = TRUE)

all(st_is_valid(bdry))                            # validate before measuring
bad  <- bdry[!st_is_valid(bdry), ]                # extract the artefacts
bdry <- st_make_valid(bdry)

bdry$area_ha <- as.numeric(st_area(bdry)) / 1e4   # hectares, by feature
```

The order of those steps is the point. Set the plane, see what arrived, validate,
pull out the invalid pieces and look at them, then measure, then break the total
down by boundary, stratum and harvest unit, then test the difference against the
project's materiality threshold.

Validation comes before measurement and never after, for a reason that catches
people once each. An invalid polygon is one whose boundary crosses itself, and
the commonest way to make one is to digitise a parcel and let the pen wander
back across a line already drawn, leaving a bow tie. Ask for the area of a bow
tie and the software returns a number. It does not warn you, it does not stop,
and the number is wrong, because one lobe of the tie is being counted with the
opposite sign. A silent wrong answer is worse than a crash, and this is a place
the software gives you one.

## Overlap and ownership

Nine submissions in ten fail this on the first pass. The test is simple. Take
every polygon in the project and intersect it with every other one, and any
positive area that comes back is land the project has claimed twice.

```{r overlap}
e  <- st_bbox(stems_utm)
x0 <- unname(e["xmin"]); y0 <- unname(e["ymin"])
a <- st_as_sfc(st_bbox(c(xmin = x0,       xmax = x0 + 250,
                        ymin = y0,        ymax = y0 + 400), crs = 32617))
b <- st_as_sfc(st_bbox(c(xmin = x0 + 200, xmax = x0 + 450,
                        ymin = y0,        ymax = y0 + 400), crs = 32617))

round(as.numeric(st_area(st_intersection(a, b))) / 1e4, 2)   # ha claimed twice
```

Two rectangles overlapping by fifty metres across four hundred hand back two
hectares counted twice. The same test is then run outward rather than inward,
against federal land, adjacent private parcels, national parks and neighbouring
ownership, because a project may not sell carbon on ground it does not control.

Then comes the trap that costs the most and is the hardest to see. A project
area is what survives after a list of exclusions has been applied, so you remove
the roads, the steep ground, the riparian buffers, the leased parcels. Each
exclusion leaves a residual, meaning the ground that survives that one test, and
the correct project area is the ground that survives every test at once, which
is the intersection of all the residuals.

Chain unions instead and you get something that looks like an answer and is not.
A union readmits any ground excluded by one test that happened to survive
another, so a pixel thrown out by the slope mask comes back in because the roads
residual kept it. Every exclusion becomes optional. The belt comes out larger,
and it comes out larger in the direction that suits the project, which is why
this error is found more often than its opposite.

## Transformation provenance

By the time a boundary reaches an auditor it has usually been through four or
five coordinate systems. It was collected on a handheld receiver in WGS 84,
brought into a desktop package in a state plane system, reprojected to a
national grid for a client, exported for a subcontractor, and reprojected back.
Each of those steps is reversible on its own. The chain as a whole is not, and
almost nobody can tell you what it was.

The tempting explanation is that the file formats cannot record it. That
explanation is wrong, and the honest version is more uncomfortable.

The shapefile genuinely cannot. It has no lineage mechanism of any kind and its
`.prj` sidecar holds the present system and nothing else. But the shapefile is
no longer the only option. GeoPackage, an encoding standard published by the
Open Geospatial Consortium, fixes the real shapefile defects, which are a two
gigabyte ceiling on file size, field names truncated to ten characters, a limit
of 255 columns, and an ambiguity between polygon and multipolygon that makes
counting features unreliable. It also carries a `gpkg_metadata` table able to
hold records in the ISO 19115 metadata standard, and ISO 19115-1:2014 defines a
lineage element intended for exactly this.

So the container can hold the chain. The chain is missing because nobody writes
it, and that is a working practice rather than a technical limit. Writing four
lines into a metadata table at each reprojection costs a minute and is the
difference between a boundary an auditor can accept and one they cannot.

## Geometry engines

Two analysts measure the same polygon and get different areas. Neither has made
a mistake, and this is not a rare event.

```{r versions}
sf_extSoftVersion()[c("GEOS", "GDAL", "proj.4")]
```

Underneath almost every geographic information system in use sit the same three
open source libraries. GEOS does the geometry, meaning intersections, unions and
validity. GDAL reads and writes the formats. PROJ does the coordinate
transformations. ArcGIS runs its own engine over GDAL. QGIS and R and Python
call the same libraries directly, and each ships whichever versions it was built
against.

The versions matter because the libraries change. PROJ 6, released in 2019,
moved the canonical description of a coordinate system from the old proj4 string
to well-known text version 2, introduced the datum ensemble mentioned earlier,
and made late binding the default, meaning the software now waits until it knows
both ends of a transformation before choosing the path between them. Every one
of those was an improvement. Every one of them means the same request can
resolve differently on either side of the change, by an amount that is usually
small and is occasionally not.

That is why a serious area check ends with a log of the runtime rather than with
a number. A number without its software environment is not reproducible, and
reproducible is the whole standard a verifier is applying.

## Plot monumentation

Everything so far has been about where a boundary is. This section is about
where a plot centre is, and it is where positional accuracy becomes carbon.

Permanent sample plots in North America are marked with rebar driven into the
ground, and the national inventory programmes use markers built to survive
decades of frost and logging and fire. That is not fussiness. A field crew
returning after five years has to stand where the last crew stood, because the
plot is defined as the trees within a fixed radius of a point, and a crew that
sets up three metres away measures a partly different set of trees.

Three metres sounds forgivable. It is not, because of what happens next. Plot
measurements are not reported as plot measurements. They are multiplied by an
expansion factor, meaning the number of hectares each plot is taken to
represent, and scaled across the whole project. A plot standing for four hundred
hectares carries its error four hundred times over into the reported total. The
error does not stay on the plot. Nothing does.

## GNSS accuracy

Report a position as a distribution, never as a single number.

The vocabulary first, because it is used loosely. GPS is one satellite
constellation, operated by the United States. *GNSS* is the family of them, and
a modern receiver listens to the American, Russian, European and Chinese systems
at once, which is why it fixes faster and holds position better under canopy
than a receiver from fifteen years ago. Real-time kinematic correction, usually
shortened to RTK, adds a second receiver at a known point that measures the
error in the signal at that instant and sends a correction to the first one. It
takes a handheld from a few metres to a few centimetres, and it costs money,
a base station and a working data link.

The way to know what your own equipment does is to measure it. Occupy a
published control point, record a few hundred fixes, and report the distribution
of the distances from the truth, including the tail. A receiver that is accurate
to two metres most of the time and to twenty metres under a closed canopy in
December is not a two metre receiver, and the plots that matter are the ones
under canopy. Source material for this section sits in `GPS-training-material`,
`survey-tools` and the KoBoToolbox deployment.

## Reference data

Interpreted ground reference points are the most valuable and least glamorous
data product in remote sensing. Every accuracy figure quoted in the rest of this
book rests on somebody having looked at an image and said what was there.

The scale of that work is worth stating, because it is invisible in the papers
that use it. The Global Land Cover Estimation training database, built at Boston
University under Curtis Woodcock and Mark Friedl and published in *Scientific
Data* in 2023, holds close to two million training units covering 1984 to 2020
at 30 metres, labelled across seven primary and twelve secondary classes. That
took years and a large team, and it exists because no model can be trained or
tested without it.

Nobody cites the reference database in the abstract. The map gets the citation.
The reference data is the reason the map can claim anything at all.

## Planar and spherical

One caveat remains, and it opens the door to the rest of the book.

```{r s2}
sf_use_s2(TRUE)
sph <- as.numeric(st_area(st_convex_hull(st_union(stems)))) / 1e4
sf_use_s2(FALSE)
ell <- as.numeric(st_area(st_convex_hull(st_union(stems)))) / 1e4

data.frame(engine = c("s2, spherical", "planar hull, geodesic area"),
           area_ha = round(c(sph, ell), 4),
           rel_diff = c(NA, round((ell - sph) / sph, 8)))
```

```{r s2-numbers, echo=FALSE}
s2_rel  <- abs((ell - sph) / sph)
box_rel <- abs(bbox_ha - hull_ha) / hull_ha
```

A conventional geographic information system represents the world as a flat
projection, the way a paper map does. S2, whose name is the mathematical
notation for the unit sphere, represents it on a sphere, the way a globe does.
The difference is not decorative. A globe has no edges, so a polygon crossing
the antimeridian is an ordinary polygon rather than a special case, and it has
no singular points, so the poles behave like everywhere else. What the library
buys in practice is fast indexing in memory, constructive operations that are
robust because coordinates are snapped to a grid before they are compared, and
efficient nearest-neighbour search.

Now the honest framing, which is the part worth teaching. The two areas above
differ by a relative amount of `r format(signif(s2_rel, 2), scientific = TRUE)`.
The bounding box and the convex hull, earlier in this chapter, differed by
`r round(100 * box_rel, 1)` per cent, which is about
`r format(round(box_rel / s2_rel), big.mark = ",")` times larger. A reader who
takes one thing from this chapter should take that ratio, because the instinct
on meeting a new source of error is to fear all of them equally, and the skill
worth having is ranking them. Choosing the wrong geometry for your boundary cost
a hectare and a half. Choosing the wrong Earth model cost a few square metres.

There is a further honesty in that code block. With s2 switched off, the hull is
constructed on a plane while the area is still computed on the ellipsoid, which
is a mixed setting and exactly the kind of thing a reader has to be able to
state about their own workflow.

The real gain from spherical predicates is not the area at all. It is that a
planar intersection of two nearly coincident boundaries manufactures *slivers*,
meaning long thin false polygons a few centimetres wide along an edge where two
versions of the same line disagree, and exact spherical arithmetic does not
manufacture them. The file `slivers layer` in the area-verification outputs is a
catalogue of them.

So the closing rule of this chapter. An area reported without its geometry
engine, its library versions and its geometry setting is not reproducible,
whatever its magnitude. And the question of which plane you compute on, and what
support the answer has, is the thread this book follows all the way to the data
cube.

## Exercises

1. Reproject the stems to a geographic coordinate system other than WGS 84 and
   back again, and report the largest distance any stem moved. State whether the
   result is within the two metre accuracy of the WGS 84 ensemble.
2. Rotate the plot outline through every angle from 0 to 45 degrees and plot the
   bounding box area against the angle. At what rotation is the box exactly a
   third larger than the plot, and what is the maximum?
3. Compute the plot area in UTM zone 17N and again in a neighbouring zone, where
   the plot sits far from the central meridian. Report the difference in
   hectares and say which of the two you would defend.
4. Take the convex hull of the stems and buffer it inward by one metre, then
   outward by one metre. Report the area of each. Say what that implies for a
   boundary digitised from imagery with a two metre positional accuracy.
5. *(Applied.)* Build two overlapping polygons and compute the project area
   twice, once as an intersection of residuals and once as a chain of unions.
   Report both, and report the ratio between them.
6. *(Applied.)* A project reports 1,240 hectares from a bounding box drawn round
   a boundary that runs 12 degrees off grid north. Using only the results in this
   chapter, estimate the area actually enclosed, and state what fraction of the
   project's credits are being claimed on ground that is not there.

::: callout-note
## Open decisions

**Does this chapter split?** Eleven sections runs 35 to 45 printed pages, at the
top of the series range. The clean line falls after "Geometry engines", leaving
coordinate systems, projection and area as chapter 1, and plot centres, GNSS and
ground reference as chapter 2.

**Which two ebooks?** The author wants the geospatial data objects section
aligned to two ebooks that present them without overwhelming the reader. Neither
was supplied. Names or paths needed.
:::
