---
title: "Coordinate Systems and the Hectare That Moves"
subtitle: "Chapter 1, restructured 6 September 2026"
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

::: callout-note
## Restructure

The prose of the 6 September redraft is kept wherever its section survives. What
changed is the order and the content, integrating
`drafts/CH01/ch01-reproducibility-notes-v3.md`. Four sections are new:
conversion against transformation, straight lines that are not, projection
distortion measured, and what an auditor can say. The area failure now arrives
second, before any theory, and the three surfaces close the chapter as its
contract with the reader.

Two claims in the notes were tested on this machine and did not survive. Both
corrections are in the text and marked.
:::

```{r setup}
library(sf)
library(dplyr)
library(lwgeom)
```

## The plot

Twenty-five point six hectares of oak and hickory sit on a hillside outside
Front Royal, Virginia, and every tree on them has been measured. The plot is a
rectangle 400 metres across and 640 metres up the slope, its corners surveyed,
its stems mapped on a local grid that starts at zero in one corner. Somebody
walked it, tagged it, and wrote down where each stem stood.

```{r first-look}
stems <- read.csv("../data/scbi_stems.csv") |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

st_crs(stems)$epsg
st_bbox(stems)
```

```{r plot-facts, echo=FALSE}
raw <- read.csv("../data/scbi_stems.csv")
n_stems <- nrow(raw)
n_spp   <- length(unique(raw$species))
w_m     <- diff(range(raw$gx, na.rm = TRUE))
h_m     <- diff(range(raw$gy, na.rm = TRUE))
claimed_ha <- 25.6
```

The file holds `r format(n_stems, big.mark = ",")` stems of `r n_spp` species.
The local grid runs `r w_m` metres in one direction and `r h_m` in the other, so
the plot covers `r claimed_ha` hectares by construction, and that is the number
the plot is known by.

Hold onto that figure. Before the next page is out the same plot will have three
different areas, none of them a survey error.

The bounding box above should land over Front Royal, near 78.1 degrees west and
38.9 north. If it had landed at zero the coordinates would never have been
georeferenced at all, and here is the trap in that. A map of the stems would
still have looked perfectly sensible, because the plotting would have used the
plot's own axes and the trees would have sat in their proper arrangement
relative to each other. Nothing on the screen tells you the whole thing is
sitting in the Gulf of Guinea.

## The hectare that moves

The plot claims `r claimed_ha` hectares. Measure it two ways and neither answer
is that.

```{r area-two-ways}
stems_utm <- st_transform(stems, 32617)   # UTM zone 17N, metres

bbox_ha <- as.numeric(st_area(st_as_sfc(st_bbox(stems_utm)))) / 1e4
hull_ha <- as.numeric(st_area(st_convex_hull(st_union(stems_utm)))) / 1e4

data.frame(
  geometry      = c("bounding box", "convex hull"),
  measured_ha   = round(c(bbox_ha, hull_ha), 2),
  difference_ha = round(c(bbox_ha, hull_ha) - claimed_ha, 2),
  pct_error     = round(100 * (c(bbox_ha, hull_ha) - claimed_ha) / claimed_ha, 1)
)
```

```{r area-numbers, echo=FALSE}
box_pct  <- 100 * (bbox_ha - claimed_ha) / claimed_ha
hull_pct <- 100 * (hull_ha - claimed_ha) / claimed_ha
rot_fit  <- lm(st_coordinates(stems_utm)[, 1] ~ raw$gx + raw$gy)
rot_deg  <- abs(atan2(coef(rot_fit)[3], coef(rot_fit)[2]) * 180 / pi)
```

The bounding box gives `r round(bbox_ha, 2)` hectares, which is
`r round(box_pct, 1)` per cent more than the plot has. The convex hull gives
`r round(hull_ha, 2)`, which is `r round(abs(hull_pct), 1)` per cent under and
is effectively the right answer.

The cause is a rotation of `r round(rot_deg, 1)` degrees. The plot was laid out
along the slope, not along the grid, so its long axis runs
`r round(rot_deg, 1)` degrees off UTM north. A bounding box is axis-aligned by
definition, so wrapping one round a tilted rectangle is like putting a tilted
book into an upright box. The box has to be bigger than the book, and the more
the book is tilted the more air you buy.

Now put a price on it, because that is what the rest of this book is for.

```{r area-cost}
c_density <- 150          # tonnes of carbon per hectare, a plausible temperate figure
co2_ratio <- 44 / 12      # molecular mass of carbon dioxide over carbon

phantom_ha  <- bbox_ha - hull_ha
phantom_co2 <- phantom_ha * c_density * co2_ratio

round(c(phantom_hectares = phantom_ha, phantom_tonnes_co2e = phantom_co2), 1)
```

```{r cost-numbers, echo=FALSE}
ph_ha <- phantom_ha; ph_co2 <- phantom_co2
```

`r round(ph_ha, 2)` hectares of ground that does not exist, and about
`r format(round(ph_co2), big.mark = ",")` tonnes of carbon dioxide equivalent
that could be claimed on it. On a plot you can walk across in ten minutes, and
before a single tree has been mismeasured.

Neither number is a survey error. Both are correct measurements of different
things. When a claimed area and a mapped area disagree, the first question is
not whether the survey is wrong. It is whether the thing being measured is the
boundary or a box drawn round it.

The rest of this chapter is about the other ways that question can be answered
wrongly, and it works upward from the coordinate itself.

## What a coordinate claims

A coordinate is not a fact about the ground. It is a claim about a model of the
Earth **and about a date**, and it is only as good as both of those.

The number 38.9 means nothing on its own. It means something once you say it is
degrees of latitude on WGS 84, a particular mathematical model of the Earth's
shape. A *coordinate reference system*, meaning the pairing of that model with
the rules for turning positions on it into numbers, is what makes a coordinate
readable. Every one in common use carries a code from the EPSG registry.

The date is the half that almost every dataset omits, and it is not optional
metadata. The ground moves. On the Pacific coast of North America the crust
drifts one to two centimetres a year relative to the whole-Earth frame, so a
latitude and longitude without a date is an incomplete statement in the same way
that a share price without a date is incomplete. Two different dates are
involved and confusing them is common. The *frame reference epoch* is the date
at which the reference frame itself is defined, and ITRF2014 has one of 2010.0.
The *coordinate epoch* is the date at which your particular measurement was
valid, so a post surveyed in June 2016 has a coordinate epoch of 2016.47. To
compare two coordinates you need both frames and both epochs. Moving a
coordinate between epochs within one frame is a *point motion operation*, done
with a velocity model.

That distinction is why two national frames drift apart. A *dynamic* frame is
fixed to the whole Earth, so the coordinates of a fixed post change over time. A
*static*, or plate-fixed, frame moves with the tectonic plate, so the
coordinates of that post stay roughly constant within the plate and drift
relative to the globe. The International Terrestrial Reference Frame is dynamic.
The Canadian realisation of the North American Datum of 1983 is plate-fixed.
They differ by more than a metre on the west coast and the gap widens every
year.

The EPSG registry holds over ten thousand definitions, which is not a useful
number. What is useful is that they come in a small number of kinds, and only
three of them appear in most forest carbon work. **Geocentric** systems give
three Cartesian numbers measured from the Earth's centre of mass, and that is
what a satellite receiver actually solves for. **Geographic** systems give
latitude and longitude on an ellipsoid, and that is the storage and exchange
convention. **Projected** systems give a flat grid in metres, and that is for
display and for legacy area computation. Two more matter as soon as height is
involved, **vertical** systems quoting height above a defined surface and
**compound** systems bolting a horizontal and a vertical together. Errors enter
at the junctions between these, not inside any one of them.

```{r crs-lookup}
codes <- c(4326, 32617, 3005, 3157)
data.frame(
  epsg = codes,
  name = vapply(codes, function(x) st_crs(x)$Name, character(1)),
  units = vapply(codes, function(x) {
    u <- st_crs(x)$units_gdal; if (is.null(u)) NA_character_ else u }, character(1))
)
```

The data above declares 4326, longitude and latitude in degrees. Degrees are a
poor unit for measuring anything, because a degree of longitude is 111
kilometres at the equator and nothing at all at the pole, so the first move in
almost any analysis is to change the system. Which brings us to the distinction
the rest of the chapter rests on.

## Conversion and transformation

These two words are used interchangeably in ordinary speech and they name
completely different operations. ISO 19111 draws the line precisely, and it is
the single most useful distinction in this chapter.

A **conversion** changes the coordinate system and not the datum. Latitude and
longitude on WGS 84 to Universal Transverse Mercator zone 17 north on WGS 84 is
a conversion. It is defined by an exact formula, it is reversible to the limits
of floating point arithmetic, and it has no uncertainty of its own. Nothing
moves on the ground.

A **transformation** changes the datum. The Canadian realisation of the North
American Datum of 1983 to WGS 84 realisation G1762 is a transformation. It is
*estimated from observations*, so it carries a published accuracy. It is *not
unique*, so several valid transformations exist between the same pair of datums
and they give different answers. It often depends on a *grid shift file*, an
external data product with its own version. Things move on the ground, by
metres.

Both are called *coordinate operations* when you need a word for the pair.

The practical test is one line. **If the operation has an accuracy figure
attached, it is a transformation, and you must record which one you used. If it
does not, it is a conversion, and it is not your problem.**

```{r conversion-vs-transformation}
# A hundred stems is enough, and st_distance needs by_element or it builds
# a full n by n matrix, which on 38,000 points does not finish.
sample <- stems[seq(1, nrow(stems), length.out = 100), ]

# A conversion. Same datum, new grid. Round-trip and see what is lost.
back <- st_transform(st_transform(sample, 32617), 4326)
conv_max_m <- max(st_distance(sample, back, by_element = TRUE))

# A transformation. Different datum. Coordinates move.
nad83 <- st_transform(sample, 4269)          # NAD83 geographic
trans_max_m <- max(st_distance(sample, st_transform(nad83, 4326),
                               by_element = TRUE))

round(c(conversion_roundtrip_m = as.numeric(conv_max_m),
        transformation_shift_m = as.numeric(trans_max_m)), 6)
```

```{r cvt-numbers, echo=FALSE}
cm <- as.numeric(conv_max_m); tm <- as.numeric(trans_max_m)
```

The conversion round-trip loses `r format(signif(cm, 2), scientific = TRUE)`
metres, which is floating point noise. Whether the transformation moves anything depends on which
operation the software selected, and that is the point of the next two sections.
Ask which one was used, not whether one was used.

## Ellipsoid, geoid and the ensemble

Every flat measurement is a curved surface pressed onto a plane, and the
flattening is a choice with consequences you can compute.

The physical Earth is lumpy, so three surfaces stand in for it and they are not
interchangeable. The *ellipsoid* is a smooth squashed sphere chosen to fit the
planet as a whole, a pure mathematical object with no physical existence. The
*geoid* is the surface at which gravity has the same value everywhere, roughly
the shape the oceans would settle into if they ran under the continents, and it
undulates by tens of metres because rock is denser in some places than others.
The *terrain* is the ground you actually stand on. Horizontal position is quoted
against the ellipsoid and height against the geoid, which is why height is the
harder coordinate and why an elevation that does not name its vertical datum is
close to meaningless.

The oldest version of this problem was solved with a stick. Eratosthenes,
measuring the shadow at Alexandria at the moment the sun stood directly over a
well at Syene, got the circumference of the Earth to within a few per cent
around 240 BC. Latitude yields to that kind of observation, because the sun and
the stars tell you how far north you are from anywhere with a clear horizon and
a table. Longitude does not. Longitude is a question about time, in that knowing
how far east you are means knowing what the hour is somewhere else at the same
instant, and that required a clock that would keep time on a rolling deck
through months of salt and temperature swing. The British Board of Longitude
offered a prize for it in 1714 and the prize took a lifetime to claim. That is
the shortest route into what a datum is, which is an agreement about where zero
is and what shape the Earth has, arrived at because the alternative was ships
hitting rocks.

A datum has two parts, and only the second one is interesting. First a
**shape**, almost always an ellipsoid with a stated semi-major axis and
flattening. Second an **anchoring** of that shape to the physical planet,
historically by a marked monument and now by the coordinates and velocities of a
global station network. The shape alone is not a datum. WGS 84 and the North
American Datum of 1983 use ellipsoids that are nearly identical in size, and
they differ because they are anchored differently. That is why the offset
between them is over a metre while the difference between their ellipsoids is
below a millimetre.

Now the part that damages a carbon inventory. WGS 84 as most software resolves
it is not one datum at all. It is an **ensemble**, a bundle of successive
realisations treated as interchangeable for approximate work, and the register
states its accuracy in the definition itself. Do not take that on trust. Print
what your own machine returns.

```{r ensemble}
wkt <- st_crs(4326)$wkt
members <- gsub('MEMBER\\["|"$', "",
                regmatches(wkt, gregexpr('MEMBER\\["[^"]+"', wkt))[[1]])
accuracy <- regmatches(wkt, regexpr("ENSEMBLEACCURACY\\[[0-9.]+\\]", wkt))

length(members)
members
accuracy
sf_extSoftVersion()[c("GEOS", "GDAL", "proj.4")]
```

```{r ensemble-numbers, echo=FALSE}
n_mem <- length(members)
proj_v <- unname(sf_extSoftVersion()["proj.4"])
```

`r n_mem` members, and an accuracy of two metres, on PROJ `r proj_v`.

Print that on a different machine and you may get a different count. The
standard's own published example listed six. Registers in circulation during
2024 listed seven. This machine lists `r n_mem`, having gained the most recent
realisation. **That is not a flaw in the demonstration, it is the demonstration.**
The definition of the most widely used coordinate reference system on Earth
changes underneath the software, and the number of realisations bundled into it
is a property of the version you happen to be running.

So what does the modern well-known text notation buy? It cannot remove the
ambiguity, because the ambiguity is real. What it does is **convert a silent
error into a declared uncertainty**, by writing the ensemble and its accuracy
into the definition where a machine can read them. A silent error cannot be
audited. A declared two metres can be tested against a materiality threshold,
reported with a confidence statement, and either accepted or raised as a
corrective action. The notation does not fix the problem. It turns the problem
into evidence.

The fix itself is a specification rather than a piece of software, and it has
five parts. Require the **realisation** and not the family, so WGS 84 (G1762)
rather than WGS 84. Require the **coordinate epoch** as a decimal year. Require
the **transformation by its EPSG code**, with its published accuracy, not by
prose. Store the result in a container that can hold all of that, which a
shapefile's `.prj` sidecar cannot and a GeoPackage can. And **pin the software**,
meaning the PROJ version, the EPSG database version and the grid file versions.

One line to carry out of this section. WGS 84 is not a datum. It is a two-metre
agreement to stop arguing, and that agreement is fine for navigation and fatal
for a hectare-accurate inventory.

## Straight lines that are not

Everything so far has been about the numbers at the corners. This section is
about what the software assumes lies between them, and it is the assumption
nobody declares.

Underneath every geographic information system sits a data model called Simple
Features, and it is a data model rather than a file format. It answers two
questions. What counts as a shape, and what may you ask about shapes. The
vocabulary is seven types, point, line, polygon and their plural forms plus a
collection. The word *simple* is technical rather than a claim about ease, and
it means a geometry that does not cross itself. The questions you may ask are a
fixed set of predicates, whether one shape intersects, contains, touches,
crosses or is within another. In a database every one of these carries the
prefix `ST_`, and when you see `ST_Area` you are looking at Simple Features.

Here is the assumption. A line is defined by its vertices **plus an implicit
rule about what lies between them**, and ISO 19125-1 fixes that rule as linear
interpolation. Straight lines. And "straight" is defined in whatever number
space the coordinates happen to live in.

The model was designed for a plane, for metres on a survey grid. Nothing stops
you putting degrees into it, and everyone does. When you do, "straight" quietly
means straight in a grid of degrees, which is an equirectangular projection you
never chose, never declared and cannot see.

Simple Features is a grammar for shapes. It tells you what a well-formed
sentence looks like. It does not tell you which language you are speaking, and
fed degrees it parses perfectly and means something other than you intended.

What that costs is measurable. Take two vertices a hundred kilometres apart at
the same latitude, and ask where the line between them actually runs.

```{r linear-interpolation}
edge <- st_sfc(st_linestring(rbind(c(-126.0, 50), c(-124.6, 50))), crs = 4326)

# the great-circle path between the same two vertices, densified
gc  <- st_cast(st_segmentize(edge, dfMaxLength = 1000), "POINT")
co  <- st_coordinates(gc)
mid <- co[round(nrow(co) / 2), c("X", "Y")]

c(simple_features_midpoint_lat = 50,
  great_circle_midpoint_lat    = round(mid[["Y"]], 5),
  divergence_metres            = round((mid[["Y"]] - 50) * 111320, 1))
```

```{r si-numbers, echo=FALSE}
sf_div <- round((mid[["Y"]] - 50) * 111320, 1)
```

The two vertices agree exactly. Halfway along, the two interpretations of the
line between them are `r sf_div` metres apart. The Simple Features edge runs due
west at a constant latitude. The path a surveyor would walk, and the path a
spherical geometry engine computes, bows north. On a boundary made of long
east-west edges, and on any boundary at high latitude, that difference changes
the enclosed area.

This is the reason the next chapters insist on spherical predicates, and the
reason is not accuracy. It is that a spherical library uses exact arithmetic for
its predicates, so the answer to whether a plot lies inside a jurisdiction is
identical on every machine and every version. Determinism is the win.

## Projection distortion, measured

A projection distorts. Every book says so and almost none of them says by how
much, in hectares, on ground the reader works on. Two demonstrations, both
computed here.

### The block that moves across a zone

Universal Transverse Mercator cuts the world into sixty north-south strips six
degrees wide. Each strip has a central meridian where the scale factor is
0.9996, deliberately less than one, so that the error at the strip's edge is
about equal and opposite to the error at its middle. Take one 500 hectare block
and slide it from the central meridian to the zone edge.

```{r utm-sweep}
square_500ha <- function(lon, lat, ha = 500) {
  d <- sqrt(ha * 1e4)
  dlat <- d / 111320
  dlon <- d / (111320 * cos(lat * pi / 180))
  st_sfc(st_polygon(list(rbind(c(lon, lat), c(lon + dlon, lat),
                               c(lon + dlon, lat + dlat), c(lon, lat + dlat),
                               c(lon, lat)))), crs = 4326)
}

sweep <- lapply(c(-123, -124, -125, -126), function(lon) {
  p <- square_500ha(lon, 49.5)
  g <- as.numeric(st_geod_area(p)) / 1e4
  u <- as.numeric(st_area(st_transform(p, 32610))) / 1e4
  data.frame(deg_from_central_meridian = abs(lon + 123),
             geodesic_ha = round(g, 3), utm_ha = round(u, 3),
             error_ha = round(u - g, 3), error_pct = round(100 * (u - g) / g, 3))
}) |> bind_rows()

sweep
```

```{r utm-numbers, echo=FALSE}
e_cm   <- sweep$error_ha[sweep$deg_from_central_meridian == 0]
e_edge <- sweep$error_ha[sweep$deg_from_central_meridian == 3]
swing  <- e_edge - e_cm
```

On the central meridian the block **shrinks** by `r sprintf('%.2f', abs(e_cm))`
hectares. At the zone edge it **grows** by `r sprintf('%.2f', e_edge)`. The full swing across one zone is
`r round(swing, 2)` hectares on 500, and the sign flips on the way. Nothing
about the block changed.

::: callout-important
## A correction to the working notes

The notes for this chapter proposed teaching this through a cutblock straddling
the boundary at 126 degrees west, measured in zone 9 and again in zone 10, on
the expectation that the two zones would disagree. They do not.

```{r zone-straddle}
p <- square_500ha(-126.02, 49.5)
z9  <- as.numeric(st_area(st_transform(p, 32609))) / 1e4
z10 <- as.numeric(st_area(st_transform(p, 32610))) / 1e4
alb <- as.numeric(st_area(st_transform(p, 3005)))  / 1e4
geo <- as.numeric(st_geod_area(p)) / 1e4

data.frame(basis = c("geodesic on the ellipsoid", "UTM zone 9N",
                     "UTM zone 10N", "BC Albers, equal area"),
           ha = round(c(geo, z9, z10, alb), 3),
           error_ha = round(c(0, z9, z10, alb) - c(0, geo, geo, geo), 3))
```

```{r straddle-numbers, echo=FALSE}
z_gap <- abs(z9 - z10)
```

The two zones differ by `r signif(z_gap, 2)` hectares, which is nothing. The
reason is worth more than the original example. A block sitting on a zone
boundary is three degrees from **both** central meridians, so both projections
distort it identically. Choosing the zone is a red herring. What matters is the
distance from whichever central meridian you chose, which is what the sweep
above measures. And the equal-area projection reproduces the geodesic area to
three decimal places, which is what an equal-area projection is for.
:::

### The buffer that is not five kilometres

Web Mercator stretches distance by one over the cosine of the latitude, so a
metre on its grid is less than a metre on the ground, and increasingly so as you
go north. Buffering is one of the most frequent operations in this field and one
of the least checked, and leakage belts, reference regions and exclusion zones
are all defined by it.

```{r buffer-latitude}
buffer_check <- lapply(c(49, 60, 68), function(lat) {
  pt <- st_sfc(st_point(c(-120, lat)), crs = 4326)
  planar   <- st_transform(st_buffer(st_transform(pt, 3857), 5000), 4326)
  geodesic <- st_buffer(pt, 5000)          # s2, 5000 metres on the ground
  data.frame(latitude = lat,
             web_mercator_ha = round(as.numeric(st_area(planar)) / 1e4, 0),
             geodesic_ha     = round(as.numeric(st_area(geodesic)) / 1e4, 0),
             fraction        = round(as.numeric(st_area(planar)) /
                                     as.numeric(st_area(geodesic)), 2))
}) |> bind_rows()

buffer_check
```

```{r buffer-numbers, echo=FALSE}
f49 <- buffer_check$fraction[buffer_check$latitude == 49]
f68 <- buffer_check$fraction[buffer_check$latitude == 68]
a68 <- buffer_check$web_mercator_ha[buffer_check$latitude == 68]
g68 <- buffer_check$geodesic_ha[buffer_check$latitude == 68]
```

At 49 degrees north the planar buffer encloses `r f49` of the ground it should.
At 68 degrees north, ordinary boreal forest, it encloses `r f68`, which is
`r format(a68, big.mark = ",")` hectares where
`r format(g68, big.mark = ",")` were intended. The error is a smooth function of
latitude rather than an edge case at the pole, which is why it is so rarely
noticed and so often wrong.

## Where a coordinate comes from

Everything above treats the coordinate as given. It is worth one section on
where it came from, because the order of operations explains why the ellipsoid
is negotiable and the measurement is not.

A navigation satellite broadcasts a signal saying, in effect, that it is at a
stated place and sent this at a stated instant. The receiver picks up several
such signals, works out how long each took to arrive, multiplies by the speed of
light to get a distance, and finds the one point in space consistent with all of
those distances at once. Four satellites are the minimum, because there are four
unknowns, three for position and one for the receiver's own clock error. The
technique is called *multilateration*, meaning position from multiple distances.

The answer comes out as three plain numbers in metres from the Earth's centre of
mass, called *Earth-centred, Earth-fixed*. There is no sphere here and no flat
map. **The measurement itself is not made on any map, on any ellipsoid or on any
sphere.** All of those come afterwards and all of them are conventions.

Latitude, longitude and height are then *computed from* those three numbers by
choosing an ellipsoid and asking where the point falls on it. Height needs one
step more, because the ellipsoid is smooth and sea level is not, so converting
an ellipsoidal height into the height above sea level a forester recognises
requires a geoid model.

So the order is measurement, then convention. The Cartesian solution is what the
physics produced. The ellipsoid is a choice made in software afterwards, and the
sphere used by a spherical geometry library is a further approximation applied
after that. Which is also why no clever choice of geometry downstream can
improve a position. It can only fail to damage it.

The vocabulary is used loosely and is worth fixing. GPS is one constellation,
operated by the United States. *GNSS* is the family, and a modern receiver
listens to the American, Russian, European and Chinese systems at once, which is
why it fixes faster and holds position better under canopy than a receiver from
fifteen years ago. All four are aligned to the International Terrestrial
Reference Frame at centimetre level, which is why mixing them works. *Real-time
kinematic* correction adds a second receiver at a known point that measures the
error at that instant and sends a correction to the first, taking a handheld
from a few metres to a few centimetres, at the cost of a base station and a
working data link.

The way to know what your own equipment does is to measure it. Occupy a
published control point, record a few hundred fixes, and report the distribution
of the distances from truth, including the tail. A receiver that is accurate to
two metres most of the time and to twenty metres under a closed canopy in
December is not a two-metre receiver, and the plots that matter are the ones
under canopy.

That matters because of what a plot centre is. Permanent sample plots in North
America are marked with rebar driven into the ground, and national inventory
programmes use markers built to survive decades of frost and logging and fire.
A field crew returning after five years has to stand where the last crew stood,
because the plot is defined as the trees within a fixed radius of a point, and a
crew that sets up three metres away measures a partly different set of trees.
Three metres sounds forgivable and it is not, because plot measurements are
multiplied by an expansion factor, meaning the number of hectares each plot is
taken to represent, and scaled across the project. A plot standing for four
hundred hectares carries its error four hundred times over into the reported
total. The error does not stay on the plot. Nothing does.

## The engines and the chain

Two analysts measure the same polygon and get different areas. Neither has made
a mistake, and this is not a rare event.

```{r versions}
sf_extSoftVersion()[c("GEOS", "GDAL", "proj.4")]
```

Underneath almost every geographic information system sit the same three open
source libraries. GEOS does the geometry, meaning intersections, unions and
validity. GDAL reads and writes the formats. PROJ does the coordinate
operations. Commercial desktop packages run their own engine over GDAL. R and
Python and QGIS call the same libraries directly, and each ships whichever
versions it was built against.

The versions matter because the architecture changed twice, and both changes
altered numbers rather than only interfaces.

Before 2018, PROJ was a projection engine with datum handling bolted on
afterwards. It used **early binding**, meaning a datum shift was baked into the
definition of each system and every operation pivoted through WGS 84 as a hub.
That is how a definition string could silently carry a seven-parameter shift
nobody had chosen. PROJ 5 introduced an explicit transformation pipeline, so an
operation became something you could write down and re-execute. PROJ 6, in 2019,
was the break that matters. The EPSG dataset was imported into PROJ itself as a
database, so definitions stopped being scattered across disagreeing files.
Well-known text version 2 arrived, so a system could be written and read without
loss. **Late binding** replaced early binding, so PROJ now looks up candidate
operations in the register, each with its own accuracy and area of validity, and
picks one at run time. And axis order began following the authority, which meant
a great deal of code that had assumed longitude first silently swapped
coordinates. PROJ 7 moved grid files to a format that can be fetched on demand
from a network, which is convenient and is bad for reproducibility unless you
pin.

**Here is the consequence, and it is the paragraph to remember.** Two analysts
running the identical command on the identical file can get different numbers,
because the operation is chosen at run time from a database whose version, and
from a grid collection whose contents, differ between their machines. That is
not a bug. It is the design. It is also why "I reprojected to WGS 84" is not a
reproducible statement.

Which is why the chain of operations a boundary has been through is worth
recording, and why almost nobody records it. By the time a boundary reaches an
auditor it has usually been through four or five systems. Collected on a
handheld in WGS 84, brought into a desktop package in a state plane system,
reprojected to a national grid for a client, exported for a subcontractor, and
reprojected back. Each step is reversible alone. The chain is not.

The tempting explanation is that the file formats cannot record it, and that
explanation is wrong. The shapefile genuinely cannot, having no lineage
mechanism of any kind and a `.prj` sidecar that holds the present system and
nothing else. But the shapefile is no longer the only option. GeoPackage fixes
the real shapefile defects, which are a two gigabyte ceiling, field names
truncated to ten characters, a limit of 255 columns, and an ambiguity between
polygon and multipolygon that makes counting features unreliable. It also
carries a metadata table able to hold ISO 19115 records, and that standard
defines a lineage element intended for exactly this. So the container can hold
the chain. The chain is missing because nobody writes it, which is a working
practice rather than a technical limit.

An audit-ready submission therefore has seven items in it, and they belong in a
data product specification as conformance criteria rather than in an email as
requests.

1. The **realisation**, not the family. WGS 84 (G1762) or NAD83(CSRS), never
   plain WGS 84.
2. The **coordinate epoch**, as a decimal year.
3. The **transformation used**, by EPSG code, with its published accuracy.
4. **Software versions**, meaning PROJ, the EPSG database, and any grid files.
5. The **edge rule**, planar or geodesic.
6. The **measurement basis**, meaning the method and the ellipsoid the area was
   computed on.
7. A **checksum** over the geometry.

## What an auditor can say

Suppose a boundary arrives without most of that. The question is not whether to
accept or reject it, which is a false choice. The question is what can be
written down.

Four questions, from the auditing standard ISO 19011, apply to the metadata and
not only to the geometry. Is the information complete, correct, consistent and
current? A projection file with no epoch fails "complete", and it fails for a
structural reason rather than because anyone withheld anything. The same
standard says that only information which can be verified to some degree counts
as audit evidence, and that where verifiability is low the auditor must use
professional judgement about how much to rely on it. A boundary whose
realisation is unknown is low-verifiability evidence. Say so in the file rather
than silently recomputing and hoping.

The vocabulary for the finding comes from ISO 19157, the standard on data
quality, and it is more useful than it looks because positional accuracy is not
one thing. *Absolute external accuracy* is how far the data are from truth.
*Relative internal accuracy* is how well the vertices agree among themselves.
The ensemble problem damages the first and leaves the second intact, so an
auditor can write "internally excellent, externally uncertain to two metres"
instead of arguing about whether the data are good.

The same standard offers a way to qualify a number you cannot fully stand
behind. *Metaquality* is information about the quality of your quality
statement, and it has three elements, confidence, representativity and
homogeneity. Report the accuracy figure, attach a confidence statement naming
the unresolved ensemble, and use homogeneity where a boundary was captured
across seasons on different base stations. That is the mechanism that lets a
verifier be honest without either accepting or rejecting outright.

Two harder facts close this out. **The standard sets no thresholds and says so
in its own scope.** Acceptable quality lives in a *conformance quality level*,
a number stated in a data product specification or in user requirements, so the
registry or the verification body has to set it because the standard has
declined to. The reproducibility contract is not a coordinate system
declaration. It is a specification with numbers in it.

And **lineage is the fallback, and lineage is the weak link.** Where positional
accuracy cannot be measured directly, the standard routes you to indirect
evaluation, meaning assessing quality from external knowledge, and gives lineage
as its example. Lineage is metadata, described in the standard's own words as
general, non-quantitative and illustrative. So when the measurement is
unavailable you are told to rely on the record of where the data has been, and
the record of where the data has been is exactly what the shapefile export
destroyed.

## The three surfaces

Everything above reduces to three rules, and they are stated last rather than
first because they are conclusions. Call them the three surfaces. Globe for
geometry, ellipsoid for measurement, plane for pictures.

**Do the geometry on the globe.** Predicates, overlays and indexing on longitude
and latitude are computed with spherical geometry, which in R is the `s2` engine
that `sf` uses by default. The reason is not that a sphere is more accurate than
a plane. It is that the spherical library uses exact arithmetic for its
predicates, so the answer to whether a plot lies inside a jurisdiction is
identical on every machine and every version. Determinism is the win.

**Do the measurement on the ellipsoid.** No number that reaches a verification
statement comes from spherical or projected geometry. Areas, lengths and
distances come from geodesic calculations on a declared ellipsoid.

**Declare the surface you used.** A spherical measurement states its sphere
radius, because there is no single right one. An ellipsoidal measurement states
its ellipsoid. A projected measurement states its projection. A number without
its surface is not a measurement, it is a rumour.

```{r three-surfaces}
hull <- st_convex_hull(st_union(stems))

sf_use_s2(TRUE);  sphere <- as.numeric(st_area(hull)) / 1e4
sf_use_s2(FALSE); ellips <- as.numeric(st_area(hull)) / 1e4
sf_use_s2(TRUE)

data.frame(surface = c("sphere, radius 6371010 m", "ellipsoid, geodesic",
                       "plane, UTM zone 17N"),
           area_ha = round(c(sphere, ellips, hull_ha), 4))
```

```{r surface-numbers, echo=FALSE}
s2_rel  <- abs((ellips - sphere) / sphere)
box_rel <- abs(bbox_ha - hull_ha) / hull_ha
```

Now the ranking, which is the single most useful thing in this chapter. Sphere
against ellipsoid differ by a relative amount of
`r format(signif(s2_rel, 2), scientific = TRUE)`. The bounding box against the
convex hull, on page one, differed by `r round(100 * box_rel, 1)` per cent,
about `r format(round(box_rel / s2_rel), big.mark = ",")` times larger.

The instinct on meeting a new source of error is to fear all of them equally.
The skill worth having is ranking them. Choosing the wrong geometry for your
boundary cost a hectare and a half. Choosing the wrong Earth model cost a few
square metres. Choosing the wrong realisation, or losing the record of which one
you used, costs an argument you cannot win.

And there is a standing caveat that belongs beside the rules rather than inside
them. Spherical geometry is not a coordinate reference system. It has no EPSG
code, no ellipsoid and no epoch. Give it a mislabelled coordinate and it returns
a fast, exact, perfectly repeatable answer in the wrong place. The three
surfaces fix *how* you compute. They do not fix *what* you were given.

The rest of this book obeys these rules and says so at the first measurement in
every chapter. The question of which surface you computed on, and what support
the answer has, is the thread that runs from here to the data cube.

## Exercises

1. Reproject the stems to a geographic system other than WGS 84 and back again,
   and report the largest distance any stem moved. State whether the result is
   within the two-metre accuracy of the ensemble, and say whether the operation
   you performed was a conversion or a transformation.
2. Rotate the plot outline through every angle from 0 to 45 degrees and plot the
   bounding box area against the angle. At what rotation is the box exactly a
   third larger than the plot, and what is the maximum?
3. Extend the zone sweep to the neighbouring zone, so the block runs from three
   degrees west of one central meridian to three degrees east of the next.
   Plot the error against longitude and mark where it crosses zero.
4. Repeat the buffer comparison at every degree from 0 to 80 north and plot the
   enclosed fraction against latitude. At what latitude does a planar buffer
   enclose less than half the ground intended?
5. Take the two vertices from the linear interpolation section and move them to
   68 degrees north, keeping the same separation in longitude. Report the
   divergence and say what it implies for a boreal project boundary digitised
   from widely spaced vertices.
6. *(Applied.)* A project reports 1,240 hectares from a bounding box drawn round
   a boundary running 12 degrees off grid north. Using only the results in this
   chapter, estimate the area actually enclosed, and state what fraction of the
   project's credits are being claimed on ground that is not there.
7. *(Applied.)* Write the seven-item submission specification from this chapter
   as a machine-readable sidecar for the stem dataset, filling in every field you
   can and marking the ones you cannot. State which of the gaps you would raise
   as a finding and which you would accept with a confidence statement.

::: callout-note
## Open decisions

**This chapter now has to split.** It ran eleven sections before the notes were
integrated and it runs eleven after, but four of them are new and the argument
is denser. The clean line falls after "Straight lines that are not", which
leaves a first chapter on what a coordinate claims and what it costs, and a
second on the audit trail, running from projection distortion through the
engines to what an auditor can say. That is the recommendation. It also relieves
the seventeen-chapter page budget problem in the wrong direction, taking the
book to eighteen before the two candidate chapters are counted.

**Two claims from the working notes did not survive testing** and both
corrections are in the text. The cutblock straddling two UTM zones does not
produce disagreeing areas, because a block on the boundary is equidistant from
both central meridians; the distance from a single central meridian is what
matters and the sweep now shows it. And the ensemble member count is eight on
this machine rather than the six or seven the notes anticipated, which the
chapter now uses as its own demonstration.

**Still pending.** The boundary verification and overlap sections were cut from
this draft and must be rebuilt on public data. No client or engagement material
may appear in this book in any form. The two ebooks the geospatial objects
section was to be aligned to have still not been supplied.
:::
