---
title: "Points, Lines, Polygons, and the Hectare That Moves"
subtitle: "Chapter 1, condensed 6 September 2026"
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

::: callout-note
## Condensed

Cut from fifteen sections to eight on 6 September 2026 and refocused on one
question, what a geospatial object is and how well its position on the Earth is
known. Three recently revised standards carry it, ISO 19125-1 on what a point, a
line and a polygon are, ISO 19111:2019 on how coordinates are referenced, and
ISO 19157 on how positional accuracy is stated.

The reader is the one the accepted proposal names. The main text is for graduate
students in forestry, environmental science and remote sensing and for analysts
arriving from a neighbouring field, and it assumes introductory statistics and no
prior geographic information systems. Registry material sits in tinted
verification tracks, and skipping every one of them still leaves a complete
chapter.
:::

```{r setup}
library(sf)
library(dplyr)
library(lwgeom)
```
## The plot

Twenty-five point six hectares of oak and hickory sit on a hillside outside
Front Royal, Virginia, and every tree on them has been measured. The plot is a
rectangle 400 metres across and 640 metres up the slope, its stems mapped on a
local grid that starts at zero in one corner. Somebody walked it, tagged it, and
wrote down where each stem stood.

```{r first-look}
stems <- read.csv("../data/scbi_stems.csv") |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

st_crs(stems)$epsg
st_bbox(stems)
```
```{r plot-facts, echo=FALSE}
raw <- read.csv("../data/scbi_stems.csv")
n_stems <- nrow(raw)
n_spp   <- length(unique(raw$species))
w_m     <- diff(range(raw$gx, na.rm = TRUE))
h_m     <- diff(range(raw$gy, na.rm = TRUE))
claimed_ha <- 25.6
```
::: callout-important
## Where these positions come from

The stems are real and so are their positions within the plot. The census
recorded each one on the plot's own grid, in metres from a corner, and those are
the `gx` and `gy` columns.

The longitude and latitude are not surveyed. They are reconstructed by
`data-raw/prepare-scbi-data.R`, which lays the plot grid onto Universal
Transverse Mercator zone 17 north with a fixed affine map, a rotation of about
1.8 degrees together with scale factors of 1.001149 and 0.997105. Anything in
this chapter that depends on where the plot sits on the Earth depends on that
reconstruction, and is representative rather than a survey of record. This is
said again wherever it bites.
:::

The file holds `r format(n_stems, big.mark = ",")` stems of `r n_spp` species.
The local grid runs `r w_m` metres one way and `r h_m` the other, so the plot
covers `r claimed_ha` hectares by construction, and that is the number it is
known by. Hold onto it.

## Points, lines, polygons

Everything in a forest carbon project is one of three shapes, and the whole
software stack agrees on what those three are.

A **point** is a single position. A plot centre is a point, and so is a tree, a
control monument and a soil pit. A **line**, or LineString, is an ordered list of
points. A road is a line, so is a stream, and so is the traverse a surveyor walks
round a parcel. A **polygon** is a closed ring of points with the inside
declared, optionally with further rings punched out of it as holes. A project
boundary is a polygon. So is a stratum, a harvest unit and a lake.

That agreement has a name and a document behind it. The model is called *Simple
Features*, standardised as ISO 19125-1, and it is not one implementation you can
choose to avoid. It sits underneath shapefiles, GeoJSON, GeoPackage, PostGIS,
QGIS, and the R package used throughout this book, whose name `sf` is the
standard's initials. When you meet a function beginning `st_`, you are looking at
Simple Features.

```{r geometry-types}
pt   <- st_sfc(st_point(c(-78.144, 38.895)), crs = 4326)
line <- st_sfc(st_linestring(rbind(c(-78.145, 38.894), c(-78.141, 38.899))),
               crs = 4326)
poly <- st_as_sfc(st_bbox(stems))

data.frame(object   = c("plot centre", "access track", "plot boundary"),
           type     = c(st_geometry_type(pt), st_geometry_type(line),
                        st_geometry_type(poly)),
           vertices = c(nrow(st_coordinates(pt)), nrow(st_coordinates(line)),
                        nrow(st_coordinates(poly))))
```

Now the part that causes trouble, and it is the reason this section comes before
the coordinate systems rather than after them.

A line is defined by its vertices **plus an implicit rule about what lies between
them**, and ISO 19125-1 fixes that rule as linear interpolation. Straight lines.
The catch is that "straight" means straight in whatever number space the
coordinates happen to occupy. The model was designed for a plane, for metres on a
survey grid. Nothing stops you putting degrees of longitude and latitude into it,
and everybody does, at which point straight quietly means straight in a grid of
degrees, which is a map projection you never chose, never declared and cannot
see.

Simple Features is a grammar for shapes. It tells you what a well-formed sentence
looks like. It does not tell you which language you are speaking, so feed it
degrees and it parses perfectly while meaning something other than you intended.

The cost is measurable. Take two vertices on the same parallel, a hundred
kilometres apart at 50 degrees north, and compare the edge Simple Features draws
between them with the path a surveyor would actually walk.

```{r linear-interpolation}
edge <- st_sfc(st_linestring(rbind(c(-126.0, 50), c(-124.6, 50))), crs = 4326)

# the great-circle path between the same two vertices, densified
gc  <- st_cast(st_segmentize(edge, dfMaxLength = 1000), "POINT")
co  <- st_coordinates(gc)
mid <- co[round(nrow(co) / 2), c("X", "Y")]

c(simple_features_midpoint_lat = 50,
  great_circle_midpoint_lat    = round(mid[["Y"]], 5),
  divergence_metres            = round((mid[["Y"]] - 50) * 111320, 1))
```
```{r si-numbers, echo=FALSE}
sf_div <- round((mid[["Y"]] - 50) * 111320, 1)
```
The two paths part company by `r abs(sf_div)` metres at the midpoint. The
vertices agree exactly. The polygon between them does not, and along a long east
to west boundary that difference changes an area. This is why a boundary quoted
as a list of corner coordinates is an incomplete description of a parcel until
you also say what the edges are.

## Two areas disagree

The plot claims `r claimed_ha` hectares. Measure it two ways and neither answer
is that.

```{r area-two-ways}
stems_utm <- st_transform(stems, 32617)   # UTM zone 17N, metres

bbox_ha <- as.numeric(st_area(st_as_sfc(st_bbox(stems_utm)))) / 1e4
hull_ha <- as.numeric(st_area(st_convex_hull(st_union(stems_utm)))) / 1e4

data.frame(
  geometry      = c("bounding box", "convex hull"),
  measured_ha   = round(c(bbox_ha, hull_ha), 2),
  difference_ha = round(c(bbox_ha, hull_ha) - claimed_ha, 2),
  pct_error     = round(100 * (c(bbox_ha, hull_ha) - claimed_ha) / claimed_ha, 1)
)
```
```{r area-numbers, echo=FALSE}
box_pct  <- 100 * (bbox_ha - claimed_ha) / claimed_ha
hull_pct <- 100 * (hull_ha - claimed_ha) / claimed_ha
```
The bounding box gives `r round(bbox_ha, 2)` hectares, `r round(box_pct, 1)` per
cent more than the plot has. The convex hull gives `r round(hull_ha, 2)`.

Two new words, both of them shapes from the previous section. A *bounding box* is
the smallest upright rectangle containing every point, upright meaning its sides
run along the grid axes. A *convex hull* is the polygon you would get by
stretching an elastic band round the outermost stems. Wrapping an upright box
round a tilted rectangle is like putting a tilted book into an upright carton.
The carton has to be bigger than the book, and the more the book is tilted the
more air you pay for.

Be exact about what has been demonstrated and what has not. The angle can be
recovered from the data.

```{r rotation}
rot_fit <- lm(st_coordinates(stems_utm)[, 1] ~ raw$gx + raw$gy)
rot_deg <- abs(atan2(coef(rot_fit)[3], coef(rot_fit)[2]) * 180 / pi)
round(rot_deg, 2)
```
That returns `r round(rot_deg, 2)` degrees, and it is not a finding. It is the
constant `data-raw/prepare-scbi-data.R` used to build the coordinates, recovered
by fitting a line through its own output. The hull's shortfall of
`r round(abs(hull_pct), 2)` per cent is the same story, because the scale factors
in that script, 1.001149 and 0.997105, predict
`r round(25.6 * 1.001149 * 0.997105, 3)` hectares against the
`r round(hull_ha, 3)` measured.

An analysis that recovers the number it was given has shown nothing about the
world, and this book will point out several more, because they are the commonest
way a carbon inventory reassures itself. What the comparison does show is real.
The gap between a box and a hull is set by how far the shape is turned relative
to the grid, which is a fact about geometry rather than about this plot, and any
boundary not running north to south loses area the same way.

Neither figure is a survey error. Both are correct measurements of different
things, so when a claimed area and a mapped area disagree the first question is
not whether the survey is wrong. It is whether the thing being measured is the
boundary or a box drawn round it.

::: callout-tip
## Verification track: what the gap is worth

Skip this box and the chapter still works. It converts the area gap into the
currency a crediting project is paid in.

```{r area-cost}
c_density <- 150          # tonnes of carbon per hectare, a plausible temperate figure
co2_ratio <- 44 / 12      # molecular mass of carbon dioxide over carbon

phantom_ha  <- bbox_ha - hull_ha
phantom_co2 <- phantom_ha * c_density * co2_ratio

round(c(phantom_hectares = phantom_ha, phantom_tonnes_co2e = phantom_co2), 1)
```
`r round(phantom_ha, 2)` hectares that are not there, and about
`r format(round(phantom_co2), big.mark = ",")` tonnes of carbon dioxide
equivalent that could be claimed on them, on a plot you can walk across in ten
minutes and before a single tree has been mismeasured.

Read that as an order of magnitude and no finer. The carbon density is a
plausible temperate figure rather than a measured one, and the hectares rest on
the reconstructed georeferencing declared at the top of this chapter.
:::

## What coordinates claim

A coordinate is not a fact about the ground. It is a claim about a model of the
Earth **and about a date**, and it is only as good as both.

The number 38.9 means nothing alone. It means something once you say it is
degrees of latitude on WGS 84, a particular mathematical model of the Earth's
shape. A *coordinate reference system* is that model paired with the rules for
turning positions on it into numbers, and every one in common use carries a code
from the EPSG registry, a public catalogue that now conforms to ISO 19111:2019.

Three kinds cover almost all forest carbon work. **Geographic** systems give
latitude and longitude on an ellipsoid, and that is the storage and exchange
convention. **Projected** systems give a flat grid in metres, and that is for
display and for area computation. **Geocentric** systems give three distances
from the Earth's centre, and that is what a satellite receiver actually solves
before anything converts it. Errors enter at the junctions between these, not
inside any one of them.

```{r crs-lookup}
codes <- c(4326, 32617, 3005, 3157)
data.frame(
  epsg = codes,
  name = vapply(codes, function(x) st_crs(x)$Name, character(1)),
  units = vapply(codes, function(x) {
    u <- st_crs(x)$units_gdal; if (is.null(u)) NA_character_ else u }, character(1))
)
```
The date is the half almost every dataset omits, and it is not decoration. The
ground moves. On the Pacific coast of North America the crust drifts one to two
centimetres a year relative to the whole Earth, so a latitude and longitude
without a date is incomplete in the way a share price without a date is
incomplete. Two dates are involved and confusing them is common. The *frame
reference epoch* is when the reference frame itself is defined, and ITRF2014 has
one of 2010.0. The *coordinate epoch* is when your particular measurement was
valid, so a post surveyed in June 2016 has a coordinate epoch of 2016.47.

That is also why two national frames drift apart. A *dynamic* frame is fixed to
the whole Earth, so the coordinates of a fixed post change over time. A *static*
frame moves with the tectonic plate, so those coordinates stay roughly constant
within the plate and drift relative to the globe. The International Terrestrial
Reference Frame is dynamic and the Canadian realisation of the North American
Datum of 1983 is plate-fixed, they differ by more than a metre on the west coast,
and the gap widens every year.

## Conversion and transformation

These two words are used interchangeably in speech and they name completely
different operations. ISO 19111:2019 draws the line precisely, and it is the most
useful distinction in this chapter.

A **conversion** changes the coordinate system and not the datum. Latitude and
longitude on WGS 84 to Universal Transverse Mercator zone 17 north on WGS 84 is a
conversion. It is defined by an exact formula, it reverses to the limits of
floating point arithmetic, and it has no uncertainty of its own. Nothing moves on
the ground.

A **transformation** changes the datum. The Canadian realisation of the North
American Datum of 1983 to WGS 84 realisation G1762 is a transformation. It is
estimated from observations, so it carries a published accuracy. It is not
unique, so several valid transformations exist between the same pair of datums
and they give different answers. It often depends on a grid shift file, an
external data product with its own version. Things move on the ground, by metres.
Both are *coordinate operations* when you need one word for the pair.

The practical test is one line. **If the operation has an accuracy figure
attached it is a transformation, and you must record which one you used. If it
does not, it is a conversion, and it is not your problem.**

```{r conversion-vs-transformation}
# A hundred stems is enough, and st_distance needs by_element or it builds
# a full n by n matrix, which on 38,000 points does not finish.
sample <- stems[seq(1, nrow(stems), length.out = 100), ]

# A conversion. Same datum, new grid. Round-trip and see what is lost.
back <- st_transform(st_transform(sample, 32617), 4326)
conv_max_m <- max(st_distance(sample, back, by_element = TRUE))

# A transformation. Different datum. Coordinates move.
nad83 <- st_transform(sample, 4269)          # NAD83 geographic
trans_max_m <- max(st_distance(sample, st_transform(nad83, 4326),
                               by_element = TRUE))

round(c(conversion_roundtrip_m = as.numeric(conv_max_m),
        transformation_shift_m = as.numeric(trans_max_m)), 6)
```
```{r cvt-numbers, echo=FALSE}
cm <- as.numeric(conv_max_m); tm <- as.numeric(trans_max_m)
```
The conversion round-trip loses `r format(signif(cm, 2), scientific = TRUE)`
metres, which is floating point noise. Whether the transformation moved anything
depends on which operation the software chose, and that is the question to ask.
Not whether one was used, but which one.

## Horizontal accuracy

Now the quantity this chapter is named for, and the reason the previous two
sections had to come first.

WGS 84, as most software resolves it, is not a datum at all. It is an
**ensemble**, defined in ISO 19111:2019 as a set of successive realisations
treated as interchangeable for approximate work, and the register states its
accuracy inside the definition. Do not take that on trust. Print what your own
machine returns, because the answer changes between library versions and that is
itself the point.

```{r ensemble}
wkt <- st_crs(4326)$wkt
members <- gsub('MEMBER\\["|"$', "",
                regmatches(wkt, gregexpr('MEMBER\\["[^"]+"', wkt))[[1]])
accuracy <- regmatches(wkt, regexpr("ENSEMBLEACCURACY\\[[0-9.]+\\]", wkt))

length(members)
members
accuracy
sf_extSoftVersion()[c("GEOS", "GDAL", "proj.4")]
```
```{r ensemble-numbers, echo=FALSE}
n_mem <- length(members)
proj_v <- unname(sf_extSoftVersion()["proj.4"])
```
`r n_mem` realisations, treated as one, with the accuracy figure printed above
and PROJ `r proj_v` doing the resolving. Every coordinate in the course dataset
inherits that.

ISO 19157, the standard on data quality, gives the vocabulary for saying so
without either accepting or rejecting the data. Positional accuracy is not one
quantity but several. **Absolute external accuracy** is how far the positions sit
from truth. **Relative internal accuracy** is how well the vertices agree among
themselves. The ensemble damages the first and leaves the second untouched, which
is why the honest statement about a boundary like this one is that it is
internally excellent and externally uncertain to about two metres, rather than
that it is good or bad.

The same standard lets you qualify a number you cannot fully stand behind.
*Metaquality* is information about the quality of your quality statement, and it
has three elements, confidence, representativity and homogeneity. Report the
accuracy figure, attach a confidence statement naming the unresolved ensemble,
and use homogeneity where a boundary was captured across several seasons on
different base stations.

What ISO 19157 will not give you is a threshold, and it says so in its scope.
Acceptable quality lives in a *conformance quality level*, a number stated in a
data product specification or in a user requirement, so the registry or the
verification body has to set it because the standard has declined to. Say that
plainly, because it is the shape of the whole problem. The reproducibility
contract is not a coordinate reference system declaration. It is a specification
with numbers in it.

::: callout-tip
## Verification track: writing it up

ISO 19011, the auditing standard, asks four questions of the metadata and not
only of the geometry. Is the information complete, correct, consistent and
current? A projection file carrying no epoch fails "complete". Clause 6.4.7
allows only information that can be verified as evidence, so a coordinate whose
realisation is unknown is low-verifiability evidence and should be recorded as
such rather than silently recomputed.

The fallback when positional accuracy cannot be measured directly is *lineage*,
the record of where the data came from and what was done to it. ISO 19157 routes
lineage to ISO 19115 as metadata and calls it general, non-quantitative and
illustrative, and it gives lineage as the example of indirect evaluation. So the
standard sends you to lineage exactly when you need it most, and lineage is the
first thing a shapefile export destroys, because that format records the system
the data is in now and nothing about where it has been.
:::

## Three surfaces

Everything above reduces to three rules, stated last because they are
conclusions. Globe for geometry, ellipsoid for measurement, plane for pictures.

```{r three-surfaces}
hull <- st_convex_hull(st_union(stems))

sf_use_s2(TRUE);  sphere <- as.numeric(st_area(hull)) / 1e4
sf_use_s2(FALSE); ellips <- as.numeric(st_area(hull)) / 1e4
sf_use_s2(TRUE)

data.frame(surface = c("sphere, radius 6371010 m", "ellipsoid, geodesic",
                       "plane, UTM zone 17N"),
           area_ha = round(c(sphere, ellips, hull_ha), 4))
```
```{r surface-numbers, echo=FALSE}
s2_rel  <- abs((ellips - sphere) / sphere)
box_rel <- abs(bbox_ha - hull_ha) / hull_ha
```
**Do the geometry on the globe.** Predicates, overlays and indexing on longitude
and latitude are computed with spherical geometry, which in R is the `s2` engine
`sf` uses by default. The reason is not that a sphere is more accurate than an
ellipsoid. It is that a sphere has no edges and no singular points, so a polygon
crossing the antimeridian is ordinary and the poles behave like everywhere else,
and exact spherical predicates do not manufacture the thin false slivers a planar
intersection of two nearly coincident boundaries produces.

**Measure on the ellipsoid.** Areas and distances reported as numbers are
computed geodesically, on the shape that fits the planet, which is the middle row
above.

**Project only to draw.** A plane is for pictures and for legacy area
computation, and it is the row that carries the distortion.

Now rank them, because the instinct on meeting a new source of error is to fear
all of them equally. The sphere and the ellipsoid differ by a relative
`r format(signif(s2_rel, 2), scientific = TRUE)`. The bounding box and the convex
hull differed by `r round(100 * box_rel, 1)` per cent, roughly
`r format(round(box_rel / s2_rel), big.mark = ",")` times larger. Choosing the
wrong shape for your boundary cost a hectare and a half. Choosing the wrong Earth
model cost a few square metres.

So the closing rule. A position is a claim about a geometry, a datum, an epoch
and an operation, and an area reported without all four is not reproducible
whatever its magnitude. The next chapter takes the same question to attributes
and joins, and asks what a value attached to one of these shapes actually refers
to.

## Exercises

1. Build a LineString between two vertices 200 kilometres apart at 60 degrees
   north and repeat the midpoint comparison in this chapter. Report the
   divergence in metres and say how it scales with latitude.
2. Reproject the stems to a geographic system other than WGS 84 and back again.
   Report the largest distance any stem moved, and say whether the operation was
   a conversion or a transformation.
3. Rotate the plot outline through every angle from 0 to 45 degrees and plot the
   bounding box area against the angle. At what rotation is the box a third
   larger than the plot, and what is the maximum?
4. Print the WGS 84 ensemble definition on your own machine and compare the
   member count and accuracy with the values printed in this chapter. State your
   PROJ and GDAL versions beside the answer.
5. Write two sentences about the course dataset's boundary in the vocabulary of
   ISO 19157, one on absolute external accuracy and one on relative internal
   accuracy, and attach a confidence statement.
6. *(Applied.)* A project submits a boundary as a list of corner coordinates in
   degrees, with no epoch and no statement of what lies between the corners.
   List what you cannot compute from it, and say which single piece of missing
   information you would ask for first.

::: callout-note
## Open decisions

**Confirm the ISO 19157 edition before submission.** The working notes cite
ISO 19157:2013 and flag the citation for checking. A later edition exists and the
clause numbering may have moved, so the reference is verified against the
standard itself rather than against the notes.

**Chapter 2 has not been checked against the accepted proposal's audience.** The
two-track structure applied here, main text free of registry dependence with
verification material in tinted boxes, has not been applied to the reflectance
chapter.
:::
