---
title: "Coordinate Systems and the Hectare That Moves"
subtitle: "Chapter 1 skeleton, 4 September 2026"
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

::: callout-important
## Skeleton, not a draft

Section headings, the argument each section has to make, and the code that
carries it. Prose is written only where the argument is settled. Two decisions
are open and are marked where they bite. Brief at
`publishing/chapter-briefs/01-positional-accuracy.md`.
:::

```{r setup}
library(sf)
library(dplyr)
```

## Finding a system

*Argument. A coordinate is a claim about a model of the Earth, not a fact about
the ground. Kept short and concrete: look one up, read one, change one, and know
why you must.*

```{r first-look}
stems <- read.csv("../data/scbi_stems.csv") |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

st_crs(stems)$epsg
st_crs(stems)$Name
st_bbox(stems)
```

*Beat: the bounding box should land over Front Royal, Virginia, near 78.1 west
and 38.9 north. If it landed at the origin the coordinates were never
georeferenced, and the map would look fine on its own axes.*

```{r reproject}
stems_utm <- st_transform(stems, 32617)   # UTM zone 17N, metres
st_bbox(stems_utm)
```

*Beat: epsg.io as the lookup. What the `.prj` sidecar holds, which is one system
in well-known text, the present one, and never a history.*

## The shape underneath

*Argument. Every planar measurement is a projection of a curved surface onto a
flat one, with a distortion you chose. The geoid is the visual aid that carries
the whole chapter. Ellipsoid, geoid and terrain as three different surfaces, and
why height is the harder coordinate.*

*Beat, one page maximum. Eratosthenes with a stick and a well. Latitude solvable
from a horizon and a table, longitude waiting on a clock that kept time at sea.
The Board of Longitude paid for that clock. This is the fastest route into what
a datum is and it earns its space, but it does not get more than a page.*

*Beat: Clarke 1866 to NAD 83 to WGS 84, and the metres that moved with each. The
WGS 84 ensemble carries a stated accuracy of about two metres that almost nobody
reads.*

## Measuring an area

*Argument. This is the applied core. A claimed area and a mapped area disagree,
and the disagreement is the lesson.*

```{r area-two-ways}
claimed_ha <- 25.6

bbox_ha <- as.numeric(st_area(st_as_sfc(st_bbox(stems_utm)))) / 1e4
hull_ha  <- as.numeric(st_area(st_convex_hull(st_union(stems_utm)))) / 1e4

data.frame(
  geometry      = c("bounding box", "convex hull"),
  measured_ha   = round(c(bbox_ha, hull_ha), 2),
  difference_ha = round(c(bbox_ha, hull_ha) - claimed_ha, 2),
  pct_error     = round(100 * (c(bbox_ha, hull_ha) - claimed_ha) / claimed_ha, 1)
)
```

*Beat: the plot grid runs about 1.8 degrees off UTM north. A bounding box is
axis-aligned by construction so it wraps a tilted rectangle inside a larger
upright one. The hull follows the stems and recovers the nominal area. Neither
figure is a survey error. When a claimed area and a mapped area disagree, the
first question is not whether the survey is wrong but whether the thing being
measured is the boundary or a box drawn round it, because that choice alone moved
the answer by more than a hectare.*

## Checking a submission

*Argument. Area non-compliance is the most common finding an auditor raises and
it is raised before any carbon is calculated.*

::: callout-warning
## Data not yet committed

This section runs on the a public dataset ACR submission in
`training/area-verification`, which is outside this repository. Before this chapter can
render offline the following have to be copied into `data/` and cleared for
publication: the submitted boundary, the strata, the harvest units, and the
`slivers` and `artifacts` outputs. Anonymise the client if TÜV SÜD clearance is
not in hand.
:::

```{r area-check, eval=FALSE}
# Sequence as run in training/area-verification/area-verification.qmd
sf_use_s2(FALSE)                                  # configure the geometric plane
st_layers("../data/submission.gpkg")              # what arrived
bdry <- st_read("../data/submission.gpkg", "boundary", quiet = TRUE)

all(st_is_valid(bdry))                            # validate before measuring
bad  <- bdry[!st_is_valid(bdry), ]                # extract the artefacts
bdry <- st_make_valid(bdry)

bdry$area_ha <- as.numeric(st_area(bdry)) / 1e4   # hectares, by feature
```

*Beat: the workflow is set the plane, unpack, summarise the layers, validate,
extract the invalid artefacts, calculate hectares, break the total down by
boundary, stratum and harvest unit, then test the difference against materiality.
Area and overlap operations on an invalid polygon return nonsense without
erroring, which is why validation precedes measurement and not the reverse.*

## Overlap and ownership

*Argument. Nine submissions in ten fail this on the first pass. Any positive
intersection is land claimed twice.*

```{r overlap}
e  <- st_bbox(stems_utm)
x0 <- unname(e["xmin"]); y0 <- unname(e["ymin"])
a <- st_as_sfc(st_bbox(c(xmin = x0,       xmax = x0 + 250,
                        ymin = y0,        ymax = y0 + 400), crs = 32617))
b <- st_as_sfc(st_bbox(c(xmin = x0 + 200, xmax = x0 + 450,
                        ymin = y0,        ymax = y0 + 400), crs = 32617))

round(as.numeric(st_area(st_intersection(a, b))) / 1e4, 2)   # ha claimed twice
```

*Beat: test against federal land, adjacent private parcels, national parks and
neighbouring ownership. Then the set-operation trap. The valid area is the ground
surviving every exclusion, which is an intersection of the residuals. Chaining
unions instead readmits every excluded parcel, because a pixel excluded by the
slope mask survives in the roads residual, and the belt comes out larger in a
direction that suits the project.*

## The inheritance path

*Argument. The chain of reprojections is lost because nobody writes it down, not
because the container cannot hold it. That is the defensible version.*

*Beat: the shapefile family has no lineage mechanism at all. GeoPackage is an OGC
encoding standard, not an ISO one, and it fixes the real shapefile defects, the
two gigabyte ceiling, the ten-character field names, the 255-column limit and
polygon against multipolygon ambiguity. Its `gpkg_metadata` table can hold ISO
19115 records, and ISO 19115-1:2014 does define a lineage element. But lineage is
only there if somebody writes it, and almost nobody does.*

## The library underneath

*Argument. Two analysts measure the same polygon and get different areas. A
reported area without its software environment stated is not reproducible.*

```{r versions}
sf_extSoftVersion()[c("GEOS", "GDAL", "proj.4")]
```

*Beat: ArcGIS runs its own engine over GDAL. GDAL ships in many versions. PROJ 6,
released 2019, moved the canonical representation from proj4 strings to WKT2,
introduced the datum ensemble and made late-binding the default, which is why the
same request can resolve differently across versions. This is why the
area-verification workflow ends on a runtime log rather than on a number.*

## Plot centres

*Argument. Positional accuracy connects to the carbon number here, and it scales.*

*Beat: permanent sample plots in North America are marked with rebar, and public
inventory programmes use markers built to survive decades. A crew that cannot
relocate a centre and sets up three metres away measures a different set of
trees. Because plot means are extrapolated across the whole project, that error
does not stay on the plot, it multiplies by the expansion factor.*

## Assessing a fix

*Argument. Report a positional accuracy as a distribution, not as a single
number.*

*Beat: GPS is one constellation, GNSS is the family. What real-time kinematic
correction does and what it costs. How to run a check against a known control
point. Source material in `GPS-training-material`, `survey-tools` and the
KoBoToolbox deployment.*

## What ground truth costs

*Argument. Interpreted ground reference points are the most valuable and least
glamorous data product in remote sensing.*

*Beat: the Global Land Cover Estimation training database holds close to two
million training units for 1984 to 2020, at 30 metres, across seven primary and
twelve secondary classes, built at Boston University under Woodcock and Friedl
and published in Scientific Data in 2023. Why it took years and a large team, and
why that effort is the foundation of every accuracy statement later in this book.*

## The plane you compute on

*Argument. Closing caveat and the hinge into the rest of the book.*

```{r s2}
sf_use_s2(TRUE)
sph <- as.numeric(st_area(st_convex_hull(st_union(stems)))) / 1e4
sf_use_s2(FALSE)
ell <- as.numeric(st_area(st_convex_hull(st_union(stems)))) / 1e4

data.frame(engine = c("s2, sphere", "GeographicLib, ellipsoid"),
           area_ha = round(c(sph, ell), 4),
           rel_diff = c(NA, round((ell - sph) / sph, 8)))
```

*Beat: a traditional geographic information system represents the world as a flat
projection, the way a paper map does. S2 represents it on a sphere, the way a
globe does, and the name is the notation for the unit sphere. No seams, so a
polygon crossing the antimeridian is ordinary. No singularities, so the poles
behave like everywhere else. What it buys is fast in-memory indexing, robust
constructive operations through snap rounding, and efficient nearest-neighbour
queries.*

*Beat, and this is the honest framing. The area difference is around 6.4e-05, so
two thirds of a hectare on ten thousand. That is small against the 6.6 per cent
the bounding box and the hull disagreed by earlier in this chapter, and saying so
teaches the reader to rank error sources rather than fear them equally. The real
gain is that a planar intersection of two nearly coincident boundaries
manufactures slivers, which is what `slivers layer` in the area-verification outputs
contains, and exact spherical predicates do not.*

*Closing rule: an area reported without its geometry engine, its library versions
and its geometry setting is not reproducible, whatever its magnitude. And the
question of which plane you compute on, and what support the answer has, is the
thread this book follows to the cube. See `publishing/book-spine.md`.*

## Exercises

*To be written. Six to eight, each closing on a number a verifier would accept.*

::: callout-note
## Open decisions

**Does this chapter split?** Ten sections runs 35 to 45 printed pages, at the top
of the series range. The clean line is after "The library underneath", leaving
projection and area as chapter 1 and plot centres, GNSS and ground truth as
chapter 2.

**Which two ebooks?** The author wants the geospatial data objects section
aligned to two ebooks that present them without overwhelming the reader. Neither
was supplied. Names or paths needed.
:::
