---
title: "Reading Reflectance, and What an Index Can Support"
subtitle: "Chapter 2 redraft, 6 September 2026"
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

::: callout-note
## Redraft

A redraft of `02-reflectance.qmd` under the `writing-scientific-books` skill,
kept beside the original for comparison. Every code chunk is unchanged, so every
computed number is identical. The prose is rewritten to introduce each term in
the sentence that first uses it and to cut what the reader does not need.

One substantive correction. The disturbance signatures section previously
reported the beetle classification of Murphy et al. (2026) as having a
commission error of 0.750 on the red-attack class and a kappa of 0.166. Table 8
of that paper reports the reverse, a kappa of 0.750 and a commission error on
red attack of 0.198, and the paper calls those error rates acceptable. The
section is rewritten against the published tables, and the argument it was
making is moved to the fire results, where the tables do support it.
:::

```{r setup}
library(terra)

# Two dates over the same ground, ten Sentinel-2 Level-2A bands each, stored as
# signed integers scaled by 10,000. The division recovers reflectance.
clear <- rast("../data/rondonia_s2_2022-07-16.tif") / 1e4
smoke <- rast("../data/rondonia_s2_2022-09-02.tif") / 1e4

legend <- read.csv("../data/rondonia_legend.csv")
bands  <- read.csv("../data/sentinel2_bands.csv")

# The published nine-class map, projected onto the tile grid. It is stored in
# EPSG:3857 at about 20.26 m and the scene is in EPSG:32720 at 20 m, so a
# resampling happens here whether or not anyone writes it down. Nearest
# neighbour, because the values are labels and a label has no mean.
lulc  <- rast("../data/rondonia_class_2022.tif")
class <- project(lulc, clear, method = "near")
names(class) <- "class"
```
## Introduction

A satellite does not measure forest. It counts photons landing in a detector
over a fixed slice of wavelength, turns that count into a number, and writes the
number to a file. Everything between that file and a reported tonne of carbon is
inference, and this chapter covers the first two steps of it, the conversion of
a count into a reflectance and the compression of several reflectances into one
index.

The chain is worth walking once in full, because the rest of the book assumes
it. Sunlight comes down through the atmosphere, and some of it scatters off air
and dust on the way. What arrives at the canopy meets leaves, bark, soil and
water. Some of it goes back up, passes through the atmosphere a second time, and
reaches the sensor. The sensor records the brightness arriving at the top of the
atmosphere. A processing chain then works backwards and asks what the ground
must have reflected for that brightness to arrive, which needs a model of the
atmosphere on that day over that place.

The output of that calculation is called surface reflectance, and the word to
hold onto is *estimate*. It is a *retrieval*, meaning a quantity worked out
backwards from something else that was measured, and it carries an uncertainty
of its own. It is not a measurement of the ground. No part of forest carbon
accounting treats it as anything else.

The reason to persevere is that optical data is the only record long enough to
matter. Landsat 1 launched in July 1972 and the record has not stopped since,
which makes it the longest continuous measurement of the land surface that
exists. Almost every forest carbon project on Earth sets its historical baseline
from that archive, because nothing else reaches back far enough. The archive is
imperfect in the specific ways this chapter measures. It is also irreplaceable.

The scene used throughout is a 24 kilometre window on the Madeira river in
Rondonia, in the Brazilian Amazon, on two dates in 2022. It holds intact forest,
riparian forest, open water, wetland, and the rectangular clearings of the
Amazon frontier, so every argument below is made on real ground rather than on a
surface built to prove a point.

```{r scene}
plotRGB(clear[[c("B04", "B03", "B02")]], stretch = "lin",
        main = "Rondonia, 16 July 2022, true colour")
```
## Spectral response

All of optical vegetation monitoring rests on one physical contrast, and the
contrast is not a convention that could have been chosen differently.

Chlorophyll absorbs strongly in the red, near 665 nanometres, because that is
where the pigment does its work. Near infrared light at 842 nanometres meets no
pigment that absorbs it at all, so it passes into the leaf and bounces around
inside. The interior of a leaf is a stack of cell walls with air pockets between
them, and light crossing from wet wall to dry air and back again scatters at
every boundary, the way a glass of water full of bubbles looks white while the
water and the glass are both clear. Most of that light comes back out.

So a living leaf is dark in the red and bright in the near infrared, and a dead
or absent one is neither. The gap between those two bands is a measurement of
living tissue.

Sentinel-2 records ten bands relevant here. The table is read from the product
specification rather than typed from memory.

```{r bands}
knitr::kable(bands, col.names = c("Band", "Name", "Centre (nm)",
                                  "Width (nm)", "Native cell (m)"))
```
The *signature* of a land cover, meaning its average reflectance in every band,
is what separates one cover from another. Three of them carry the whole
argument.

```{r signatures}
d <- as.data.frame(c(clear, class), na.rm = FALSE)
d <- d[complete.cases(d), ]
d$cover <- legend$cover[match(d$class, legend$id)]

sig <- aggregate(d[names(clear)], list(cover = d$cover), mean)
knitr::kable(sig, digits = 3)
```
```{r signature-numbers, echo=FALSE}
g <- function(cov, b) sig[[b]][sig$cover == cov]
f_red <- g("Forest", "B04");  f_nir <- g("Forest", "B08")
s_red <- g("Clear cut, bare soil", "B04"); s_nir <- g("Clear cut, bare soil", "B08")
w_red <- g("Water", "B04");   w_nir <- g("Water", "B08")
```
Intact forest reflects `r round(f_red, 3)` in the red and `r round(f_nir, 3)` in
the near infrared, a ratio of `r round(f_nir / f_red, 1)`. Cleared ground with
bare soil reflects `r round(s_red, 3)` and `r round(s_nir, 3)`, a ratio of
`r round(s_nir / s_red, 1)`, because soil is brighter in the red and there is no
longer a canopy scattering the near infrared. Open water reflects
`r round(w_red, 3)` in the red and `r round(w_nir, 3)` in the near infrared, so
its near infrared sits *below* its red, because water swallows near infrared
almost completely. That inversion is why water takes negative values in every
vegetation index, and it is the only reason water is easy.

Read across the rest of the table and the two shortwave infrared bands do
something the visible bands do not. Forest reflects `r round(g("Forest","B11"), 3)`
at 1,610 nanometres and clear-cut regrowth reflects
`r round(g("Clear cut, regrowth","B11"), 3)`, a gap of
`r round(g("Clear cut, regrowth","B11") - g("Forest","B11"), 3)`, while their red
reflectances differ by only
`r round(g("Clear cut, regrowth","B04") - f_red, 3)`. The shortwave infrared is
absorbed by liquid water, so it reports on how wet the canopy is, and a young
regenerating stand holds less water than a closed mature one. Hold that
observation. The section on class separability turns it into the chapter's first
result.

## Reflectance retrieval

The files above are Level-2A products, which means an atmospheric correction has
already run on them. It is worth knowing what it corrected and what it did not.

A Level-1C product holds top of atmosphere reflectance, which is everything the
atmosphere did on the way down and on the way up, mixed in with the ground. The
correction to Level-2A estimates the atmosphere's contribution and subtracts it.
Doing that needs two quantities for the day and place in question, the amount of
haze in the air and the amount of water vapour, either retrieved from the scene
itself or taken from a weather forecast. It is like reading a page through
frosted glass and trying to subtract the glass. Where the estimate of the glass
is good, the result approximates what an instrument held just above the canopy
would have read. Where it is not, and the smoky scene later in this chapter is
one such case, the leftover atmosphere stays in the number and cannot be told
apart from the ground.

There is a second and duller way the same number goes wrong, and it has cost
projects more than the physics has.

Sentinel-2 processing baseline 04.00 came into operation on 25 January 2022.
Among other changes it shifted the stored range of values by a constant that
differs from band to band, so that a negative surface reflectance, which is
physically impossible but statistically ordinary over dark water, could be
stored rather than clipped to zero. From that date the correct recovery is

$$\rho = \frac{\mathrm{DN} + \mathrm{BOA\_ADD\_OFFSET}}{\mathrm{QUANTIFICATION\_VALUE}}$$

and not $\mathrm{DN} / 10000$. The offset is carried in the product metadata and
is read, never assumed. Code written before 2022 and run afterwards subtracts
nothing, and the consequence is not subtle.

```{r offset}
dn <- rast("../data/rondonia_s2_2022-07-16.tif")

nd <- function(r, a, b) (r[[a]] - r[[b]]) / (r[[a]] + r[[b]])
forest_only <- function(r) mask(r, class == 4, maskvalues = FALSE)

applied <- dn / 1e4              # the offset was subtracted upstream
missed  <- (dn + 1000) / 1e4     # an offset of 1,000 DN never subtracted

off <- sapply(list(NDVI = c("B08", "B04"), NDMI = c("B08", "B11"),
                   NBR  = c("B08", "B12")),
  function(p) c(
    applied = global(forest_only(nd(applied, p[1], p[2])), "mean", na.rm = TRUE)[1, 1],
    missed  = global(forest_only(nd(missed,  p[1], p[2])), "mean", na.rm = TRUE)[1, 1]))

off <- rbind(off, difference = off["missed", ] - off["applied", ])
round(off, 3)
```
Mean forest vegetation index falls from `r round(off["applied","NDVI"], 3)` to
`r round(off["missed","NDVI"], 3)`, a shift of
`r round(abs(off["difference","NDVI"]), 3)`, on ground where nothing happened.
That is larger than most of the real differences a disturbance study is looking
for.

The lesson is not the number, which depends on an offset read from metadata. The
lesson is that a processing convention changed on a stated date, so any time
series crossing that date without accounting for it carries a step that looks
exactly like an event.

## Band support

The previous chapter asked which plane the arithmetic happens on. Asked of a
sensor, the same question becomes what area of ground one number describes. That
area is the *support* of the value, and it is the second half of this book's
spine.

Sentinel-2 does not answer it once. It collects the blue, green, red and broad
near infrared bands at a *ground sample distance* of 10 metres, meaning each
number covers a 10 metre square of ground, and it collects the three red edge
bands, the narrow near infrared band and the two shortwave infrared bands at 20
metres. The files here sit on a single 20 metre grid, so six bands were
coarsened to get there and none was sharpened. An index mixing a 10 metre band
with a 20 metre band is not comparing like with like until somebody resamples,
and once somebody resamples the mismatch is hidden rather than fixed.

```{r support-table}
knitr::kable(aggregate(band ~ gsd_m, bands, paste, collapse = ", "),
             col.names = c("Native cell (m)", "Bands acquired at it"))
```
The consequence at a boundary can be measured. Aggregate the scene to 60 metre
cells, three by three, and ask how many of those cells are entirely forest,
entirely something else, or partly both.

```{r mixed-cells}
is_forest <- as.numeric(class == 4)
frac60 <- aggregate(is_forest, fact = 3, fun = "mean")
v <- values(frac60); v <- v[!is.na(v)]

round(100 * c(pure_forest = mean(v == 1),
              pure_other  = mean(v == 0),
              mixed       = mean(v > 0 & v < 1)), 1)
```
```{r mixed-numbers, echo=FALSE}
pct_mixed <- round(100 * mean(v > 0 & v < 1), 1)
```
`r pct_mixed` per cent of 60 metre cells straddle the forest boundary. Their
reflectance describes neither forest nor clearing. Stir a tin of green paint
into a tin of brown and the result is a real colour that matches neither tin,
and no amount of looking at it tells you the proportions you started with.
Whatever value an index returns for those cells is an average of two things that
were never averaged in nature.

This is the whole of the mixed pixel problem, and it is why a 30 metre Landsat
cell and a 10 metre Sentinel-2 cell are not two measurements of the same ground
at different quality. They are measurements of different things.

The order of operations matters for the same reason. An index is a ratio, and
the ratio of averages is not the average of ratios.

```{r order-of-operations}
ndvi20 <- nd(clear, "B08", "B04")

first_average <- nd(aggregate(clear, fact = 3, fun = "mean"), "B08", "B04")
first_index   <- aggregate(ndvi20, fact = 3, fun = "mean")

difference <- first_average - first_index
names(difference) <- "NDVI difference"
global(abs(difference), quantile, probs = c(0.5, 0.95, 0.99, 1), na.rm = TRUE)
```
```{r order-numbers, echo=FALSE}
q <- global(abs(difference), quantile, probs = c(0.5, 1), na.rm = TRUE)
```
The median disagreement is `r format(round(q[1, 1], 4), scientific = FALSE)`,
which is nothing, and the maximum is `r round(q[1, 2], 3)`, which is a great
deal. The obvious guess is that the large disagreements sit at boundaries, where
the two source bands move in opposite directions across the cell. Test it rather
than assert it.

```{r order-where}
a <- values(abs(difference)); m <- values(frac60 > 0 & frac60 < 1)
ok <- !is.na(a) & !is.na(m); a <- a[ok]; m <- as.logical(m[ok])

worst <- a >= quantile(a, 0.99)
round(c(mixed_share_of_scene = 100 * mean(m),
        mixed_share_of_worst = 100 * mean(m[worst]),
        median_mixed         = median(a[m]),
        median_pure          = median(a[!m])), 4)
```
```{r order-where-numbers, echo=FALSE}
share  <- 100 * mean(m)
worsts <- 100 * mean(m[worst])
```
On the cells where both quantities are defined, which is why the share differs
slightly from the figure above, mixed cells are `r round(share, 1)` per cent of
the scene and `r round(worsts, 1)` per cent of the worst one per cent of
disagreements, so they are over-represented by a factor of
`r round(worsts / share, 1)`. They do not account for all of it, so the obvious
guess was too strong. The typical disagreement in a mixed cell is about twice
the typical disagreement in a pure one.

Which order is correct depends on what the number is for. Whether it matters is
settled. It matters most exactly where you are most likely to be looking.

## Index construction

A spectral index compresses several bands into one number. The normalised
difference is the standard form,

$$\mathrm{ND}(a, b) = \frac{\rho_a - \rho_b}{\rho_a + \rho_b}$$

which is bounded between -1 and 1 and, being a ratio, cancels part of any effect
that raises or lowers both bands together. Weigh two stones on a scale reading
ten per cent heavy and each weight is wrong, but their ratio is right. That
cancellation is the reason indices are used at all, and the section on
atmospheric sensitivity shows where it stops working.

The family is one function applied to different band pairs, and writing it that
way makes the point that its members differ only in which physics they ask
about.

```{r indices}
# Naming the arguments of c() would return a list rather than a SpatRaster, so
# the layers are combined unnamed and named afterwards.
indices <- function(r) {
  out <- c(nd(r, "B08", "B04"),   # chlorophyll absorption in the red
           nd(r, "B08", "B11"),   # liquid water in the canopy
           nd(r, "B08", "B12"),   # char and water loss after fire
           nd(r, "B08", "B05"),   # the red edge, sensitive where NDVI saturates
           nd(r, "B08", "B03"),   # chlorophyll absorption in the green

           # The enhanced vegetation index is not a normalised difference. Its
           # blue term is there to resist aerosol scattering and its soil term
           # to resist background brightness.
           2.5 * (r[["B08"]] - r[["B04"]]) /
             (r[["B08"]] + 6 * r[["B04"]] - 7.5 * r[["B02"]] + 1))

  names(out) <- c("NDVI", "NDMI", "NBR", "NDRE1", "GNDVI", "EVI")
  out
}

i_clear <- indices(clear)
plot(i_clear[[c("NDVI", "NDMI", "NBR", "NDRE1")]], range = c(-0.6, 1))
```
Look at the four panels together rather than at any one of them.

The vegetation index is nearly flat across everything green, because it
saturates. Once a canopy has closed, the red reflectance is already close to
zero, so adding more leaves changes almost nothing, the way pouring water onto a
sponge that is already full does not make it wetter. The moisture index and the
burn ratio keep pulling the clearings away from the forest long after the
vegetation index has stopped. That is not a trick of the display. It is the next
section's result.

## Class separability

Take the two classes a deforestation project has to tell apart and ask whether
an index can do it. Intact forest and clear cut with regrowth are, respectively,
`r round(100 * mean(d$class == 4), 1)` and
`r round(100 * mean(d$class == 6), 1)` per cent of the scene.

```{r separability-dist}
# na.rm is applied after the join, not inside as.data.frame(). The index layers
# carry a few more missing cells than the band stack does, where a normalised
# difference divides by zero, and dropping per layer would misalign the rows.
iv <- as.data.frame(c(i_clear, class), na.rm = FALSE)
iv <- iv[complete.cases(iv), ]
iv$cover <- legend$cover[match(iv$class, legend$id)]

quantiles <- do.call(rbind, lapply(split(iv$NDVI, iv$cover), quantile,
                                   probs = c(0.05, 0.25, 0.5, 0.75, 0.95)))
knitr::kable(round(quantiles, 3))
```
```{r separability-numbers, echo=FALSE}
fnd <- iv$NDVI[iv$cover == "Forest"]
gnd <- iv$NDVI[iv$cover == "Clear cut, regrowth"]
pctile <- round(100 * mean(fnd < median(gnd)), 1)
```
Regrowth has a median vegetation index of `r round(median(gnd), 3)`. Forest has
a median of `r round(median(fnd), 3)` and a fifth percentile of
`r round(quantile(fnd, 0.05), 3)`, meaning one forest cell in twenty falls below
that value. The regrowth median therefore sits at the `r pctile`th percentile of
the forest distribution.

There is no threshold that separates them, because there is no gap to put one
in. A two-year-old clearing grown back to scrub sits spectrally inside the
forest distribution while being, in every sense a project cares about, not
forest.

That can be made precise. Sweep every threshold and take the best.

```{r threshold-sweep}
best_threshold <- function(a, b, n = 400) {
  th <- seq(min(c(a, b)), max(c(a, b)), length.out = n)
  ba <- vapply(th, function(t) (mean(a >= t) + mean(b < t)) / 2, numeric(1))
  i  <- which.max(ba)
  c(threshold = th[i], balanced_accuracy = ba[i],
    regrowth_called_forest = mean(b >= th[i]),
    forest_called_other    = mean(a <  th[i]))
}

sub <- iv[iv$class %in% c(4, 6), ]
sweep <- t(sapply(c("NBR", "NDMI", "NDRE1", "NDVI"), function(nm)
  best_threshold(sub[[nm]][sub$class == 4], sub[[nm]][sub$class == 6])))
round(sweep, 3)
```
```{r sweep-numbers, echo=FALSE}
ba <- round(sweep[, "balanced_accuracy"], 3)
tv <- round(sweep["NDVI", "threshold"], 3)
```
Two things in that table are worth more than the rest of the chapter.

The first is that the best the vegetation index can manage is a *balanced
accuracy* of `r ba["NDVI"]`, meaning the average of how often it gets forest
right and how often it gets regrowth right, counted separately so the larger
class cannot flatter it. That best comes at a threshold of `r tv`, and even
there it calls `r round(100 * sweep["NDVI", "regrowth_called_forest"], 1)` per
cent of regrowth forest. Better than nothing is not the same as fit for purpose.
A project mapping forest by thresholding a vegetation index has a systematic
error it cannot see, because the errors sit in exactly the class the project is
paid to find.

The second is the ranking. The burn ratio reaches `r ba["NBR"]`, the moisture
index `r ba["NDMI"]`, the red edge index `r ba["NDRE1"]` and the vegetation
index `r ba["NDVI"]`. The two shortwave infrared indices beat the two visible
ones, and they beat them because regrowth differs from forest more in canopy
water than in greenness, which is what the band signatures showed earlier. The
index the whole field reaches for first is the worst of the four at the job it
is reached for.

None of this argues against the vegetation index. It argues for choosing a band
pair by the physics of the difference you are trying to see, which is a
decision, and for reporting which one you chose, which almost nobody does.

## Atmospheric sensitivity

Now the second date. On 2 September 2022 the same ground lay under smoke from
biomass burning, which is an ordinary condition in Rondonia in September and not
an exotic failure case.

```{r smoke-rgb}
par(mfrow = c(1, 2))
plotRGB(clear[[c("B04", "B03", "B02")]], stretch = "lin", main = "16 July 2022")
plotRGB(smoke[[c("B04", "B03", "B02")]], stretch = "lin", main = "2 September 2022")
```
Both scenes are Level-2A, so an atmospheric correction ran on both and neither
was withheld. What the correction left behind can be measured band by band.

```{r smoke-bands}
# global() returns a data frame whose row names carry the band, so the means are
# renamed here rather than being indexed by position later.
m1 <- setNames(global(clear, "mean", na.rm = TRUE)[, 1], names(clear))
m2 <- setNames(global(smoke, "mean", na.rm = TRUE)[, 1], names(smoke))

knitr::kable(data.frame(band = names(clear), centre_nm = bands$centre_nm,
                        july = round(m1, 4), september = round(m2, 4),
                        ratio = round(m2 / m1, 2)), row.names = FALSE)
```
```{r smoke-numbers, echo=FALSE}
r_ratio <- round(m2 / m1, 2)
```
Scene-mean blue reflectance is `r r_ratio["B02"]` times its clear-date value.
Green is `r r_ratio["B03"]`, red `r r_ratio["B04"]`, near infrared
`r r_ratio["B08"]` and the longer shortwave infrared band `r r_ratio["B12"]`.

The pattern runs steadily with wavelength and it has a physical cause. Fine
particles scatter short wavelengths far more than long ones, so smoke brightens
the blue end of the spectrum and barely touches the infrared. This is the same
physics that makes the sky blue, working on an instrument that was trying to see
through it.

What that does to an index depends entirely on which bands the index uses.

```{r smoke-indices}
i_smoke <- indices(smoke)

shift <- sapply(names(i_clear), function(nm) c(
  july      = global(forest_only(i_clear[[nm]]), "mean", na.rm = TRUE)[1, 1],
  september = global(forest_only(i_smoke[[nm]]), "mean", na.rm = TRUE)[1, 1]))
shift <- rbind(shift, change = shift["september", ] - shift["july", ])
round(shift, 3)
```
```{r smoke-index-numbers, echo=FALSE}
ch <- round(shift["change", ], 3)
fs <- values(forest_only(i_smoke[["NDVI"]])); fs <- fs[!is.na(fs)]
fc <- values(forest_only(i_clear[["NDVI"]])); fc <- fc[!is.na(fc)]
below_clear <- round(100 * mean(fc < sweep["NDVI", "threshold"]), 1)
below_smoke <- round(100 * mean(fs < sweep["NDVI", "threshold"]), 1)
```
Mean forest vegetation index falls by `r abs(ch["NDVI"])`. The burn ratio, both
of whose bands sit beyond 800 nanometres, moves `r abs(ch["NBR"])`. On identical
ground, where nothing happened between the two dates, one index moved
`r round(abs(ch["NDVI"]) / abs(ch["NBR"]), 0)` times as far as the other.

Put that next to the threshold from the separability sweep. On the clear date,
`r below_clear` per cent of forest cells fall below the best threshold of
`r tv`. On the smoky date, `r below_smoke` per cent do.

```{r threshold-collapse}
c(clear_date = below_clear, smoky_date = below_smoke)
```
A forest map made by thresholding the vegetation index would have called almost
the entire forest non-forest, and it would have done so with no error, no
warning and a plausible-looking output. The index that separated worst also
breaks worst, and the pairing of those two results is what this chapter has to
say.

The enhanced vegetation index deserves its own sentence, because it is the one
in the set built specifically to resist this. Its blue term, with a coefficient
of -7.5, is there to subtract the haze.

```{r evi-failure}
round(shift[, "EVI"], 3)
```
It moved the wrong way, rising `r round(shift["change","EVI"], 3)` on forest.
The correction was calibrated for a modest amount of haze, and under heavy smoke
a large blue value drives the denominator down and inflates the whole ratio. A
correction used outside the range it was built for is not a weaker correction.
It is a different error, and a worse one, because the first kind announces
itself and the second does not.

The operational answer to all of this is masking cloud and haze and then
compositing over a window of dates, which is the subject of the time series
chapter. The answer this chapter can give is narrower and comes first. Know
which bands your index uses, and know that the state of the atmosphere is a
condition on the value in the same way the cell size is. The number describes the
ground and the column of air above it, and the retrieval is the attempt to tell
those two apart.

## Disturbance signatures

Two disturbances dominate the forest carbon landscapes this book is about, and
both were mapped with the indices computed above long before anyone modelled
them. This section says what those signatures are and where they fail. Turning a
signature into a dated, attributed, area-adjusted estimate belongs to later
chapters.

The evidence throughout is Murphy et al. (2026), who worked in the Darkwoods
Conservation Area in the southern Selkirk Mountains of British Columbia and
tested both disturbances against field plots on the same ground. Their tables
are quoted rather than paraphrased, because the useful part is in the per-class
numbers rather than in the headline.

### Insect attack

Mountain pine beetle, *Dendroctonus ponderosae* Hopkins, kills lodgepole pine by
tunnelling into the layer of living tissue just under the bark, and the crown
then fades over the seasons that follow. What matters for a sensor is that the
fading runs through three named stages, and each is a different physical state.

In the green stage the tree has been attacked and its needles are still green.
That is the stage an operational detector would most like to catch and the stage
none of them catches reliably. In the red stage, the first twelve to twenty-four
months after attack, the tree holds red needles while the water in its sapwood
and foliage drains away, so a visible-band index and a shortwave infrared one
both have something to register. In the grey stage the needles and then the
standing dead stems have fallen, and what the sensor sees is the understorey and
the ground.

Which index to use in the red stage is an empirical question rather than a
deduction, and it was answered by measurement. Murphy et al. mapped red-stage
attack from an annual Landsat 5 series with a pre-outbreak scene of 20 August
2003 and a post-outbreak scene of 31 August 2013. They tested the moisture index
against the wetness, greenness and brightness components of the tasselled cap
transform, which is a standard recombination of Landsat bands into three axes
meant to track soil, vegetation and moisture. The moisture index won on
regression against field plots, explaining 0.905 of the variance in the basal
area fraction of beetle-killed trees at a root mean square error of 4.83, with a
cross-validation error ratio of 1.09 that sits below the threshold above which a
model is treated as overfitted.

The classification that followed reached an overall accuracy of 0.824 and a
kappa of 0.750, kappa being the agreement left once the agreement expected by
chance is taken out. Errors on the red-attack class itself were a commission
error of 0.198, meaning about one cell in five called red attack was not, and an
omission error of 0.199, meaning about one in five of the real red attack was
missed. The paper describes those rates as acceptable and selects the moisture
index on them.

That is a good result, and it is worth being clear about what a good result buys
you. One cell in five is fine for mapping where an outbreak is. It is not fine
for a project claiming payment on the difference between two years, because a
twenty per cent false alarm rate on the disturbed class is comparable in size to
the change many projects are trying to detect. The number that decides whether a
map is usable is never the overall accuracy. It is the error on the class you
are being paid to find.

### Fire

Fire is the mirror case and the easier one. Combustion drives the water out of
the canopy and leaves char behind, and both of those push the shortwave infrared
up hard while the near infrared falls, so the normalised burn ratio moves
further and more reliably than any visible-band index.

Murphy et al. tested five candidate indices against 94 burn plots, 9 of them
field plots established immediately after the fire in 2015 and resampled in
2020, and 85 scored from aerial photographs. Every plot was rated with a
composite burn index, a field protocol that scores fire effects across four
layers, the substrate, the ground shrubs, the understorey and the overstorey, on
a scale from 0.00 to 10.00, with below 2.50 counted as no change and above 7.50
as high severity. The burn ratio had the strongest relationship with those field
scores, explaining 0.879 of the variance where the next best index explained
0.584.

Its classification reached an overall accuracy of 0.872 and a kappa of 0.792,
which are the best numbers anywhere in the study. Now look inside them. On the
high severity class the burn ratio missed 4 per cent. On the low severity class
it missed 55 per cent, and of the cells it did call low severity, 22 per cent
were something else.

So the headline figure of 0.872 is carried almost entirely by the classes that
were easy, the unburned ground and the severe burns, and it conceals a class the
index can barely see. Low severity fire is where more than half the events are,
and it is where an index built on canopy water has the least to work with,
because a fire that scorches the understorey and leaves the canopy standing
changes very little of what the sensor is measuring. An overall accuracy is an
average over classes weighted by how common they are, so it rewards a
classifier for getting the easy majority right. Read the per-class errors first
and the overall accuracy last.

One detail of the field design belongs to this book's spine rather than to fire
ecology. Plots were placed at the centre of areas of relatively uniform
disturbance within a four-cell Landsat quadrant of 60 by 60 metres, and where
severity classes were mixed a minimum plot diameter of 20 metres and a minimum
spacing of 45 metres between classes were enforced. The plot was sized to the
sensor's support, not to the ecology. That is the correct decision, it is almost
never written down, and a reader who understands why it was taken has understood
most of what this book is about.

### Where this leads

A disturbance is not only a patch on a map. It is an event in a system that has
states, and the interesting question about a beetle outbreak or a fire regime is
not where it was but whether the system it happened in is still the same system
afterwards.

That question is the closing argument of this book, and it is named here so that
the index just computed is understood as the measurement the argument will rest
on. Insect outbreaks are the classic case of a population with a threshold, in
that the same forest and the same beetle and a slightly warmer winter give a
different trajectory altogether. The mathematics of that kind of shift is a live
and unsettled debate in the disturbance ecology literature rather than a settled
tool. Whether a forest carbon landscape can cross such a threshold, and whether a
monitoring system built from the indices in this chapter could tell if it did, is
a question about the length and the regularity of the record as much as about the
ecology. The final part of the book takes it up with a data cube as the data
structure and Markov chain Monte Carlo as the inference machinery, and it returns
to these same two disturbances.

None of that is attempted here. What this chapter delivers is the measurement,
its physical basis, and an honest account of the three ways it goes wrong, the
wrong band pair, an uncorrected atmosphere, and a support the analyst never
stated.

## Conclusions

A spectral index is a ratio of two estimated quantities, and every property of it
follows from that. It is bounded, which makes it saturate. It is a ratio, which
makes it cancel some effects and not others. It is built from estimates, which
means the retrieval that produced them is part of the measurement.

Three results from this scene stand behind that. Choosing the band pair by the
physics of the difference matters more than choosing the index by convention, and
on separating forest from regrowth the shortwave infrared indices beat the
visible ones by `r round(ba["NBR"] - ba["NDVI"], 3)` in balanced accuracy. An
uncorrected atmosphere moves a visible-band index
`r round(abs(ch["NDVI"]) / abs(ch["NBR"]), 0)` times as far as an infrared one,
far enough to reclassify an entire forest without an error being raised. And a
processing convention that changed on a stated date moves the same index by
`r round(abs(off["difference","NDVI"]), 3)` if the metadata is not read.

A fourth stands behind the section on disturbance. The burn ratio reached an
overall accuracy of 0.872 in the Selkirk Mountains and still missed 55 per cent
of the low severity class. An accuracy figure quoted without its per-class errors
is not a result, it is a summary of which classes were common.

None of those are exotic. All of them sit in the first hour of an ordinary
workflow, which is where the errors that reach an audit are made.

## Exercises

1. Compute the normalised difference between the narrow near infrared band (B8A)
   and the red band, and compare it with the standard vegetation index built on
   the broad band (B08). The two bands differ by 23 nanometres in centre
   wavelength and by 124 nanometres in width. How much does the index change, and
   which class changes most?
2. The band support section aggregated to 60 metre cells. Repeat it at 100 and
   200 metres and plot the mixed fraction against cell size. At what cell size
   does more than a quarter of the scene straddle a boundary?
3. Repeat the threshold sweep for riparian forest against wetland, two classes
   adjacent on the ground and in the legend. Which index does best, and is any of
   them usable?
4. Take the confusion between low and moderate severity described in the fire
   section. Given a class that is missed 55 per cent of the time and covers half
   the burned area, compute the range within which the true burned area could lie
   if only the classified map were available.
5. *(Applied.)* The scene on 2 September is smoky everywhere. Build a crude haze
   screen by flagging cells whose blue reflectance exceeds twice the scene median
   for the clear date, apply it to the smoky scene, and report what fraction of
   the scene survives. State whether the surviving fraction is spread evenly
   across the scene, and what that implies for a composite built from it.
6. *(Applied.)* A project reports a forest area from a vegetation index threshold
   and does not state the acquisition date, the processing baseline or the cell
   size. Using only the results in this chapter, put a range on how far that
   reported area could be from the truth, and say which of the three omissions
   you would ask about first.
