---
title: "Reading Reflectance, and What an Index Can Support"
subtitle: "Chapter 2 draft, 5 September 2026"
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

::: callout-note
## Draft, not a skeleton

Every number in the prose is computed by a chunk on this page. Brief at
`publishing/chapter-briefs/02-reflectance.md`. The disturbance signatures
section plants the closing arc and is the one place where the argument is
deliberately unfinished.
:::

```{r setup}
library(terra)

# Two dates over the same ground, ten Sentinel-2 Level-2A bands each, stored as
# signed integers scaled by 10,000. The division recovers reflectance.
clear <- rast("../data/rondonia_s2_2022-07-16.tif") / 1e4
smoke <- rast("../data/rondonia_s2_2022-09-02.tif") / 1e4

legend <- read.csv("../data/rondonia_legend.csv")
bands  <- read.csv("../data/sentinel2_bands.csv")

# The published nine-class map, projected onto the tile grid. It is stored in
# EPSG:3857 at about 20.26 m and the scene is in EPSG:32720 at 20 m, so a
# resampling happens here whether or not anyone writes it down. Nearest
# neighbour, because the values are labels and a label has no mean.
lulc  <- rast("../data/rondonia_class_2022.tif")
class <- project(lulc, clear, method = "near")
names(class) <- "class"
```

## Introduction

A satellite does not measure forest. It counts photons arriving in a detector
over a fixed interval of wavelength, converts that count to a digital number,
and writes the digital number to a file. Everything between that file and a
reported tonne of carbon is inference, and this chapter is about the first two
steps of it, the conversion from a count to a reflectance and the compression of
several reflectances into one index.

The chain is worth stating once in full, because the rest of the book assumes
it. Sunlight passes down through the atmosphere, some of it scatters and some is
absorbed. What reaches the canopy interacts with leaves, bark, soil and water.
Some of that returns upward, passes through the atmosphere a second time, and
reaches the sensor. The sensor records radiance at the top of the atmosphere. A
processing chain then estimates what the surface must have reflected for that
radiance to arrive, which requires a model of the atmosphere on that day over
that place. The output is called surface reflectance, and the word to hold onto
is *estimate*. A surface reflectance product is a retrieval with its own
uncertainty, not a measurement, and no part of forest carbon accounting treats
it that way.

The reason to persevere is that optical data is the only record long enough to
matter. Landsat 1 launched in July 1972 and the record has not stopped since,
which makes it the longest continuous measurement of the land surface that
exists. Almost every forest carbon project on Earth sets its historical baseline
from that archive, because nothing else can reach back far enough. The archive
is imperfect in the specific ways this chapter measures. It is also
irreplaceable.

The scene used throughout is a 24 kilometre window on the Madeira river in
Rondonia, in the Brazilian Amazon, on two dates in 2022. It carries intact
forest, riparian forest, open water, wetland, and the rectangular clearings of
the Amazon frontier, so every argument below can be made on real ground rather
than on a constructed surface.

```{r scene}
plotRGB(clear[[c("B04", "B03", "B02")]], stretch = "lin",
        main = "Rondonia, 16 July 2022, true colour")
```

## Spectral Response

The contrast that all of optical vegetation monitoring rests on is physical, not
conventional. Chlorophyll absorbs strongly in the red, near 665 nanometres,
because that is where the pigment does its work. The internal structure of a
leaf, the cell walls and the air spaces between them, scatters near infrared
light near 842 nanometres almost without absorbing it, because no pigment in the
leaf has an absorption feature there. A living leaf is therefore dark in the red
and bright in the near infrared, and a dead or absent one is not. The difference
between those two bands is a measurement of living tissue.

Sentinel-2 records ten bands relevant here. The table below is read from the
product specification rather than typed from memory.

```{r bands}
knitr::kable(bands, col.names = c("Band", "Name", "Centre (nm)",
                                  "Width (nm)", "Native cell (m)"))
```

The signature of each land cover is the mean reflectance of its cells in every
band. Three of them carry the whole argument.

```{r signatures}
d <- as.data.frame(c(clear, class), na.rm = FALSE)
d <- d[complete.cases(d), ]
d$cover <- legend$cover[match(d$class, legend$id)]

sig <- aggregate(d[names(clear)], list(cover = d$cover), mean)
knitr::kable(sig, digits = 3)
```

```{r signature-numbers, echo=FALSE}
g <- function(cov, b) sig[[b]][sig$cover == cov]
f_red <- g("Forest", "B04");  f_nir <- g("Forest", "B08")
s_red <- g("Clear cut, bare soil", "B04"); s_nir <- g("Clear cut, bare soil", "B08")
w_red <- g("Water", "B04");   w_nir <- g("Water", "B08")
```

Intact forest reflects `r round(f_red, 3)` in the red and `r round(f_nir, 3)` in
the near infrared, a ratio of `r round(f_nir / f_red, 1)`. Cleared ground with
bare soil reflects `r round(s_red, 3)` and `r round(s_nir, 3)`, a ratio of
`r round(s_nir / s_red, 1)`, because soil is brighter in the red and no longer
has a canopy scattering the near infrared. Open water reflects `r round(w_red, 3)`
in the red and `r round(w_nir, 3)` in the near infrared, so its near infrared is
*below* its red, because water absorbs near infrared almost completely. That
inversion is why water takes negative values in every vegetation index, and it
is the only reason water is easy.

Read across the rest of the table and the two shortwave infrared bands do
something the visible bands do not. Forest reflects `r round(g("Forest","B11"), 3)`
at 1,610 nanometres and clear-cut regrowth reflects
`r round(g("Clear cut, regrowth","B11"), 3)`, a gap of
`r round(g("Clear cut, regrowth","B11") - g("Forest","B11"), 3)`, while their red
reflectances differ by only
`r round(g("Clear cut, regrowth","B04") - f_red, 3)`. The shortwave infrared
responds to liquid water in the canopy, and a young regenerating stand holds
less of it than a closed mature one. Hold that observation. The section on class
separability turns it into the chapter's first result.

## Reflectance Retrieval

The files above are Level-2A products, meaning an atmospheric correction has
already been applied. It is worth knowing what was corrected and what was not.

A Level-1C product carries top of atmosphere reflectance, which includes
everything the atmosphere did on the way down and on the way up. The correction
to Level-2A estimates and removes the atmospheric contribution, using aerosol
optical thickness and water vapour either retrieved from the scene itself or
taken from a forecast. Where the retrieval works, the result approximates what a
spectrometer held above the canopy would have read. Where it does not, and
the smoky scene later in this chapter is one such case, the residual atmosphere stays in the number and
is indistinguishable from the ground.

There is a second, duller way the same number goes wrong, and it has cost
projects more than the physics has. Sentinel-2 processing baseline 04.00 entered
operations on 25 January 2022. Among other changes it shifted the stored dynamic
range by a band-dependent constant, so that a negative surface reflectance,
which is physically meaningless but statistically ordinary over dark water,
could be stored rather than clamped away. From that date the correct recovery is

$$\rho = \frac{\mathrm{DN} + \mathrm{BOA\_ADD\_OFFSET}}{\mathrm{QUANTIFICATION\_VALUE}}$$

and not $\mathrm{DN} / 10000$. The offset is carried in the product metadata and
is read, never assumed. Code written before 2022 and run afterwards subtracts
nothing, and the consequence is not subtle.

```{r offset}
dn <- rast("../data/rondonia_s2_2022-07-16.tif")

nd <- function(r, a, b) (r[[a]] - r[[b]]) / (r[[a]] + r[[b]])
forest_only <- function(r) mask(r, class == 4, maskvalues = FALSE)

applied <- dn / 1e4              # the offset was subtracted upstream
missed  <- (dn + 1000) / 1e4     # an offset of 1,000 DN never subtracted

off <- sapply(list(NDVI = c("B08", "B04"), NDMI = c("B08", "B11"),
                   NBR  = c("B08", "B12")),
  function(p) c(
    applied = global(forest_only(nd(applied, p[1], p[2])), "mean", na.rm = TRUE)[1, 1],
    missed  = global(forest_only(nd(missed,  p[1], p[2])), "mean", na.rm = TRUE)[1, 1]))

off <- rbind(off, difference = off["missed", ] - off["applied", ])
round(off, 3)
```

Mean forest vegetation index falls from `r round(off["applied","NDVI"], 3)` to
`r round(off["missed","NDVI"], 3)`, a shift of
`r round(abs(off["difference","NDVI"]), 3)`, on ground where nothing happened.
That is larger than most of the real differences a disturbance study is looking
for. The lesson is not the number, which depends on an offset value read from
metadata. The lesson is that a processing convention changed on a stated date,
and any time series that crosses that date without accounting for it carries a
step change that looks exactly like an event.

## Band Support

The previous chapter asked which plane the arithmetic happens on. The same question at the level
of a sensor is what area of ground one number describes.

Sentinel-2 does not answer that once. It acquires the blue, green, red and broad
near infrared bands at a native ground sample distance of 10 metres, and the
three red edge bands, the narrow near infrared band and the two shortwave
infrared bands at 20 metres. The files here sit on a single 20 metre grid, which
means six bands were coarsened to reach it and none was sharpened. An index that
mixes a 10 metre band with a 20 metre band is not comparing like with like
before somebody resamples, and after somebody resamples the mismatch is hidden
rather than resolved.

```{r support-table}
knitr::kable(aggregate(band ~ gsd_m, bands, paste, collapse = ", "),
             col.names = c("Native cell (m)", "Bands acquired at it"))
```

The consequence at a boundary is measurable. Aggregate the scene to 60 metre
cells, three by three, and ask how many of those cells are entirely forest,
entirely something else, or partly both.

```{r mixed-cells}
is_forest <- as.numeric(class == 4)
frac60 <- aggregate(is_forest, fact = 3, fun = "mean")
v <- values(frac60); v <- v[!is.na(v)]

round(100 * c(pure_forest = mean(v == 1),
              pure_other  = mean(v == 0),
              mixed       = mean(v > 0 & v < 1)), 1)
```

```{r mixed-numbers, echo=FALSE}
pct_mixed <- round(100 * mean(v > 0 & v < 1), 1)
```

`r pct_mixed` per cent of 60 metre cells straddle the forest boundary. Their
reflectance describes neither forest nor clearing, and whatever value an index
returns for them is a weighted average of two things that were never averaged in
nature. This is the whole of the mixed pixel problem, and it is why a 30 metre
Landsat cell and a 10 metre Sentinel-2 cell are not two measurements of the same
ground at different quality. They are measurements of different things.

The order of operations matters for the same reason. An index is a ratio, and
the ratio of averages is not the average of ratios.

```{r order-of-operations}
ndvi20 <- nd(clear, "B08", "B04")

first_average <- nd(aggregate(clear, fact = 3, fun = "mean"), "B08", "B04")
first_index   <- aggregate(ndvi20, fact = 3, fun = "mean")

difference <- first_average - first_index
names(difference) <- "NDVI difference"
global(abs(difference), quantile, probs = c(0.5, 0.95, 0.99, 1), na.rm = TRUE)
```

```{r order-numbers, echo=FALSE}
q <- global(abs(difference), quantile, probs = c(0.5, 1), na.rm = TRUE)
```

The median disagreement is `r format(round(q[1, 1], 4), scientific = FALSE)`,
which is nothing, and the maximum is `r round(q[1, 2], 3)`, which is a great
deal. The obvious guess is that the large disagreements sit at boundaries, where
the two source bands vary in opposite directions across the cell, so test it
rather than assert it.

```{r order-where}
a <- values(abs(difference)); m <- values(frac60 > 0 & frac60 < 1)
ok <- !is.na(a) & !is.na(m); a <- a[ok]; m <- as.logical(m[ok])

worst <- a >= quantile(a, 0.99)
round(c(mixed_share_of_scene = 100 * mean(m),
        mixed_share_of_worst = 100 * mean(m[worst]),
        median_mixed         = median(a[m]),
        median_pure          = median(a[!m])), 4)
```

```{r order-where-numbers, echo=FALSE}
share  <- 100 * mean(m)
worsts <- 100 * mean(m[worst])
```

On the cells where both quantities are defined, which is why the share differs
slightly from the figure above, mixed cells are `r round(share, 1)` per cent of
the scene and `r round(worsts, 1)` per cent of the worst one per cent of
disagreements, so they are over-represented by a factor of
`r round(worsts / share, 1)`. They do
not account for all of it, and the obvious guess was therefore too strong. The
typical disagreement in a mixed cell is about twice the typical disagreement in
a pure one. So the answer to which order is correct is that it depends on what
the number is for, and the answer to whether it matters is that it matters most
where you are most likely to be looking.

## Index Construction

A spectral index compresses several bands into one number. The normalised
difference is the standard form,

$$\mathrm{ND}(a, b) = \frac{\rho_a - \rho_b}{\rho_a + \rho_b}$$

which is bounded between -1 and 1 and, because it is a ratio, cancels part of
any multiplicative effect that raises or lowers both bands together. That
cancellation is the reason indices are used at all, and the section on
atmospheric sensitivity shows the limits of it.

The family is one function applied to different band pairs, and writing it that
way makes the point that the members differ only in which physics they ask
about.

```{r indices}
# Naming the arguments of c() would return a list rather than a SpatRaster, so
# the layers are combined unnamed and named afterwards.
indices <- function(r) {
  out <- c(nd(r, "B08", "B04"),   # chlorophyll absorption in the red
           nd(r, "B08", "B11"),   # liquid water in the canopy
           nd(r, "B08", "B12"),   # char and water loss after fire
           nd(r, "B08", "B05"),   # the red edge, sensitive where NDVI saturates
           nd(r, "B08", "B03"),   # chlorophyll absorption in the green

           # The enhanced vegetation index is not a normalised difference. Its
           # blue term is there to resist aerosol scattering and its soil term
           # to resist background brightness.
           2.5 * (r[["B08"]] - r[["B04"]]) /
             (r[["B08"]] + 6 * r[["B04"]] - 7.5 * r[["B02"]] + 1))

  names(out) <- c("NDVI", "NDMI", "NBR", "NDRE1", "GNDVI", "EVI")
  out
}

i_clear <- indices(clear)
plot(i_clear[[c("NDVI", "NDMI", "NBR", "NDRE1")]], range = c(-0.6, 1))
```

Look at the four panels together rather than at any one of them. The vegetation
index is nearly flat across everything that is green, because it saturates, in
that once a canopy is closed, adding leaf area barely changes the red
reflectance, which is already near zero. The moisture index and the burn ratio keep separating the
clearings from the forest well after the vegetation index has stopped. That is
not an artefact of the display. It is the next section's result.

## Class Separability

Take the two classes a deforestation project has to tell apart, and ask whether
an index can do it. Intact forest and clear cut with regrowth are, respectively,
`r round(100 * mean(d$class == 4), 1)` and
`r round(100 * mean(d$class == 6), 1)` per cent of the scene.

```{r separability-dist}
# na.rm is applied after the join, not inside as.data.frame(). The index layers
# carry a few more missing cells than the band stack does, where a normalised
# difference divides by zero, and dropping per layer would misalign the rows.
iv <- as.data.frame(c(i_clear, class), na.rm = FALSE)
iv <- iv[complete.cases(iv), ]
iv$cover <- legend$cover[match(iv$class, legend$id)]

quantiles <- do.call(rbind, lapply(split(iv$NDVI, iv$cover), quantile,
                                   probs = c(0.05, 0.25, 0.5, 0.75, 0.95)))
knitr::kable(round(quantiles, 3))
```

```{r separability-numbers, echo=FALSE}
fnd <- iv$NDVI[iv$cover == "Forest"]
gnd <- iv$NDVI[iv$cover == "Clear cut, regrowth"]
pctile <- round(100 * mean(fnd < median(gnd)), 1)
```

Regrowth has a median vegetation index of `r round(median(gnd), 3)`. Forest has
a median of `r round(median(fnd), 3)` and a fifth percentile of
`r round(quantile(fnd, 0.05), 3)`. The regrowth median therefore sits at the
`r pctile`th percentile of the forest distribution. There is no threshold that
separates them, because there is no gap to put one in, and a two-year-old
clearing that has grown back to scrub is spectrally inside the forest
distribution while being, in every sense a project cares about, not forest.

That statement can be made precise. Sweep every threshold and take the best.

```{r threshold-sweep}
best_threshold <- function(a, b, n = 400) {
  th <- seq(min(c(a, b)), max(c(a, b)), length.out = n)
  ba <- vapply(th, function(t) (mean(a >= t) + mean(b < t)) / 2, numeric(1))
  i  <- which.max(ba)
  c(threshold = th[i], balanced_accuracy = ba[i],
    regrowth_called_forest = mean(b >= th[i]),
    forest_called_other    = mean(a <  th[i]))
}

sub <- iv[iv$class %in% c(4, 6), ]
sweep <- t(sapply(c("NBR", "NDMI", "NDRE1", "NDVI"), function(nm)
  best_threshold(sub[[nm]][sub$class == 4], sub[[nm]][sub$class == 6])))
round(sweep, 3)
```

```{r sweep-numbers, echo=FALSE}
ba <- round(sweep[, "balanced_accuracy"], 3)
tv <- round(sweep["NDVI", "threshold"], 3)
```

Two things in that table are worth more than the rest of the chapter.

The first is that the best the vegetation index can do is a balanced accuracy of
`r ba["NDVI"]`, at a threshold of `r tv`, and at that threshold it still calls
`r round(100 * sweep["NDVI", "regrowth_called_forest"], 1)` per cent of regrowth
forest. Better than nothing is not the same as fit for purpose, and a project
that maps forest by thresholding a vegetation index has a systematic error it
cannot see, because the errors are concentrated in exactly the class the project
is paid to find.

The second is the ranking. The burn ratio reaches `r ba["NBR"]`, the moisture
index `r ba["NDMI"]`, the red edge index `r ba["NDRE1"]` and the vegetation
index `r ba["NDVI"]`. The two shortwave infrared indices beat the two visible
ones, and they beat them because regrowth differs from forest more in canopy
water than in greenness, exactly as the band signatures earlier in this chapter showed.
The index the whole field reaches for first is the worst of the four at the job
it is reached for.

None of this is an argument against the vegetation index. It is an argument for
choosing a band pair by the physics of the difference you are trying to see,
which is a decision, and for reporting which one you chose, which almost nobody
does.

## Atmospheric Sensitivity

Now the second date. On 2 September 2022 the same ground was under smoke from
biomass burning, which is an ordinary condition in Rondonia in September and not
an exotic failure case.

```{r smoke-rgb}
par(mfrow = c(1, 2))
plotRGB(clear[[c("B04", "B03", "B02")]], stretch = "lin", main = "16 July 2022")
plotRGB(smoke[[c("B04", "B03", "B02")]], stretch = "lin", main = "2 September 2022")
```

Both scenes are Level-2A, so an atmospheric correction ran on both and neither
was withheld. What the correction left behind is measurable band by band.

```{r smoke-bands}
# global() returns a data frame whose row names carry the band, so the means are
# renamed here rather than being indexed by position later.
m1 <- setNames(global(clear, "mean", na.rm = TRUE)[, 1], names(clear))
m2 <- setNames(global(smoke, "mean", na.rm = TRUE)[, 1], names(smoke))

knitr::kable(data.frame(band = names(clear), centre_nm = bands$centre_nm,
                        july = round(m1, 4), september = round(m2, 4),
                        ratio = round(m2 / m1, 2)), row.names = FALSE)
```

```{r smoke-numbers, echo=FALSE}
r_ratio <- round(m2 / m1, 2)
```

Scene-mean blue reflectance is `r r_ratio["B02"]` times its clear-date value.
Green is `r r_ratio["B03"]`, red `r r_ratio["B04"]`, near infrared
`r r_ratio["B08"]` and the longer shortwave infrared band `r r_ratio["B12"]`.
The pattern is monotonic in wavelength and it has a physical cause, because
scattering by fine aerosol particles falls off steeply as wavelength rises, so smoke
brightens the blue end of the spectrum and barely touches the infrared. This is
the same physics that makes the sky blue, operating on an instrument that was
trying to look through it.

What that does to an index depends entirely on which bands the index uses.

```{r smoke-indices}
i_smoke <- indices(smoke)

shift <- sapply(names(i_clear), function(nm) c(
  july      = global(forest_only(i_clear[[nm]]), "mean", na.rm = TRUE)[1, 1],
  september = global(forest_only(i_smoke[[nm]]), "mean", na.rm = TRUE)[1, 1]))
shift <- rbind(shift, change = shift["september", ] - shift["july", ])
round(shift, 3)
```

```{r smoke-index-numbers, echo=FALSE}
ch <- round(shift["change", ], 3)
fs <- values(forest_only(i_smoke[["NDVI"]])); fs <- fs[!is.na(fs)]
fc <- values(forest_only(i_clear[["NDVI"]])); fc <- fc[!is.na(fc)]
below_clear <- round(100 * mean(fc < sweep["NDVI", "threshold"]), 1)
below_smoke <- round(100 * mean(fs < sweep["NDVI", "threshold"]), 1)
```

Mean forest vegetation index falls by `r abs(ch["NDVI"])`. The burn ratio, both
of whose bands sit beyond 800 nanometres, moves `r abs(ch["NBR"])`. On identical
ground, on which nothing happened between the two dates, one index moved
`r round(abs(ch["NDVI"]) / abs(ch["NBR"]), 0)` times as far as the other.

Put that next to the threshold from the separability sweep. On the clear date,
`r below_clear` per cent of forest cells fall below the best threshold of
`r tv`. On the smoky date, `r below_smoke` per cent do.

```{r threshold-collapse}
c(clear_date = below_clear, smoky_date = below_smoke)
```

A forest map made by thresholding the vegetation index would have classified
essentially the entire forest as non-forest, and it would have done so with no
error, no warning and a plausible-looking output. The index that separated
worst also breaks worst, and the pairing of those two results is what this
chapter has to say.

The enhanced vegetation index deserves its own sentence, because it is the one
in the set explicitly designed to resist this. Its blue term, with a coefficient
of -7.5, exists to subtract the aerosol contribution.

```{r evi-failure}
round(shift[, "EVI"], 3)
```

It moved the wrong way, rising `r round(shift["change","EVI"], 3)` on forest.
The correction was calibrated for a modest aerosol perturbation, and under heavy
smoke a large blue value drives the denominator down and inflates the whole
ratio. A correction applied outside the range it was designed for is not a
weaker correction. It is a different error, and it is a worse one, because the
first kind announces itself and the second does not.

The operational answer to all of this is cloud and aerosol masking followed by
compositing over a window, which is the subject of the time series chapter. The
answer this chapter can give is narrower and comes first. Know which bands your
index uses, and know that an atmospheric state is a condition on the value in
the same way a cell size is. The number describes the ground and the column of
air above it, and the retrieval is the attempt to tell them apart.

## Disturbance Signatures

Two disturbances dominate the forest carbon landscapes this book is about, and
both were mapped with the indices computed above long before anyone modelled
them. This section is a survey of what those signatures are and where they fail.
The methods that turn a signature into a dated, attributed, area-adjusted
estimate belong to later chapters and are not attempted here.

### Insect Attack

Mountain pine beetle, *Dendroctonus ponderosae* Hopkins, kills lodgepole pine by
attacking the phloem, and the crown then fades over the following seasons. What
matters for a sensor is that the fading runs through three named phases, and each
one is a different physical state.

In the green phase the tree has been attacked and its needles are still green,
which is the phase every operational detector would most like to catch and the
phase none of them catches reliably. In the red stage, the first twelve to
twenty-four months after attack, the tree retains red needles while sapwood and
foliar moisture decline together, so both a visible-band index and a shortwave
infrared one have something to register. In the grey stage the remaining needles
and the snags have fallen, and what the sensor sees is the understorey and the
ground.

Which index to use in the red stage is therefore an empirical question rather
than a deduction, and it was answered by measurement. Murphy et al. (2026)
mapped red-stage attack across the Darkwoods Conservation Area in the southern
Selkirk Mountains of British Columbia from an annual Landsat series running 2003
to 2013, tested the moisture index against the wetness, greenness and brightness
components of the tasselled cap transform, and selected the moisture index on
its regression diagnostics against field reference plots.

Their accuracy assessment is the most useful thing in this chapter for a reader
who has to decide whether to believe a map. The moisture index classification
reached an overall accuracy of 0.824 and a kappa of 0.750, kappa being the
agreement left once the agreement expected by chance is removed. On the
red-attack class itself the commission error was 0.198 and the omission error
0.199, so about one cell in five called red attack was not, and about one in
five of the real red attack was missed. The study describes those rates as
acceptable and selects the moisture index on them.

That is a good result, and it is worth being exact about what a good result
buys. One cell in five is fine for showing where an outbreak is. It is not fine
for a project claiming payment on the difference between two years, because a
twenty per cent false alarm rate on the disturbed class is comparable in size to
the change many projects are trying to detect. The figure that decides whether a
map is usable is never the overall accuracy. It is the error on the class the
project is paid to find, and that is why the classification chapter exists.

### Fire

Fire is the mirror case and the easier one. Combustion removes canopy water and
leaves char, and both effects push the shortwave infrared up hard while the near
infrared falls, so the normalised burn ratio moves further and more reliably than
any visible-band index. The same study tested five candidate indices against 94 burn
plots, 9 of them field plots established immediately after the fire in 2015 and
resampled in 2020, and 85 scored from aerial photographs. Every plot carried a
composite burn index, a field protocol that rates fire effects across substrate,
ground shrubs, understorey and overstorey on a scale from 0.00 to 10.00, with
below 2.50 counted as no change and above 7.50 as high severity. The burn ratio
had the highest correlation with the field scores, at an R-squared of 0.879
against 0.584 for the next best index.

Its classification reached an overall accuracy of 0.872 and a kappa of 0.792,
the best figures anywhere in the study, and the per-class errors are where the
lesson is. On the high severity class the burn ratio missed 4 per cent. On the
low severity class it missed 55 per cent, and 22 per cent of the cells it called
low severity were something else. The headline figure is carried by the classes
that were easy, the unburned ground and the severe burns, and it conceals a
class the index can barely see, because a fire that scorches the understorey and
leaves the canopy standing changes very little of what the sensor measures. An
overall accuracy is an average over classes weighted by how common they are, so
it rewards a classifier for getting the easy majority right. Read the per-class
errors first and the overall accuracy last.

One detail of that field design belongs to this book's spine rather than to fire
ecology. Plots were located at the centre of areas of uniform disturbance within
a four-cell Landsat quadrant of 60 by 60 metres, with a minimum plot diameter of
20 metres and a minimum spacing of 45 metres between plots of different classes.
The plot was sized to the sensor's support and not to the ecology. That is the
correct decision, it is almost never written down, and a reader who understands
why it was taken has understood most of what this book is about.

### Where This Leads

A disturbance is not only a patch on a map. It is an event in a system that has
states, and the interesting question about a beetle outbreak or a fire regime is
not where it was but whether the system it happened in is still in the same
regime afterwards.

That question is the closing argument of this book and it is named here so that
the index just computed is understood as the measurement the argument will rest
on. Insect outbreaks are the classic case of a population with a threshold, in that
the same forest, the same beetle and a slightly warmer winter give a
qualitatively different trajectory, and the mathematics of that shift, bifurcation and tipping
behaviour, is a live and unsettled debate in the disturbance ecology literature
rather than a settled tool. Whether a forest carbon landscape can pass such a
threshold, and whether a monitoring system built from the indices in this
chapter could detect it if it did, is a question about the length and the
regularity of the record as much as about the ecology. The final part of the
book takes it up with a spatio-temporal data cube as the data structure and
Markov chain Monte Carlo as the inference machinery, and it returns to these
same two disturbances.

Nothing of that is attempted here. What this chapter delivers is the
measurement, its physical basis, and an honest account of the three ways it goes
wrong, the wrong band pair, an uncorrected atmosphere, and a support the analyst
never stated.

## Conclusions

A spectral index is a ratio of two estimated quantities, and every property of it
follows from that. It is bounded, which makes it saturate. It is a ratio, which
makes it cancel some effects and not others. It is built from estimates, which
means the retrieval that produced them is part of the measurement.

Three results from this scene stand behind that. Choosing the band pair by the
physics of the difference matters more than choosing the index by convention,
and on separating forest from regrowth the shortwave infrared indices beat the
visible ones by `r round(ba["NBR"] - ba["NDVI"], 3)` in balanced accuracy. An
uncorrected atmosphere moves a visible-band index `r round(abs(ch["NDVI"]) / abs(ch["NBR"]), 0)`
times as far as an infrared one, far enough to reclassify an entire forest
without an error being raised. And a processing convention that changed on a
stated date moves the same index by `r round(abs(off["difference","NDVI"]), 3)`
if the metadata is not read.

None of those are exotic. All three are in the first hour of an ordinary
workflow, which is where the errors that reach an audit are made.

## Exercises

1. Compute the normalised difference between the narrow near infrared band
   (B8A) and the red band, and compare it with the standard vegetation index
   built on the broad band (B08). The two bands differ by 23 nanometres in
   centre wavelength and by 124 nanometres in width. How much does the index
   change, and which class changes most?
2. The band support section aggregated to 60 metre cells. Repeat it at 100 and 200
   metres and plot the mixed fraction against cell size. At what cell size does
   more than a quarter of the scene straddle a boundary?
3. Repeat the threshold sweep for riparian forest against wetland, two classes that are adjacent on the ground and in the legend. Which
   index does best, and is any of them usable?
4. *(Applied.)* The scene on 2 September is smoky everywhere. Build a crude
   aerosol screen by flagging cells whose blue reflectance exceeds twice the
   scene median for the clear date, apply it to the smoky scene, and report what
   fraction of the scene survives. State whether the surviving fraction is
   spatially random, and what that implies for a composite built from it.
5. *(Applied.)* A project reports a forest area from a vegetation index
   threshold and does not state the acquisition date, the processing baseline or
   the cell size. Using only the results in this chapter, put a range on how far
   that reported area could be from the truth, and say which of the three
   omissions you would ask about first.
