---
title: "Deriving Canopy Height from Airborne Laser Scanning"
subtitle: "Chapter 4 draft, 6 September 2026"
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

::: callout-note
## Draft

Every number in the prose is computed by a chunk on this page. The working
example is `MixedConifer.laz`, which ships inside the `lidR` package, so this
chapter needs no committed data and no network. Brief at
`publishing/chapter-briefs/04-airborne-laser.md`. The chapter ends at the height
metric; all plot-to-metric modelling moved to chapter 9 on 6 September 2026.

The reader is the one the accepted proposal names. The main text is for graduate
students in forestry, environmental science and remote sensing and for analysts
arriving from a neighbouring field, and it assumes introductory statistics and no
prior geographic information systems. Registry material sits in tinted
verification tracks, and a reader who skips every one of them still gets a
complete chapter.
:::

```{r setup}
library(lidR)
library(terra)
library(sf)
library(dplyr)
```

## Introduction

A laser scanner mounted under an aircraft sends out a pulse and listens. The
pulse spreads as it travels, so by the time it reaches a forest it is a cone
rather than a line, and different parts of that cone hit different things. Some
of it strikes the top of a crown and comes straight back. Some passes through a
gap in the foliage, hits a branch lower down, and returns later. Some reaches
the ground. One pulse, several returns, each arriving at a different time and
therefore reporting a different height.

That is the whole physical basis of airborne laser scanning in forestry, and
everything in this chapter follows from one consequence of it. **The instrument
measures ranges. It does not measure trees.** Turning a cloud of ranges into a
canopy height requires deciding which returns hit the ground, and turning a
canopy height into a stem count requires deciding how wide to look for a
treetop. Both are decisions, both are made in software, and this chapter is
about how much they move the answer.

The question it settles is how much of a laser-derived height is the sensor and
how much is the analyst. By the end you will have seen the same returns produce
stem densities differing by a factor of twelve, with no change to the data at
all.

```{r load}
las <- readLAS(system.file("extdata", "MixedConifer.laz", package = "lidR"))
las
```

```{r load-facts, echo=FALSE}
area_ha <- as.numeric(st_area(st_as_sfc(st_bbox(las)))) / 1e4
n_pts   <- npoints(las)
dens    <- round(n_pts / (area_ha * 1e4), 2)
```

`r format(n_pts, big.mark = ",")` returns over `r round(area_ha, 2)` hectares of
mixed conifer, a density of `r dens` returns per square metre. That is modest by
current standards and typical of what a project actually receives.

## Provenance and vertical accuracy

Before any processing, ask what the product claims about itself. This is the
first dataset in this book that can answer.

Airborne lidar delivered through a national programme carries a stated quality
level. The United States 3D Elevation Program grades its collections, and its
Quality Level 1 specification requires an absolute vertical accuracy of 9.8
centimetres or better at 95 per cent confidence in open terrain, at a nominal
point density of at least eight returns per square metre. Canada's High
Resolution Digital Elevation Model programme and New Zealand's national
elevation holdings publish comparable statements.

Three words in that specification do the work. **Absolute** means against
surveyed control, not against itself. **In open terrain** means the figure does
not apply under canopy, which is where forestry works. And **95 per cent** means
one return in twenty is worse than the stated figure, by an unstated amount.

The previous chapters argued that a dataset should state its realisation, its
epoch and its accuracy. A national lidar collection does. Almost nothing else a
forest carbon project touches does, and that contrast is worth noticing before
the complaining starts about how difficult lidar is.

Now turn the same question on the file this chapter actually runs on, because a
chapter that tells you to interrogate a product and then does not interrogate its
own is worth nothing.

```{r header}
h <- las@header

data.frame(
  declares = c("coordinate reference system", "LAS version",
               "generating software", "file creation year", "day of year",
               "acquisition date", "sensor", "stated vertical accuracy"),
  value = c(st_crs(las)$Name,
            paste0(h@PHB[["Version Major"]], ".", h@PHB[["Version Minor"]]),
            h@PHB[["Generating Software"]],
            as.character(h@PHB[["File Creation Year"]]),
            as.character(h@PHB[["File Creation Day of Year"]]),
            "not recorded", "not recorded", "not recorded"))
```

It declares a coordinate system and nothing else that matters. There is no
acquisition date, no sensor, no flying height, no quality level and no vertical
accuracy. The single date it carries is a file creation year of 2017 with a day
of year of 0, which is not a date at all, and the generating software is
`las2las`, a format conversion tool. So 2017 records when somebody last rewrote
the file, not when an aircraft flew.

That is the ordinary condition of a lidar tile arriving at a project, and it is
why the vertical accuracy in the specification above cannot simply be assumed to
apply to whatever you were sent. A quality level is a claim a programme makes
about a collection, and the tile is one file out of that collection. The header
has a field for the software that last wrote the file and none for the aircraft
that flew it, so the accuracy statement travels in a separate document that is
easily parted from the data. Ask for that document by name, and ask which
collection the tile came from, because the tile itself will not tell you.

## Two ways in

Most people who work with lidar in forestry never see a point cloud. They are
handed two rasters, a **digital surface model** giving the height of whatever is
topmost at each cell, and a **digital elevation model** giving the height of the
ground, and they subtract one from the other to get canopy height. National
programmes distribute both, and the arithmetic is one line.

That route is legitimate and it is not equivalent. Subtracting two supplied
surfaces means the vendor's own ground classification is already baked into the
answer and cannot be inspected, adjusted or reported. It is a measurement whose
largest single decision was made by somebody else and is not recorded anywhere
in the file you were given.

This chapter takes the point cloud route, because that is the one where the
decisions are visible. Everything it demonstrates applies to the differenced
route as well, with the difference that you cannot see it happening.

## Ground classification

A return is ground or it is not, and no instrument reports which. Classifying
them is an inference from geometry, and two standard algorithms make that
inference differently.

The **cloth simulation filter** imagines the terrain inverted and a stiff cloth
dropped onto it from below, then calls ground whatever the cloth touches. Its
parameters are the cloth's rigidness and its resolution. A **progressive
morphological filter** instead opens the surface with a window that grows in
steps, removing anything that rises too steeply above its surroundings for its
size.

```{r ground}
g_csf <- classify_ground(las, algorithm = csf(sloop_smooth = FALSE))
g_pmf <- classify_ground(las, algorithm = pmf(ws = seq(3, 12, 3),
                                              th = seq(0.1, 1.5, length.out = 4)))

n_csf <- sum(g_csf$Classification == 2L)
n_pmf <- sum(g_pmf$Classification == 2L)
agree <- sum(g_csf$Classification == 2L & g_pmf$Classification == 2L)

data.frame(
  algorithm = c("cloth simulation", "progressive morphological", "both agree"),
  ground_returns = c(n_csf, n_pmf, agree),
  pct_of_cloud = round(100 * c(n_csf, n_pmf, agree) / npoints(las), 1)
)
```

```{r ground-numbers, echo=FALSE}
disagree <- (n_csf - agree) + (n_pmf - agree)
pct_dis <- round(100 * disagree / max(n_csf, n_pmf), 1)
```

The two disagree on `r format(disagree, big.mark = ",")` returns, which is
`r pct_dis` per cent of the larger ground set. That is not a rounding
difference. More than a quarter of what one algorithm calls ground the other
does not, and the distribution of the disagreement matters more than its size. Disagreements are not scattered
at random. They concentrate under dense low vegetation, where neither algorithm
can see the soil, and on breaks of slope, where a steep face looks to a
morphological filter exactly like a large object. Those are the same places
where the terrain model most needs to be right, because an error in ground
height propagates into every canopy height above it, one for one and with the
opposite sign.

## Normalisation, and an order that cannot be reversed

Canopy height is height above ground, so the terrain surface has to be
subtracted from every return. In `lidR` that is one call, and the cloud used
here arrives already normalised, which the height distribution shows.

```{r normalised}
round(c(min_Z = min(las$Z), median_Z = median(las$Z), max_Z = max(las$Z)), 2)
```

A minimum of zero and a maximum near `r round(max(las$Z))` metres is a canopy
height distribution, not an elevation. An unnormalised cloud in this landscape
would start somewhere above a thousand metres above sea level.

The order of operations here cannot be reversed, and getting it wrong fails
silently. Build a terrain model from a cloud that has already been normalised
and you interpolate a surface through points whose ground is already at zero.
The result is a flat plane. It is a valid raster, it has a coordinate reference
system, it plots without complaint, and it is useless. Nothing errors, nothing
warns, and every height computed against it is simply the height that was
already there.

That is the cheapest trap in this chapter and the one most often sprung, because
a workflow assembled from fragments of other people's scripts has no way of
knowing which stage its input has already been through.

## The canopy height model

Rasterising the returns into a canopy height model turns an irregular cloud into
a grid, which is the same support decision chapter 2 made for reflectance,
arriving now in three dimensions.

```{r chm}
chm <- rasterize_canopy(las, res = 0.5, algorithm = p2r(subcircle = 0.15))
chm
plot(chm, main = "Canopy height model, 0.5 m cells")
```

The `subcircle` argument replaces each return with a small disc before
rasterising, which fills the pits that appear where a pulse slipped between
crowns and reached the ground. Pit filling is cosmetic in a picture and is not
cosmetic in what follows, because a pit inside a crown can be detected as the
edge between two trees.

## The window function

Now the chapter's subject. Finding treetops means finding local maxima, and
finding a local maximum means choosing how large a neighbourhood counts as
local. Choose a window too small and one crown yields several tops. Too large
and several crowns yield one.

Because tall trees generally have wider crowns than short ones, the standard
approach makes the window a function of height. In the great majority of
published and operational work that function is a straight line with two
coefficients,

$$w(h) = a\,h + b$$

where $w$ is the window width in metres and $h$ is the canopy height at that
cell. Two numbers. They are the largest single lever on a stem count anywhere in
forest remote sensing, and they are very often borrowed.

```{r window-shape}
wf <- function(a, b) function(x) pmax(a * x + b, 1)

heights <- seq(0, 35, 0.5)
plot(heights, wf(0.05, 0.6)(heights), type = "l", ylim = c(0, 6), lwd = 2,
     xlab = "canopy height (m)", ylab = "window width (m)",
     main = "Variable window functions")
lines(heights, wf(0.03, 0.3)(heights), lty = 2)
lines(heights, wf(0.10, 1.5)(heights), lty = 3)
legend("topleft", bty = "n", lty = c(1, 2, 3),
       legend = c("0.05h + 0.6", "0.03h + 0.3", "0.10h + 1.5"))
```

Now vary the two coefficients across the range that appears in practice, and
count the trees each choice produces on identical returns.

```{r window-sweep}
grid <- expand.grid(a = c(0.03, 0.05, 0.07, 0.10),
                    b = c(0.3, 0.6, 1.0, 1.5))

grid$stems_ha <- vapply(seq_len(nrow(grid)), function(i) {
  tops <- locate_trees(chm, lmf(ws = wf(grid$a[i], grid$b[i]), hmin = 2))
  round(nrow(tops) / area_ha)
}, numeric(1))

knitr::kable(grid)
```

```{r window-numbers, echo=FALSE}
lo <- min(grid$stems_ha); hi <- max(grid$stems_ha)
borrowed <- grid$stems_ha[grid$a == 0.05 & grid$b == 0.6]
borrowed_rank <- sum(grid$stems_ha <= borrowed)
```

Stem density ranges from `r format(lo, big.mark = ",")` to
`r format(hi, big.mark = ",")` stems per hectare, a spread of
`r round(hi / lo, 1)`-fold, on one cloud, with one canopy height model, and no
field measurement involved anywhere. The choice of two coefficients moved the
answer by more than an order of magnitude.

That should be read alongside the allometry chapter, where five defensible
published equations applied to the same eighty-four oaks returned totals
spanning 4.75-fold. The window function is a larger lever than the allometric
equation, and it receives a small fraction of the scrutiny.

A fixed window, which is the simpler alternative, is no safer.

```{r fixed-window}
fixed <- data.frame(window_m = c(2, 3, 5, 8))
fixed$stems_ha <- vapply(fixed$window_m, function(w)
  round(nrow(locate_trees(chm, lmf(ws = w, hmin = 2))) / area_ha), numeric(1))
fixed
```

```{r fixed-numbers, echo=FALSE}
fr <- round(max(fixed$stems_ha) / min(fixed$stems_ha), 1)
```

A `r fr`-fold spread from one parameter.

### The function that travelled

The coefficients used above are not arbitrary. The pairing $a = 0.05$,
$b = 0.6$ is by some distance the most widely copied in operational work, and
its history is worth following because it is instructive rather than
embarrassing.

The same two-line function appears in two of this author's own project
repositories. In a boreal study of tree height variability around La Ronge,
Saskatchewan, it is defined as `wf_Popescu` and introduced as "Popescu and
Wynne's function (2004) which was developed in pine forests". In a stocking
density remapping of a plantation forest near Gisborne, New Zealand, the
identical function is defined as `wf_plowright`, its plot is labelled
"Plowright, 2018", and the code comment beside it reads "Used Plowright's window
function as temporary fix here", followed by a link to a package vignette.

Same slope. Same intercept. Two attributions. Boreal spruce and jack pine on one
continent, plantation radiata pine on another. Neither calibrated to its own
stand. Both produced operational outputs, and the second was compared against an
official stand-level inventory and used to propose a remapping.

Nothing was concealed. Both files say plainly what they did, which is why they
can be used this way. The point is what the sweep above already established. On
this cloud the borrowed pairing returns `r format(borrowed, big.mark = ",")`
stems per hectare, which is the `r borrowed_rank`th of the sixteen combinations
tested from the bottom. It is not a neutral default. It is a choice, and it
behaves like one.

::: callout-tip
## Verification track: what to ask of a stem density

A stem density derived from lidar is evidence, and evidence has to be traceable
to a method. Ask for four things and none of them is onerous.

The window function, written out, with its coefficients. The source those
coefficients came from, by citation, not by package vignette. Whether they were
calibrated on this forest or transferred from another, and if transferred, from
what species and what structure. And the minimum height threshold below which
returns were ignored, since that alone changes what counts as a tree.

A submission that cannot answer the second and third questions has not given you
a measurement. It has given you an output.
:::

## Height metrics and their support

Individual tree detection is the glamorous product and it is not what most
carbon inventories run on. Area-based metrics are. Summarise the returns inside
a cell, use the summaries as predictors, and never identify a tree at all.

```{r metrics}
m <- pixel_metrics(las, ~list(
  p95  = quantile(Z, 0.95),
  mean = mean(Z),
  max  = max(Z),
  cover = 100 * sum(Z > 2) / length(Z)), res = 20)

round(global(m, "mean", na.rm = TRUE), 2)
```

```{r metric-numbers, echo=FALSE}
mm <- global(m, "mean", na.rm = TRUE)[, 1]
names(mm) <- c("p95", "mean", "max", "cover")
```

Read those three heights together. Averaged over 20 metre cells, the 95th
percentile of return height is `r round(mm[["p95"]], 1)` metres, the mean return
height is `r round(mm[["mean"]], 1)` metres, and the maximum is
`r round(mm[["max"]], 1)`. The 95th percentile sits at
`r round(100 * mm[["p95"]] / mm[["max"]])` per cent of the maximum and
`r round(mm[["p95"]] / mm[["mean"]], 1)` times the mean.

That gap is the whole reason upper percentiles are used to predict biomass. A
cell holding several crowns and a great deal of empty air between them has a
mean return height dragged down by every pulse that passed through a gap. The
95th percentile ignores those and reports something close to the tallest crown
present, which is where the mass is.

And it is why an upper percentile **is not stand height**. It is a statistic of
the returns in a cell, its support is the cell, and it belongs to no tree. A
project that reports a 95th percentile as though it were a mensurational top
height has changed what the number means without changing the number, which is
the failure this book keeps returning to.

## Where the metric decides the fieldwork

The last section runs the argument backwards. Everything so far has used the
sensor to describe the forest. This uses it to decide how much fieldwork the
forest needs, and it is where the chapter hands over to plot design.

The height variation hypothesis holds that structural variability, not just
mean structure, carries ecological information. Its practical corollary in
inventory is simpler. Where canopy height varies sharply over short distances, a
fixed sampling intensity buys less precision than it does in a uniform stand,
because each plot represents a more heterogeneous neighbourhood.

Lidar measures that variability directly and cheaply, before anyone walks in.

```{r heterogeneity}
sd_h <- pixel_metrics(las, ~sd(Z), res = 20)
names(sd_h) <- "height_sd"

v <- values(sd_h); v <- v[!is.na(v)]
round(c(min = min(v), median = median(v), max = max(v),
        ratio_max_min = max(v) / min(v)), 2)
```

```{r het-numbers, echo=FALSE}
hr <- round(max(v) / min(v), 1)
```

Across this stand the standard deviation of return height varies `r hr`-fold
between cells, from `r round(min(v), 1)` to `r round(max(v), 1)` metres. The
range is modest here only because the extent is `r round(area_ha, 1)` hectares
of one forest type; across a management unit spanning burned, regenerating and
intact stands it is far wider. Under a sampling design whose precision depends
on within-unit variance, cells like these do not deserve equal effort. Classifying a
landscape this way, into what the La Ronge work calls height heterogeneity
areas, tells a crew where to add plots or shrink sampling units before the
season starts.

That is the support argument arriving from the opposite direction. Chapter 1
asked what area a number describes. This asks how many numbers an area needs.
The chapter on field plot design takes it from here.

## Conclusions

An airborne laser measures ranges, and every forestry quantity derived from it
passes through decisions that are not in the data.

Which returns are ground is inferred, and two standard algorithms disagree on
`r pct_dis` per cent of them, concentrated exactly where terrain error matters
most. The order of normalisation cannot be reversed and fails without erroring.
The window function that turns a height surface into a stem count is two
coefficients, and moving them across their ordinary published range changed
density `r round(hi / lo, 1)`-fold on identical returns, which is a larger lever
than the choice of allometric equation. And an upper height percentile, the
workhorse predictor of biomass, is a property of a cell rather than of a tree.

None of that argues against lidar, which remains the best measurement of forest
structure available at landscape scale by a wide margin. It argues for reporting
the decisions alongside the number, because a stem density without its window
function is not reproducible, and a canopy height without its ground
classification is not a measurement.

## Exercises

1. Rerun the ground classification with the cloth simulation filter's
   `sloop_smooth` argument set to `TRUE` and report how many returns change
   class. State whether the change is larger or smaller than the disagreement
   between the two algorithms.
2. Build the canopy height model at 0.25, 0.5 and 1 metre cells and rerun the
   window sweep on each. Report whether the coefficient sensitivity gets larger
   or smaller as cells get coarser, and explain why.
3. The sweep used a minimum height of 2 metres. Repeat it at 1, 2 and 5 metres
   and report how much of the spread in stem density is attributable to that
   threshold rather than to the two coefficients.
4. Compute the 95th percentile of return height at 10, 20 and 40 metre cells.
   Report how the mean of that statistic changes with cell size and say which of
   the three you would use to predict biomass, and why.
5. *(Applied.)* A project reports 340 stems per hectare from a lidar stem map
   and does not state its window function. Using only the results in this
   chapter, give the range of window coefficients consistent with that figure on
   a stand like this one, and say what you would ask for.
6. *(Applied.)* Take the height standard deviation surface from the last section
   and propose a two-class split into ordinary and heterogeneous ground. State
   the threshold you chose, the fraction of the stand in each class, and how you
   would allocate a fixed budget of twenty plots between them.

::: callout-note
## Open decisions

**The attribution needs checking.** Popescu and Wynne (2004) published
species-specific window functions for pine and deciduous stands, and the linear
form used throughout this chapter appears to originate in a package vignette
rather than in that paper. Open the paper before print. If the attribution is
wrong it should be corrected in both source repositories, and the correction is
worth a sentence here, because a misattributed parameter is harder to audit than
an uncalibrated one.

**A units check is outstanding.** The Gisborne workflow converts a treetop count
on a ten metre summary grid to stems per hectare by multiplying by ten, where
ten metre cells would require one hundred. `ForestTools::sp_summarise` is no
longer exported by the installed version, so this could not be settled by
running it and is deliberately not stated in the chapter. If it holds it belongs
in section 4.6, because that layer was compared against official stand
attributes and passed.

**Data.** The chapter currently runs entirely on `MixedConifer.laz` from `lidR`,
which is public, tiny and needs no commit. That is a virtue for reproducibility
and a limitation for the argument, since the stand is 0.8 hectares of one forest
type. Committing a one hectare clip of a national programme tile would let
section 4.2 read a real quality level off a real file rather than describing
one.
:::
