---
title: "Conditioning a Digital Elevation Model for Hydrological Routing"
subtitle: "Chapter 5 draft, 6 September 2026"
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

::: callout-note
## Draft

A restructure of `04-terrain-hydrology.qmd`, which retires when this is
promoted. What is new is the opening, which leads with the failure and its
price, a section on the elevation model as an observed product with a precision
you can measure, and a comparison of every conditioning algorithm the engine
offers. That comparison overturned the previous chapter's conclusion, which held
that a terminal sink must be lost to make a surface routable. It need not be.
Every number in the prose is computed by a chunk on this page.

The reader is the one the accepted proposal names. The main text is for graduate
students in forestry, environmental science and remote sensing and for analysts
arriving from a neighbouring field, and it assumes introductory statistics and no
prior geographic information systems. Registry material sits in tinted
verification tracks, and a reader who skips every one of them still gets a
complete chapter.
:::

```{r setup}
library(terra)
library(sf)
library(whitebox)

sf_use_s2(FALSE)

dem    <- rast("../data/chilwa_dem.tif")
lake   <- st_read("../data/chilwa_hydro.gpkg", layer = "lake",   quiet = TRUE)
basin  <- st_read("../data/chilwa_hydro.gpkg", layer = "basin",  quiet = TRUE)
rivers <- st_read("../data/chilwa_hydro.gpkg", layer = "rivers", quiet = TRUE)
```

## One metre

Lake Chilwa sits in southern Malawi. Water enters from the surrounding hills and
leaves only by evaporation, so it is *endorheic*, meaning it has no outlet. Its
area swings by a factor of several between wet and dry years and it has dried out
completely three times in living memory, in 1968, 1995 and 2012.

A wetland carbon project on ground like this reports an inundated area. Here is
what that number is worth.

```{r stage-area}
# equal-area frame, and only the lake's own catchment
demu  <- project(dem, "EPSG:32736", method = "bilinear")
basu  <- vect(st_transform(basin, "EPSG:32736"))
laku  <- vect(st_transform(lake,  "EPSG:32736"))

d   <- mask(crop(demu, basu), basu)
csz <- cellSize(d, unit = "ha")

# the lake bed, taken as the fifth percentile of elevation under the polygon
bed <- as.numeric(global(mask(crop(demu, laku), laku),
                         quantile, probs = 0.05, na.rm = TRUE))

# the connected depression, not every low cell in the basin
patch_at_3m <- patches(ifel(d <= bed + 3, 1, NA), directions = 8)
main <- patch_at_3m %in% unique(na.omit(extract(patch_at_3m, laku)[, 2]))

stage <- data.frame(stage_m = 0:3)
stage$flooded_ha <- vapply(stage$stage_m, function(s)
  as.numeric(global(ifel(d <= bed + s & main, csz, 0), "sum", na.rm = TRUE)),
  numeric(1))
stage$ratio <- round(stage$flooded_ha / stage$flooded_ha[1], 1)
stage$flooded_ha <- round(stage$flooded_ha)
stage
```

```{r stage-numbers, echo=FALSE}
a0 <- stage$flooded_ha[1]; a1 <- stage$flooded_ha[2]; a2 <- stage$flooded_ha[3]
jump <- round(a1 / a0, 1)
next_step <- round(100 * (a2 - a1) / a1, 1)
hydro_ha <- round(sum(expanse(laku, unit = "ha")))
```

Raise the water level by one metre and the flooded area goes from
`r format(a0, big.mark = ",")` hectares to `r format(a1, big.mark = ",")`, a
factor of `r jump`. Raise it by a second metre and the area grows by
`r next_step` per cent. The curve is not gentle and it is not symmetric. Nearly
everything happens in the first metre.

Now the part that matters. **One metre is the finest distinction this elevation
model can make.**

```{r precision}
v <- values(dem); v <- v[!is.na(v)]

c(storage = datatype(dem),
  unique_values = length(unique(v)),
  all_integers = all(v == round(v)),
  smallest_step_m = min(diff(sort(unique(v)))))
```

The file holds signed integers. Every elevation in this basin is a whole number
of metres, and the smallest difference it can express is exactly the increment
that moves the lake by a factor of `r jump`.

So the chapter's question is not how to model inundation. It is what an
inundated area can mean when the surface it is measured on cannot resolve the
step that determines it. Everything below works toward an answer, and the answer
is not that the data are unusable.

## What the elevation model is

The previous chapter read a vertical accuracy off a lidar product: 9.8
centimetres or better at 95 per cent confidence, stated by the programme that
flew it. Ask this file the same question.

```{r provenance}
dem
```

It is a 100 metre grid in EPSG:3857 covering the basin plus margin. It came from
the Amazon Web Services Terrain Tiles mosaic through the `elevatr` package at
zoom level 10, which returns roughly five arc-second cells at this latitude.
That mosaic is composited from several global and national sources whose
identity varies by location and is not recorded per cell.

Nowhere in the file, and nowhere in the service that supplied it, is there a
vertical accuracy figure for this basin. There is no quality level, no
confidence statement, no separation of open ground from vegetated. What can be
established is what the file itself demonstrates, which is a one metre
quantisation, and that is a precision rather than an accuracy. A value can be
recorded to the metre and be wrong by ten.

That is the ordinary condition of terrain data in this field, and it is worth
stating plainly rather than working around quietly. A lidar collection tells you
how wrong it might be. A global elevation mosaic tells you the number.

## The grid it sits on

One property has to be settled before any area is quoted, because it silently
corrupts every hydrological figure that follows.

EPSG:3857 is a Mercator projection. It is conformal rather than equal-area, and
its scale factor grows with latitude, so a cell that is nominally 100 metres
square does not cover a hectare of ground anywhere except the equator.

```{r cell-area}
cs <- cellSize(dem, unit = "ha")
round(unlist(global(cs, c("min", "mean", "max"), na.rm = TRUE)), 4)
```

```{r cell-numbers, echo=FALSE}
cs_mean <- as.numeric(global(cs, "mean", na.rm = TRUE))
cs_err  <- round(100 * (1 / cs_mean - 1), 1)
```

A cell covers `r round(cs_mean, 3)` hectares, not one. Treating each cell as a
hectare, which is the standard shortcut for converting a flow accumulation count
into contributing area, **overstates area by `r cs_err` per cent**.

::: callout-warning
## A contradiction in this book's own data script

`data-raw/prepare-chilwa-data.R` states in its header that "at 100 m each cell
is exactly one hectare, so flow accumulation counts convert to contributing area
by multiplying by 0.01 to get square kilometres" and adds that "that identity is
used in the chapter's prose". Two paragraphs later the same header states that
the projection inflates area by about seven per cent and that the chapter
therefore reports in UTM.

Both cannot be true. The measurement above settles it: the identity is wrong by
`r cs_err` per cent and the second paragraph is right. The header must be
corrected, and it is recorded here rather than fixed silently because a
contradiction between a data script and the chapter that reads it is exactly the
kind of thing that survives review.
:::

This chapter uses `cellSize()` throughout and reports areas after reprojection
to UTM zone 36S, which is close enough to equal-area at this extent to serve as
truth.

## Terrain derivatives

Elevation is rarely the variable of interest. What it implies is.

```{r derivatives}
slope <- terrain(dem, "slope", unit = "degrees")
tri   <- terrain(dem, "TRI")

round(unlist(global(c(slope, tri), "mean", na.rm = TRUE)), 2)
```

```{r hillshade}
#| fig-cap: "Relief of the Lake Chilwa basin. The lake occupies the flat floor at centre; the Zomba and Mulanje massifs rise on the western and southern margins."
hs <- shade(terrain(dem, "slope",  unit = "radians"),
            terrain(dem, "aspect", unit = "radians"),
            angle = 40, direction = 315)

plot(hs, col = grey(0:100 / 100), legend = FALSE, mar = c(2, 2, 2, 2))
plot(st_geometry(lake),  add = TRUE, border = "steelblue", lwd = 2)
plot(st_geometry(basin), add = TRUE, border = "orangered", lwd = 1.5)
```

Slope governs operability and erosion risk, and ruggedness is a useful
productivity covariate, because sheltered gentler terrain often carries taller
and denser stands. Both are computed from differences between neighbouring
cells, so both inherit the one metre quantisation, and on ground this flat that
matters. A slope computed from integer elevations across 100 metre cells cannot
express anything shallower than about half a degree.

## Depressions, real and artefact

Water moves downhill, so an elevation model already encodes drainage. It does not
encode it cleanly. Every model contains cells with no lower neighbour, produced
by interpolation error, vegetation returns and finite vertical precision. A
routing algorithm entering one has nowhere to send the water, and the drainage
network fragments.

```{r sinks}
dir.create("../data/chilwa-hydro", showWarnings = FALSE)
wd <- "../data/chilwa-hydro/"
file.copy("../data/chilwa_dem.tif", paste0(wd, "dem.tif"), overwrite = TRUE)

wbt_depth_in_sink(dem = "dem.tif", output = "sink_depth.tif",
                  wd = wd, zero_background = TRUE, verbose_mode = FALSE)

sink_depth <- rast(paste0(wd, "sink_depth.tif"))
sd_v <- values(sink_depth); sd_v <- sd_v[!is.na(sd_v) & sd_v > 0]

data.frame(
  cells_in_a_depression = length(sd_v),
  within_1m  = round(100 * mean(sd_v <= 1), 1),
  within_2m  = round(100 * mean(sd_v <= 2), 1),
  max_depth_m = round(max(sd_v), 1))
```

```{r sink-numbers, echo=FALSE}
n_sink <- length(sd_v)
w1 <- round(100 * mean(sd_v <= 1), 1)
w2 <- round(100 * mean(sd_v <= 2), 1)
maxd <- round(max(sd_v))
```

`r format(n_sink, big.mark = ",")` cells sit in some depression, and the deepest
is `r maxd` metres below its spillover point.

Read the middle two columns against the previous section. **`r w1` per cent of
those depressions are one metre deep or less**, which is a single quantisation
step, and `r w2` per cent are within two. Those cannot be distinguished from
rounding. They are not necessarily artefacts, but nothing in the data can say
that they are not, and a rule that removes depressions below a depth threshold
is really a rule about the storage format.

No threshold on depth alone separates real basins from interpolation pits,
because the two overlap in size. The deepest single pit in this grid is an
artefact, not the lake.

## Conditioning, and the parameter that does the damage

Two families of algorithm exist. **Filling** raises the floor of a depression
until it overflows, which is simple and destroys the depression. **Breaching**
carves a channel out through the lowest point of the rim, which changes far
fewer cells and preserves the basin's shape.

WhiteboxTools implements Lindsay's hybrid method, which breaches what it can
within a depth limit and fills whatever remains. Two parameters do the work.
`max_depth` caps how deep a channel may be carved. `flat_increment` adds a small
monotonic gradient along the carved channel so flow directions stay unambiguous
across ground the model records as perfectly level.

The obvious way to protect a genuine 400 metre basin is to cap the carving, so
that shallow artefacts are breached and the real basin is left alone. That
reasoning is exactly backwards, and the way to see it is to run every algorithm
the engine offers on the same surface.

```{r algorithms}
run_one <- function(label, fn) {
  t   <- system.time(fn())
  out <- rast(paste0(wd, "cond.tif"))
  ch  <- out - dem
  lk  <- ifel(!is.na(rasterize(vect(lake), dem)), ch, NA)
  data.frame(algorithm = label,
             seconds       = round(t[["elapsed"]], 1),
             cells_changed = global(ch != 0, "sum", na.rm = TRUE)[1, 1],
             max_carve_m   = round(-global(ch, "min", na.rm = TRUE)[1, 1], 1),
             max_fill_m    = round(global(ch, "max", na.rm = TRUE)[1, 1], 1),
             lake_raise_m  = round(global(lk, "mean", na.rm = TRUE)[1, 1], 1))
}

algos <- rbind(
  run_one("breach, Lindsay, max_depth 3", function()
    wbt_breach_depressions(dem = "dem.tif", output = "cond.tif", wd = wd,
      max_depth = 3, flat_increment = 0.001, verbose_mode = FALSE)),
  run_one("breach, Lindsay, unconstrained", function()
    wbt_breach_depressions(dem = "dem.tif", output = "cond.tif", wd = wd,
      flat_increment = 0.001, verbose_mode = FALSE)),
  run_one("breach, least cost", function()
    wbt_breach_depressions_least_cost(dem = "dem.tif", output = "cond.tif",
      wd = wd, dist = 100, fill = TRUE, verbose_mode = FALSE)),
  run_one("fill, Wang and Liu", function()
    wbt_fill_depressions_wang_and_liu(dem = "dem.tif", output = "cond.tif",
      wd = wd, verbose_mode = FALSE)),
  run_one("fill, Planchon and Darboux", function()
    wbt_fill_depressions_planchon_and_darboux(dem = "dem.tif",
      output = "cond.tif", wd = wd, verbose_mode = FALSE)))

knitr::kable(algos)
```

```{r algo-numbers, echo=FALSE}
g <- function(lab, col) algos[[col]][algos$algorithm == lab]
r_con <- g("breach, Lindsay, max_depth 3", "lake_raise_m")
r_unc <- g("breach, Lindsay, unconstrained", "lake_raise_m")
c_unc <- g("breach, Lindsay, unconstrained", "max_carve_m")
f_con <- g("breach, Lindsay, max_depth 3", "max_fill_m")
r_lc  <- g("breach, least cost", "lake_raise_m")
r_fil <- g("fill, Wang and Liu", "lake_raise_m")
```

Read the last two columns together, because the result is the opposite of what
the parameter's name suggests.

**Constrained to three metres, the algorithm fills the lake by
`r r_con` metres on average and by up to `r f_con` metres.** It refuses to carve
a channel deeper than three metres, and the only way out of a four hundred metre
basin is a channel deeper than three metres, so it fills the basin instead.

**Unconstrained, it does not raise a single cell.** The lake is untouched, and
the largest carve is `r c_unc` metres, which is one channel out of the basin.

The parameter intended to protect deep basins is the parameter that destroys
them. Capping the carving does not preserve the depression. It removes the
algorithm's only means of preserving it.

Both filling algorithms do the same damage as the constrained breach, at
`r r_fil` metres, which tells you the choice that matters is not breaching
against filling. The least-cost breach is worse than either at `r r_lc` metres.

The unconstrained surface is also fully routable, which is the claim that has to
be checked rather than assumed.

```{r routable}
wbt_breach_depressions(dem = "dem.tif", output = "breached.tif", wd = wd,
                       flat_increment = 0.001, verbose_mode = FALSE)
wbt_depth_in_sink(dem = "breached.tif", output = "check_sink.tif", wd = wd,
                  zero_background = TRUE, verbose_mode = FALSE)

remaining <- values(rast(paste0(wd, "check_sink.tif")))
breached  <- rast(paste0(wd, "breached.tif"))
change    <- breached - dem
carve     <- values(change); carve <- carve[!is.na(carve) & carve != 0]

c(depressions_remaining = sum(remaining > 0, na.rm = TRUE),
  cells_altered         = length(carve),
  any_cell_raised       = any(carve > 0),
  median_carve_m        = round(quantile(-carve, 0.5), 2),
  p90_carve_m           = round(quantile(-carve, 0.9), 2))
```

Zero depressions remain, so the surface routes. No cell was raised. The median
altered cell was carved by nothing at all and nine in ten by less than two
metres, because almost all of the work is removing single-cell pits. One channel
carries the whole `r c_unc` metre carve.

## What conditioning destroys, and what it need not

```{r endorheic}
lake_mask <- rasterize(vect(lake), dem)

constrained <- local({
  wbt_breach_depressions(dem = "dem.tif", output = "con3.tif", wd = wd,
    max_depth = 3, flat_increment = 0.001, verbose_mode = FALSE)
  rast(paste0(wd, "con3.tif"))
})

beds <- data.frame(
  surface = c("raw", "breached, max_depth 3", "breached, unconstrained"),
  lake_bed_median_m = round(c(
    as.numeric(global(ifel(!is.na(lake_mask), dem, NA),
                      quantile, probs = 0.5, na.rm = TRUE)),
    as.numeric(global(ifel(!is.na(lake_mask), constrained, NA),
                      quantile, probs = 0.5, na.rm = TRUE)),
    as.numeric(global(ifel(!is.na(lake_mask), breached, NA),
                      quantile, probs = 0.5, na.rm = TRUE))), 1))
beds
```

```{r bed-numbers, echo=FALSE}
b_raw <- beds$lake_bed_median_m[1]
b_con <- beds$lake_bed_median_m[2]
b_unc <- beds$lake_bed_median_m[3]
```

The raw lake bed sits at `r b_raw` metres. Under the constrained breach it sits
at `r b_con`. Under the unconstrained breach it sits at `r b_unc`, which is the
raw value. Chilwa survives one of these and not the other.

That changes the standing advice, which has been that flow routing requires
removing terminal sinks and the loss is unavoidable. It is avoidable here. A
breach that is allowed to carve as deep as the terrain requires produces a
routable surface and an intact basin at the same time, because it gives the
water an outlet instead of raising the floor until there is no basin left.

Two rules still follow and they are now weaker than they were.

**Anything depending on the shape of the depression must be computed on a
surface that still has the depression.** That means depression depth, inundated
extent and any stage-area relationship, including the calculation this chapter
opened with. On a constrained-breach or filled surface the basin is already
gone, and running an inundation model on one is a silent and total error.

**And the conditioning parameters belong in the report.** Two analysts running
the same function on the same file, one with a depth cap and one without, get a
lake and no lake. Neither is warned.

## Flow routing, and a maximum in the wrong place

```{r routing}
wbt_d8_pointer(dem = "breached.tif", output = "d8.tif", wd = wd,
               verbose_mode = FALSE)
wbt_d8_flow_accumulation(input = "d8.tif", output = "acc.tif", wd = wd,
                         pntr = TRUE, verbose_mode = FALSE)

acc <- rast(paste0(wd, "acc.tif"))
global(acc, "max", na.rm = TRUE)[1, 1]
```

```{r routing-check}
# where is that maximum, and is it in the lake?
peak_cell <- which.max(values(acc))
peak_xy   <- xyFromCell(acc, peak_cell)
in_lake   <- !is.na(extract(lake_mask, peak_xy)[, 1])

lake_acc <- global(ifel(!is.na(lake_mask), acc, NA), "max", na.rm = TRUE)[1, 1]

data.frame(global_max_cells = global(acc, "max", na.rm = TRUE)[1, 1],
           global_max_in_lake = in_lake,
           lake_max_cells = lake_acc)
```

```{r acc-numbers, echo=FALSE}
gmax <- global(acc, "max", na.rm = TRUE)[1, 1]
lmax <- lake_acc
```

The largest flow accumulation in the grid is
`r format(gmax, big.mark = ",")` cells, and it is **not in the lake**. It belongs
to a neighbouring catchment draining off the edge of the raster. Chilwa's own
convergence peaks at `r format(lmax, big.mark = ",")` cells.

That distinction is easy to lose. A script that reports the maximum accumulation
in a grid as the catchment's outlet is right whenever the grid contains one
catchment and wrong the rest of the time, and nothing about the output announces
which case you are in. The check is two lines and almost nobody writes them.

Multiplying either figure by a nominal hectare per cell would then compound the
error by the `r cs_err` per cent established earlier.

## Stage and area

Return to the opening calculation with everything the chapter has established.

```{r rating}
stage_full <- data.frame(stage_m = seq(0, 4, 0.5))
stage_full$flooded_ha <- vapply(stage_full$stage_m, function(s)
  round(as.numeric(global(ifel(d <= bed + s & main, csz, 0), "sum", na.rm = TRUE))),
  numeric(1))

plot(stage_full$stage_m, stage_full$flooded_ha / 1000, type = "b", pch = 19,
     xlab = "stage above lake bed (m)", ylab = "flooded area (thousand ha)",
     main = "Stage-area curve, Lake Chilwa basin")
abline(h = hydro_ha / 1000, lty = 2, col = "steelblue")
text(3, hydro_ha / 1000, "HydroLAKES maximum extent", pos = 3, cex = 0.8,
     col = "steelblue")
```

```{r rating-numbers, echo=FALSE}
at1 <- stage_full$flooded_ha[stage_full$stage_m == 1]
agree_pct <- round(100 * abs(at1 - hydro_ha) / hydro_ha, 1)
```

Two things are visible. The curve is nearly vertical in its first metre and
nearly flat afterwards, which is what a shallow basin with a wide flat floor
looks like. And the one metre stage encloses
`r format(at1, big.mark = ",")` hectares against the HydroLAKES published
maximum extent of `r format(hydro_ha, big.mark = ",")`, a difference of
`r agree_pct` per cent.

That agreement is a genuine result and it is easy to over-read. The published
polygon is a long-run maximum extent and the model surface is an instantaneous
threshold, so the two are not the same quantity and this is not a validation. It
does say that a stage one metre above the lake bed reproduces the published
maximum to within a couple of per cent, which is a useful anchor for a rating
curve and a poor substitute for a gauge.

::: callout-tip
## Verification track: accepting an inundated area

A wetland methodology asks for an inundated area, usually as an annual figure or
a class transition. Four questions decide whether the figure can be accepted.

Which surface was the extent computed on, the raw model or the conditioned one.
A conditioned surface has had the basin removed and cannot produce an inundated
extent at all.

What is the vertical precision of that surface, and how does it compare with the
stage difference the project is claiming. Here the precision is one metre and
one metre is a factor of `r jump` in area.

What was the stage, and where did it come from. A gauge, a satellite altimeter,
or a threshold chosen to reproduce a published polygon are three different
levels of evidence.

And in what projection was the area computed. A `r cs_err` per cent error is
available for free to anyone who counts cells on a Mercator grid.
:::

## Which soil, and therefore which method

An inundated area is worth knowing because of what lies under it. This section
is about the layer that decides how the carbon beneath a wetland is accounted
for, where that layer comes from, and why one binary in it changes the method
rather than the number.

::: callout-tip
## Verification track: this whole section

Everything from here to the conclusions rests on the IPCC Guidelines and their
2013 Wetlands Supplement, so it is registry material rather than method. A
reader who wants the spatial analysis alone can go straight to the conclusions
without losing the thread.

It is set as a section rather than a box because it runs to four parts and
carries a table, and because the binary it turns on decides which method applies
rather than only which number is reported. Whether it stays in the main text or
moves to the appendix is an open decision, recorded at the foot of this chapter.
:::

### The binary that decides everything

The IPCC divides every soil in the world into two classes and gives them
different accounting.

An **organic soil**, meaning a peat or Histosol, is defined by the 2006
Guidelines after FAO. The organic horizon must be at least ten centimetres
thick; a horizon under twenty centimetres must hold twelve per cent or more
organic carbon when mixed to twenty centimetres; a soil never saturated for more
than a few days must exceed twenty per cent organic carbon by weight. Everything
that is not an organic soil is a **mineral soil**.

That single classification changes the method.

| | Organic soil | Mineral soil |
|---|---|---|
| Method | Area-based annual **emission factor** | **Stock change** against a reference |
| Quantity | Tonnes per hectare **per year** | Tonnes per hectare, differenced over a period |
| Reference needed | None | Reference stock, plus land use, management and input factors |
| Default period | None. At Tier 1 there is **no transition period** | Twenty years by default |
| Recently against long drained | Tier 1 does not distinguish them | Tier 1 does not distinguish them |

The consequence is that the two are not comparable and cannot be summed
casually. A flux is a rate and a stock change is a difference, and a project
that reports both has to be explicit about which pools each covers.

A second rule follows from the definition and is easy to miss. **All organic
soils are assumed to have originally been wet**, so in the IPCC's language a dry
organic soil is always a *drained* organic soil, whatever anyone did or did not
do to it. Drainage is not established by evidence of ditching. It is inferred
from the soil being dry.

### The gases are not the same either

Drained organic soils emit carbon dioxide, and the Wetlands Supplement gives
Tier 1 factors for it in Table 2.1. They also emit methane, which the 2006
Guidelines had assumed negligible, and the Supplement gives that in Table 2.3
for the land and separately in Table 2.4 **for the drainage ditches**, which are
a small fraction of the area and a large fraction of the methane. Nitrous oxide
comes from the nitrogen in the organic matter and from any fertiliser. And
carbon leaves in the water as dissolved organic carbon, which the Supplement
gives a Tier 1 method for, and as particulate and dissolved inorganic carbon,
for which it gives none.

Rewetting inverts the pattern rather than reversing it. Carbon dioxide falls,
nitrous oxide drops to almost nothing, and methane rises, because the oxygen
level falls and methanogenesis restarts.

Coastal wetlands, meaning mangrove, tidal marsh and seagrass, are a separate
chapter again, and there methane and nitrous oxide are **not significant except
where the wetland receives agricultural run-off or sewage**. Saline inland
wetlands, the playas and pans and sabkhas, get no factors at all, because the
Supplement records that the literature cannot support them.

### The one number that matters most, and its interval

```{r ipcc-defaults}
# IPCC 2013 Wetlands Supplement, Table 2.1, transcribed from
# references/standards/IPCC-md/IPCC-2013-Wetlands-Supplement.md
tier1 <- data.frame(
  land_use = c("Forest, nutrient-poor", "Forest, nutrient-rich",
               "Forest", "Forest and cleared forest"),
  zone     = c("Boreal", "Boreal", "Temperate", "Tropical"),
  ef_tCO2C_ha_yr = c(0.37, 0.93, 2.6, 5.3),
  ci_low   = c(-0.11, 0.54, 2.0, -0.7),
  ci_high  = c(0.84, 1.30, 3.3, 9.5),
  sites    = c(63L, 62L, 8L, 21L))

tier1$spans_zero <- tier1$ci_low < 0 & tier1$ci_high > 0
knitr::kable(tier1)
```

```{r ipcc-numbers, echo=FALSE}
trop <- tier1[tier1$zone == "Tropical", ]
temp <- tier1[tier1$zone == "Temperate", ]
```

The tropical factor is `r trop$ef_tCO2C_ha_yr` tonnes of carbon dioxide carbon
per hectare per year, and its ninety-five per cent interval runs from
`r trop$ci_low` to `r trop$ci_high`. **It spans zero.** On `r trop$sites` sites,
the published Tier 1 default cannot establish whether a tropical drained organic
forest soil is emitting or removing.

Set that beside the temperate row, whose interval is tight and rests on
`r temp$sites` sites, and the boreal rows, whose intervals are wider on more
than sixty sites each. The width of a confidence interval here is telling you
about the spread of the underlying sites, not about how much evidence there is,
and the two are easy to confuse.

### Where the layer comes from

Two scales of answer exist, and a project needs both.

**Globally**, the reference soil group probabilities from SoilGrids give a
Histosol layer at 250 metres, and the Harmonized World Soil Database gives a
coarser classed alternative. Both answer the organic-or-mineral question
directly. The layer below was fetched from the ISRIC coverage service over this
basin and is committed with the book.

```{r histosol}
histosol <- rast("../data/chilwa_histosol_pct.tif")
names(histosol) <- "histosol_pct"

hb <- mask(crop(histosol, vect(st_transform(basin, crs(histosol)))),
           vect(st_transform(basin, crs(histosol))))
hv <- values(hb); hv <- hv[!is.na(hv)]

c(cells = length(hv),
  max_probability_pct  = max(hv),
  mean_probability_pct = round(mean(hv), 2),
  cells_above_10pct    = sum(hv > 10))
```

```{r hist-numbers, echo=FALSE}
h_max <- max(hv); h_n10 <- sum(hv > 10)
```

Across the whole Chilwa basin the highest Histosol probability anywhere is
`r h_max` per cent, and `r h_n10` cells exceed ten per cent.

**Chilwa is not a peatland.** That is a result rather than a disappointment, and
it settles the method. The organic soil apparatus above, Table 2.1 and its
methane companions, does not apply here. What applies is Chapter 5 of the
Supplement, on inland wetland mineral soils, which covers riparian wetlands,
forested swamps and marshes, works by stock change against a reference stock,
and defaults to a twenty year period.

Anyone who had reached for the tropical drained organic soil factor of
`r trop$ef_tCO2C_ha_yr` because the map said wetland would have applied a rate
to a soil that has no such rate, and would have done it on a factor whose sign
is unresolved.

**At country level** the global layers are a screening tool and not the answer.
National soil surveys, and in many countries a national wetland or peatland
inventory, carry the mapping the inventory compiler is expected to use, and
their class boundaries will not match SoilGrids. The global layer tells you
which question to ask. The national layer answers it.

One caution belongs with any organic soil figure outside Southeast Asia. The
Supplement records that organic soil fire emissions are validated for that
region only, and that FAOSTAT sets them to zero beyond Indonesia, Malaysia and
Brunei. An organic soil fire estimate for an African or American basin is an
extrapolation past the evidence its own default rests on, and should say so.

## Conclusions

An elevation model is an observation with a precision, and on low-relief ground
that precision governs the answer.

This basin's model resolves one metre, and one metre moves the lake by a factor
of `r jump`. `r w1` per cent of its depressions are a single quantisation step
deep and cannot be told from rounding. Conditioning it for flow routing raised the
lake bed by `r r_con` metres on average when the carve was capped at three
metres, and by nothing at all when it was not, because the cap removes the only
way out of a deep basin and the algorithm fills instead. The largest flow
accumulation in the grid belongs to a different catchment. And a Mercator cell
here is `r round(cs_mean, 3)` hectares, so the convenient identity overstates
area by `r cs_err` per cent.

None of that makes the data unusable. It makes the reporting conditional, and
the condition is always the same. State which surface, state its precision, and
state whether the quantity you are reporting depends on the depression shape or
only on the convergence pattern. Those three sentences are the difference
between an inundated area a verifier can accept and one they cannot.

## Exercises

1. Rerun the breaching step with `max_depth` at 1, 3 and 10 metres. Report the
   maximum carve and the maximum fill in each case, and confirm that only the
   first responds to the parameter.
2. Compute the stage-area curve at 0.1 metre steps between 0 and 2 metres. The
   elevation model stores integers. Explain what the intermediate points
   represent and whether they carry information.
3. Reproject the elevation model to UTM zone 36S before conditioning rather than
   after, and report whether the flow accumulation maximum lands in the same
   place. State which order you would defend.
4. Take the sink depth surface and classify depressions into those deeper and
   shallower than two metres. Map both. Say what fraction of the deep class lies
   inside the HydroLAKES polygon and what that implies about using depth alone
   as a filter.
5. *(Applied.)* A project reports 96,000 hectares of inundated wetland from this
   basin and does not state its stage or its surface. Using only the results in
   this chapter, give the range of stages consistent with that figure and say
   what you would ask for first.
6. *(Applied.)* Compute contributing area at the lake's convergence point twice,
   once by multiplying the cell count by 0.01 square kilometres and once with
   `cellSize()`. Report both and the difference in square kilometres.

::: callout-note
## Open decisions

**Does the soil section stay in the main text?** It runs from "Which soil, and
therefore which method" to the conclusions, rests entirely on the IPCC
Guidelines and the 2013 Wetlands Supplement, and the accepted proposal commits
that the main narrative can be read straight through without reference to any
crediting standard. It is marked as a track at its head rather than moved,
because moving 150 lines and a table into the appendix is a decision about the
book rather than about this chapter.

**The data script contradicts itself** and must be corrected.
`data-raw/prepare-chilwa-data.R` asserts the one-cell-one-hectare identity in
one paragraph and refutes it two paragraphs later. The measurement in this
chapter settles it at `r cs_err` per cent.

**No vertical accuracy is available for this surface.** The chapter states the
precision, which is measurable, and says plainly that no accuracy figure exists.
If a published assessment of the AWS Terrain Tiles mosaic over southern Africa
can be found, section two should cite it and the conclusions should be
rewritten around it.

**The stage-area anchor is not a validation** and the draft says so. Whether a
gauge record for Chilwa can be obtained is worth an hour. If it can, the rating
curve becomes a real one and this chapter gains its strongest number.
:::
