---
title: "Conditioning a Digital Elevation Model for Hydrological Routing"
subtitle: "Chapter 5 draft, 6 September 2026"
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

::: callout-note
## Draft

A restructure of `04-terrain-hydrology.qmd`, which retires when this is
promoted. The conditioning argument is unchanged because it was already right.
What is new is the opening, which now leads with the failure and its price, and
a section on the elevation model as an observed product with a precision you can
measure. Every number in the prose is computed by a chunk on this page.
:::

```{r setup}
library(terra)
library(sf)
library(whitebox)

sf_use_s2(FALSE)

dem    <- rast("../data/chilwa_dem.tif")
lake   <- st_read("../data/chilwa_hydro.gpkg", layer = "lake",   quiet = TRUE)
basin  <- st_read("../data/chilwa_hydro.gpkg", layer = "basin",  quiet = TRUE)
rivers <- st_read("../data/chilwa_hydro.gpkg", layer = "rivers", quiet = TRUE)
```

## One metre

Lake Chilwa sits in southern Malawi. Water enters from the surrounding hills and
leaves only by evaporation, so it is *endorheic*, meaning it has no outlet. Its
area swings by a factor of several between wet and dry years and it has dried out
completely three times in living memory, in 1968, 1995 and 2012.

A wetland carbon project on ground like this reports an inundated area. Here is
what that number is worth.

```{r stage-area}
# equal-area frame, and only the lake's own catchment
demu  <- project(dem, "EPSG:32736", method = "bilinear")
basu  <- vect(st_transform(basin, "EPSG:32736"))
laku  <- vect(st_transform(lake,  "EPSG:32736"))

d   <- mask(crop(demu, basu), basu)
csz <- cellSize(d, unit = "ha")

# the lake bed, taken as the fifth percentile of elevation under the polygon
bed <- as.numeric(global(mask(crop(demu, laku), laku),
                         quantile, probs = 0.05, na.rm = TRUE))

# the connected depression, not every low cell in the basin
patch_at_3m <- patches(ifel(d <= bed + 3, 1, NA), directions = 8)
main <- patch_at_3m %in% unique(na.omit(extract(patch_at_3m, laku)[, 2]))

stage <- data.frame(stage_m = 0:3)
stage$flooded_ha <- vapply(stage$stage_m, function(s)
  as.numeric(global(ifel(d <= bed + s & main, csz, 0), "sum", na.rm = TRUE)),
  numeric(1))
stage$ratio <- round(stage$flooded_ha / stage$flooded_ha[1], 1)
stage$flooded_ha <- round(stage$flooded_ha)
stage
```

```{r stage-numbers, echo=FALSE}
a0 <- stage$flooded_ha[1]; a1 <- stage$flooded_ha[2]; a2 <- stage$flooded_ha[3]
jump <- round(a1 / a0, 1)
next_step <- round(100 * (a2 - a1) / a1, 1)
hydro_ha <- round(sum(expanse(laku, unit = "ha")))
```

Raise the water level by one metre and the flooded area goes from
`r format(a0, big.mark = ",")` hectares to `r format(a1, big.mark = ",")`, a
factor of `r jump`. Raise it by a second metre and the area grows by
`r next_step` per cent. The curve is not gentle and it is not symmetric. Nearly
everything happens in the first metre.

Now the part that matters. **One metre is the finest distinction this elevation
model can make.**

```{r precision}
v <- values(dem); v <- v[!is.na(v)]

c(storage = datatype(dem),
  unique_values = length(unique(v)),
  all_integers = all(v == round(v)),
  smallest_step_m = min(diff(sort(unique(v)))))
```

The file holds signed integers. Every elevation in this basin is a whole number
of metres, and the smallest difference it can express is exactly the increment
that moves the lake by a factor of `r jump`.

So the chapter's question is not how to model inundation. It is what an
inundated area can mean when the surface it is measured on cannot resolve the
step that determines it. Everything below works toward an answer, and the answer
is not that the data are unusable.

## What the elevation model is

The previous chapter read a vertical accuracy off a lidar product: 9.8
centimetres or better at 95 per cent confidence, stated by the programme that
flew it. Ask this file the same question.

```{r provenance}
dem
```

It is a 100 metre grid in EPSG:3857 covering the basin plus margin. It came from
the Amazon Web Services Terrain Tiles mosaic through the `elevatr` package at
zoom level 10, which returns roughly five arc-second cells at this latitude.
That mosaic is composited from several global and national sources whose
identity varies by location and is not recorded per cell.

Nowhere in the file, and nowhere in the service that supplied it, is there a
vertical accuracy figure for this basin. There is no quality level, no
confidence statement, no separation of open ground from vegetated. What can be
established is what the file itself demonstrates, which is a one metre
quantisation, and that is a precision rather than an accuracy. A value can be
recorded to the metre and be wrong by ten.

That is the ordinary condition of terrain data in this field, and it is worth
stating plainly rather than working around quietly. A lidar collection tells you
how wrong it might be. A global elevation mosaic tells you the number.

## The grid it sits on

One property has to be settled before any area is quoted, because it silently
corrupts every hydrological figure that follows.

EPSG:3857 is a Mercator projection. It is conformal rather than equal-area, and
its scale factor grows with latitude, so a cell that is nominally 100 metres
square does not cover a hectare of ground anywhere except the equator.

```{r cell-area}
cs <- cellSize(dem, unit = "ha")
round(unlist(global(cs, c("min", "mean", "max"), na.rm = TRUE)), 4)
```

```{r cell-numbers, echo=FALSE}
cs_mean <- as.numeric(global(cs, "mean", na.rm = TRUE))
cs_err  <- round(100 * (1 / cs_mean - 1), 1)
```

A cell covers `r round(cs_mean, 3)` hectares, not one. Treating each cell as a
hectare, which is the standard shortcut for converting a flow accumulation count
into contributing area, **overstates area by `r cs_err` per cent**.

::: callout-warning
## A contradiction in this book's own data script

`data-raw/prepare-chilwa-data.R` states in its header that "at 100 m each cell
is exactly one hectare, so flow accumulation counts convert to contributing area
by multiplying by 0.01 to get square kilometres" and adds that "that identity is
used in the chapter's prose". Two paragraphs later the same header states that
the projection inflates area by about seven per cent and that the chapter
therefore reports in UTM.

Both cannot be true. The measurement above settles it: the identity is wrong by
`r cs_err` per cent and the second paragraph is right. The header must be
corrected, and it is recorded here rather than fixed silently because a
contradiction between a data script and the chapter that reads it is exactly the
kind of thing that survives review.
:::

This chapter uses `cellSize()` throughout and reports areas after reprojection
to UTM zone 36S, which is close enough to equal-area at this extent to serve as
truth.

## Terrain derivatives

Elevation is rarely the variable of interest. What it implies is.

```{r derivatives}
slope <- terrain(dem, "slope", unit = "degrees")
tri   <- terrain(dem, "TRI")

round(unlist(global(c(slope, tri), "mean", na.rm = TRUE)), 2)
```

```{r hillshade}
#| fig-cap: "Relief of the Lake Chilwa basin. The lake occupies the flat floor at centre; the Zomba and Mulanje massifs rise on the western and southern margins."
hs <- shade(terrain(dem, "slope",  unit = "radians"),
            terrain(dem, "aspect", unit = "radians"),
            angle = 40, direction = 315)

plot(hs, col = grey(0:100 / 100), legend = FALSE, mar = c(2, 2, 2, 2))
plot(st_geometry(lake),  add = TRUE, border = "steelblue", lwd = 2)
plot(st_geometry(basin), add = TRUE, border = "orangered", lwd = 1.5)
```

Slope governs operability and erosion risk, and ruggedness is a useful
productivity covariate, because sheltered gentler terrain often carries taller
and denser stands. Both are computed from differences between neighbouring
cells, so both inherit the one metre quantisation, and on ground this flat that
matters. A slope computed from integer elevations across 100 metre cells cannot
express anything shallower than about half a degree.

## Depressions, real and artefact

Water moves downhill, so an elevation model already encodes drainage. It does not
encode it cleanly. Every model contains cells with no lower neighbour, produced
by interpolation error, vegetation returns and finite vertical precision. A
routing algorithm entering one has nowhere to send the water, and the drainage
network fragments.

```{r sinks}
dir.create("../data/chilwa-hydro", showWarnings = FALSE)
wd <- "../data/chilwa-hydro/"
file.copy("../data/chilwa_dem.tif", paste0(wd, "dem.tif"), overwrite = TRUE)

wbt_depth_in_sink(dem = "dem.tif", output = "sink_depth.tif",
                  wd = wd, zero_background = TRUE, verbose_mode = FALSE)

sink_depth <- rast(paste0(wd, "sink_depth.tif"))
sd_v <- values(sink_depth); sd_v <- sd_v[!is.na(sd_v) & sd_v > 0]

data.frame(
  cells_in_a_depression = length(sd_v),
  within_1m  = round(100 * mean(sd_v <= 1), 1),
  within_2m  = round(100 * mean(sd_v <= 2), 1),
  max_depth_m = round(max(sd_v), 1))
```

```{r sink-numbers, echo=FALSE}
n_sink <- length(sd_v)
w1 <- round(100 * mean(sd_v <= 1), 1)
w2 <- round(100 * mean(sd_v <= 2), 1)
maxd <- round(max(sd_v))
```

`r format(n_sink, big.mark = ",")` cells sit in some depression, and the deepest
is `r maxd` metres below its spillover point.

Read the middle two columns against the previous section. **`r w1` per cent of
those depressions are one metre deep or less**, which is a single quantisation
step, and `r w2` per cent are within two. Those cannot be distinguished from
rounding. They are not necessarily artefacts, but nothing in the data can say
that they are not, and a rule that removes depressions below a depth threshold
is really a rule about the storage format.

No threshold on depth alone separates real basins from interpolation pits,
because the two overlap in size. The deepest single pit in this grid is an
artefact, not the lake.

## Conditioning, and the asymmetry inside it

Two families of algorithm exist. **Filling** raises the floor of a depression
until it overflows, which is simple and destroys the depression. **Breaching**
carves a channel out through the lowest point of the rim, which changes far
fewer cells and preserves the basin's shape.

WhiteboxTools implements Lindsay's hybrid method, which breaches what it can
within a depth limit and fills whatever remains. Two parameters do the work.
`max_depth` caps how deep a channel may be carved. `flat_increment` adds a small
monotonic gradient along the carved channel so flow directions stay unambiguous
across ground the model records as perfectly level.

```{r breach}
wbt_breach_depressions(dem = "dem.tif", output = "breached.tif", wd = wd,
                       max_depth = 3, flat_increment = 0.001,
                       verbose_mode = FALSE)

breached <- rast(paste0(wd, "breached.tif"))
change   <- breached - dem

data.frame(
  cells_changed = global(change != 0, "sum", na.rm = TRUE)[1, 1],
  max_carved_m  = round(-global(change, "min", na.rm = TRUE)[1, 1], 2),
  max_raised_m  = round(global(change, "max", na.rm = TRUE)[1, 1], 2))
```

```{r breach-numbers, echo=FALSE}
max_carve <- round(-global(change, "min", na.rm = TRUE)[1, 1], 1)
max_raise <- round(global(change, "max", na.rm = TRUE)[1, 1], 1)
```

Read both columns together, because the asymmetry between them is the single
most consequential behaviour in this chapter.

The carving obeys its limit exactly. No cell is lowered by more than the
`r max_carve` metres imposed. The raising does not obey it, because **the limit
does not apply to filling at all**. The deepest fill is `r max_raise` metres,
`r round(max_raise / max_carve)` times the carving limit that was set.

A published description of this method as one that preserves deeper basin
structures while breaching shallow artefacts is therefore only half right. The
constraint bounds the carving. It does not bound what the algorithm does to
everything it declines to carve.

```{r fill}
wbt_fill_depressions_wang_and_liu(dem = "breached.tif", output = "filled.tif",
                                  wd = wd, verbose_mode = FALSE)

filled <- rast(paste0(wd, "filled.tif"))
global((filled - breached) != 0, "sum", na.rm = TRUE)[1, 1]
```

Zero cells change in the second pass. That is the correct result and it is worth
pausing on, because it is easy to run a fill step, see a new file appear, and
assume it did something. The breaching step already filled every depression it
declined to breach, so its output was depression-free before the filling
algorithm ever saw it.

## What conditioning destroys

If the conditioned surface has no depressions left, then it has no Lake Chilwa.

```{r endorheic}
lake_mask <- rasterize(vect(lake), dem)
raised    <- ifel(!is.na(lake_mask), breached - dem, NA)

data.frame(
  mean_raise_m = round(global(raised, "mean", na.rm = TRUE)[1, 1], 1),
  max_raise_m  = round(global(raised, "max",  na.rm = TRUE)[1, 1], 1),
  raw_bed_m    = round(as.numeric(global(ifel(!is.na(lake_mask), dem, NA),
                                         quantile, probs = 0.5, na.rm = TRUE)), 1),
  conditioned_bed_m = round(as.numeric(global(ifel(!is.na(lake_mask), breached, NA),
                                              quantile, probs = 0.5, na.rm = TRUE)), 1))
```

```{r endo-numbers, echo=FALSE}
mr <- round(global(raised, "mean", na.rm = TRUE)[1, 1], 1)
xr <- round(global(raised, "max",  na.rm = TRUE)[1, 1], 1)
```

The lake bed has been raised by `r mr` metres on average and by `r xr` at its
deepest. In the conditioned model, Chilwa is a gentle ramp that water crosses on
its way somewhere else.

This is not a bug and it cannot be tuned away. Flow routing requires that every
cell have a downhill neighbour, and a terminal sink by definition does not. Any
algorithm that produces a routable surface must remove the sink. The question is
not how to avoid that but which surface each downstream calculation is entitled
to use, and there are only two rules.

**Flow routing uses the conditioned model**, and the endorheic basin is expected
to be gone from it. That is acceptable, because what routing is for is the
convergence pattern, and filling a flat sink with a monotonic gradient preserves
where water arrives even while inventing where it leaves.

**Anything depending on the shape of the depression uses the raw model.** That
means depression depth, inundated extent and any stage-area relationship,
including the calculation this chapter opened with. Running an inundation model
on a conditioned surface is a silent and total error, because the basin whose
flooding is being modelled has already been filled.

## Flow routing, and a maximum in the wrong place

```{r routing}
wbt_d8_pointer(dem = "breached.tif", output = "d8.tif", wd = wd,
               verbose_mode = FALSE)
wbt_d8_flow_accumulation(input = "d8.tif", output = "acc.tif", wd = wd,
                         pntr = TRUE, verbose_mode = FALSE)

acc <- rast(paste0(wd, "acc.tif"))
global(acc, "max", na.rm = TRUE)[1, 1]
```

```{r routing-check}
# where is that maximum, and is it in the lake?
peak_cell <- which.max(values(acc))
peak_xy   <- xyFromCell(acc, peak_cell)
in_lake   <- !is.na(extract(lake_mask, peak_xy)[, 1])

lake_acc <- global(ifel(!is.na(lake_mask), acc, NA), "max", na.rm = TRUE)[1, 1]

data.frame(global_max_cells = global(acc, "max", na.rm = TRUE)[1, 1],
           global_max_in_lake = in_lake,
           lake_max_cells = lake_acc)
```

```{r acc-numbers, echo=FALSE}
gmax <- global(acc, "max", na.rm = TRUE)[1, 1]
lmax <- lake_acc
```

The largest flow accumulation in the grid is
`r format(gmax, big.mark = ",")` cells, and it is **not in the lake**. It belongs
to a neighbouring catchment draining off the edge of the raster. Chilwa's own
convergence peaks at `r format(lmax, big.mark = ",")` cells.

That distinction is easy to lose. A script that reports the maximum accumulation
in a grid as the catchment's outlet is right whenever the grid contains one
catchment and wrong the rest of the time, and nothing about the output announces
which case you are in. The check is two lines and almost nobody writes them.

Multiplying either figure by a nominal hectare per cell would then compound the
error by the `r cs_err` per cent established earlier.

## Stage and area

Return to the opening calculation with everything the chapter has established.

```{r rating}
stage_full <- data.frame(stage_m = seq(0, 4, 0.5))
stage_full$flooded_ha <- vapply(stage_full$stage_m, function(s)
  round(as.numeric(global(ifel(d <= bed + s & main, csz, 0), "sum", na.rm = TRUE))),
  numeric(1))

plot(stage_full$stage_m, stage_full$flooded_ha / 1000, type = "b", pch = 19,
     xlab = "stage above lake bed (m)", ylab = "flooded area (thousand ha)",
     main = "Stage-area curve, Lake Chilwa basin")
abline(h = hydro_ha / 1000, lty = 2, col = "steelblue")
text(3, hydro_ha / 1000, "HydroLAKES maximum extent", pos = 3, cex = 0.8,
     col = "steelblue")
```

```{r rating-numbers, echo=FALSE}
at1 <- stage_full$flooded_ha[stage_full$stage_m == 1]
agree_pct <- round(100 * abs(at1 - hydro_ha) / hydro_ha, 1)
```

Two things are visible. The curve is nearly vertical in its first metre and
nearly flat afterwards, which is what a shallow basin with a wide flat floor
looks like. And the one metre stage encloses
`r format(at1, big.mark = ",")` hectares against the HydroLAKES published
maximum extent of `r format(hydro_ha, big.mark = ",")`, a difference of
`r agree_pct` per cent.

That agreement is a genuine result and it is easy to over-read. The published
polygon is a long-run maximum extent and the model surface is an instantaneous
threshold, so the two are not the same quantity and this is not a validation. It
does say that a stage one metre above the lake bed reproduces the published
maximum to within a couple of per cent, which is a useful anchor for a rating
curve and a poor substitute for a gauge.

::: {.verification}
#### Under Wetland Methodologies {.unnumbered .unlisted}

A wetland methodology asks for an inundated area, usually as an annual figure or
a class transition. Four questions decide whether the figure can be accepted.

Which surface was the extent computed on, the raw model or the conditioned one.
A conditioned surface has had the basin removed and cannot produce an inundated
extent at all.

What is the vertical precision of that surface, and how does it compare with the
stage difference the project is claiming. Here the precision is one metre and
one metre is a factor of `r jump` in area.

What was the stage, and where did it come from. A gauge, a satellite altimeter,
or a threshold chosen to reproduce a published polygon are three different
levels of evidence.

And in what projection was the area computed. A `r cs_err` per cent error is
available for free to anyone who counts cells on a Mercator grid.
:::

## Conclusions

An elevation model is an observation with a precision, and on low-relief ground
that precision governs the answer.

This basin's model resolves one metre, and one metre moves the lake by a factor
of `r jump`. `r w1` per cent of its depressions are a single quantisation step
deep and cannot be told from rounding. Conditioning it for flow routing raised
the lake bed by `r mr` metres on average and `r xr` at its deepest, because a
depth limit bounds carving and does not bound filling. The largest flow
accumulation in the grid belongs to a different catchment. And a Mercator cell
here is `r round(cs_mean, 3)` hectares, so the convenient identity overstates
area by `r cs_err` per cent.

None of that makes the data unusable. It makes the reporting conditional, and
the condition is always the same. State which surface, state its precision, and
state whether the quantity you are reporting depends on the depression shape or
only on the convergence pattern. Those three sentences are the difference
between an inundated area a verifier can accept and one they cannot.

## Exercises

1. Rerun the breaching step with `max_depth` at 1, 3 and 10 metres. Report the
   maximum carve and the maximum fill in each case, and confirm that only the
   first responds to the parameter.
2. Compute the stage-area curve at 0.1 metre steps between 0 and 2 metres. The
   elevation model stores integers. Explain what the intermediate points
   represent and whether they carry information.
3. Reproject the elevation model to UTM zone 36S before conditioning rather than
   after, and report whether the flow accumulation maximum lands in the same
   place. State which order you would defend.
4. Take the sink depth surface and classify depressions into those deeper and
   shallower than two metres. Map both. Say what fraction of the deep class lies
   inside the HydroLAKES polygon and what that implies about using depth alone
   as a filter.
5. *(Applied.)* A project reports 96,000 hectares of inundated wetland from this
   basin and does not state its stage or its surface. Using only the results in
   this chapter, give the range of stages consistent with that figure and say
   what you would ask for first.
6. *(Applied.)* Compute contributing area at the lake's convergence point twice,
   once by multiplying the cell count by 0.01 square kilometres and once with
   `cellSize()`. Report both and the difference in square kilometres.

::: callout-note
## Open decisions

**The data script contradicts itself** and must be corrected.
`data-raw/prepare-chilwa-data.R` asserts the one-cell-one-hectare identity in
one paragraph and refutes it two paragraphs later. The measurement in this
chapter settles it at `r cs_err` per cent.

**No vertical accuracy is available for this surface.** The chapter states the
precision, which is measurable, and says plainly that no accuracy figure exists.
If a published assessment of the AWS Terrain Tiles mosaic over southern Africa
can be found, section two should cite it and the conclusions should be
rewritten around it.

**The stage-area anchor is not a validation** and the draft says so. Whether a
gauge record for Chilwa can be obtained is worth an hour. If it can, the rating
curve becomes a real one and this chapter gains its strongest number.
:::
