---
title: "Designing a Field Plot Network, and What the Next Plot Buys"
subtitle: "Chapter 6 draft, 8 September 2026"
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

::: callout-note
## Draft

Drafted 8 September 2026 on the Smithsonian Conservation Biology Institute
ForestGEO census, which is committed to this repository and needs no network.
Every number in the prose below is computed by a chunk on this page. The chapter
ends where the field measurement ends. Converting a diameter into mass belongs
to the allometry chapter, scaling a plot mean to a landscape belongs to the
chapter after it, and fitting a remotely sensed predictor to these plots belongs
to the calibration chapter.
:::

```{r setup}
library(spatstat.geom)
library(ggplot2)

set.seed(2026)

stems <- read.csv("../data/scbi_stems.csv")
stems$ba <- pi * (stems$dbh / 200)^2   # basal area in square metres, dbh in cm

W <- 400      # plot width in metres, along the census grid
H <- 640      # plot height in metres
census_ha <- W * H / 1e4
```

## A forest where the answer is already known

The Smithsonian Conservation Biology Institute plot sits in Front Royal,
Virginia, and every woody stem inside its 400 by 640 metre rectangle was
located and measured in the 2008 census. That is `r format(nrow(stems), big.mark = ",")` stems over
`r sprintf("%.1f", census_ha)` hectares, each one carrying a surveyed position
on the plot grid and a diameter at breast height. A forest inventory exists to
estimate quantities that nobody can afford to measure completely, and this is a
forest where somebody did, which makes it the one place a sampling design can be
marked rather than argued about.

Every claim in this chapter was therefore checked the same way. A design was
written down, plots were placed under it several thousand times, and the
estimates those plots produced were compared against the quantity the census
already knew. Where a design is described as more precise, the spread of its
estimates was measured. Where an interval is described as honest, the fraction
of intervals that actually contained the true value was counted. Nothing below
rests on a formula that was not also run.

The chapter proceeds in five steps. It defines what a plot is and what
population it is drawn from, since the second is where most designs go wrong. It
measures what a sample of forty plots buys and shows that the answer is a
distribution rather than a number. It compares three ways of placing those plots
against each other on the same ground. It asks what a diameter threshold and a
plot size cost in fieldwork and buy in precision. It closes on positional
accuracy, where the measured result contradicts the reason most projects give
for wanting it.

## The trees that carry the wood, and the trees that carry the work

An inventory does not measure every stem it can see, because a protocol sets a
minimum diameter and the crew ignores anything below it. The consequence of that
threshold is visible in the census and it is larger than most crews expect.

```{r population}
big <- stems[stems$dbh >= 10, ]

pop <- data.frame(
  Population       = c("every stem, 1 cm and above", "stems 10 cm and above"),
  Stems            = c(nrow(stems), nrow(big)),
  `Stems per ha`   = round(c(nrow(stems), nrow(big)) / census_ha),
  `Basal area m2/ha` = round(c(sum(stems$ba), sum(big$ba)) / census_ha, 2),
  check.names = FALSE
)
knitr::kable(pop)

small_stems <- 100 * sum(stems$dbh < 10) / nrow(stems)
small_ba    <- 100 * sum(stems$ba[stems$dbh < 10]) / sum(stems$ba)
lindera_n   <- 100 * sum(stems$genus == "Lindera") / nrow(stems)
lindera_ba  <- 100 * sum(stems$ba[stems$genus == "Lindera"]) / sum(stems$ba)
```

Stems under 10 centimetres were `r sprintf("%.1f", small_stems)` per cent of
everything standing in the plot and carried
`r sprintf("%.1f", small_ba)` per cent of its basal area. Spicebush,
*Lindera benzoin* (L.) Blume, an understorey shrub, was
`r sprintf("%.1f", lindera_n)` per cent of the stems on its own and
`r sprintf("%.1f", lindera_ba)` per cent of the basal area. A crew that measures
to 1 centimetre spends most of its day on wood that will not appear in the
answer, which is why the rest of this chapter works on the
`r format(nrow(big), big.mark = ",")` stems of 10 centimetres and above, and why the cost of that
decision is measured later rather than assumed.

Basal area was used as the response throughout, being the cross-sectional area
of the stems at breast height summed over a unit of ground. It was chosen
because it is measured rather than modelled, so the sampling behaviour below is
the sampling behaviour of the design and not of an allometric equation. The
equation enters in the next chapter and it enters carrying its own error.

## The plot, and the population it is drawn from

A fixed-area plot is a circle of stated size centred on a located point, and
every stem whose centre falls inside it is measured. The plots used here were
400 square metres, which is a radius of
`r sprintf("%.2f", sqrt(400/pi))` metres and a common size in temperate carbon
inventory work.

```{r frame}
r_plot   <- sqrt(400 / pi)
plot_ha  <- 400 / 1e4
xrange   <- c(r_plot, W - r_plot)
yrange   <- c(r_plot, H - r_plot)
frame_ha <- diff(xrange) * diff(yrange) / 1e4

P <- ppp(big$gx, big$gy, window = owin(c(0, W), c(0, H)),
         marks = big$ba, checkdup = FALSE)

# basal area per hectare on plots centred at (cx, cy)
plot_ba <- function(cx, cy) {
  Q  <- ppp(cx, cy, window = owin(c(0, W), c(0, H)), checkdup = FALSE)
  cp <- crosspairs(P, Q, rmax = r_plot, what = "indices")
  v  <- numeric(length(cx))
  s  <- tapply(marks(P)[cp$i], cp$j, sum)
  v[as.integer(names(s))] <- s
  v / plot_ha
}

stems_per_plot <- nrow(big) / census_ha * plot_ha
```

At `r round(nrow(big)/census_ha)` stems per hectare of 10 centimetres and above,
a 400 square metre plot held about `r sprintf("%.0f", stems_per_plot)` trees, so
one plot was roughly a quarter of an hour of a two-person crew's time once the
centre had been found.

Plot centres were restricted to the interior of the census rectangle, at least
one plot radius from its edge, so that no plot was clipped by the boundary. That
restriction is not a technicality and it is the chapter's first debt to the
book's spine. Restricting the centres shrank the sampling frame from
`r sprintf("%.2f", census_ha)` hectares to
`r sprintf("%.2f", frame_ha)`, and the mean the sample estimates is therefore the
mean of that interior, not the mean of the whole census.

```{r truth}
gx <- seq(xrange[1], xrange[2], length.out = 220)
gy <- seq(yrange[1], yrange[2], length.out = 340)
grid_v <- plot_ba(expand.grid(cx = gx, cy = gy)$cx,
                  expand.grid(cx = gx, cy = gy)$cy)

target    <- mean(grid_v)          # what a randomly placed plot estimates
plot_sd   <- sd(grid_v)
whole     <- sum(big$ba) / census_ha
plot_cv   <- 100 * plot_sd / target
```

The two quantities differ. A randomly placed 400 square metre plot inside the
buffered frame returned `r sprintf("%.2f", target)` square metres per hectare on
average, while the census as a whole held
`r sprintf("%.2f", whole)`. The gap of
`r sprintf("%.2f", target - whole)` square metres per hectare is the edge of the
plot, which is thinner than its middle and which the design excluded. An
estimate of `r sprintf("%.2f", target)` is correct for the ground the design
sampled and wrong for the ground the project owns, and no amount of extra
fieldwork under that design will close the difference.

Plot to plot, basal area varied a great deal. The standard deviation across all
positions in the frame was `r sprintf("%.2f", plot_sd)` square metres per
hectare, a coefficient of variation of
`r sprintf("%.1f", plot_cv)` per cent, and that number is the forest rather than
the sample. It does not fall when more plots are installed. What falls is the
standard error, which describes the mean of the sample and not the spread of the
forest, and which carries a square root in its denominator that governs the
economics of every inventory ever commissioned.

## Forty plots, two thousand times

The clearest way to see what a sample buys is to draw it repeatedly on ground
whose answer is known. Simple random sampling was run at six sample sizes, two
thousand independent draws at each, and for every draw the estimate, its
standard error and its 90 per cent confidence interval were recorded.

```{r srs, cache = TRUE}
srs <- function(n, reps = 2000) {
  est <- se <- numeric(reps)
  for (i in seq_len(reps)) {
    cx <- runif(n, xrange[1], xrange[2])
    cy <- runif(n, yrange[1], yrange[2])
    v  <- plot_ba(cx, cy)
    est[i] <- mean(v)
    se[i]  <- sd(v) / sqrt(n)
  }
  half <- qt(0.95, n - 1) * se
  data.frame(
    n          = n,
    mean       = mean(est),
    bias       = mean(est) - target,
    sd_est     = sd(est),
    mean_se    = mean(se),
    coverage   = 100 * mean(abs(est - target) <= half),
    rel90      = median(100 * half / est),
    passing    = 100 * mean(100 * half / est <= 10)
  )
}

R <- do.call(rbind, lapply(c(10, 20, 40, 55, 80, 160), srs))
knitr::kable(R, digits = c(0, 2, 3, 3, 3, 1, 2, 1), row.names = FALSE)
```

Three things in that table matter and the third is the chapter's argument.

The estimator was unbiased at every sample size, the largest departure from the
target across all six being
`r sprintf("%.3f", max(abs(R$bias)))` square metres per hectare against a target
of `r sprintf("%.2f", target)`. The standard error computed from the sample
tracked the actual spread of the estimates closely, and the nominal 90 per cent
interval contained the true value between
`r sprintf("%.1f", min(R$coverage))` and
`r sprintf("%.1f", max(R$coverage))` per cent of the time. The arithmetic in
every inventory textbook does what it claims to do.

Precision improved with the square root of effort, so that going from
`r R$n[2]` plots to `r R$n[5]` plots, four times the fieldwork, halved the
standard error from `r sprintf("%.2f", R$mean_se[2])` to
`r sprintf("%.2f", R$mean_se[5])` square metres per hectare. That relationship
is the reason inventory budgets are what they are. Halving an uncertainty costs
four times as much, and halving it again costs sixteen.

## A forty-plot inventory is a coin toss

Registries do not ask for a standard error, they ask whether an estimate clears
a threshold. A common form of that test requires the half-width of the 90 per
cent confidence interval to be no more than 10 per cent of the estimate, and a
project that fails it is either deducted or sent back to the field.

The median inventory of `r R$n[3]` plots on this ground reported a half-width of
`r sprintf("%.2f", R$rel90[3])` per cent, which sits almost exactly on that
threshold. The word median is doing all the work in that sentence. Across two
thousand draws of the same design on the same forest by the same crew,
`r sprintf("%.0f", R$passing[3])` per cent of them passed and the rest failed,
on nothing whatever except which forty positions the random number generator
chose.

```{r threshold, cache = TRUE}
Rn <- do.call(rbind, lapply(c(45, 50, 55, 60, 70), srs))
knitr::kable(Rn[, c("n", "rel90", "passing")], digits = c(0, 2, 1),
             row.names = FALSE)
```

Reaching the threshold reliably took considerably more fieldwork than reaching
it on average. Forty plots passed
`r sprintf("%.0f", R$passing[R$n == 40])` per cent of the time and
`r Rn$n[which(Rn$passing >= 95)[1]]` plots passed
`r sprintf("%.0f", Rn$passing[which(Rn$passing >= 95)[1]])` per cent of the
time, so buying a
`r sprintf("%.0f", Rn$passing[which(Rn$passing >= 95)[1]])` per cent chance of
clearing the test rather than a coin toss cost
`r sprintf("%.0f", 100 * (Rn$n[which(Rn$passing >= 95)[1]] - 40) / 40)` per cent
more plots. A sample size computed from the standard formula targets the middle
of that distribution, which means a project that installs exactly the number the
formula returns has designed an inventory that fails about half the time.

::: callout-tip
## Verification track: what a sampling design must state

A design that cannot be re-run cannot be checked, and four things were needed
here before any of the numbers above could be reproduced.

The population, meaning the diameter threshold, the species included and the
pools measured. The frame, meaning the ground a plot centre could have landed
on, which here was `r sprintf("%.2f", frame_ha)` hectares and not the
`r sprintf("%.2f", census_ha)` the project owns. The selection rule, meaning how
centres were drawn and whether the draw was random, systematic or stratified,
because the variance estimator follows from it. And the plot itself, meaning
size, shape and how a stem on the boundary was decided.

A verifier reading an inventory report that states a mean, a standard error and
a plot count, but not the frame, cannot tell whether the estimate applies to the
project area. The correct question is not how many plots were installed. It is
what population the mean is a mean of.
:::

## The grid that is more precise than its own arithmetic admits

Plots are rarely thrown at a forest at random. A systematic grid with a random
start is easier to navigate, spreads the crew evenly, and cannot by accident
leave a whole corner of the property unvisited. It was compared here against
simple random sampling at the same forty plots on the same ground.

```{r systematic, cache = TRUE}
n <- 40; REPS <- 2000
kx <- diff(xrange) / 5; ky <- diff(yrange) / 8

est_sys <- se_sys <- numeric(REPS)
for (i in seq_len(REPS)) {
  g <- expand.grid(cx = xrange[1] + runif(1, 0, kx) + kx * (0:4),
                   cy = yrange[1] + runif(1, 0, ky) + ky * (0:7))
  v <- plot_ba(g$cx, g$cy)
  est_sys[i] <- mean(v)
  se_sys[i]  <- sd(v) / sqrt(n)
}

sys <- data.frame(
  Design   = c("simple random", "systematic 5 by 8, random start"),
  Bias     = c(R$bias[R$n == 40], mean(est_sys) - target),
  `True SD of the estimate` = c(R$sd_est[R$n == 40], sd(est_sys)),
  `SD the sample reported`  = c(R$mean_se[R$n == 40], mean(se_sys)),
  check.names = FALSE
)
knitr::kable(sys, digits = 3, row.names = FALSE)

gain <- 100 * (1 - sd(est_sys) / R$sd_est[R$n == 40])
```

The grid was genuinely the better design. Its estimates scattered
`r sprintf("%.0f", gain)` per cent less widely around the truth than the random
ones did, because a regular lattice cannot repeat itself the way an independent
draw can and therefore samples the plot's spatial structure more evenly.

The grid was not able to say so. The standard error each systematic sample
computed for itself was
`r sprintf("%.2f", mean(se_sys))` square metres per hectare, essentially the
same figure the random samples reported, because the formula treats the forty
values as independent draws and has no way of knowing they were laid out on a
lattice. There is no unbiased variance estimator for a systematic sample of one
random start, so the standard practice is to compute the interval as though the
sample were random and accept that the result is conservative. The project pays
for the extra precision in a better design and receives no credit for it in the
reported interval, which is worth knowing before a design is defended to a
verifier as more precise.

## Stratifying on something visible everywhere

Stratification splits the population into groups, samples within each, and
combines the group means with weights equal to the group areas. It reduces
variance when the strata differ from one another, and it can only be built from
something known for every hectare before any plot is installed, which in
practice means a map.

The stratifier used here was a deliberately imperfect map. True basal area was
computed on a 20 metre quadrat grid, Gaussian noise was added to it, and the
result was cut at its terciles into three strata. Varying the noise varied the
quality of the map, which is exactly the variable a project controls when it
decides how much to spend on remote sensing before the field season.

```{r stratified, cache = TRUE}
qs <- 20; nx <- W / qs; ny <- H / qs
qi <- pmin(floor(big$gx / qs), nx - 1)
qj <- pmin(floor(big$gy / qs), ny - 1)
qba <- tapply(big$ba, factor(qi + nx * qj, levels = 0:(nx * ny - 1)), sum)
qba[is.na(qba)] <- 0
qba <- as.numeric(qba) / (qs^2 / 1e4)
qcx <- (0:(nx * ny - 1) %%  nx) * qs + qs / 2
qcy <- (0:(nx * ny - 1) %/% nx) * qs + qs / 2
keep <- qcx >= xrange[1] & qcx <= xrange[2] & qcy >= yrange[1] & qcy <= yrange[2]

strat_run <- function(rho_quadrat, reps = 1500, per = 13) {
  sdn <- sd(qba) * sqrt(1 / rho_quadrat^2 - 1)
  est <- se <- rr <- numeric(reps)
  for (i in seq_len(reps)) {
    proxy <- qba + rnorm(length(qba), 0, sdn)
    br <- quantile(proxy[keep], c(0, 1/3, 2/3, 1))
    st <- cut(proxy, br, include.lowest = TRUE, labels = FALSE)
    e <- 0; v_num <- 0; allv <- allp <- numeric(0)
    for (k in 1:3) {
      idx <- which(keep & st == k)
      wk  <- length(idx) / sum(keep)
      pick <- sample(idx, per, replace = TRUE)
      cx <- pmin(pmax(qcx[pick] + runif(per, -qs/2, qs/2), xrange[1]), xrange[2])
      cy <- pmin(pmax(qcy[pick] + runif(per, -qs/2, qs/2), yrange[1]), yrange[2])
      v  <- plot_ba(cx, cy)
      e     <- e + wk * mean(v)
      v_num <- v_num + wk^2 * var(v) / per
      allv <- c(allv, v); allp <- c(allp, proxy[pick])
    }
    est[i] <- e; se[i] <- sqrt(v_num); rr[i] <- cor(allv, allp)
  }
  half <- qt(0.95, 3 * per - 1) * se
  data.frame(`Map correlation at plot scale` = round(mean(rr), 2),
             Bias = mean(est) - target,
             `True SD` = sd(est),
             Coverage = 100 * mean(abs(est - target) <= half),
             `Draws passing` = 100 * mean(100 * half / est <= 10),
             check.names = FALSE)
}

S <- do.call(rbind, lapply(c(0.9, 0.7, 0.5), strat_run))
S <- cbind(`Map correlation at quadrat scale` = c(0.9, 0.7, 0.5), S)
knitr::kable(S, digits = c(2, 2, 3, 3, 1, 1), row.names = FALSE)
```

The payoff was real and it scaled with the quality of the map. With the best
stratifier, thirty-nine plots split evenly across three strata passed the 10 per
cent test in `r sprintf("%.0f", S[1, "Draws passing"])` per cent of draws
against `r sprintf("%.0f", R$passing[R$n == 40])` per cent for forty randomly
placed plots. With the weakest, the pass rate was
`r sprintf("%.0f", S[3, "Draws passing"])` per cent. Stratification bought
between roughly ten and thirty percentage points of pass rate for no extra
fieldwork at all, and the entire difference between those two outcomes was the
quality of a map made before anyone went to the field.

The intervals were honest while it happened. Nominal 90 per cent coverage came
out between `r sprintf("%.1f", min(S$Coverage))` and
`r sprintf("%.1f", max(S$Coverage))` per cent, so the narrower interval was
narrower because the design was better and not because the estimator was
flattering itself.

One number in that table deserves attention because it is the trap in this
method. The map's correlation with basal area was set at the 20 metre quadrat
scale, but a 400 square metre plot placed somewhere inside a 400 square metre
quadrat does not see the quadrat's value, and the correlation that actually
governed the variance reduction was measured at
`r sprintf("%.2f", S[1, "Map correlation at plot scale"])` where the quadrat
figure was 0.90. A stratifier validated at one support does not carry that
validation to another, which is the same argument the optical chapter made about
a reflectance value and the laser chapter made about a height metric. The
chapter on calibration returns to it, because a model fitted at plot support and
applied at pixel support fails in exactly this way.

## Where the fix does not matter, and where it decides everything

Field crews are told that plot centres must be located accurately, and the
reason usually given is that an inaccurate centre moves which trees fall inside
the plot. That reason was tested here by displacing every plot centre by a
Gaussian error before the plot was measured.

```{r position, cache = TRUE}
pos_run <- function(sigma, reps = 1500) {
  est <- se <- numeric(reps)
  for (i in seq_len(reps)) {
    cx <- runif(n, xrange[1], xrange[2]); cy <- runif(n, yrange[1], yrange[2])
    cx <- pmin(pmax(cx + rnorm(n, 0, sigma), xrange[1]), xrange[2])
    cy <- pmin(pmax(cy + rnorm(n, 0, sigma), yrange[1]), yrange[2])
    v <- plot_ba(cx, cy); est[i] <- mean(v); se[i] <- sd(v) / sqrt(n)
  }
  data.frame(`Fix error, m` = sigma, Bias = mean(est) - target,
             `SD of the estimate` = sd(est), check.names = FALSE)
}
Pos <- do.call(rbind, lapply(c(0, 5, 10, 20), pos_run))
knitr::kable(Pos, digits = 3, row.names = FALSE)
```

The reason usually given is wrong, at least for this purpose. Displacing every
centre by a standard deviation of 20 metres, nearly twice the plot radius, left
the estimate unbiased and moved the spread of the estimates from
`r sprintf("%.3f", Pos[1, 3])` to `r sprintf("%.3f", Pos[4, 3])` square metres
per hectare, a change of `r sprintf("%.1f", 100*(Pos[4,3]/Pos[1,3]-1))` per
cent. The trees inside the plot did indeed change, and it did not matter,
because a randomly placed plot that is then moved at random is still a randomly
placed plot, and the estimator only ever asked for that.

Positional accuracy earns its cost somewhere else. A permanent plot exists to be
found again, and growth is estimated from the difference between two visits to
the same trees, which is a paired comparison whose precision comes entirely from
the pairing. Losing the location breaks the pair.

```{r remeasure, cache = TRUE}
growth_run <- function(sigma, reps = 1500, rate = 0.03) {
  ch <- numeric(reps)
  for (i in seq_len(reps)) {
    cx <- runif(n, xrange[1], xrange[2]); cy <- runif(n, yrange[1], yrange[2])
    v1 <- plot_ba(cx, cy)
    rx <- pmin(pmax(cx + rnorm(n, 0, sigma), xrange[1]), xrange[2])
    ry <- pmin(pmax(cy + rnorm(n, 0, sigma), yrange[1]), yrange[2])
    v2 <- plot_ba(rx, ry) * (1 + rate)
    ch[i] <- mean(v2 - v1)
  }
  data.frame(`Relocation error, m` = sigma,
             `Mean change` = mean(ch), `SD of the change` = sd(ch),
             `Signal to noise` = rate * target / sd(ch), check.names = FALSE)
}
Gr <- do.call(rbind, lapply(c(0, 3, 5, 10), growth_run))
knitr::kable(Gr, digits = 3, row.names = FALSE)
```

Growth was imposed rather than observed, every stem gaining 3 per cent of basal
area, so that the true change was exactly
`r sprintf("%.3f", 0.03 * target)` square metres per hectare and the only thing
varying between the rows of that table was the crew's ability to find the plot
again. That is the one place in this chapter where a simulated quantity was used
and the reason is that no real dataset can hold both a known growth increment
and a controlled relocation error.

A plot relocated perfectly recovered the increment with a standard deviation of
`r sprintf("%.3f", Gr[1, 3])` square metres per hectare, a signal-to-noise ratio
of `r sprintf("%.1f", Gr[1, 4])`. A relocation error of three metres, which is
an ordinary consumer handheld receiver under canopy, raised that standard
deviation to `r sprintf("%.2f", Gr[2, 3])` and dropped the ratio to
`r sprintf("%.2f", Gr[2, 4])`, which is to say the noise became larger than the
thing being measured. At ten metres the ratio was
`r sprintf("%.2f", Gr[4, 4])`.

The rule that follows is not the one crews are usually given. For a one-off
estimate of a stand mean the fix hardly matters, and money spent on survey-grade
positioning would be better spent on more plots. For a permanent plot network
measuring change, the fix is the measurement, and a network whose centres cannot
be reoccupied to within a metre is not measuring growth however carefully its
diameters are read. That is why permanent sample plots are monumented with
rebar and witness trees rather than recorded as a coordinate, and the
justification is the paired comparison rather than the tree list.

## What a threshold and a plot size cost

Two design choices remain and both trade fieldwork against precision. The first
is the diameter threshold, already introduced, and the second is the size of the
plot itself.

```{r effort, cache = TRUE}
P_all <- ppp(stems$gx, stems$gy, window = owin(c(0, W), c(0, H)),
             marks = stems$ba, checkdup = FALSE)
ba_all <- function(cx, cy) {
  Q <- ppp(cx, cy, window = owin(c(0, W), c(0, H)), checkdup = FALSE)
  cp <- crosspairs(P_all, Q, rmax = r_plot, what = "indices")
  v <- numeric(length(cx)); s <- tapply(marks(P_all)[cp$i], cp$j, sum)
  v[as.integer(names(s))] <- s; v / plot_ha
}
gg <- expand.grid(cx = gx, cy = gy)
v_all <- ba_all(gg$cx, gg$cy)

thr <- data.frame(
  Threshold        = c("1 cm", "10 cm"),
  `Stems per plot` = round(c(nrow(stems), nrow(big)) / census_ha * plot_ha, 1),
  `Plot CV, %`     = round(c(100 * sd(v_all) / mean(v_all),
                             100 * plot_sd / target), 1),
  check.names = FALSE
)
knitr::kable(thr)
```

Dropping the threshold from 10 centimetres to 1 raised the trees measured on an
average plot from `r thr[2, 2]` to `r thr[1, 2]`, roughly
`r sprintf("%.1f", thr[1,2]/thr[2,2])` times the fieldwork, and moved the
coefficient of variation from `r thr[2, 3]` per cent to `r thr[1, 3]`. Roughly
`r sprintf("%.0f", thr[1,2] - thr[2,2])` extra stems on every plot, which is
four fifths of the crew's measuring time, took
`r sprintf("%.1f", thr[2,3] - thr[1,3])` percentage points off the coefficient
of variation, and that is the same argument the chapter makes about plots taken
one scale further down. The small stems are worth measuring when regeneration is
the question, and they are not worth measuring when standing stock is.

```{r plotsize, cache = TRUE}
size_run <- function(area_m2, reps = 800) {
  rr <- sqrt(area_m2 / pi); a <- area_m2 / 1e4
  xr2 <- c(rr, W - rr); yr2 <- c(rr, H - rr); nn <- round(16000 / area_m2)
  f <- function(cx, cy) {
    Q <- ppp(cx, cy, window = owin(c(0, W), c(0, H)), checkdup = FALSE)
    cp <- crosspairs(P, Q, rmax = rr, what = "indices")
    v <- numeric(length(cx)); s <- tapply(marks(P)[cp$i], cp$j, sum)
    v[as.integer(names(s))] <- s; v / a
  }
  est <- se <- numeric(reps)
  for (i in seq_len(reps)) {
    cx <- runif(nn, xr2[1], xr2[2]); cy <- runif(nn, yr2[1], yr2[2])
    v <- f(cx, cy); est[i] <- mean(v); se[i] <- sd(v) / sqrt(nn)
  }
  half <- qt(0.95, nn - 1) * se
  data.frame(`Plot size, m2` = area_m2, Plots = nn,
             `SD of the estimate` = sd(est),
             `Draws passing` = 100 * mean(100 * half / est <= 10),
             check.names = FALSE)
}
Sz <- do.call(rbind, lapply(c(100, 400, 1000, 2000), size_run))
knitr::kable(Sz, digits = c(0, 0, 3, 1), row.names = FALSE)
```

Every row of that table measured 1.6 hectares of ground, so the crew walked
further on the first row and measured longer on the last, and the total area
inside the tapes was held constant. The 400 square metre plot was the best of
the four, passing the test in `r sprintf("%.0f", Sz[2, 4])` per cent of draws
against `r sprintf("%.0f", Sz[1, 4])` for a hundred small ones and
`r sprintf("%.0f", Sz[4, 4])` for eight large ones. Small plots suffer because
a single large tree either falls inside or does not, and at 100 square metres
that one decision moves the plot's value violently. Large plots suffer because
eight of them cannot cover the variation in a
`r sprintf("%.0f", frame_ha)` hectare stand however carefully each is measured.

The differences in the true standard deviation across the first three rows were
small, from `r sprintf("%.2f", min(Sz[1:3, 3]))` to
`r sprintf("%.2f", max(Sz[1:3, 3]))`, while the pass rates spanned
`r sprintf("%.0f", min(Sz[1:3, 4]))` to
`r sprintf("%.0f", max(Sz[1:3, 4]))` per cent. That sensitivity is a property of
the threshold rather than of the forest, because a design sitting on a pass or
fail boundary converts a small change in precision into a large change in
outcome, and every design in this chapter was placed deliberately near that
boundary for exactly that reason.

## Conclusions

The census at Front Royal was complete, so every design in this chapter could be
marked against an answer rather than defended by an argument, and four results
came out of that marking.

A sample size computed from the standard formula targets the middle of a
distribution of outcomes, so forty plots that return a median half-width of
`r sprintf("%.1f", R$rel90[R$n == 40])` per cent against a 10 per cent threshold
passed `r sprintf("%.0f", R$passing[R$n == 40])` per cent of the time and
`r Rn$n[which(Rn$passing >= 95)[1]]` plots were needed before passing became
reliable. Eligibility is a random variable and a project that budgets for the
median has budgeted for a coin toss.

A systematic grid was `r sprintf("%.0f", gain)` per cent more precise than
simple random sampling and could not report the fact, because no unbiased
variance estimator exists for a single random start and the conventional
practice of computing the interval as though the sample were random is
conservative by construction.

Stratifying on a map raised the pass rate from
`r sprintf("%.0f", R$passing[R$n == 40])` per cent to
`r sprintf("%.0f", S[1, "Draws passing"])` for no extra fieldwork, with honest
coverage throughout, and the size of that gain was set by the map's correlation
with basal area at plot support rather than at the support the map was validated
on.

Positional accuracy did almost nothing for a one-off estimate of a stand mean
and everything for a remeasurement, a three metre relocation error dropping the
signal-to-noise ratio on a growth estimate from
`r sprintf("%.1f", Gr[1, 4])` to `r sprintf("%.2f", Gr[2, 4])`. The common
justification for accurate plot centres, that the fix decides which trees are
in the plot, is true and is not the reason the fix matters.

Underneath all four sits the frame. Restricting plot centres to the interior of
the census moved the quantity being estimated from
`r sprintf("%.2f", whole)` to `r sprintf("%.2f", target)` square metres per
hectare, a difference no sample size can remove, because a plot has the support
of its own area and a stand does not inherit that support for free. The next
chapter takes the diameters these plots recorded and converts them into mass,
where the error introduced by that conversion is larger than anything measured
here.

## Exercises

1. The plot centres in this chapter were restricted to the interior of the
   census rectangle. Re-run the simple random sampling with centres drawn from
   the full rectangle and plots clipped at the boundary, and report the bias
   that appears and its sign.

2. The stratified design allocated thirteen plots to each of three strata.
   Reallocate them in proportion to the standard deviation within each stratum,
   which is Neyman allocation, and report how much of the remaining variance it
   removes.

3. The map used for stratification was built from true quadrat basal area plus
   noise. Replace it with a stratifier that is spatially smooth but only weakly
   related to basal area, and show that a map can be visually convincing and
   worth nothing as a stratifier.

4. The remeasurement demonstration imposed uniform 3 per cent growth on every
   stem. Impose instead a growth increment that depends on diameter, and report
   whether relocation error then biases the growth estimate as well as
   inflating its variance.

5. Basal area was used throughout because it is measured rather than modelled.
   Repeat the forty-plot sampling using stem count per hectare as the response
   and explain, from the coefficient of variation of each, why the two need
   different sample sizes on the same ground.

::: callout-note
## Open decisions

The plot sizes compared here were circular and fixed-area. Variable-radius
sampling with an angle gauge, which selects trees with probability proportional
to basal area and is the dominant field method in North American operational
inventory, is not covered and would fit either this chapter or the one on
scaling.

The stratifier is synthetic, being true quadrat basal area with noise added.
It is honest about being synthetic and it makes the argument cleanly, but a real
canopy height model over these hectares would make the same argument without the
caveat and would connect this chapter to the laser chapter directly. No lidar
coverage of the Front Royal plot is committed to this repository.

The 3 per cent growth increment is imposed rather than measured. The 2013 and
2018 SCBI censuses exist and would supply a real increment on the same stems,
which would remove the only simulated quantity in the chapter. Only the 2008
census is committed.
:::
