---
title: "Spatial R Basics"
format:
  docx:
    reference-doc: ./references/style.docx
    highlight-style: github
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Reproducibility

An audit stands or falls on whether someone else can reproduce the number. A spreadsheet that took forty manual clicks to build cannot be re-run, cannot be reviewed line by line, and cannot be trusted a year later when the inputs have changed. A script can. This is the central reason to do spatial work in R: every decision, from the projection you chose to the buffer distance you applied, is written down and repeatable.

R is a free, open-source statistical language with a mature ecosystem for geographic data. In a single session you can read a shapefile a surveyor sent you, pull a digital elevation model from an open server, overlay the two, extract terrain values at your field plots, fit a model, and export a map, without ever leaving the environment or paying for a licence. For a forest-carbon or agriculture auditor, that means the entire chain from raw data to reported estimate lives in one reviewable document.

This course uses the modern R-spatial stack, built around two packages that have become the standard tools of the field [@pebesma2018sf; @hijmans2024terra].

## The R-spatial stack

You will meet these packages throughout the book. Install them once before you begin.

| Package | Role |
|---|---|
| `sf` | Vector data: points, lines, polygons |
| `terra` | Raster data: imagery, elevation, gridded surfaces |
| `dplyr` / `tidyverse` | Data wrangling and pipes |
| `tmap` | Thematic maps, static and interactive |
| `ggplot2` | Statistical graphics |
| `gstat` | Geostatistics and interpolation |
| `spdep` | Spatial autocorrelation and neighbours |

To install everything the course needs, run this once. It is set not to execute when the book is rendered, so run it yourself in the console.

```{r}
#| eval: false
install.packages(c(
  "sf", "terra", "dplyr", "tidyverse", "tmap",
  "ggplot2", "gstat", "spdep", "elevatr", "rnaturalearth"
))
```

Load the core three at the top of any spatial script:

```{r}
library(sf)
library(terra)
library(dplyr)
```

The messages `sf` and `terra` print on loading tell you which versions of the underlying geospatial libraries (GDAL, GEOS, PROJ) you are linked against. It is good practice to note them, because a projection or geometry result can occasionally differ between library versions.

## Reproducible projects

Before loading any data, put your work in an **RStudio Project**. A project is simply a folder with an `.Rproj` marker file, but it fixes two things that otherwise cause endless confusion: it sets the working directory to the project root, and it lets you refer to data with short relative paths like `data/forest_plots.csv` instead of a full path tied to your machine. Open RStudio, choose *File ▸ New Project*, and point it at a folder that contains a `data/` subfolder.

A tidy layout for a course like this looks like:

```
training-rspatial/
├── data/              # input datasets
├── 01-r-for-spatial.qmd
├── 02-data-models.qmd
├── ...
└── training-rspatial.Rproj
```

With a project open, the working directory is the project root, so every path in this book that begins with `data/` will resolve correctly.

## The example data

The running example is a forest inventory from the Alberni Valley on Vancouver Island. Three files in `data/` describe it:

- `forest_plots.csv` — one row per inventory plot, with coordinates and stand-level summaries.
- `tree_measurements.csv` — one row per measured tree, linked to a plot by `plot_id`.
- `forest_plots.shp` — the same plots as a spatial point layer (an ESRI shapefile).

As noted in the preface, these measurements are simulated for teaching, but they are built to behave like a real fixed-area plot cruise.

### Data provenance

Before any of this reaches R, someone stood in the forest and recorded it. The credibility of every downstream number rests on how those coordinates and measurements were captured: the GPS receiver's accuracy under canopy, whether positions were averaged or differentially corrected, and whether the survey followed a documented protocol. In operational work I capture plots with standardised digital forms, such as ESRI FieldMaps or KoboToolbox, that enforce the schema at the point of entry and log an accuracy estimate with every fix, and I assess receiver accuracy against known control before a campaign. An auditor should always ask for that provenance: a plot table is only as trustworthy as the survey that produced it. We treat the columns in `forest_plots.csv` as the clean output of such a protocol.

## Reading data

Start with the plot table. It is an ordinary CSV, so base R reads it without any spatial machinery:

```{r}
plots <- read.csv("data/forest_plots.csv")
head(plots)
```

Two functions tell you almost everything about a new dataset. `str()` shows the structure, the type of each column and a few values; `summary()` shows the distribution of each column:

```{r}
str(plots)
summary(plots[, c("stems_ha", "basal_area", "agb", "carbon")])
```

Already we can read the landscape: stem densities in the hundreds per hectare, basal areas around 30 m² ha⁻¹, and aboveground carbon densities that range from young riparian stands to carbon-dense old growth. The carbon column is expressed in megagrams of carbon per hectare (MgC ha⁻¹), applying the IPCC default carbon fraction of 0.47 to aboveground biomass [@ipcc2006gl].

Now the tree table, which is much longer because it holds every stem:

```{r}
trees <- read.csv("data/tree_measurements.csv")
nrow(trees)
head(trees)
```

We can already do useful forestry with plain `dplyr`. Which species carry the most measured biomass across all plots?

```{r}
trees |>
  group_by(species) |>
  summarise(n_stems = n(),
            mean_dbh = round(mean(dbh_cm), 1),
            total_agb_Mg = round(sum(agb_kg) / 1000, 1)) |>
  arrange(desc(total_agb_Mg))
```

## A first map

The CSV knows nothing about geography yet: `lon` and `lat` are just numbers. The shapefile, by contrast, carries geometry and a coordinate reference system, and `sf` reads it into a spatial object:

```{r}
plots_sf <- st_read("data/forest_plots.shp", quiet = TRUE)
plots_sf
```

Notice what `sf` prints: the geometry type (`POINT`), the bounding box, the coordinate reference system (`Geodetic CRS: WGS 84`), and then the attribute table with a `geometry` column tacked on the end. An `sf` object is a data frame that also knows where each row sits on Earth.

Plotting one attribute draws a map coloured by that value:

```{r}
plot(plots_sf["carbon"], pch = 19, main = "Aboveground carbon (MgC/ha)")
```

There is our study area: 140 plots scattered across the valley, shaded by carbon density. From here the rest of the course is about doing this properly, understanding the coordinate system, bringing in terrain and imagery, and turning these points into landscape estimates with defensible uncertainty.

## Exercises

1. Read `forest_plots.csv` and report the mean and standard deviation of `carbon` (in MgC/ha) for each `stand_type`. Which stand type is most variable, and why might that be?
2. Using `tree_measurements.csv`, find the ten largest trees by diameter. What species are they, and which plots do they fall in?
3. The `agb_kg` column is aboveground biomass per tree. Add a new column giving each tree's carbon in kilograms (biomass × 0.47), then compute total plot carbon by summing over `plot_id`. Compare your plot totals to the `agb` column (Mg/ha) in `forest_plots.csv` after accounting for the 0.1-hectare plot area.
4. Plot the `forest_plots` shapefile coloured by `stand_type` instead of `carbon`. Do the stand types cluster spatially?
