---
title: "Spherical Geometry"
format:
  docx:
    reference-doc: ./references/style.docx
    highlight-style: github
execute:
  eval: true
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    prefer-html: true
    dev: "png"
    dpi: 300
bibliography: ./references/references.bib
csl: ./references/apa.csl
---

## Spherical geometry

Longitude and latitude are angles on a curved surface, not positions on a flat sheet. Most of the time we quietly ignore this, we plot lon/lat on a rectangular axis and it looks fine, but the moment we *measure* on those coordinates, the curvature matters. A degree of longitude is about 111 km at the equator and shrinks to zero at the poles, so treating lon/lat as if they were metres on a plane distorts every distance and area. This chapter is about doing geometry correctly on a round Earth.

```{r}
library(sf)
```

## Planar and spherical

There are two ways to compute with geographic coordinates. **Planar** geometry treats lon/lat as Cartesian x/y and applies flat-earth formulae; it is fast but wrong for anything beyond a small area. **Spherical** geometry computes on the surface of the sphere (or ellipsoid), so distances follow great circles and areas account for convergence toward the poles. The straight line between two points on a map is generally *not* the shortest path on the globe, which is the geodesic.

## The s2 library

Modern `sf` handles this automatically. When data are in a geographic CRS (lon/lat), `sf` uses Google's **s2** spherical-geometry library by default, so measurements and predicates are computed on the sphere without you doing anything. You can see and toggle this:

```{r}
sf_use_s2()               # TRUE: spherical geometry is active for lon/lat
```

## Geodesic measures

Because s2 is active, an area or distance computed on lon/lat data is already geodesic, returned in metres and square metres even though the coordinates are degrees:

```{r}
plots <- st_as_sf(read.csv("data/forest_plots.csv"),
                  coords = c("lon", "lat"), crs = 4326)

st_distance(plots[1:2, ])          # geodesic distance in metres, from degrees
```

The difference between spherical and planar becomes visible if we switch s2 off and compare. With s2 disabled, `sf` falls back to treating degrees as planar coordinates, and the "distance" comes back in meaningless degree units:

```{r}
sf_use_s2(FALSE)
st_distance(plots[1:2, ])          # planar: nonsense units (degrees)
sf_use_s2(TRUE)                    # restore the correct default
```

This is why simply reading a distance off lon/lat coordinates, without either s2 or a projection, gives numbers that look plausible but are wrong.

## Local and global

The size of the error depends on extent. Over a single valley the curvature is negligible and a local projection and the sphere agree to within centimetres. Over a province or a continent the difference is large, and near the poles or across the antimeridian (the ±180° line) planar formulae fail completely. The working guidance is simple: for **local** analysis, project to a metric CRS and work on the plane, as we do for the forest study area; for **global or continental** extents, keep the data in lon/lat and let spherical geometry do the work. Knowing which regime you are in, and that `sf` defaults to the correct one for geographic data, is the foundational point.

## Exercises

1. Confirm `sf_use_s2()` is `TRUE`, then compute the geodesic distance between the two most widely separated plots. Switch s2 off, repeat, and explain why the second result is not a distance at all.
2. Compare the area of the study-area bounding box computed on lon/lat (spherical) with the same box reprojected to BC Albers (planar). How close are they, and why so close at this extent?
3. Explain, for a dataset spanning all of Canada, why an equal-area projection or spherical geometry is essential and a single UTM zone is not.
4. The shortest path between two points on the globe is a great circle. Sketch why a straight line drawn between two cities on a Mercator map overstates the true travel distance at high latitudes.
